---
author: Mallory L. Barnes
abstract-title: Chapter notes
abstract: Adapted to environmental science by Mallory Barnes. Portions adapted from Chapters 1 and 2 in Navarro, D. J. "Learning Statistics with R." <https://compcogscisydney.org/learning-statistics-with-r/>
aliases: [why-statistics.html]
---

```{r, include = FALSE}
source("global_stuff.R")
```

# Why Statistics?

> To call in statisticians after the experiment is done may be no more than asking them to perform a post-mortem examination: They may be able to say what the experiment died of. ---Sir Ronald Fisher

## On the Importance of Statistics in Environmental Science

Many students come into statistics classes with a mix of nerves and low expectations. It is not always the most eagerly anticipated part of an environmental science degree. But statistics is one of the most important tools we have for understanding complex systems. Environmental data are messy: ecosystems shift, climate varies, and our judgments are often biased. Without statistical methods we risk telling convenient stories. With them, we can detect real patterns, test competing explanations, and make stronger arguments for environmental decisions.

### Why we need Statistics: Bias and Simpson's Paradox

Common sense is useful, but it is prone to bias. When we already believe something to be true, we are more likely to interpret new evidence as supporting it, even if it does not. This tendency can lead us astray in environmental science.

For example, if you see sick fish downstream of a factory, you might conclude that the factory caused it. For example, if you are convinced that industrial farming is the main driver of bee declines, you might see every new drop in bee abundance as confirmation of that belief. Maybe you are right, but maybe disease or weather are stronger contributors. Statistics gives us tools to test these ideas systematically. It helps us separate patterns from noise, avoid being misled by our own expectations, and draw conclusions that are more likely to hold up under scrutiny. It's not magic, but it makes our conclusions far more reliable.

Another challenge is that data can tell very different stories depending on how they are aggregated. *Simpson’s paradox* occurs when a trend seen in combined data reverses after you split the data into meaningful groups.

Here's a recent example. During the COVID-19 pandemic, early data showed that Italy’s overall case fatality rate was higher than China’s. But, once researchers *disaggregated by age group,* the trend flipped: within every age group, the fatality rate was actually higher in China. The apparent contradiction was explained by differences in the age distribution of cases [@vonkügelgen2021].

Using case fatality rate (CFR), here are some illustrative numbers (made up for teaching).

```{r}
#| label: fig-simpson-covid-overall
#| fig-cap: "Crude case fatality rates. Italy appears to have a higher fatality rate than China when all cases are combined."
library(ggplot2)
library(tibble)

age_levels <- c("0–9","10–19","20–29","30–39","40–49","50–59","60–69","70–79","80+")

dat <- tibble::tribble(
  ~Country, ~Age,     ~Cases, ~Deaths,  ~CFR_band_pct,
  "Italy",  "0–9",       500,   0.05,     0.01,
  "Italy",  "10–19",     500,   0.10,     0.02,
  "Italy",  "20–29",     800,   0.32,     0.04,
  "Italy",  "30–39",    1000,   1.50,     0.15,
  "Italy",  "40–49",    2000,   8.00,     0.40,
  "Italy",  "50–59",    2000,  20.00,     1.00,
  "Italy",  "60–69",    1500,  37.50,     2.50,
  "Italy",  "70–79",    1200,  72.00,     6.00,
  "Italy",  "80+",      1500, 225.00,    15.00,

  "China",  "0–9",      8000,   1.60,     0.02,
  "China",  "10–19",    8000,   2.40,     0.03,
  "China",  "20–29",    8000,   4.00,     0.05,
  "China",  "30–39",    7000,  14.00,     0.20,
  "China",  "40–49",    7000,  42.00,     0.60,
  "China",  "50–59",    3000,  45.00,     1.50,
  "China",  "60–69",    1000,  35.00,     3.50,
  "China",  "70–79",     700,  56.00,     8.00,
  "China",  "80+",       300,  54.00,    18.00
) %>%
  mutate(
    Age = factor(Age, levels = age_levels),
    CFR_band_calc = 100 * Deaths / Cases  # sanity check, should match CFR_band_pct
  )

# Quick checks
dat %>% group_by(Country) %>% summarise(Total_cases = sum(Cases), Total_deaths = sum(Deaths),
                                        CFR_crude_pct = 100*Total_deaths/Total_cases)
# Italy ~ 3.31%, China ~ 0.59%

fig1_df <- dat %>%
  group_by(Country) %>%
  summarise(
    Cases = sum(Cases),
    Deaths = sum(Deaths),
    CFR_crude_pct = 100 * Deaths / Cases
  ) %>%
  invisible()

ggplot(fig1_df, aes(x = Country, y = CFR_crude_pct, fill = Country)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", CFR_crude_pct)),
            vjust = -0.4, size = 4.2) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Crude fatality rate (all ages combined)",
       y = "Fatality rate (%)", x = NULL) +
  theme_classic(base_size = 13) +
  theme(legend.position = "none")
```

At face value, Italy looks worse. But what happens when we stratify by age group? Because age strongly affects COVID fatality, splitting cases into 10-year bands reveals a different picture. When we do, China’s fatality rate is higher in every band.

```{r}
#| label: fig-simpson-by-age-group
#| fig-cap: "Fatality rates by age group (same made-up numbers). Within each age band, China's CFR exceeds Italy's."

fig2_df <- dat %>%
  select(Country, Age, CFR_band_pct)

ggplot(fig2_df, aes(x = Age, y = CFR_band_pct, fill = Country)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  labs(title = "Within every age band, China’s fatality rate is higher",
       x = "Age band", y = "Fatality rate (%)", fill = NULL) +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")
```

Why the reversal? Italy’s cases skew older while China’s skew younger, and baseline risk rises steeply with age. The differing age distributions weight the overall averages in opposite ways. Aggregating without age adjustment hid the within-group pattern.

```{r}
#| label: fig-age-distributions
#| fig-cap: "Case age distribution by country. Italy’s cases are older on average; China’s are younger."

fig3_df <- dat %>%
  group_by(Country, Age) %>%
  summarise(Cases = sum(Cases), .groups = "drop") %>%
  group_by(Country) %>%
  mutate(Prop = Cases / sum(Cases))


ggplot(fig3_df, aes(x = Age, y = Prop, fill = Country, group=Country)) +
  geom_col(position = position_dodge(width = 0.4), width = 0.7) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~Country)+
  labs(title = "Case age mix: Italy skewed older, China skewed younger",
       x = "Age band", y = "Share of cases", fill = NULL) +
  theme_classic(base_size = 13) +
  theme(legend.position = "top", panel.border = element_rect(size = 1, fill=NA), axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

Statistics will not answer every scientific question, but it gives us a disciplined way to sort signal from bias so our inferences age well.

## Statistics in Environmental Science: Why All the Numbers?

Environmental science produces a *lot* of data. Every day satellites measure surface temperature, field stations log rainfall, and sensors track air quality. A single study can generate thousands of rows of numbers. Without statistical tools, these numbers would be overwhelming. With them, we can summarize, compare, and test ideas systematically.

Statistics matters here because environmental systems are messy. River flow changes by the hour, forests grow unevenly, and human actions layer on additional variability. Patterns are rarely obvious by eye. Statistics helps us sort signal from noise and ask: is this apparent change real, or just random variation?

You might think, “can’t a statistician just handle the math?” But knowing the basics yourself is essential for three reasons:

1.  **Design and analysis go together.** A good study starts long before you run a statistical test. If you want to study how fertilizer affects crop yields, your sampling plan and your analysis are inseparable. Poor design, like measuring only in unusually wet fields, can’t be rescued by sophisticated analysis.

2.  **Understanding the Science:** Scientific papers on climate change, biodiversity, or pollution are built on statistical results. To interpret them, you need to know what the numbers mean.

3.  **Practicality:** Hiring a specialist for every question isn’t realistic. A working knowledge of statistics makes you more self-sufficient as a scientist.

> *"We are drowning in information,\
> but we are starved for knowledge"*\
>
> -- Various authors, original probably John Naisbitt

Finally, statistics is not only for researchers. Weather forecasts, air quality alerts, and wildlife population trends are all communicated through numbers. A basic knowledge of statistics helps you tell whether those numbers support the claims being made—or whether someone is stretching the truth. In that sense, statistics is part of being an informed citizen in a data-saturated world.

## Introduction to measurements

Every dataset begins with measurement. Measurement means assigning numbers, labels, or categories to aspects of the world so they can be recorded and analyzed.

**Why measurement matters**

Environmental science often deals with broad ideas that need to be defined before they can be studied. For example, *soil health* or *forest change* are meaningful, but vague. To analyze them, we have to decide exactly what we mean and how to capture it. That process is called **operationalization**: turning a general concept into something measurable.

Before moving on, let’s clarify how operationalization fits with three related terms:

-   **Operationalization**: the step where you define exactly how a broad idea will be captured with a measure
-   **Measure**: the tool or method used to make observations
-   **Variable**: the actual values you record once the measure is applied. Variables are the actual data that end up in our dataset.

**Examples:**

-   Concept of interest: *Soil carbon sequestration*
    -   **Operationalization:** concentration of organic carbon in the top 30 cm
    -   **Measure:** laboratory analysis of soil samples
    -   **Variable:** recorded carbon concentration values for each plot
-   Concept of interestt: *Deforestation*
    -   **Operationalization**: change in forest cover between the current year and five years ago
    -   **Measure:** aerial image classification
    -   **Variable:** change in forest cover (%) per unit area

Operationalization is rarely straightforward, and there is no single correct way to do it. The best choice depends on the question you are asking and how the data will be used. In many fields, scientists have developed common practices, but every project still requires case-by-case judgment. Even so, some principles of good operationalization apply across studies: be precise about what you mean, how you will measure it, and what values are possible.

## Scales of measurement

As the previous section indicates, the outcome of a measurement is called a **variable**. Not all variables are the same type, and knowing the type matters because it determines which statistical tools make sense. The four classic scales are **nominal, ordinal, interval,** and **ratio**.

### Nominal scale

A **nominal scale** variable (also referred to as a **categorical** variable) is one in which is one in which the values are just names. They do not have an inherent order, and it makes no sense to average them.

*Example:* eye color. Eyes can be blue, green, or brown, but none is “greater” than another, and there’s no such thing as an “average eye color.”

### Ordinal scale

**Ordinal scale** variables have a meaningful order, but the spacing between values is not defined mathematically. You can rank the values, but you can’t assume equal steps between them, nor calculate a meaningful average.

*Example:* finishing position in a race. First place comes before second, and second before third, but you don’t know how much faster one runner was than another. Saying the “average place” of the group is 2.3 doesn’t mean anything.

### Interval scale

Interval variables have equal intervals between values, so differences are meaningful. However, zero is arbitrary, so multiplication and division are not valid.

*Example:* temperature in degrees Celsius. A difference of 3$^\circ$ means the same regardless of whether it is 7 → 10 or 15 → 18. But 0$^\circ$ does not mean “no temperature.” That zero is defined by the freezing point of water. So while you can say today is 3$^\circ$ warmer than yesterday, you cannot say 20$^\circ$ is “twice as hot” as 10$^\circ$.

(Alternate example: latitude. A difference of 10° latitude is meaningful, but zero latitude—the equator—doesn’t mean “no latitude.” Twice 45° is 90°, but that doesn’t mean one place has “twice as much latitude” as another.)In contrast to nominal and ordinal scale variables, **interval scale** and ratio scale variables are variables for which the numerical value is genuinely meaningful. In the case of interval scale variables, the *differences* between the numbers are interpretable, but the variable doesn't have a "natural" zero value. You can add and subtract, but ratios don’t make sense.

### Ratio scale

Ratio scale variables have all the properties of interval variables, plus a true zero. This makes multiplication and division valid, as well as addition and subtraction.

*Example:* age in years. Zero means no age at all, and someone who is 20 years old really is twice as old as someone who is 10. Differences (20 – 10 = 10 years) and ratios (20 ÷ 10 = 2) are both meaningful.

|          | can rank | can subtract/add | can multiply/divide | example          |
|--------------|:------------:|:------------:|:------------:|----------------|
| nominal  |          |                  |                     | eye color        |
| ordinal  |    x     |                  |                     | race position    |
| interval |    x     |        x         |                     | temperature (°C) |
| ratio    |    x     |        x         |          x          | age              |

### Continuous versus discrete variables

Another useful distinction is whether a variable can take on values in between others.

-   A **continuous variable** can, in principle, take on any value within a range. For example, consider height. If you are 72 inches tall and your friend Cameron is 71 inches tall, Alan could be 71.4 inches and David 71.49 inches. Because we can always imagine a new value in between two others, height is continuous.

-   A **discrete variable** is, in effect, a variable that isn't continuous. Discrete variables have separate, distinct values with nothing in between. Nominal variables are always discrete: there is no eye color that falls “between” green and blue in the same way that 71.4 falls between 71 and 72. Ordinal variables are also discrete: although second place falls between first and third, nothing can logically fall between first and second.

Interval and ratio variables can be either. Height (a ratio variable) is continuous. But the number of people living in a household (a ratio variable) is discrete: you cannot have 4.2 people. Temperature in degrees Celsius (an interval variable) is also continuous. But the year you started college (an interval variable) is discrete: there is no year between 2022 and 2023.

The table below shows how the scales of measurement relate to this distinction. Cells with an “x” mark what is possible.

|          | continuous | discrete |
|----------|:----------:|:--------:|
| nominal  |            |    x     |
| ordinal  |            |    x     |
| interval |     x      |    x     |
| ratio    |     x      |    x     |

### A note on real data

These categories are guides, not hard rules. For example, survey responses on a 1–5 “strongly disagree” to “strongly agree” scale are technically **ordinal**, but researchers often treat them as “quasi-interval” because the spacing is assumed to be roughly equal.

## Assessing the reliability of a measurement

When we measure something, two questions matter: *is the measurement consistent, and is it accurate?*

-   **Reliability** means consistency. If you measure the same thing again, do you get the same result?

-   **Validity** means accuracy. Does the measurement reflect the real thing you care about?

These aren't the same. A soil moisture probe that is miscalibrated might always read 5% too high. It is highly **reliable**, you get the same number each time, but not **valid**, since it doesn’t reflect true soil moisture. On the other hand, a set of volunteer bird counts might fluctuate from one observer to the next (**low reliability**), but when averaged across many counts, the results may still approximate true abundance (**reasonable validity**).

In practice, a measure that is very unreliable usually ends up being invalid too, because we can’t tell which result is right. This is why reliability is often considered a prerequisite for validity.

-   Reliability can show up in different ways:

    -   **Over time** (*test–retest*): if you sample soil moisture today and tomorrow under the same conditions, do you get the same value?

    -   **Across people** (*inter-rater*): if two field crews count birds at the same site, do their tallies agree?

    -   **Across tools** (*parallel forms*): if you use two different rain gauges in the same spot, do they record the same rainfall?

    -   **Within a test** (*internal consistency*): if multiple survey questions are meant to capture the same attitude, do they give similar answers?

Not every measurement needs (or can even possibly have) every form of reliability. The point is that **reliability is necessary but not sufficient for validity**: a method can be consistently wrong, but if it isn’t consistent at all, it’s almost impossible to know whether it’s right.

## The role of variables: predictors and outcomes

One last piece of terminology before we leave variables behind. In most studies we have many variables, but when we analyze them, we usually split them into two roles: the **thing we’re trying to explain** and the **thing doing the explaining**. To keep it straight, we use $Y$ for the variable being explained, a $X_1$, $X_2$, etc. for the variables used to explain it.

Traditionally, $X$ is called the **indepdendent variable (IV)** and $Y$ is the **dependent variable (DV).** The logic is that if there's a relationship, $Y$ depends on $X$. These terms can be clunky and can be confusing: (a) IVs are rarely actually "independent of everything else" and (b) if there's no relationship, then the DV doesn't actually "depend" on the IV at all.

Alternative terminology is often clearer. In experiments, IVs are **manipulations**, and DVs are **measurements:**

| role of the variable         | classical name            | alternative  |
|:-----------------------------|:--------------------------|:-------------|
| "to be explained" ($Y$)      | dependent variable (DV)   | Measurement  |
| "to do the explaining" ($X$) | independent variable (IV) | Manipulation |

: Another useful pair is **predictors** and **outcomes**. The idea here is that what we use $X$ to make predict or guess about $Y$:

| role of the variable         | classical name            | alternative |
|:-----------------------------|:--------------------------|:------------|
| "to be explained" ($Y$)      | dependent variable (DV)   | outcome     |
| "to do the explaining" ($X$) | independent variable (IV) | predictor   |

## Experimental and non-experimental research

A big distinction in research is between experimental and non-experimental studies. What matters here is the degree of control the researcher has.

The key features of **experimental research** is that the researcher controls as many aspects of the study as possible. The researcher deliberately manipulates something (the predictor/IV) and measures its effect on outcomes (outcome/DV). The goal is to isolate causal effects. To avoid the problem of “something else” influencing the outcome, researchers hold other factors constant. In practice, it's almost impossible to *think* of everything else that might have an influence on the outcome of an experiment, much less keep it constant. The standard solution to this is **randomization** (explain). Randomization doesn’t eliminate confounds, but it makes them less likely to systematically bias results.

For instance, suppose we wanted to know if smoking causes lung cancer. Observing smokers and non-smokers can only get us so far, because those groups differ in many ways besides smoking, like occupation, income, or diet. A true experiment would require randomly assigning people to smoke or not — you can see how that would be deeply unethical. The same problem comes up in medicine: we know surprisingly little about how certain drugs or exposures affect pregnant people, precisely because we cannot ethically assign them to risky conditions.

This is where **non-experimental research** comes in. In environmental science, the limitation isn’t usually ethics but feasibility: there’s no “control planet” we can assign as a baseline. Instead, researchers rely on quasi-experiments, time series, or detailed case studies to tease out effects in complex, uncontrolled systems.

**Non-experimental research** in environmental science often encompasses studies in which researchers cannot exert full control over variables as they might in a typical experimental setup. In environmental science, this issue is particularly pronounced, as there is no "control planet" to allow for perfect experimental conditions. This unique challenge leads to a reliance on alternative approaches, such as time series analysis, to understand complex environmental phenomena.

One distinction worth making between two types of non-experimental research is the difference between **quasi-experimental research** and **case studies**. Much of the research in environmental science fits into a quasi-experimental design. Here, researchers might wish to study the effects of industrial pollution on a river system but cannot directly control all the variables, such as the quantity or type of pollutants being emitted by various industries. Instead, they must observe existing conditions and make careful comparisons, possibly using statistical tools to account for confounding variables.

**Time series analysis** becomes vital in this context, as many environmental processes unfold over extended periods. For example, tracking changes in global temperatures or sea level requires years of consistent data collection. The complexity and variability inherent in environmental systems make drawing definitive conclusions difficult, but sophisticated statistical methods can help isolate specific effects.

Environmental science also frequently relies on detailed **case studies**. These investigations focus on particular events or locations to provide in-depth insights into environmental processes and their impacts on ecosystems, human health, and societal structures. Case studies might explore the aftermath of a natural disaster, the unique ecology of a threatened habitat, or the social and economic consequences of environmental policies.

Though they often lack the broad generalizability of more controlled experimental research, case studies in environmental science offer valuable opportunities to understand the complexity of real-world situations. They can uncover nuances and subtleties that might be overlooked in a broader experimental or quasi-experimental approach.

## Assessing the validity of a study

More than any other thing, a scientist wants their research to be "valid". The conceptual idea behind **validity** is very simple: can you trust the results of your study? If not, the study is invalid. However, while it's easy to state, in practice it's much harder to check validity than it is to check reliability. And in all honesty, there's no precise, clearly agreed upon notion of what validity actually is. In fact, there's lots of different kinds of validity, each of which raises it's own issues, and not all forms of validity are relevant to all studies. I'm going to talk about five different types:

-   Internal validity

-   External validity

-   Construct validity

-   Face validity

-   Ecological validity

To give you a quick guide as to what matters here...(1) Internal and external validity are the most important, since they tie directly to the fundamental question of whether your study really works. (2) Construct validity asks whether you're measuring what you think you are. (3) Face validity isn't terribly important except insofar as you care about "appearances". (4) Ecological validity is a special case of face validity that corresponds to a kind of appearance that you might care about a lot.

### Internal validity

**Internal validity** refers to the extent to which you are able draw the correct conclusions about the causal relationships between variables. It's called "internal" because it refers to the relationships between things "inside" the study. Let's illustrate the concept with a simple example. Suppose you're interested in finding out whether a university education makes you write better. To do so, you get a group of first year students, ask them to write a 1000 word essay, and count the number of spelling and grammatical errors they make. Then you find some third-year students, who obviously have had more of a university education than the first-years, and repeat the exercise. And let's suppose it turns out that the third-year students produce fewer errors. And so you conclude that a university education improves writing skills. Right? Except... the big problem that you have with this experiment is that the third-year students are older, and they've had more experience with writing things. So it's hard to know for sure what the causal relationship is: Do older people write better? Or people who have had more writing experience? Or people who have had more education? Which of the above is the true *cause* of the superior performance of the third-years? Age? Experience? Education? You can't tell. This is an example of a failure of internal validity, because your study doesn't properly tease apart the *causal* relationships between the different variables.

### External validity

**External validity** relates to the **generalizability** of your findings. That is, to what extent do you expect to see the same pattern of results in "real life" as you saw in your study. To put it a bit more precisely, any study that you do in psychology will involve a fairly specific set of questions or tasks, will occur in a specific environment, and will involve participants that are drawn from a particular subgroup. So, if it turns out that the results don't actually generalize to people and situations beyond the ones that you studied, then what you've got is a lack of external validity.

The classic example of this issue is the fact that a very large proportion of studies in psychology will use undergraduate psychology students as the participants. Obviously, however, the researchers don't care *only* about psychology students; they care about people in general. Given that, a study that uses only psych students as participants always carries a risk of lacking external validity. That is, if there's something "special" about psychology students that makes them different to the general populace in some *relevant* respect, then we may start worrying about a lack of external validity.

That said, it is absolutely critical to realize that a study that uses only psychology students does not necessarily have a problem with external validity. I'll talk about this again later, but it's such a common mistake that I'm going to mention it here. The external validity is threatened by the choice of population if (a) the population from which you sample your participants is very narrow (e.g., psych students), and (b) the narrow population that you sampled from is systematically different from the general population, *in some respect that is relevant to the psychological phenomenon that you intend to study*. The italicized part is the bit that lots of people forget: it is true that psychology undergraduates differ from the general population in lots of ways, and so a study that uses only psych students *may* have problems with external validity. However, if those differences aren't very relevant to the phenomenon that you're studying, then there's nothing to worry about. To make this a bit more concrete, here's two extreme examples:

-   You want to measure "attitudes of the general public towards psychotherapy", but all of your participants are psychology students. This study would almost certainly have a problem with external validity.

-   You want to measure the effectiveness of a visual illusion, and your participants are all psychology students. This study is very unlikely to have a problem with external validity

Having just spent the last couple of paragraphs focusing on the choice of participants (since that's the big issue that everyone tends to worry most about), it's worth remembering that external validity is a broader concept. The following are also examples of things that might pose a threat to external validity, depending on what kind of study you're doing:

-   People might answer a "psychology questionnaire" in a manner that doesn't reflect what they would do in real life.

-   Your lab experiment on (say) "human learning" has a different structure to the learning problems people face in real life.

### Construct validity

**Construct validity** is basically a question of whether you're measuring what you want to be measuring. A measurement has good construct validity if it is actually measuring the correct theoretical construct, and bad construct validity if it doesn't. To give very simple (if ridiculous) example, suppose I'm trying to investigate the rates with which university students cheat on their exams. And the way I attempt to measure it is by asking the cheating students to stand up in the lecture theater so that I can count them. When I do this with a class of 300 students, 0 people claim to be cheaters. So I therefore conclude that the proportion of cheaters in my class is 0%. Clearly this is a bit ridiculous. But the point here is not that this is a very deep methodological example, but rather to explain what construct validity is. The problem with my measure is that while I'm *trying* to measure "the proportion of people who cheat" what I'm actually measuring is "the proportion of people stupid enough to own up to cheating, or bloody minded enough to pretend that they do". Obviously, these aren't the same thing! So my study has gone wrong, because my measurement has very poor construct validity.

### Ecological validity

**Ecological validity** is a different notion of validity, which is similar to external validity, but less important. The idea is that, in order to be ecologically valid, the entire set up of the study should closely approximate the real world scenario that is being investigated. In a sense, ecological validity is a kind of face validity -- it relates mostly to whether the study "looks" right, but with a bit more rigor to it. To be ecologically valid, the study has to look right in a fairly specific way. The idea behind it is the intuition that a study that is ecologically valid is more likely to be externally valid. It's no guarantee, of course. But the nice thing about ecological validity is that it's much easier to check whether a study is ecologically valid than it is to check whether a study is externally valid. An simple example would be eyewitness identification studies. Most of these studies tend to be done in a university setting, often with fairly simple array of faces to look at rather than a line up. The length of time between seeing the "criminal" and being asked to identify the suspect in the "line up" is usually shorter. The "crime" isn't real, so there's no chance that the witness being scared, and there's no police officers present, so there's not as much chance of feeling pressured. These things all mean that the study *definitely* lacks ecological validity. They might (but might not) mean that it also lacks external validity.

## Confounds, artifacts and other threats to validity

If we look at the issue of validity in the most general fashion, the two biggest worries that we have are *confounds* and *artifact*. These two terms are defined in the following way:

-   **Confound**: A confound is an additional, often unmeasured variable that turns out to be related to both the predictors and the outcomes. The existence of confounds threatens the internal validity of the study because you can't tell whether the predictor causes the outcome, or if the confounding variable causes it, etc.

-   **Artifact**: A result is said to be "artifactual" if it only holds in the special situation that you happened to test in your study. The possibility that your result is an artifact describes a threat to your external validity, because it raises the possibility that you can't generalize your results to the actual population that you care about.

As a general rule confounds are a bigger concern for non-experimental studies, precisely because they're not proper experiments: by definition, you're leaving lots of things uncontrolled, so there's a lot of scope for confounds working their way into your study. Experimental research tends to be much less vulnerable to confounds: the more control you have over what happens during the study, the more you can prevent confounds from appearing.

However, there's always swings and roundabouts, and when we start thinking about artifacts rather than confounds, the shoe is very firmly on the other foot. For the most part, artifactual results tend to be a concern for experimental studies than for non-experimental studies. To see this, it helps to realize that the reason that a lot of studies are non-experimental is precisely because what the researcher is trying to do is examine human behavior in a more naturalistic context. By working in a more real-world context, you lose experimental control (making yourself vulnerable to confounds) but because you tend to be studying human psychology "in the wild" you reduce the chances of getting an artifactual result. Or, to put it another way, when you take psychology out of the wild and bring it into the lab (which we usually have to do to gain our experimental control), you always run the risk of accidentally studying something different than you wanted to study: which is more or less the definition of an artifact.

Be warned though: the above is a rough guide only. It's absolutely possible to have confounds in an experiment, and to get artifactual results with non-experimental studies. This can happen for all sorts of reasons, not least of which is researcher error. In practice, it's really hard to think everything through ahead of time, and even very good researchers make mistakes. But other times it's unavoidable, simply because the researcher has ethics (e.g., see "differential attrition").

Okay. There's a sense in which almost any threat to validity can be characterized as a confound or an artifact: they're pretty vague concepts. So let's have a look at some of the most common examples...

### History effects

**History effects** refer to the possibility that specific events may occur during the study itself that might influence the outcomes. For instance, something might happen in between a pre-test and a post-test. Or, in between testing participant 23 and participant 24. Alternatively, it might be that you're looking at an older study, which was perfectly valid for its time, but the world has changed enough since then that the conclusions are no longer trustworthy. Examples of things that would count as history effects:

-   You're interested in how people think about risk and uncertainty. You started your data collection in December 2010. But finding participants and collecting data takes time, so you're still finding new people in February 2011. Unfortunately for you (and even more unfortunately for others), the Queensland floods occurred in January 2011, causing billions of dollars of damage and killing many people. Not surprisingly, the people tested in February 2011 express quite different beliefs about handling risk than the people tested in December 2010. Which (if any) of these reflects the "true" beliefs of participants? I think the answer is probably both: the Queensland floods genuinely changed the beliefs of the Australian public, though possibly only temporarily. The key thing here is that the "history" of the people tested in February is quite different to people tested in December.

-   You're testing the psychological effects of a new anti-anxiety drug. So what you do is measure anxiety before administering the drug (e.g., by self-report, and taking physiological measures, let's say), then you administer the drug, and then you take the same measures afterwards. In the middle, however, because your labs are in Los Angeles, there's an earthquake, which increases the anxiety of the participants.

### Maturation effects

As with history effects, **maturational effects** are fundamentally about change over time. However, maturation effects aren't in response to specific events. Rather, they relate to how people change on their own over time: we get older, we get tired, we get bored, etc. Some examples of maturation effects:

-   When doing developmental psychology research, you need to be aware that children grow up quite rapidly. So, suppose that you want to find out whether some educational trick helps with vocabulary size among 3 year olds. One thing that you need to be aware of is that the vocabulary size of children that age is growing at an incredible rate (multiple words per day), all on its own. If you design your study without taking this maturational effect into account, then you won't be able to tell if your educational trick works.

-   When running a very long experiment in the lab (say, something that goes for 3 hours), it's very likely that people will begin to get bored and tired, and that this maturational effect will cause performance to decline, regardless of anything else going on in the experiment

### Repeated testing effects

An important type of history effect is the effect of **repeated testing**. Suppose I want to take two measurements of some psychological construct (e.g., anxiety). One thing I might be worried about is if the first measurement has an effect on the second measurement. In other words, this is a history effect in which the "event" that influences the second measurement is the first measurement itself! This is not at all uncommon. Examples of this include:

-   *Learning and practice*: e.g., "intelligence" at time 2 might appear to go up relative to time 1 because participants learned the general rules of how to solve "intelligence-test-style" questions during the first testing session.

-   *Familiarity with the testing situation*: e.g., if people are nervous at time 1, this might make performance go down; after sitting through the first testing situation, they might calm down a lot precisely because they've seen what the testing looks like.

-   *Auxiliary changes caused by testing*: e.g., if a questionnaire assessing mood is boring, then mood at measurement at time 2 is more likely to become "bored", precisely because of the boring measurement made at time 1.

### Selection bias

**Selection bias** is a pretty broad term. Suppose that you're running an experiment with two groups of participants, where each group gets a different "treatment", and you want to see if the different treatments lead to different outcomes. However, suppose that, despite your best efforts, you've ended up with a gender imbalance across groups (say, group A has 80% females and group B has 50% females). It might sound like this could never happen, but trust me, it can. This is an example of a selection bias, in which the people "selected into" the two groups have different characteristics. If any of those characteristics turns out to be relevant (say, your treatment works better on females than males) then you're in a lot of trouble.

### Differential attrition

within the study itself, impacting both the internal and external validity of the research. This phenomenon can occur in environmental studies, where participants' commitment to certain practices might change over time, such as in multi-year studies on cover cropping.

When thinking about the effects of differential attrition, it is sometimes helpful to distinguish between two different types. The first is **homogeneous attrition**, in which the attrition effect is the same for all groups, treatments or conditions. Homogeneous attrition occurs when the dropout rate is roughly the same across different groups or conditions within the study. For instance, if many farmers participating in a cover cropping study discontinue the practice at similar rates, the sample may become unrepresentative. Though the internal validity (accuracy of the study's conclusions) might remain intact, the external validity (generalizability to the broader population) may suffer. In the realm of environmental science, this could mean the study's findings are less applicable to the broader community of farmers or agricultural systems.

The second type of differential attrition is **heterogeneous attrition**, in which the attrition effect is different for different groups. This is a much bigger problem: not only do you have to worry about your external validity, you also have to worry about your internal validity too. In a study on cover cropping, if certain groups, perhaps motivated by specific incentives or pressures, drop out at different rates, this can introduce a confounding variable. The comparison between the remaining participants may no longer be valid, as the groups have become fundamentally different due to the non-random dropout. This not only affects the generalizability of the findings but also the internal validity, as the study's conclusions within the sample itself may be compromised.

### Non-response bias

**Non-response bias** is closely related to selection bias, and to differential attrition. The simplest version of the problem goes like this. You mail out a survey to 1000 people, and only 300 of them reply. The 300 people who replied are almost certainly not a random sub-sample. People who respond to surveys are systematically different to people who don't. This introduces a problem when trying to generalize from those 300 people who replied, to the population at large; since you now have a very non-random sample. The issue of non-response bias is more general than this, though. Among the (say) 300 people that did respond to the survey, you might find that not everyone answers every question. If (say) 80 people chose not to answer one of your questions, does this introduce problems? As always, the answer is maybe. If the question that wasn't answered was on the last page of the questionnaire, and those 80 surveys were returned with the last page missing, there's a good chance that the missing data isn't a big deal: probably the pages just fell off. However, if the question that 80 people didn't answer was the most confrontational or invasive personal question in the questionnaire, then almost certainly you've got a problem. In essence, what you're dealing with here is what's called the problem of **missing data**. If the data that is missing was "lost" randomly, then it's not a big problem. If it's missing systematically, then it can be a big problem.

### Regression to the mean

**Regression to the mean** is a curious variation on selection bias. It refers to any situation where you select data based on an extreme value on some measure. Because the measure has natural variation, it almost certainly means that when you take a subsequent measurement, that later measurement will be less extreme than the first one, purely by chance.

Here's an example. Suppose we are interested in examining the effects of specific climatic conditions on the growth of a particular plant species. We identify the 20 locations with the most significant growth during a year of favorable weather conditions, such as ideal rainfall and temperature, and decide to study these areas further.

After another year, we observe that the plant growth in these 20 locations is still above average but not as exceptional as during the first year. The immediate reaction might be to conclude that a change in environmental practices or some other factor has adversely affected growth in these areas.

However, this might simply be an example of regression to the mean. Consider what is required for remarkable plant growth: optimal soil, hard work in maintenance, favorable weather conditions, and perhaps a bit of luck with other uncontrollable environmental factors.

While the soil quality and maintenance practices may be consistent from one year to the next, luck with weather is not. Certain areas that were fortunate with ideal weather conditions in the first year may not experience the same fortune in the following year. This inconsistency in luck leads to a less extreme measurement the second time.

Regression to the mean is surprisingly common. For instance, if two very tall people have kids, their children will tend to be taller than average, but not as tall as the parents. The reverse happens with very short parents: two very short parents will tend to have short children, but nevertheless those kids will tend to be taller than the parents. It can also be extremely subtle. For instance, there have been studies done that suggested that people learn better from negative feedback than from positive feedback. However, the way that people tried to show this was to give people positive reinforcement whenever they did good, and negative reinforcement when they did bad. And what you see is that after the positive reinforcement, people tended to do worse; but after the negative reinforcement they tended to do better. But! Notice that there's a selection bias here: when people do very well, you're selecting for "high" values, and so you should *expect* (because of regression to the mean) that performance on the next trial should be worse, regardless of whether reinforcement is given. Similarly, after a bad trial, people will tend to improve all on their own. The apparent superiority of negative feedback is an artifact caused by regression to the mean [@Kahneman1973]

### Experimenter bias

**Experimenter bias** can occur when the researcher, despite conscientious efforts, inadvertently influences the results of a study. This may not involve communication with human participants as in other fields but could still manifest in several ways within environmental research.

In environmental science, the experimenter might have **expectations or preconceived notions about what the data should reveal**. For instance, a researcher studying the effects of a certain agricultural practice on soil health may unconsciously select or emphasize data that align with their expectations or theoretical commitments. This can lead to a bias in data collection or interpretation, subtly steering the results toward the expected outcome.

Experimenter bias can also arise from **methodological choices**. A researcher may design an experiment, select sites for observation, or choose measurement techniques in a way that favors a particular outcome. For example, if studying the effect of pollution on a water source, the researcher might unintentionally choose sampling times or locations that are more likely to confirm their hypotheses, overlooking variations that could provide a more comprehensive and unbiased view.

In some cases, the researcher's bias may inadvertently **influence other field or laboratory personnel** involved in the study. A team member who understands the desired outcome of the study may unconsciously alter procedures, handling of samples, or recording of data to align with the researcher's expectations.

The classic example of experimenter bias is the case study of "Clever Hans", which dates back to 1907, [@Pfungst1911; @Hothersall2004]. Clever Hans was a horse that apparently was able to read and count, and perform other human like feats of intelligence. After Clever Hans became famous, psychologists started examining his behavior more closely. It turned out that -- not surprisingly -- Hans didn't know how to do maths. Rather, Hans was responding to the human observers around him. Because they did know how to count, and the horse had learned to change its behavior when people changed theirs.

The general solution to the problem of experimenter bias is to engage in double blind studies, where neither the experimenter nor the participant knows which condition the participant is in, or knows what the desired behavior is. This provides a very good solution to the problem, but it's important to recognize that it's not quite ideal, and typically not feasible in Environmental Science. Measures to mitigate this bias include rigorous peer review, careful documentation of methods, and maintaining an awareness of potential biases.

### Reactivity and Demand Effects

Even in environmental studies, the knowledge that a process or phenomenon is being observed can impact the behavior of the researchers or those involved in data collection. The basic idea is captured by the Hawthorne effect: people alter their performance because of the attention that the study focuses on them. The effect takes its name from a the "Hawthorne Works" factory outside of Chicago [@Adair1984]. A study done in the 1920s looking at the effects of lighting on worker productivity at the factory turned out to be an effect of the fact that the workers knew they were being studied, rather than the lighting. This so-called "Hawthorne effect" might influence how a field team conducts measurements if they know the specific goal of the study.

### Placebo effects

While typically associated with clinical trials, the **placebo effect** can have an analogous impact in environmental research. For instance, a belief in the efficacy of a certain conservation practice might lead to subjective reporting or unintentional data manipulation.

### Situation, measurement and subpopulation effects

In some respects, these terms are a catch-all term for "all other threats to external validity". The choice of location, time, measurement tools, and even who collects the data can all influence the results. Ensuring robust methodologies and considering the potential influences of these factors can help in achieving results that generalize more widely.

### Fraud, deception and self-deception

> *It is difficult to get a man to understand something, when his salary depends on his not understanding it.*\
>
> -- Upton Sinclair

One final thing that I feel like I should mention. While reading what the textbooks often have to say about assessing the validity of the study, I couldn't help but notice that they seem to make the assumption that the researcher is honest. I find this hilarious. While the vast majority of scientists are honest, in my experience at least, some are not. Not only that, as I mentioned earlier, scientists are not immune to belief bias -- it's easy for a researcher to end up deceiving themselves into believing the wrong thing, and this can lead them to conduct subtly flawed research, and then hide those flaws when they write it up. So you need to consider not only the (probably unlikely) possibility of outright fraud, but also the (probably quite common) possibility that the research is unintentionally "slanted". I opened a few standard textbooks and didn't find much of a discussion of this problem, so here's my own attempt to list a few ways in which these issues can arise are:

-   **Data fabrication**. Sometimes, people just make up the data. This is occasionally done with "good" intentions. For instance, the researcher believes that the fabricated data do reflect the truth, and may actually reflect "slightly cleaned up" versions of actual data. On other occasions, the fraud is deliberate and malicious. Some high-profile examples where data fabrication has been alleged or shown include Cyril Burt (a psychologist who is thought to have fabricated some of his data), Andrew Wakefield (who has been accused of fabricating his data connecting the MMR vaccine to autism) and Hwang Woo-suk (who falsified a lot of his data on stem cell research).

-   **Hoaxes**. Hoaxes share a lot of similarities with data fabrication, but they differ in the intended purpose. A hoax is often a joke, and many of them are intended to be (eventually) discovered. Often, the point of a hoax is to discredit someone or some field. There's quite a few well known scientific hoaxes that have occurred over the years (e.g., Piltdown man) some of were deliberate attempts to discredit particular fields of research (e.g., the Sokal affair).

-   **Data misrepresentation**. While fraud gets most of the headlines, it's much more common in my experience to see data being misrepresented. When I say this, I'm not referring to newspapers getting it wrong (which they do, almost always). I'm referring to the fact that often, the data don't actually say what the researchers think they say. My guess is that, almost always, this isn't the result of deliberate dishonesty, it's due to a lack of sophistication in the data analyses. For instance, think back to the example of Simpson's paradox that I discussed in the beginning of these notes. It's very common to see people present "aggregated" data of some kind; and sometimes, when you dig deeper and find the raw data yourself, you find that the aggregated data tell a different story to the disaggregated data. Alternatively, you might find that some aspect of the data is being hidden, because it tells an inconvenient story (e.g., the researcher might choose not to refer to a particular variable). There's a lot of variants on this; many of which are very hard to detect.

-   **Study "misdesign"**. Okay, this one is subtle. Basically, the issue here is that a researcher designs a study that has built-in flaws, and those flaws are never reported in the paper. The data that are reported are completely real, and are correctly analysed, but they are produced by a study that is actually quite wrongly put together. The researcher really wants to find a particular effect, and so the study is set up in such a way as to make it "easy" to (artifactually) observe that effect. One sneaky way to do this -- in case you're feeling like dabbling in a bit of fraud yourself -- is to design an experiment in which it's obvious to the participants what they're "supposed" to be doing, and then let reactivity work its magic for you. If you want, you can add all the trappings of double blind experimentation etc. It won't make a difference, since the study materials themselves are subtly telling people what you want them to do. When you write up the results, the fraud won't be obvious to the reader: what's obvious to the participant when they're in the experimental context isn't always obvious to the person reading the paper. Of course, the way I've described this makes it sound like it's always fraud: probably there are cases where this is done deliberately, but in my experience the bigger concern has been with unintentional misdesign. The researcher *believes* ...and so the study just happens to end up with a built in flaw, and that flaw then magically erases itself when the study is written up for publication.

-   **Data mining & post hoc hypothesizing**. Another way in which the authors of a study can more or less lie about what they found is by engaging in what's referred to as "data mining". As we'll discuss later in the class, if you keep trying to analyse your data in lots of different ways, you'll eventually find something that "looks" like a real effect but isn't. This is referred to as "data mining". It used to be quite rare because data analysis used to take weeks, but now that everyone has very powerful statistical software on their computers, it's becoming very common. Data mining per se isn't "wrong", but the more that you do it, the bigger the risk you're taking. The thing that is wrong, and I suspect is very common, is *unacknowledged* data mining. That is, the researcher run every possible analysis known to humanity, finds the one that works, and then pretends that this was the only analysis that they ever conducted. Worse yet, they often "invent" a hypothesis after looking at the data, to cover up the data mining. To be clear: it's not wrong to change your beliefs after looking at the data, and to reanalyze your data using your new "post hoc" hypotheses. What is wrong (and, I suspect, common) is failing to acknowledge that you've done so. If you acknowledge that you did it, then other researchers are able to take your behavior into account. If you don't, then they can't. And that makes your behavior deceptive. Bad!

-   **Publication bias & self-censoring**. Finally, a pervasive bias is "non-reporting" of negative results. This is almost impossible to prevent. Journals don't publish every article that is submitted to them: they prefer to publish articles that find "something". So, if 20 people run an experiment looking at whether reading *Finnegans Wake* causes insanity in humans, and 19 of them find that it doesn't, which one do you think is going to get published? Obviously, it's the one study that did find that *Finnegans Wake* causes insanity. This is an example of a *publication bias*: since no-one ever published the 19 studies that didn't find an effect, a naive reader would never know that they existed. Worse yet, most researchers "internalize" this bias, and end up *self-censoring* their research. Knowing that negative results aren't going to be accepted for publication, they never even try to report them. As a friend of mine says "for every experiment that you get published, you also have 10 failures". And she's right. The catch is, while some (maybe most) of those studies are failures for boring reasons (e.g. you stuffed something up) others might be genuine "null" results that you ought to acknowledge when you write up the "good" experiment. And telling which is which is often hard to do. A good place to start is a paper by @Ioannidis2005 with the depressing title "Why most published research findings are false". I'd also suggest taking a look at work by @Kuhberger2014 presenting statistical evidence that this actually happens in psychology.

There's probably a lot more issues like this to think about, but that'll do to start with. What I really want to point out is the blindingly obvious truth that real world science is conducted by actual humans, and only the most gullible of people automatically assumes that everyone else is honest and impartial. Actual scientists aren't usually *that* naive, but for some reason the world likes to pretend that we are, and the textbooks we usually write seem to reinforce that stereotype.

## Summary

In this chapter, we have explored essential aspects of research methodology pertinent to environmental statistics:

-   **Operationalization and Measurement**: Understanding how to define and measure theoretical constructs and recognizing the distinctions between variables are foundational to research.

-   **Scales of measurement and types of variables**: This section addressed the differences between discrete and continuous data, and the various scale types including nominal, ordinal, interval, and ratio scales.

-   **Reliability of a measurement**: If I measure the "same"" thing twice, should I expect to see the same result? Only if my measure is reliable. But what does it mean to talk about doing the "same" thing? Well, that's why we have different types of reliability. Make sure you remember what they are.

-   **Terminology: predictors and outcomes**: What roles do variables play in an analysis? Can you remember the difference between predictors and outcomes? Dependent and independent variables? Etc.

-   **Experimental and non-experimental research designs**: Clarification of the roles that variables play in an analysis, including the distinctions between predictors and outcomes, and dependent and independent variables.

-   **Research Designs**: Highlighting what constitutes an experiment within environmental research, the chapter delves into the differentiation between experimental and non-experimental research designs.

-   **Validity and Threats**: A comprehensive look at whether a study measures what it aims to, understanding potential pitfalls, and recognizing the myriad ways things can go awry.

Study design is a paramount component of research methodology, with numerous textbooks and resources available to further explore these concepts. Drawing on established texts like @Campbell1963, this chapter serves as an introduction and guide to these key topics, recognizing the intrinsic connection between statistics and study design within the field of environmental science.

## Videos

### Terms of Statistics

<iframe width="560" height="315" src="https://www.youtube.com/embed/HgtnpOtk350" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen>

</iframe>
