---
author: Mallory L. Barnes
abstract-title: Chapter notes
abstract: Adapted to environmental science by Mallory Barnes. Portions adapted nearly verbatim from Chapters 1 and 2 in Navarro, D. J. "Learning Statistics with R." <https://compcogscisydney.org/learning-statistics-with-r/>
aliases: [why-statistics.html]
---

```{r, include = FALSE}
source("global_stuff.R")
```

# Why Statistics?

> To call in statisticians after the experiment is done may be no more than asking them to perform a post-mortem examination: They may be able to say what the experiment died of. ---Sir Ronald Fisher

## On the Importance of Statistics in Environmental Science

Many students come into statistics classes with a mix of nerves and low expectations. It is not always the most eagerly anticipated part of an environmental science degree. But statistics is one of the most important tools we have for understanding complex systems. Data are noisy, ecosystems are variable, and our own judgments are often biased. Without statistical methods, we would be left with anecdotes and guesses. With them, we can detect real patterns, test competing explanations, and make stronger arguments for environmental decisions.\
\
Common sense is useful, but it is prone to bias. If you see sick fish downstream of a factory, you might conclude that the factory caused it. Maybe you are right, but maybe the fish were already stressed by temperature, parasites, or another pollutant. Statistics gives us tools to test these ideas systematically. It helps us separate patterns from noise, avoid being misled by our own expectations, and draw conclusions that are more likely to hold up under scrutiny.

Environmental science is especially messy. Ecosystems shift, climate varies, and data are rarely clean. Without statistical tools we risk telling convenient stories instead of uncovering real patterns. With them we can keep ourselves honest, which is why statistics is essential for environmental science.

### Why we need Statistics: Bias and Simpson's Paradox

Human judgment is powerful but far from perfect. When we already believe something to be true, we are more likely to interpret new evidence as supporting it, even if it does not. This tendency can lead us astray in environmental science. For example, if you are convinced that industrial farming is the main driver of bee declines, you might see every new drop in bee abundance as confirmation of that belief, even when disease or weather may be stronger contributors. Statistics gives us a systematic way to test these competing explanations, rather than relying on intuition that can be biased.

Think of it this way: relying only on our intuition is a bit like trying to identify trends in a messy field notebook by eye. We might be right more often than chance, but our judgments are still shaped by what we expect to see. With statistics, we move from a “better than guessing” success rate to a much higher level of reliability. It is not magic, but it dramatically increases the chance that we draw the correct conclusion. That is reason number one why scientists use statistics. It helps us check our assumptions and keep our conclusions grounded in evidence rather than belief.

## Simpson's paradox

Another challenge is that data can tell very different stories depending on how they are aggregated. A phenomenon known as *Simpson’s paradox* occurs when trends in combined data reverse once you separate the data into groups.

Here's a recent example from the early COVID-19 era. During the COVID-19 pandemic, early data showed that Italy’s overall case fatality rate was higher than China’s. Once researchers *disaggregated by age group,* the trend flipped: within every age group, the fatality rate was actually higher in China. The apparent contradiction was explained by differences in the age distribution of cases [@vonkügelgen2021] .

```{r}
#| label: fig-simpson-covid-overall
#| fig-cap: "Crude case fatality rates (made-up numbers). Italy appears to have a higher fatality rate than China when all cases are combined."
library(ggplot2)
library(tibble)

age_levels <- c("0–9","10–19","20–29","30–39","40–49","50–59","60–69","70–79","80+")

dat <- tibble::tribble(
  ~Country, ~Age,     ~Cases, ~Deaths,  ~CFR_band_pct,
  "Italy",  "0–9",       500,   0.05,     0.01,
  "Italy",  "10–19",     500,   0.10,     0.02,
  "Italy",  "20–29",     800,   0.32,     0.04,
  "Italy",  "30–39",    1000,   1.50,     0.15,
  "Italy",  "40–49",    2000,   8.00,     0.40,
  "Italy",  "50–59",    2000,  20.00,     1.00,
  "Italy",  "60–69",    1500,  37.50,     2.50,
  "Italy",  "70–79",    1200,  72.00,     6.00,
  "Italy",  "80+",      1500, 225.00,    15.00,

  "China",  "0–9",      8000,   1.60,     0.02,
  "China",  "10–19",    8000,   2.40,     0.03,
  "China",  "20–29",    8000,   4.00,     0.05,
  "China",  "30–39",    7000,  14.00,     0.20,
  "China",  "40–49",    7000,  42.00,     0.60,
  "China",  "50–59",    3000,  45.00,     1.50,
  "China",  "60–69",    1000,  35.00,     3.50,
  "China",  "70–79",     700,  56.00,     8.00,
  "China",  "80+",       300,  54.00,    18.00
) %>%
  mutate(
    Age = factor(Age, levels = age_levels),
    CFR_band_calc = 100 * Deaths / Cases  # sanity check, should match CFR_band_pct
  )

# Quick checks
dat %>% group_by(Country) %>% summarise(Total_cases = sum(Cases), Total_deaths = sum(Deaths),
                                        CFR_crude_pct = 100*Total_deaths/Total_cases)
# Italy ~ 3.31%, China ~ 0.59%

# ------------------------------------------------------------------------------
# 2) Figure 1: crude (overall) fatality rate by country
#    Knit this, then add a short paragraph before the next chunk.
# ------------------------------------------------------------------------------

fig1_df <- dat %>%
  group_by(Country) %>%
  summarise(
    Cases = sum(Cases),
    Deaths = sum(Deaths),
    CFR_crude_pct = 100 * Deaths / Cases
  )

ggplot(fig1_df, aes(x = Country, y = CFR_crude_pct, fill = Country)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", CFR_crude_pct)),
            vjust = -0.4, size = 4.2) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Crude fatality rate (all ages combined)",
       y = "Fatality rate (%)", x = NULL) +
  theme_classic(base_size = 13) +
  theme(legend.position = "none")
```

At face value, Italy looks worse. But what happens when we disaggregate by age group?

```{r}
fig2_df <- dat %>%
  select(Country, Age, CFR_band_pct)

ggplot(fig2_df, aes(x = Age, y = CFR_band_pct, fill = Country)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  labs(title = "Within every age band, China’s fatality rate is higher",
       x = "Age band", y = "Fatality rate (%)", fill = NULL) +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")

# ------------------------------------------------------------------------------
# 4) Figure 3: case age distribution by country
#    Shows why the crude rates reverse: Italy is older, China is younger.
#    Plot proportions so the skew is obvious.
# ------------------------------------------------------------------------------

fig3_df <- dat %>%
  group_by(Country, Age) %>%
  summarise(Cases = sum(Cases), .groups = "drop") %>%
  group_by(Country) %>%
  mutate(Prop = Cases / sum(Cases))

ggplot(fig3_df, aes(x = Age, y = Prop, fill = Country)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Case age mix: Italy skewed older, China skewed younger",
       x = "Age band", y = "Share of cases", fill = NULL) +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")


```

-   he crude rate is higher in Italy, which could tempt a quick conclusion if you stop there.

-   

    Before **Figure 2**: explain that age is a confounder, so you stratify by 10-year bands. Point out that within every band China’s CFR is higher.

-   

    Before **Figure 3**: show that Italy’s cases skew older while China’s skew younger, which flips the crude comparison. Then give the moral: check stratified results and adjust for key covariates when you can.Within every age group, China’s fatality rate is higher. The reversal arises because China had many more young cases, and younger groups have much lower fatality rates. Aggregating the data without age adjustment hid this reality.

In short there are a lot of critical questions that you can't answer with statistics, but the answers to those questions will have a huge impact on how you analyse and interpret data. And this is the reason why you should always think of statistics as a tool to help you learn about your data, no more and no less. It's a powerful tool to that end, but there's no substitute for careful thought.

## Statistics in Environmental Science: Why All the Numbers?

You may wonder why your environmental science class is suddenly filled with graphs, equations, and statistical terms. Why does environmental science have so much statistics? Let me break it down for you!

**Why Does Environmental Science Need Statistics?**

It's a bit of a complex story, but let's face it: nature is complex too! Unlike physics, where you might be studying straightforward things like electrons, in environmental science, we're dealing with living ecosystems, unpredictable weather, and human behavior.

Think of nature as a giant puzzle with pieces constantly changing shape. We use statistics to piece it together. We're not dealing with simple objects; we're dealing with the flow of rivers, the growth of forests, and the migration of animals. It's a whole different ballgame.

In a sense, environmental scientists have to be part detective, part mathematician. We need to understand statistics because we're solving harder problems than some other fields. We're trying to make sense of the messy, beautiful, and intricate web of life.

**Can't Someone Else Handle the Statistics?**

Sure, you could ask someone else to do the math, but here's the catch: understanding statistics is crucial for understanding environmental problems and solutions.

1.  **The Design Connection:** Want to research how pollution affects fish? You'll need to understand the statistics behind your study. Statistics and research design are like two sides of the same coin.

2.  **Understanding the Science:** Want to read up on climate change? You'll find a lot of statistics in those papers. To truly get what's going on, you need to know what those numbers are saying.

3.  **The Money Matter:** Hiring a statistician for every project? That'll cost you. Being self-sufficient in statistics means you can tackle more projects with fewer resources.

**But What if I Don't Care About Jobs, Research, or Conservation? Why Do I Need Statistics?**

Well, first of all, you've got me curious! But even if you're just an everyday citizen, statistics matter. You're surrounded by data, whether it's weather forecasts, pollution reports, or wildlife trends.

Knowing a bit about statistics is like having a decoder ring for the modern world. It helps you understand the news, make informed decisions, and even argue a point more convincingly.

So yes, statistics might feel like a strange guest in your environmental science class. But trust me, it's a guest you want to get to know. Because understanding statistics isn't just about crunching numbers; it's about understanding our world and how we can make it better.

## Statistics in everyday life

> *"We are drowning in information,\
> but we are starved for knowledge"*\
>
> -- Various authors, original probably John Naisbitt

When I started writing up my lecture notes I took the 20 most recent news articles posted to the ABC news website. Of those 20 articles, it turned out that 8 of them involved a discussion of something that I would call a statistical topic; 6 of those made a mistake. The most common error, if you're curious, was failing to report baseline data (e.g., the article mentions that 5% of people in situation X have some characteristic Y, but doesn't say how common the characteristic is for everyone else!) The point I'm trying to make here isn't that journalists are bad at statistics (though they almost always are), it's that a basic knowledge of statistics is very helpful for trying to figure out when someone else is either making a mistake or even lying to you. Perhaps, one of the biggest things that a knowledge of statistics does to you is cause you to get angry at the newspaper or the internet on a far more frequent basis :).

## There's more to research methods than statistics

So far, most of what I've talked about is statistics, and so you'd be forgiven for thinking that statistics is all I care about in life. To be fair, you wouldn't be far wrong, but research methodology is a broader concept than statistics. So most research methods courses will cover a lot of topics that relate much more to the pragmatics of research design, and in particular the issues that you encounter when trying to do research with humans. However, about 99% of student *fears* relate to the statistics part of the course, so I've focused on the stats in this discussion, and hopefully I've convinced you that statistics matters, and more importantly, that it's not to be feared. That being said, it's pretty typical for introductory research methods classes to be very stats-heavy. This is not (usually) because the lecturers are evil people. Quite the contrary, in fact. Introductory classes focus a lot on the statistics because you almost always find yourself needing statistics before you need the other research methods training. Why? Because almost all of your assignments in other classes will rely on statistical training, to a much greater extent than they rely on other methodological tools. It's not common for undergraduate assignments to require you to design your own study from the ground up (in which case you would need to know a lot about research design), but it *is* common for assignments to ask you to analyse and interpret data that were collected in a study that someone else designed (in which case you need statistics). In that sense, from the perspective of allowing you to do well in all your other classes, the statistics is more urgent.

But note that "urgent" is different from "important" -- they both matter. I really do want to stress that research design is just as important as data analysis, and this book does spend a fair amount of time on it. However, while statistics has a kind of universality, and provides a set of core tools that are useful for most types of psychological research, the research methods side isn't quite so universal. There are some general principles that everyone should think about, but a lot of research design is very idiosyncratic, and is specific to the area of research that you want to engage in. To the extent that it's the details that matter, those details don't usually show up in an introductory stats and research methods class.

## A brief introduction to research design

In this chapter, we're going to start thinking about the basic ideas that go into designing a study, collecting data, checking whether your data collection works, and so on. It won't give you enough information to allow you to design studies of your own, but it will give you a lot of the basic tools that you need to assess the studies done by other people. However, since the focus of this book is much more on data analysis than on data collection, I'm only giving a very brief overview. Note that this chapter is "special" in two ways. Firstly, it's much more psychology-specific than the later chapters. Secondly, it focuses much more heavily on the scientific problem of research methodology, and much less on the statistical problem of data analysis. Nevertheless, the two problems are related to one another, so it's traditional for stats textbooks to discuss the problem in a little detail. This chapter relies heavily on @Campbell1963 for the discussion of study design, and @Stevens1946 for the discussion of scales of measurement. Later versions will attempt to be more precise in the citations.

## Introduction to measurements

The first thing to understand is data collection can be thought of as a kind of **measurement**. That is, what we're trying to do here is measure something about human behavior or the human mind. What do I mean by "measurement"?

### Some thoughts about measurements

Measurement itself is a subtle concept, but basically it comes down to finding some way of assigning numbers, or labels, or some other kind of well-defined descriptions to "stuff". So, any of the following would count as a measurement:

-   My **age** is *33 years*.

-   I *do not* **like anchovies**.

-   My **chromosomal gender** is *male*.

-   My **self-identified gender** is *male*.

In the short list above, the **bolded part** is "the thing to be measured", and the *italicized part* is "the measurement itself". In fact, we can expand on this a little bit, by thinking about the set of possible measurements that could have arisen in each case:

-   My **age** (in years) could have been *0, 1, 2, 3 ...*, etc. The upper bound on what my age could possibly be is a bit fuzzy, but in practice you'd be safe in saying that the largest possible age is *150*, since no human has ever lived that long.

-   When asked if I **like anchovies**, I might have said that *I do*, or *I do not*, or *I have no opinion*, or *I sometimes do*.

-   My **chromosomal gender** is almost certainly going to be *male (XY)* or *female (XX)*, but there are a few other possibilities. I could also have *Klinfelter's syndrome (XXY)*, which is more similar to male than to female. And I imagine there are other possibilities too.

-   My **self-identified gender** is also likely to be *male* or *female*, but it doesn't have to agree with my chromosomal gender. I may also choose to identify with *neither*, or to explicitly call myself *transgender*.

As you can see, for some things (like age) it seems fairly obvious what the set of possible measurements should be, whereas for other things it gets a bit tricky. But I want to point out that even in the case of someone's age, it's much more subtle than this. For instance, in the example above, I assumed that it was okay to measure age in years. But if you're a developmental psychologist, that's way too crude, and so you often measure age in *years and months* (if a child is 2 years and 11 months, this is usually written as "2;11"). If you're interested in newborns, you might want to measure age in *days since birth*, maybe even *hours since birth*. In other words, the way in which you specify the allowable measurement values is important.

Looking at this a bit more closely, you might also realize that the concept of "age" isn't actually all that precise. In general, when we say "age" we implicitly mean "the length of time since birth". But that's not always the right way to do it. Suppose you're interested in how newborn babies control their eye movements. If you're interested in kids that young, you might also start to worry that "birth" is not the only meaningful point in time to care about. If Baby Alice is born 3 weeks premature and Baby Bianca is born 1 week late, would it really make sense to say that they are the "same age" if we encountered them "2 hours after birth"? In one sense, yes: by social convention, we use birth as our reference point for talking about age in everyday life, since it defines the amount of time the person has been operating as an independent entity in the world, but from a scientific perspective that's not the only thing we care about. When we think about the biology of human beings, it's often useful to think of ourselves as organisms that have been growing and maturing since conception, and from that perspective Alice and Bianca aren't the same age at all. So you might want to define the concept of "age" in two different ways: the length of time since conception, and the length of time since birth. When dealing with adults, it won't make much difference, but when dealing with newborns it might.

Moving beyond these issues, there's the question of methodology. What specific "measurement method" are you going to use to find out someone's age? As before, there are lots of different possibilities:

-   You could just ask people "how old are you?" The method of self-report is fast, cheap and easy, but it only works with people old enough to understand the question, and some people lie about their age.

-   You could ask an authority (e.g., a parent) "how old is your child?" This method is fast, and when dealing with kids it's not all that hard since the parent is almost always around. It doesn't work as well if you want to know "age since conception", since a lot of parents can't say for sure when conception took place. For that, you might need a different authority (e.g., an obstetrician).

-   You could look up official records, like birth certificates. This is time consuming and annoying, but it has its uses (e.g., if the person is now dead).

### Operationalization: defining your measurement

All of the ideas discussed in the previous section all relate to the concept of **operationalization**. To be a bit more precise about the idea, operationalization is the process by which we take a meaningful but somewhat vague concept, and turn it into a precise measurement. The process of operationalization can involve several different things:

-   Being precise about what you are trying to measure: For instance, does "age" mean "time since birth" or "time since conception" in the context of your research?

-   Determining what method you will use to measure it: Will you use self-report to measure age, ask a parent, or look up an official record? If you're using self-report, how will you phrase the question?

-   Defining the set of the allowable values that the measurement can take: Note that these values don't always have to be numerical, though they often are. When measuring age, the values are numerical, but we still need to think carefully about what numbers are allowed. Do we want age in years, years and months, days, hours? Etc. For other types of measurements (e.g., gender), the values aren't numerical. But, just as before, we need to think about what values are allowed. If we're asking people to self-report their gender, what options to we allow them to choose between? Is it enough to allow only "male" or "female"? Do you need an "other" option? Or should we not give people any specific options, and let them answer in their own words? And if you open up the set of possible values to include all verbal response, how will you interpret their answers?

Operationalization is a tricky business, and there's no "one, true way" to do it. The way in which you choose to operationalize the informal concept of "age" or "gender" into a formal measurement depends on what you need to use the measurement for. Often you'll find that the community of scientists who work in your area have some fairly well-established ideas for how to go about it. In other words, operationalization needs to be thought through on a case by case basis. Nevertheless, while there a lot of issues that are specific to each individual research project, there are some aspects to it that are pretty general.

Before moving on, I want to take a moment to clear up our terminology, and in the process introduce one more term. Here are four different things that are closely related to each other:

-   **A theoretical construct**. This is the thing that you're trying to take a measurement of, like "age", "gender" or an "opinion". A theoretical construct can't be directly observed, and often they're actually a bit vague.

-   **A measure**. The measure refers to the method or the tool that you use to make your observations. A question in a survey, a behavioral observation or a brain scan could all count as a measure.

-   **An operationalization**. The term "operationalization" refers to the logical connection between the measure and the theoretical construct, or to the process by which we try to derive a measure from a theoretical construct.

-   **A variable**. Finally, a new term. A variable is what we end up with when we apply our measure to something in the world. That is, variables are the actual "data" that we end up with in our data sets.

In practice, even scientists tend to blur the distinction between these things, but it's very helpful to try to understand the differences.

## Scales of measurement

As the previous section indicates, the outcome of a psychological measurement is called a variable. But not all variables are of the same qualitative type, and it's very useful to understand what types there are. A very useful concept for distinguishing between different types of variables is what's known as **scales of measurement**.

### Nominal scale

A **nominal scale** variable (also referred to as a **categorical** variable) is one in which there is no particular relationship between the different possibilities: for these kinds of variables it doesn't make any sense to say that one of them is \`\`bigger' or "better" than any other one, and it absolutely doesn't make any sense to average them. The classic example for this is "eye color". Eyes can be blue, green and brown, among other possibilities, but none of them is any "better" than any other one. As a result, it would feel really weird to talk about an "average eye color". Similarly, gender is nominal too: male isn't better or worse than female, neither does it make sense to try to talk about an "average gender". In short, nominal scale variables are those for which the only thing you can say about the different possibilities is that they are different. That's it.

Let's take a slightly closer look at this. Suppose I was doing research on how people commute to and from work. One variable I would have to measure would be what kind of transportation people use to get to work. This "transport type" variable could have quite a few possible values, including: "train", "bus", "car", "bicycle", etc. For now, let's suppose that these four are the only possibilities, and suppose that when I ask 100 people how they got to work today, and I get this:

| Transportation | Number of people |
|:---------------|:----------------:|
| \(1\) Train    |        12        |
| \(2\) Bus      |        30        |
| \(3\) Car      |        48        |
| \(4\) Bicycle  |        10        |

So, what's the average transportation type? Obviously, the answer here is that there isn't one. It's a silly question to ask. You can say that travel by car is the most popular method, and travel by train is the least popular method, but that's about all. Similarly, notice that the order in which I list the options isn't very interesting. I could have chosen to display the data like this and nothing really changes.

| Transportation | Number of people |
|:---------------|:----------------:|
| \(3\) Car      |        48        |
| \(1\) Train    |        12        |
| \(4\) Bicycle  |        10        |
| \(2\) Bus      |        30        |

### Ordinal scale

**Ordinal scale** variables have a bit more structure than nominal scale variables, but not by a lot. An ordinal scale variable is one in which there is a natural, meaningful way to order the different possibilities, but you can't do anything else. The usual example given of an ordinal variable is "finishing position in a race". You *can* say that the person who finished first was faster than the person who finished second, but you *don't* know how much faster. As a consequence we know that 1st $>$ 2nd, and we know that 2nd $>$ 3rd, but the difference between 1st and 2nd might be much larger than the difference between 2nd and 3rd.

Here's an more psychologically interesting example. Suppose I'm interested in people's attitudes to climate change, and I ask them to pick one of these four statements that most closely matches their beliefs:

> (1) Temperatures are rising, because of human activity\
> (2) Temperatures are rising, but we don't know why\
> (3) Temperatures are rising, but not because of humans\
> (4) Temperatures are not rising

Notice that these four statements actually do have a natural ordering, in terms of "the extent to which they agree with the current science". Statement 1 is a close match, statement 2 is a reasonable match, statement 3 isn't a very good match, and statement 4 is in strong opposition to the science. So, in terms of the thing I'm interested in (the extent to which people endorse the science), I can order the items as $1 > 2 > 3 > 4$. Since this ordering exists, it would be very weird to list the options like this...

> (3) Temperatures are rising, but not because of humans\
> (4) Temperatures are rising, because of human activity\
> (5) Temperatures are not rising\
> (6) Temperatures are rising, but we don't know why

...because it seems to violate the natural "structure" to the question.

So, let's suppose I asked 100 people these questions, and got the following answers:

|                                                          | Number |
|----------------------------------------------------------|:------:|
| \(1\) Temperatures are rising, because of human activity |   51   |
| \(2\) Temperatures are rising, but we don't know why     |   20   |
| \(3\) Temperatures are rising, but not because of humans |   10   |
| \(4\) Temperatures are not rising                        |   19   |

When analyzing these data, it seems quite reasonable to try to group (1), (2) and (3) together, and say that 81 of 100 people were willing to *at least partially* endorse the science. And it's *also* quite reasonable to group (2), (3) and (4) together and say that 49 of 100 people registered *at least some disagreement* with the dominant scientific view. However, it would be entirely bizarre to try to group (1), (2) and (4) together and say that 90 of 100 people said...what? There's nothing sensible that allows you to group those responses together at all.

That said, notice that while we *can* use the natural ordering of these items to construct sensible groupings, what we *can't* do is average them. For instance, in my simple example here, the "average" response to the question is 1.97. If you can tell me what that means, I'd love to know. Because that sounds like gibberish to me!

### Interval scale

In contrast to nominal and ordinal scale variables, **interval scale** and ratio scale variables are variables for which the numerical value is genuinely meaningful. In the case of interval scale variables, the *differences* between the numbers are interpretable, but the variable doesn't have a "natural" zero value. A good example of an interval scale variable is measuring temperature in degrees Celsius. For instance, if it was 15$^\circ$ yesterday and 18$^\circ$ today, then the 3$^\circ$ difference between the two is genuinely meaningful. Moreover, that 3$^\circ$ difference is *exactly the same* as the 3$^\circ$ difference between $7^\circ$ and $10^\circ$. In short, addition and subtraction are meaningful for interval scale variables.

However, notice that the $0^\circ$ does not mean "no temperature at all": it actually means "the temperature at which water freezes", which is pretty arbitrary. As a consequence, it becomes pointless to try to multiply and divide temperatures. It is wrong to say that $20^\circ$ is *twice as hot* as $10^\circ$, just as it is weird and meaningless to try to claim that $20^\circ$ is negative two times as hot as $-10^\circ$.

Again, lets look at a more psychological example. Suppose I'm interested in looking at how the attitudes of first-year university students have changed over time. Obviously, I'm going to want to record the year in which each student started. This is an interval scale variable. A student who started in 2003 did arrive 5 years before a student who started in 2008. However, it would be completely insane for me to divide 2008 by 2003 and say that the second student started "1.0024 times later" than the first one. That doesn't make any sense at all.

### Ratio scale

The fourth and final type of variable to consider is a **ratio scale** variable, in which zero really means zero, and it's okay to multiply and divide. A good psychological example of a ratio scale variable is response time (RT). In a lot of tasks it's very common to record the amount of time somebody takes to solve a problem or answer a question, because it's an indicator of how difficult the task is. Suppose that Alan takes 2.3 seconds to respond to a question, whereas Ben takes 3.1 seconds. As with an interval scale variable, addition and subtraction are both meaningful here. Ben really did take $3.1 - 2.3 = 0.8$ seconds longer than Alan did. However, notice that multiplication and division also make sense here too: Ben took $3.1 / 2.3 = 1.35$ times as long as Alan did to answer the question. And the reason why you can do this is that, for a ratio scale variable such as RT, "zero seconds" really does mean "no time at all".

### Continuous versus discrete variables

There's a second kind of distinction that you need to be aware of, regarding what types of variables you can run into. This is the distinction between continuous variables and discrete variables. The difference between these is as follows:

-   A **continuous variable** is one in which, for any two values that you can think of, it's always logically possible to have another value in between.

-   A **discrete variable** is, in effect, a variable that isn't continuous. For a discrete variable, it's sometimes the case that there's nothing in the middle.

These definitions probably seem a bit abstract, but they're pretty simple once you see some examples. For instance, response time is continuous. If Alan takes 3.1 seconds and Ben takes 2.3 seconds to respond to a question, then it's possible for Cameron's response time to lie in between, by taking 3.0 seconds. And of course it would also be possible for David to take 3.031 seconds to respond, meaning that his RT would lie in between Cameron's and Alan's. And while in practice it might be impossible to measure RT that precisely, it's certainly possible in principle. Because we can always find a new value for RT in between any two other ones, we say that RT is continuous.

Discrete variables occur when this rule is violated. For example, nominal scale variables are always discrete: there isn't a type of transportation that falls "in between" trains and bicycles, not in the strict mathematical way that 2.3 falls in between 2 and 3. So transportation type is discrete. Similarly, ordinal scale variables are always discrete: although "2nd place" does fall between "1st place" and "3rd place", there's nothing that can logically fall in between "1st place" and "2nd place". Interval scale and ratio scale variables can go either way. As we saw above, response time (a ratio scale variable) is continuous. Temperature in degrees Celsius (an interval scale variable) is also continuous. However, the year you went to school (an interval scale variable) is discrete. There's no year in between 2002 and 2003. The number of questions you get right on a true-or-false test (a ratio scale variable) is also discrete: since a true-or-false question doesn't allow you to be "partially correct", there's nothing in between 5/10 and 6/10. The table summarizes the relationship between the scales of measurement and the discrete/continuity distinction. Cells with a tick mark correspond to things that are possible. I'm trying to hammer this point home, because (a) some textbooks get this wrong, and (b) people very often say things like "discrete variable" when they mean "nominal scale variable". It's very unfortunate.

|          | continuous | discrete |
|----------|:----------:|:--------:|
| nominal  |            |    x     |
| ordinal  |            |    x     |
| interval |     x      |    x     |
| ratio    |     x      |    x     |

: The relationship between the scales of measurement and the discrete/continuity distinction. Cells with an x correspond to things that are possible.

### Some complexities

Okay, I know you're going to be shocked to hear this, but ...the real world is much messier than this little classification scheme suggests. Very few variables in real life actually fall into these nice neat categories, so you need to be kind of careful not to treat the scales of measurement as if they were hard and fast rules. It doesn't work like that: they're guidelines, intended to help you think about the situations in which you should treat different variables differently. Nothing more.

So let's take a classic example, maybe *the* classic example, of a psychological measurement tool: the **Likert scale**. The humble Likert scale is the bread and butter tool of all survey design. You yourself have filled out hundreds, maybe thousands of them, and odds are you've even used one yourself. Suppose we have a survey question that looks like this:

> Which of the following best describes your opinion of the statement that "all pirates are freaking awesome" ...

and then the options presented to the participant are these:

> \(1\) Strongly disagree\
> (2) Disagree\
> (3) Neither agree nor disagree\
> (4) Agree\
> (5) Strongly agree

This set of items is an example of a 5-point Likert scale: people are asked to choose among one of several (in this case 5) clearly ordered possibilities, generally with a verbal descriptor given in each case. However, it's not necessary that all items be explicitly described. This is a perfectly good example of a 5-point Likert scale too:

> \(1\) Strongly disagree\
> (2)\
> (3)\
> (4)\
> (5) Strongly agree

Likert scales are very handy, if somewhat limited, tools. The question is, what kind of variable are they? They're obviously discrete, since you can't give a response of 2.5. They're obviously not nominal scale, since the items are ordered; and they're not ratio scale either, since there's no natural zero.

But are they ordinal scale or interval scale? One argument says that we can't really prove that the difference between "strongly agree" and "agree" is of the same size as the difference between "agree" and "neither agree nor disagree". In fact, in everyday life it's pretty obvious that they're not the same at all. So this suggests that we ought to treat Likert scales as ordinal variables. On the other hand, in practice most participants do seem to take the whole "on a scale from 1 to 5" part fairly seriously, and they tend to act as if the differences between the five response options were fairly similar to one another. As a consequence, a lot of researchers treat Likert scale data as if it were interval scale. It's not interval scale, but in practice it's close enough that we usually think of it as being **quasi-interval scale**.

## Assessing the reliability of a measurement

At this point we've thought a little bit about how to operationalize a theoretical construct and thereby create a psychological measure; and we've seen that by applying psychological measures we end up with variables, which can come in many different types. At this point, we should start discussing the obvious question: is the measurement any good? We'll do this in terms of two related ideas: *reliability* and *validity*. Put simply, the **reliability** of a measure tells you how *precisely* you are measuring something, whereas the validity of a measure tells you how *accurate* the measure is.

Reliability is actually a very simple concept: it refers to the repeatability or consistency of your measurement. The measurement of my weight by means of a "bathroom scale" is very reliable: if I step on and off the scales over and over again, it'll keep giving me the same answer. Measuring my intelligence by means of "asking my mom" is very unreliable: some days she tells me I'm a bit thick, and other days she tells me I'm a complete moron. Notice that this concept of reliability is different to the question of whether the measurements are correct (the correctness of a measurement relates to it's validity). If I'm holding a sack of potatoes when I step on and off of the bathroom scales, the measurement will still be reliable: it will always give me the same answer. However, this highly reliable answer doesn't match up to my true weight at all, therefore it's wrong. In technical terms, this is a *reliable but invalid* measurement. Similarly, while my mom's estimate of my intelligence is a bit unreliable, she might be right. Maybe I'm just not too bright, and so while her estimate of my intelligence fluctuates pretty wildly from day to day, it's basically right. So that would be an *unreliable but valid* measure. Of course, to some extent, notice that if my mum's estimates are too unreliable, it's going to be very hard to figure out which one of her many claims about my intelligence is actually the right one. To some extent, then, a very unreliable measure tends to end up being invalid for practical purposes; so much so that many people would say that reliability is necessary (but not sufficient) to ensure validity.

Okay, now that we're clear on the distinction between reliability and validity, let's have a think about the different ways in which we might measure reliability:

-   **Test-retest reliability**. This relates to consistency over time: if we repeat the measurement at a later date, do we get a the same answer?

-   **Inter-rater reliability**. This relates to consistency across people: if someone else repeats the measurement (e.g., someone else rates my intelligence) will they produce the same answer?

-   **Parallel forms reliability**. This relates to consistency across theoretically-equivalent measurements: if I use a different set of bathroom scales to measure my weight, does it give the same answer?

-   **Internal consistency reliability**. If a measurement is constructed from lots of different parts that perform similar functions (e.g., a personality questionnaire result is added up across several questions) do the individual parts tend to give similar answers.

Not all measurements need to possess all forms of reliability. For instance, educational assessment can be thought of as a form of measurement. One of the subjects that I teach, *Inroduction to Environmental Science*, has an assessment structure that has a presentation component and an exam component (plus other things). The exam component is *intended* to measure something different from the presentation component, so the assessment as a whole has low internal consistency. However, within the exam there are several questions that are intended to (approximately) measure the same things, and those tend to produce similar outcomes; so the exam on its own has a fairly high internal consistency. Which is as it should be. You should only demand reliability in those situations where you want to be measure the same thing!

## The role of variables: predictors and outcomes

Okay, I've got one last piece of terminology that I need to explain to you before moving away from variables. Normally, when we do some research we end up with lots of different variables. Then, when we analyse our data we usually try to explain some of the variables in terms of some of the other variables. It's important to keep the two roles "thing doing the explaining" and "thing being explained" distinct. So let's be clear about this now. Firstly, we might as well get used to the idea of using mathematical symbols to describe variables, since it's going to happen over and over again. Let's denote the "to be explained" variable $Y$, and denote the variables "doing the explaining" as $X_1$, $X_2$, etc.

Now, when we doing an analysis, we have different names for $X$ and $Y$, since they play different roles in the analysis. The classical names for these roles are **independent variable** (IV) and **dependent variable** (DV). The IV is the variable that you use to do the explaining (i.e., $X$) and the DV is the variable being explained (i.e., $Y$). The logic behind these names goes like this: if there really is a relationship between $X$ and $Y$ then we can say that $Y$ depends on $X$, and if we have designed our study "properly" then $X$ isn't dependent on anything else. However, I personally find those names horrible: they're hard to remember and they're highly misleading, because (a) the IV is never actually "independent of everything else" and (b) if there's no relationship, then the DV doesn't actually depend on the IV. And in fact, because I'm not the only person who thinks that IV and DV are just awful names, there are a number of alternatives that I find more appealing.

For example, in an experiment the IV refers to the **manipulation**, and the DV refers to the **measurement**. So, we could use **manipulated variable** (independent variable) and **measured variable** (dependent variable).

| role of the variable   | classical name            | modern name  |
|:-----------------------|:--------------------------|:-------------|
| "to be explained"      | dependent variable (DV)   | Measurement  |
| "to do the explaining" | independent variable (IV) | Manipulation |

: The terminology used to distinguish between different roles that a variable can play when analyzing a data set.

We could also use **predictors** and **outcomes**. The idea here is that what you're trying to do is use $X$ (the predictors) to make guesses about $Y$ (the outcomes). This is summarized in the table:

| role of the variable   | classical name            | modern name |
|:-----------------------|:--------------------------|:------------|
| "to be explained"      | dependent variable (DV)   | outcome     |
| "to do the explaining" | independent variable (IV) | predictor   |

: The terminology used to distinguish between different roles that a variable can play when analyzing a data set.

## Experimental and non-experimental research

One of the big distinctions that you should be aware of is the distinction between "experimental research" and "non-experimental research". When we make this distinction, what we're really talking about is the degree of control that the researcher exercises over the people and events in the study.

### Experimental research

The key features of **experimental research** is that the researcher controls all aspects of the study, especially what participants experience during the study. In particular, the researcher manipulates or varies something (IVs), and then allows the outcome variable (DV) to vary naturally. The idea here is to deliberately vary the something in the world (IVs) to see if it has any causal effects on the outcomes. Moreover, in order to ensure that there's no chance that something other than the manipulated variable is causing the outcomes, everything else is kept constant or is in some other way "balanced" to ensure that they have no effect on the results. In practice, it's almost impossible to *think* of everything else that might have an influence on the outcome of an experiment, much less keep it constant. The standard solution to this is **randomization**: that is, we randomly assign people to different groups, and then give each group a different treatment (i.e., assign them different values of the predictor variables). We'll talk more about randomization later in this course, but for now, it's enough to say that what randomization does is minimize (but not eliminate) the chances that there are any systematic difference between groups.

Let's consider a very simple, completely unrealistic and grossly unethical example. Suppose you wanted to find out if smoking causes lung cancer. One way to do this would be to find people who smoke and people who don't smoke, and look to see if smokers have a higher rate of lung cancer. This is *not* a proper experiment, since the researcher doesn't have a lot of control over who is and isn't a smoker. And this really matters: for instance, it might be that people who choose to smoke cigarettes also tend to have poor diets, or maybe they tend to work in asbestos mines, or whatever. The point here is that the groups (smokers and non-smokers) actually differ on lots of things, not *just* smoking. So it might be that the higher incidence of lung cancer among smokers is caused by something else, not by smoking per se. In technical terms, these other things (e.g. diet) are called "confounds", and we'll talk about those in just a moment.

In the meantime, let's now consider what a proper experiment might look like. Recall that our concern was that smokers and non-smokers might differ in lots of ways. The solution, as long as you have no ethics, is to *control* who smokes and who doesn't. Specifically, if we randomly divide participants into two groups, and force half of them to become smokers, then it's very unlikely that the groups will differ in any respect other than the fact that half of them smoke. That way, if our smoking group gets cancer at a higher rate than the non-smoking group, then we can feel pretty confident that (a) smoking does cause cancer and (b) we're murderers.

### Non-experimental research

**Non-experimental research** in environmental science often encompasses studies in which researchers cannot exert full control over variables as they might in a typical experimental setup. In environmental science, this issue is particularly pronounced, as there is no "control planet" to allow for perfect experimental conditions. This unique challenge leads to a reliance on alternative approaches, such as time series analysis, to understand complex environmental phenomena.

One distinction worth making between two types of non-experimental research is the difference between **quasi-experimental research** and **case studies**. Much of the research in environmental science fits into a quasi-experimental design. Here, researchers might wish to study the effects of industrial pollution on a river system but cannot directly control all the variables, such as the quantity or type of pollutants being emitted by various industries. Instead, they must observe existing conditions and make careful comparisons, possibly using statistical tools to account for confounding variables.

**Time series analysis** becomes vital in this context, as many environmental processes unfold over extended periods. For example, tracking changes in global temperatures or sea level requires years of consistent data collection. The complexity and variability inherent in environmental systems make drawing definitive conclusions difficult, but sophisticated statistical methods can help isolate specific effects.

Environmental science also frequently relies on detailed **case studies**. These investigations focus on particular events or locations to provide in-depth insights into environmental processes and their impacts on ecosystems, human health, and societal structures. Case studies might explore the aftermath of a natural disaster, the unique ecology of a threatened habitat, or the social and economic consequences of environmental policies.

Though they often lack the broad generalizability of more controlled experimental research, case studies in environmental science offer valuable opportunities to understand the complexity of real-world situations. They can uncover nuances and subtleties that might be overlooked in a broader experimental or quasi-experimental approach.

The absence of a control planet and the complexity of environmental systems necessitates flexible and often non-traditional approaches to scientific inquiry in environmental science. While these methods, including quasi-experimental designs and case studies, may not offer the same level of control as experimental designs, they are indispensable tools in the field. By leveraging time series data, employing careful statistical analysis, and embracing the rich insights offered by focused case studies, researchers can continue to expand our understanding of the biosphere and its interactions with human activities.

## Assessing the validity of a study

More than any other thing, a scientist wants their research to be "valid". The conceptual idea behind **validity** is very simple: can you trust the results of your study? If not, the study is invalid. However, while it's easy to state, in practice it's much harder to check validity than it is to check reliability. And in all honesty, there's no precise, clearly agreed upon notion of what validity actually is. In fact, there's lots of different kinds of validity, each of which raises it's own issues, and not all forms of validity are relevant to all studies. I'm going to talk about five different types:

-   Internal validity

-   External validity

-   Construct validity

-   Face validity

-   Ecological validity

To give you a quick guide as to what matters here...(1) Internal and external validity are the most important, since they tie directly to the fundamental question of whether your study really works. (2) Construct validity asks whether you're measuring what you think you are. (3) Face validity isn't terribly important except insofar as you care about "appearances". (4) Ecological validity is a special case of face validity that corresponds to a kind of appearance that you might care about a lot.

### Internal validity

**Internal validity** refers to the extent to which you are able draw the correct conclusions about the causal relationships between variables. It's called "internal" because it refers to the relationships between things "inside" the study. Let's illustrate the concept with a simple example. Suppose you're interested in finding out whether a university education makes you write better. To do so, you get a group of first year students, ask them to write a 1000 word essay, and count the number of spelling and grammatical errors they make. Then you find some third-year students, who obviously have had more of a university education than the first-years, and repeat the exercise. And let's suppose it turns out that the third-year students produce fewer errors. And so you conclude that a university education improves writing skills. Right? Except... the big problem that you have with this experiment is that the third-year students are older, and they've had more experience with writing things. So it's hard to know for sure what the causal relationship is: Do older people write better? Or people who have had more writing experience? Or people who have had more education? Which of the above is the true *cause* of the superior performance of the third-years? Age? Experience? Education? You can't tell. This is an example of a failure of internal validity, because your study doesn't properly tease apart the *causal* relationships between the different variables.

### External validity

**External validity** relates to the **generalizability** of your findings. That is, to what extent do you expect to see the same pattern of results in "real life" as you saw in your study. To put it a bit more precisely, any study that you do in psychology will involve a fairly specific set of questions or tasks, will occur in a specific environment, and will involve participants that are drawn from a particular subgroup. So, if it turns out that the results don't actually generalize to people and situations beyond the ones that you studied, then what you've got is a lack of external validity.

The classic example of this issue is the fact that a very large proportion of studies in psychology will use undergraduate psychology students as the participants. Obviously, however, the researchers don't care *only* about psychology students; they care about people in general. Given that, a study that uses only psych students as participants always carries a risk of lacking external validity. That is, if there's something "special" about psychology students that makes them different to the general populace in some *relevant* respect, then we may start worrying about a lack of external validity.

That said, it is absolutely critical to realize that a study that uses only psychology students does not necessarily have a problem with external validity. I'll talk about this again later, but it's such a common mistake that I'm going to mention it here. The external validity is threatened by the choice of population if (a) the population from which you sample your participants is very narrow (e.g., psych students), and (b) the narrow population that you sampled from is systematically different from the general population, *in some respect that is relevant to the psychological phenomenon that you intend to study*. The italicized part is the bit that lots of people forget: it is true that psychology undergraduates differ from the general population in lots of ways, and so a study that uses only psych students *may* have problems with external validity. However, if those differences aren't very relevant to the phenomenon that you're studying, then there's nothing to worry about. To make this a bit more concrete, here's two extreme examples:

-   You want to measure "attitudes of the general public towards psychotherapy", but all of your participants are psychology students. This study would almost certainly have a problem with external validity.

-   You want to measure the effectiveness of a visual illusion, and your participants are all psychology students. This study is very unlikely to have a problem with external validity

Having just spent the last couple of paragraphs focusing on the choice of participants (since that's the big issue that everyone tends to worry most about), it's worth remembering that external validity is a broader concept. The following are also examples of things that might pose a threat to external validity, depending on what kind of study you're doing:

-   People might answer a "psychology questionnaire" in a manner that doesn't reflect what they would do in real life.

-   Your lab experiment on (say) "human learning" has a different structure to the learning problems people face in real life.

### Construct validity

**Construct validity** is basically a question of whether you're measuring what you want to be measuring. A measurement has good construct validity if it is actually measuring the correct theoretical construct, and bad construct validity if it doesn't. To give very simple (if ridiculous) example, suppose I'm trying to investigate the rates with which university students cheat on their exams. And the way I attempt to measure it is by asking the cheating students to stand up in the lecture theater so that I can count them. When I do this with a class of 300 students, 0 people claim to be cheaters. So I therefore conclude that the proportion of cheaters in my class is 0%. Clearly this is a bit ridiculous. But the point here is not that this is a very deep methodological example, but rather to explain what construct validity is. The problem with my measure is that while I'm *trying* to measure "the proportion of people who cheat" what I'm actually measuring is "the proportion of people stupid enough to own up to cheating, or bloody minded enough to pretend that they do". Obviously, these aren't the same thing! So my study has gone wrong, because my measurement has very poor construct validity.

### Face validity

**Face validity** simply refers to whether or not a measure "looks like" it's doing what it's supposed to, nothing more. If I design a test of intelligence, and people look at it and they say "no, that test doesn't measure intelligence", then the measure lacks face validity. It's as simple as that. Obviously, face validity isn't very important from a pure scientific perspective. After all, what we care about is whether or not the measure *actually* does what it's supposed to do, not whether it *looks like* it does what it's supposed to do. As a consequence, we generally don't care very much about face validity. That said, the concept of face validity serves three useful pragmatic purposes:

-   Sometimes, an experienced scientist will have a "hunch" that a particular measure won't work. While these sorts of hunches have no strict evidentiary value, it's often worth paying attention to them. Because often times people have knowledge that they can't quite verbalize, so there might be something to worry about even if you can't quite say why. In other words, when someone you trust criticizes the face validity of your study, it's worth taking the time to think more carefully about your design to see if you can think of reasons why it might go awry. Mind you, if you don't find any reason for concern, then you should probably not worry: after all, face validity really doesn't matter much.

-   Often (very often), completely uninformed people will also have a "hunch" that your research is crap. And they'll criticize it on the internet or something. On close inspection, you'll often notice that these criticisms are actually focused entirely on how the study "looks", but not on anything deeper. The concept of face validity is useful for gently explaining to people that they need to substantiate their arguments further.

-   Expanding on the last point, if the beliefs of untrained people are critical (e.g., this is often the case for applied research where you actually want to convince policy makers of something or other) then you *have* to care about face validity. Simply because -- whether you like it or not -- a lot of people will use face validity as a proxy for real validity. If you want the government to change a law on scientific, psychological grounds, then it won't matter how good your studies "really" are. If they lack face validity, you'll find that politicians ignore you. Of course, it's somewhat unfair that policy often depends more on appearance than fact, but that's how things go.

### Ecological validity

**Ecological validity** is a different notion of validity, which is similar to external validity, but less important. The idea is that, in order to be ecologically valid, the entire set up of the study should closely approximate the real world scenario that is being investigated. In a sense, ecological validity is a kind of face validity -- it relates mostly to whether the study "looks" right, but with a bit more rigor to it. To be ecologically valid, the study has to look right in a fairly specific way. The idea behind it is the intuition that a study that is ecologically valid is more likely to be externally valid. It's no guarantee, of course. But the nice thing about ecological validity is that it's much easier to check whether a study is ecologically valid than it is to check whether a study is externally valid. An simple example would be eyewitness identification studies. Most of these studies tend to be done in a university setting, often with fairly simple array of faces to look at rather than a line up. The length of time between seeing the "criminal" and being asked to identify the suspect in the "line up" is usually shorter. The "crime" isn't real, so there's no chance that the witness being scared, and there's no police officers present, so there's not as much chance of feeling pressured. These things all mean that the study *definitely* lacks ecological validity. They might (but might not) mean that it also lacks external validity.

## Confounds, artifacts and other threats to validity

If we look at the issue of validity in the most general fashion, the two biggest worries that we have are *confounds* and *artifact*. These two terms are defined in the following way:

-   **Confound**: A confound is an additional, often unmeasured variable that turns out to be related to both the predictors and the outcomes. The existence of confounds threatens the internal validity of the study because you can't tell whether the predictor causes the outcome, or if the confounding variable causes it, etc.

-   **Artifact**: A result is said to be "artifactual" if it only holds in the special situation that you happened to test in your study. The possibility that your result is an artifact describes a threat to your external validity, because it raises the possibility that you can't generalize your results to the actual population that you care about.

As a general rule confounds are a bigger concern for non-experimental studies, precisely because they're not proper experiments: by definition, you're leaving lots of things uncontrolled, so there's a lot of scope for confounds working their way into your study. Experimental research tends to be much less vulnerable to confounds: the more control you have over what happens during the study, the more you can prevent confounds from appearing.

However, there's always swings and roundabouts, and when we start thinking about artifacts rather than confounds, the shoe is very firmly on the other foot. For the most part, artifactual results tend to be a concern for experimental studies than for non-experimental studies. To see this, it helps to realize that the reason that a lot of studies are non-experimental is precisely because what the researcher is trying to do is examine human behavior in a more naturalistic context. By working in a more real-world context, you lose experimental control (making yourself vulnerable to confounds) but because you tend to be studying human psychology "in the wild" you reduce the chances of getting an artifactual result. Or, to put it another way, when you take psychology out of the wild and bring it into the lab (which we usually have to do to gain our experimental control), you always run the risk of accidentally studying something different than you wanted to study: which is more or less the definition of an artifact.

Be warned though: the above is a rough guide only. It's absolutely possible to have confounds in an experiment, and to get artifactual results with non-experimental studies. This can happen for all sorts of reasons, not least of which is researcher error. In practice, it's really hard to think everything through ahead of time, and even very good researchers make mistakes. But other times it's unavoidable, simply because the researcher has ethics (e.g., see "differential attrition").

Okay. There's a sense in which almost any threat to validity can be characterized as a confound or an artifact: they're pretty vague concepts. So let's have a look at some of the most common examples...

### History effects

**History effects** refer to the possibility that specific events may occur during the study itself that might influence the outcomes. For instance, something might happen in between a pre-test and a post-test. Or, in between testing participant 23 and participant 24. Alternatively, it might be that you're looking at an older study, which was perfectly valid for its time, but the world has changed enough since then that the conclusions are no longer trustworthy. Examples of things that would count as history effects:

-   You're interested in how people think about risk and uncertainty. You started your data collection in December 2010. But finding participants and collecting data takes time, so you're still finding new people in February 2011. Unfortunately for you (and even more unfortunately for others), the Queensland floods occurred in January 2011, causing billions of dollars of damage and killing many people. Not surprisingly, the people tested in February 2011 express quite different beliefs about handling risk than the people tested in December 2010. Which (if any) of these reflects the "true" beliefs of participants? I think the answer is probably both: the Queensland floods genuinely changed the beliefs of the Australian public, though possibly only temporarily. The key thing here is that the "history" of the people tested in February is quite different to people tested in December.

-   You're testing the psychological effects of a new anti-anxiety drug. So what you do is measure anxiety before administering the drug (e.g., by self-report, and taking physiological measures, let's say), then you administer the drug, and then you take the same measures afterwards. In the middle, however, because your labs are in Los Angeles, there's an earthquake, which increases the anxiety of the participants.

### Maturation effects

As with history effects, **maturational effects** are fundamentally about change over time. However, maturation effects aren't in response to specific events. Rather, they relate to how people change on their own over time: we get older, we get tired, we get bored, etc. Some examples of maturation effects:

-   When doing developmental psychology research, you need to be aware that children grow up quite rapidly. So, suppose that you want to find out whether some educational trick helps with vocabulary size among 3 year olds. One thing that you need to be aware of is that the vocabulary size of children that age is growing at an incredible rate (multiple words per day), all on its own. If you design your study without taking this maturational effect into account, then you won't be able to tell if your educational trick works.

-   When running a very long experiment in the lab (say, something that goes for 3 hours), it's very likely that people will begin to get bored and tired, and that this maturational effect will cause performance to decline, regardless of anything else going on in the experiment

### Repeated testing effects

An important type of history effect is the effect of **repeated testing**. Suppose I want to take two measurements of some psychological construct (e.g., anxiety). One thing I might be worried about is if the first measurement has an effect on the second measurement. In other words, this is a history effect in which the "event" that influences the second measurement is the first measurement itself! This is not at all uncommon. Examples of this include:

-   *Learning and practice*: e.g., "intelligence" at time 2 might appear to go up relative to time 1 because participants learned the general rules of how to solve "intelligence-test-style" questions during the first testing session.

-   *Familiarity with the testing situation*: e.g., if people are nervous at time 1, this might make performance go down; after sitting through the first testing situation, they might calm down a lot precisely because they've seen what the testing looks like.

-   *Auxiliary changes caused by testing*: e.g., if a questionnaire assessing mood is boring, then mood at measurement at time 2 is more likely to become "bored", precisely because of the boring measurement made at time 1.

### Selection bias

**Selection bias** is a pretty broad term. Suppose that you're running an experiment with two groups of participants, where each group gets a different "treatment", and you want to see if the different treatments lead to different outcomes. However, suppose that, despite your best efforts, you've ended up with a gender imbalance across groups (say, group A has 80% females and group B has 50% females). It might sound like this could never happen, but trust me, it can. This is an example of a selection bias, in which the people "selected into" the two groups have different characteristics. If any of those characteristics turns out to be relevant (say, your treatment works better on females than males) then you're in a lot of trouble.

### Differential attrition

within the study itself, impacting both the internal and external validity of the research. This phenomenon can occur in environmental studies, where participants' commitment to certain practices might change over time, such as in multi-year studies on cover cropping.

When thinking about the effects of differential attrition, it is sometimes helpful to distinguish between two different types. The first is **homogeneous attrition**, in which the attrition effect is the same for all groups, treatments or conditions. Homogeneous attrition occurs when the dropout rate is roughly the same across different groups or conditions within the study. For instance, if many farmers participating in a cover cropping study discontinue the practice at similar rates, the sample may become unrepresentative. Though the internal validity (accuracy of the study's conclusions) might remain intact, the external validity (generalizability to the broader population) may suffer. In the realm of environmental science, this could mean the study's findings are less applicable to the broader community of farmers or agricultural systems.

The second type of differential attrition is **heterogeneous attrition**, in which the attrition effect is different for different groups. This is a much bigger problem: not only do you have to worry about your external validity, you also have to worry about your internal validity too. In a study on cover cropping, if certain groups, perhaps motivated by specific incentives or pressures, drop out at different rates, this can introduce a confounding variable. The comparison between the remaining participants may no longer be valid, as the groups have become fundamentally different due to the non-random dropout. This not only affects the generalizability of the findings but also the internal validity, as the study's conclusions within the sample itself may be compromised.

### Non-response bias

**Non-response bias** is closely related to selection bias, and to differential attrition. The simplest version of the problem goes like this. You mail out a survey to 1000 people, and only 300 of them reply. The 300 people who replied are almost certainly not a random sub-sample. People who respond to surveys are systematically different to people who don't. This introduces a problem when trying to generalize from those 300 people who replied, to the population at large; since you now have a very non-random sample. The issue of non-response bias is more general than this, though. Among the (say) 300 people that did respond to the survey, you might find that not everyone answers every question. If (say) 80 people chose not to answer one of your questions, does this introduce problems? As always, the answer is maybe. If the question that wasn't answered was on the last page of the questionnaire, and those 80 surveys were returned with the last page missing, there's a good chance that the missing data isn't a big deal: probably the pages just fell off. However, if the question that 80 people didn't answer was the most confrontational or invasive personal question in the questionnaire, then almost certainly you've got a problem. In essence, what you're dealing with here is what's called the problem of **missing data**. If the data that is missing was "lost" randomly, then it's not a big problem. If it's missing systematically, then it can be a big problem.

### Regression to the mean

**Regression to the mean** is a curious variation on selection bias. It refers to any situation where you select data based on an extreme value on some measure. Because the measure has natural variation, it almost certainly means that when you take a subsequent measurement, that later measurement will be less extreme than the first one, purely by chance.

Here's an example. Suppose we are interested in examining the effects of specific climatic conditions on the growth of a particular plant species. We identify the 20 locations with the most significant growth during a year of favorable weather conditions, such as ideal rainfall and temperature, and decide to study these areas further.

After another year, we observe that the plant growth in these 20 locations is still above average but not as exceptional as during the first year. The immediate reaction might be to conclude that a change in environmental practices or some other factor has adversely affected growth in these areas.

However, this might simply be an example of regression to the mean. Consider what is required for remarkable plant growth: optimal soil, hard work in maintenance, favorable weather conditions, and perhaps a bit of luck with other uncontrollable environmental factors.

While the soil quality and maintenance practices may be consistent from one year to the next, luck with weather is not. Certain areas that were fortunate with ideal weather conditions in the first year may not experience the same fortune in the following year. This inconsistency in luck leads to a less extreme measurement the second time.

Regression to the mean is surprisingly common. For instance, if two very tall people have kids, their children will tend to be taller than average, but not as tall as the parents. The reverse happens with very short parents: two very short parents will tend to have short children, but nevertheless those kids will tend to be taller than the parents. It can also be extremely subtle. For instance, there have been studies done that suggested that people learn better from negative feedback than from positive feedback. However, the way that people tried to show this was to give people positive reinforcement whenever they did good, and negative reinforcement when they did bad. And what you see is that after the positive reinforcement, people tended to do worse; but after the negative reinforcement they tended to do better. But! Notice that there's a selection bias here: when people do very well, you're selecting for "high" values, and so you should *expect* (because of regression to the mean) that performance on the next trial should be worse, regardless of whether reinforcement is given. Similarly, after a bad trial, people will tend to improve all on their own. The apparent superiority of negative feedback is an artifact caused by regression to the mean [@Kahneman1973]

### Experimenter bias

**Experimenter bias** can occur when the researcher, despite conscientious efforts, inadvertently influences the results of a study. This may not involve communication with human participants as in other fields but could still manifest in several ways within environmental research.

In environmental science, the experimenter might have **expectations or preconceived notions about what the data should reveal**. For instance, a researcher studying the effects of a certain agricultural practice on soil health may unconsciously select or emphasize data that align with their expectations or theoretical commitments. This can lead to a bias in data collection or interpretation, subtly steering the results toward the expected outcome.

Experimenter bias can also arise from **methodological choices**. A researcher may design an experiment, select sites for observation, or choose measurement techniques in a way that favors a particular outcome. For example, if studying the effect of pollution on a water source, the researcher might unintentionally choose sampling times or locations that are more likely to confirm their hypotheses, overlooking variations that could provide a more comprehensive and unbiased view.

In some cases, the researcher's bias may inadvertently **influence other field or laboratory personnel** involved in the study. A team member who understands the desired outcome of the study may unconsciously alter procedures, handling of samples, or recording of data to align with the researcher's expectations.

The classic example of experimenter bias is the case study of "Clever Hans", which dates back to 1907, [@Pfungst1911; @Hothersall2004]. Clever Hans was a horse that apparently was able to read and count, and perform other human like feats of intelligence. After Clever Hans became famous, psychologists started examining his behavior more closely. It turned out that -- not surprisingly -- Hans didn't know how to do maths. Rather, Hans was responding to the human observers around him. Because they did know how to count, and the horse had learned to change its behavior when people changed theirs.

The general solution to the problem of experimenter bias is to engage in double blind studies, where neither the experimenter nor the participant knows which condition the participant is in, or knows what the desired behavior is. This provides a very good solution to the problem, but it's important to recognize that it's not quite ideal, and typically not feasible in Environmental Science. Measures to mitigate this bias include rigorous peer review, careful documentation of methods, and maintaining an awareness of potential biases.

### Reactivity and Demand Effects

Even in environmental studies, the knowledge that a process or phenomenon is being observed can impact the behavior of the researchers or those involved in data collection. The basic idea is captured by the Hawthorne effect: people alter their performance because of the attention that the study focuses on them. The effect takes its name from a the "Hawthorne Works" factory outside of Chicago [@Adair1984]. A study done in the 1920s looking at the effects of lighting on worker productivity at the factory turned out to be an effect of the fact that the workers knew they were being studied, rather than the lighting. This so-called "Hawthorne effect" might influence how a field team conducts measurements if they know the specific goal of the study.

### Placebo effects

While typically associated with clinical trials, the **placebo effect** can have an analogous impact in environmental research. For instance, a belief in the efficacy of a certain conservation practice might lead to subjective reporting or unintentional data manipulation.

### Situation, measurement and subpopulation effects

In some respects, these terms are a catch-all term for "all other threats to external validity". The choice of location, time, measurement tools, and even who collects the data can all influence the results. Ensuring robust methodologies and considering the potential influences of these factors can help in achieving results that generalize more widely.

### Fraud, deception and self-deception

> *It is difficult to get a man to understand something, when his salary depends on his not understanding it.*\
>
> -- Upton Sinclair

One final thing that I feel like I should mention. While reading what the textbooks often have to say about assessing the validity of the study, I couldn't help but notice that they seem to make the assumption that the researcher is honest. I find this hilarious. While the vast majority of scientists are honest, in my experience at least, some are not. Not only that, as I mentioned earlier, scientists are not immune to belief bias -- it's easy for a researcher to end up deceiving themselves into believing the wrong thing, and this can lead them to conduct subtly flawed research, and then hide those flaws when they write it up. So you need to consider not only the (probably unlikely) possibility of outright fraud, but also the (probably quite common) possibility that the research is unintentionally "slanted". I opened a few standard textbooks and didn't find much of a discussion of this problem, so here's my own attempt to list a few ways in which these issues can arise are:

-   **Data fabrication**. Sometimes, people just make up the data. This is occasionally done with "good" intentions. For instance, the researcher believes that the fabricated data do reflect the truth, and may actually reflect "slightly cleaned up" versions of actual data. On other occasions, the fraud is deliberate and malicious. Some high-profile examples where data fabrication has been alleged or shown include Cyril Burt (a psychologist who is thought to have fabricated some of his data), Andrew Wakefield (who has been accused of fabricating his data connecting the MMR vaccine to autism) and Hwang Woo-suk (who falsified a lot of his data on stem cell research).

-   **Hoaxes**. Hoaxes share a lot of similarities with data fabrication, but they differ in the intended purpose. A hoax is often a joke, and many of them are intended to be (eventually) discovered. Often, the point of a hoax is to discredit someone or some field. There's quite a few well known scientific hoaxes that have occurred over the years (e.g., Piltdown man) some of were deliberate attempts to discredit particular fields of research (e.g., the Sokal affair).

-   **Data misrepresentation**. While fraud gets most of the headlines, it's much more common in my experience to see data being misrepresented. When I say this, I'm not referring to newspapers getting it wrong (which they do, almost always). I'm referring to the fact that often, the data don't actually say what the researchers think they say. My guess is that, almost always, this isn't the result of deliberate dishonesty, it's due to a lack of sophistication in the data analyses. For instance, think back to the example of Simpson's paradox that I discussed in the beginning of these notes. It's very common to see people present "aggregated" data of some kind; and sometimes, when you dig deeper and find the raw data yourself, you find that the aggregated data tell a different story to the disaggregated data. Alternatively, you might find that some aspect of the data is being hidden, because it tells an inconvenient story (e.g., the researcher might choose not to refer to a particular variable). There's a lot of variants on this; many of which are very hard to detect.

-   **Study "misdesign"**. Okay, this one is subtle. Basically, the issue here is that a researcher designs a study that has built-in flaws, and those flaws are never reported in the paper. The data that are reported are completely real, and are correctly analysed, but they are produced by a study that is actually quite wrongly put together. The researcher really wants to find a particular effect, and so the study is set up in such a way as to make it "easy" to (artifactually) observe that effect. One sneaky way to do this -- in case you're feeling like dabbling in a bit of fraud yourself -- is to design an experiment in which it's obvious to the participants what they're "supposed" to be doing, and then let reactivity work its magic for you. If you want, you can add all the trappings of double blind experimentation etc. It won't make a difference, since the study materials themselves are subtly telling people what you want them to do. When you write up the results, the fraud won't be obvious to the reader: what's obvious to the participant when they're in the experimental context isn't always obvious to the person reading the paper. Of course, the way I've described this makes it sound like it's always fraud: probably there are cases where this is done deliberately, but in my experience the bigger concern has been with unintentional misdesign. The researcher *believes* ...and so the study just happens to end up with a built in flaw, and that flaw then magically erases itself when the study is written up for publication.

-   **Data mining & post hoc hypothesizing**. Another way in which the authors of a study can more or less lie about what they found is by engaging in what's referred to as "data mining". As we'll discuss later in the class, if you keep trying to analyse your data in lots of different ways, you'll eventually find something that "looks" like a real effect but isn't. This is referred to as "data mining". It used to be quite rare because data analysis used to take weeks, but now that everyone has very powerful statistical software on their computers, it's becoming very common. Data mining per se isn't "wrong", but the more that you do it, the bigger the risk you're taking. The thing that is wrong, and I suspect is very common, is *unacknowledged* data mining. That is, the researcher run every possible analysis known to humanity, finds the one that works, and then pretends that this was the only analysis that they ever conducted. Worse yet, they often "invent" a hypothesis after looking at the data, to cover up the data mining. To be clear: it's not wrong to change your beliefs after looking at the data, and to reanalyze your data using your new "post hoc" hypotheses. What is wrong (and, I suspect, common) is failing to acknowledge that you've done so. If you acknowledge that you did it, then other researchers are able to take your behavior into account. If you don't, then they can't. And that makes your behavior deceptive. Bad!

-   **Publication bias & self-censoring**. Finally, a pervasive bias is "non-reporting" of negative results. This is almost impossible to prevent. Journals don't publish every article that is submitted to them: they prefer to publish articles that find "something". So, if 20 people run an experiment looking at whether reading *Finnegans Wake* causes insanity in humans, and 19 of them find that it doesn't, which one do you think is going to get published? Obviously, it's the one study that did find that *Finnegans Wake* causes insanity. This is an example of a *publication bias*: since no-one ever published the 19 studies that didn't find an effect, a naive reader would never know that they existed. Worse yet, most researchers "internalize" this bias, and end up *self-censoring* their research. Knowing that negative results aren't going to be accepted for publication, they never even try to report them. As a friend of mine says "for every experiment that you get published, you also have 10 failures". And she's right. The catch is, while some (maybe most) of those studies are failures for boring reasons (e.g. you stuffed something up) others might be genuine "null" results that you ought to acknowledge when you write up the "good" experiment. And telling which is which is often hard to do. A good place to start is a paper by @Ioannidis2005 with the depressing title "Why most published research findings are false". I'd also suggest taking a look at work by @Kuhberger2014 presenting statistical evidence that this actually happens in psychology.

There's probably a lot more issues like this to think about, but that'll do to start with. What I really want to point out is the blindingly obvious truth that real world science is conducted by actual humans, and only the most gullible of people automatically assumes that everyone else is honest and impartial. Actual scientists aren't usually *that* naive, but for some reason the world likes to pretend that we are, and the textbooks we usually write seem to reinforce that stereotype.

## Summary

In this chapter, we have explored essential aspects of research methodology pertinent to environmental statistics:

-   **Operationalization and Measurement**: Understanding how to define and measure theoretical constructs and recognizing the distinctions between variables are foundational to research.

-   **Scales of measurement and types of variables**: This section addressed the differences between discrete and continuous data, and the various scale types including nominal, ordinal, interval, and ratio scales.

-   **Reliability of a measurement**: If I measure the "same"" thing twice, should I expect to see the same result? Only if my measure is reliable. But what does it mean to talk about doing the "same" thing? Well, that's why we have different types of reliability. Make sure you remember what they are.

-   **Terminology: predictors and outcomes**: What roles do variables play in an analysis? Can you remember the difference between predictors and outcomes? Dependent and independent variables? Etc.

-   **Experimental and non-experimental research designs**: Clarification of the roles that variables play in an analysis, including the distinctions between predictors and outcomes, and dependent and independent variables.

-   **Research Designs**: Highlighting what constitutes an experiment within environmental research, the chapter delves into the differentiation between experimental and non-experimental research designs.

-   **Validity and Threats**: A comprehensive look at whether a study measures what it aims to, understanding potential pitfalls, and recognizing the myriad ways things can go awry.

Study design is a paramount component of research methodology, with numerous textbooks and resources available to further explore these concepts. Drawing on established texts like @Campbell1963, this chapter serves as an introduction and guide to these key topics, recognizing the intrinsic connection between statistics and study design within the field of environmental science.

## Videos

### Terms of Statistics

<iframe width="560" height="315" src="https://www.youtube.com/embed/HgtnpOtk350" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen>

</iframe>
