---
author:
  - Mallory L. Barnes
aliases: [redirects/foundations-for-inference.html]
---

```{r, include = FALSE}
source("global_stuff.R")
set.seed(538)
```

# Inference and Hypothesis Testing

> *Data and data sets are not objective; they are creations of human design. We give numbers their voice, draw inferences from them, and define their meaning through our interpretations.*
>
> Katie Crawford

**Inferential statistics** is the branch of statistics concerned with drawing conclusions about populations or causal relationships from sample data. It provides the tools to answer questions like: did our manipulation actually cause a change, or could the pattern we observed have arisen by random chance alone?

This chapter does that in two parts. The first half builds the logic by simulation, with no formulas: we work out what chance alone can produce, then compare a real result against it. The second half replaces the simulation with the standard hypothesis testing machinery, because once you can assume a known reference distribution you no longer have to rebuild the chance window from scratch every time.

By the end of this chapter you should be able to:

- **Explain why two groups can differ when nothing caused it**, and describe what chance alone is capable of producing.
- **Define the key players** in a hypothesis test: null and alternative, population and sampling distributions.
- **Test** $H_0$ three equivalent ways: a test statistic against a critical value, a confidence interval, and a $p$-value.
- **Choose between a one-tailed and two-tailed test** and justify the choice.

```{r, include=FALSE}
library(ggplot2)
library(dplyr)
library(gridExtra)
library(kableExtra)

# COLOUR SCHEME (decided 2026-08-18): keep this bespoke six-colour scheme, do not
# convert to e538_palette. This figure carries six semantic roles (null, sampling
# distribution, estimated population, true population, mu_0, rejection region) and the
# house palette has five, one of which (beige) is unusable as a data colour. The palette
# literally cannot express this figure without collapsing two distinctions.
# Consistency with Ch 3 was handled from the other side instead: Ch 3's CLT figures were
# recoloured to match this scheme where the roles overlap (population = orange,
# sampling distribution of the mean = dark green). Before that, Ch 3 used red for the
# population and blue for the sampling distribution, which collided with red = rejection
# region and blue = null here.

plot_distributions <- function(null_mean, null_sd, sample_mean, sample_sd, inferred_mean, inferred_sd, true_mean, true_sd) {
  x_vals <- seq(-6, 6, by=0.01)
  
  h0 <- dnorm(x_vals, mean=null_mean, sd=null_sd)
  sampling_mean <- dnorm(x_vals, mean=sample_mean, sd=sample_sd)
  inferred_parent <- dnorm(x_vals, mean=inferred_mean, sd=inferred_sd)
  true_parent <- dnorm(x_vals, mean=true_mean, sd=true_sd)
  CI_left <- qnorm(0.025, mean=sample_mean, sd=sample_sd)
  CI_right <- qnorm(0.975, mean=sample_mean, sd=sample_sd)
  y_intersect = dnorm(null_mean, mean=null_mean, sd=null_sd)

  
  p <- ggplot() + 
    geom_line(aes(x=x_vals, y=h0, color="Null Distribution"), size=1.5) +
    geom_line(aes(x=x_vals, y=sampling_mean, color="Sampling Distribution of the Sample Mean"), linetype="dotdash", size=1) +
    geom_line(aes(x=x_vals, y=inferred_parent, color="Estimated Population Distribution"), linetype="dashed", size=1.5) +
    geom_line(aes(x=x_vals, y=true_parent, color="True Population Distribution"), size=1.5) +
    geom_segment(aes(x = CI_left, xend = CI_right, y = 0., yend = 0), color="darkgreen", size=1.5) +
    geom_segment(aes(x = null_mean, xend = null_mean, y = y_intersect, yend = 0), linetype="dashed", color="magenta", size=1) +
    theme_classic() + 
    theme(legend.position='bottom', 
          axis.text = element_blank(), 
          axis.title = element_blank(), 
          axis.ticks = element_blank(), 
          legend.text = element_text(size = 12)) + 
    guides(color=guide_legend(title=NULL, nrow=2)) +
    scale_color_manual(values=c("Null Distribution"="blue", 
                                "Sampling Distribution of the Sample Mean"="darkgreen", 
                                "Estimated Population Distribution"="green3", 
                                "True Population Distribution"="orange"))

  return(p)
}

plot_distributions_no_error <- function(
  null_mean, null_sd,
  sample_mean, sample_sd,     # sample_sd here is the SE used for the sampling dist + CI
  inferred_mean, inferred_sd,
  show_ci = TRUE
){
  x_vals <- seq(-6, 6, by = 0.01)

  h0              <- dnorm(x_vals, mean = null_mean,    sd = null_sd)
  sampling_mean   <- dnorm(x_vals, mean = sample_mean,  sd = sample_sd)
  inferred_parent <- dnorm(x_vals, mean = inferred_mean, sd = inferred_sd)

  # 95% CI for the sample mean (normal approx)
  CI_left  <- qnorm(0.025, mean = sample_mean, sd = sample_sd)
  CI_right <- qnorm(0.975, mean = sample_mean, sd = sample_sd)

  # height at mu0 for the magenta reference drop line
  y_intersect <- dnorm(null_mean, mean = null_mean, sd = null_sd)

  p <- ggplot() +
    geom_line(aes(x = x_vals, y = h0,             color = "Null Distribution"), size = 1.5) +
    geom_line(aes(x = x_vals, y = sampling_mean,  color = "Sampling Distribution of the Sample Mean"),
              linetype = "dotdash", size = 1) +
    geom_line(aes(x = x_vals, y = inferred_parent, color = "Estimated Population Distribution"),
              linetype = "dashed", size = 1.5) +
    # mu0 (magenta dashed)
    geom_segment(aes(x = null_mean, xend = null_mean, y = y_intersect, yend = 0),
                 linetype = "dashed", color = "magenta", size = 1) +
    # CI bar (thick dark green): keep this!
    { if (show_ci) geom_segment(aes(x = CI_left, xend = CI_right, y = 0, yend = 0),
                                color = "darkgreen", size = 1.8) } +
    theme_classic() +
    theme(
      legend.position = "bottom",
      axis.text = element_blank(),
      axis.title = element_blank(),
      axis.ticks = element_blank(),
      legend.text = element_text(size = 12)
    ) +
    guides(color = guide_legend(title = NULL, nrow = 2)) +
    scale_color_manual(values = c(
      "Null Distribution" = "blue",
      "Sampling Distribution of the Sample Mean" = "darkgreen",
      "Estimated Population Distribution" = "green3"
    ))

  p
}

```

## Brief review of Experiments

A controlled experiment is a structured method of data collection designed to support inferences about causality. Experiments involve deliberately manipulating one condition while holding others constant, then measuring the response. Every experiment has two core components: the **independent variable**, the condition under experimental control, and the **dependent variable**, what is measured. Causal forces produce detectable change. If the independent variable truly affects the dependent variable, then a well-designed experiment should detect that change.

The central question becomes: how do we know if a measured difference between conditions reflects a real causal effect? Data always contain variability; measurements fluctuate even when nothing has changed. To decide whether a difference is real, we need tools that can distinguish meaningful change from background noise.

## Is there a difference?

Back to experiments. We want to know if an independent variable (our manipulation) causes a change in a dependent variable (measurement); if it does, we should see differences in the measurement as a function of the manipulation.

Some manipulations are so overwhelming that no statistics are required to see them. Flip a light switch up and the light comes on; flip it down and it goes off, every time. You don't need a $t$-test to tell you that flipping the light switch is the cause of the change in brightness.

Nothing in environmental science behaves like this. Effects are real signals buried in noisy, variable measurements, where flipping the switch changes the outcome *on average*. Not in every instance, and not by the same amount every time.

### Chance can produce differences

Random chance can produce apparent differences between groups even when no real difference exists. We have already seen this with sampling: two samples from the same distribution come out differently. The same principle applies when we compare group means. This is a fundamental problem for inference.

Consider a scenario where we expect to find no real difference. A researcher surveys 10 caves for white-nose syndrome, a fungal disease that has killed millions of North American bats, capturing 20 bats at each cave and counting how many show visible signs of infection. @fig-4batdesign lays out the design.

```{r}
#| label: fig-4batdesign
#| fig-cap: "The bat-cave scenario study design. The split into two groups is arbitrary, which is the whole point: there is no real difference for the test to find."
#| fig-alt: "Study design schematic. IV: group assignment, arbitrary. Two boxes labelled Group A and Group B, each with n = 5 caves. DV: infected bats out of 20 captured."
#| fig-asp: 0.34
#| out-width: "100%"

design_groups("Group assignment (arbitrary)", c("Group A", "Group B"), 5, "caves",
              "Infected bats (of 20 captured)",
              "With no real difference, what can chance alone produce?")
```

The key feature is that all 10 caves are drawn from the same population. There is no treatment, no environmental gradient, nothing systematic separating them, so the true infection rate is identical everywhere. Splitting them into two groups of 5 is arbitrary. Any difference we see between those groups therefore *has* to be sampling error, which makes this the cleanest possible look at what chance alone can manufacture.

Here is data from one run of this scenario:

```{r,echo=F}
group <- rep(c("A", "B"), each = 5)
infected <- rbinom(10, size = 20, prob = 0.40)
df <- data.frame(group, infected)

knitr::kable(df, col.names = c("Group", "Infected bats (of 20)"))
```

Each row in the table is one site. The first column shows the group assignment, and the second shows how many of the 20 captured bats were infected. Did Group A have more infected bats on average than Group B?

It is easier to see the comparison in a bar graph (@fig-4gumballA), which shows the mean count for each group.

```{r}
#| label: fig-4gumballA
#| fig-cap: "Mean number of infected bats per site for two groups drawn from the same population. Note the zoomed y-axis, which does not start at 0."
#| fig-alt: "Bar chart with two bars, Group A and Group B, showing the mean number of infected bats. The y-axis is zoomed to run from 6 to 10, so a modest difference between the two bars looks larger than it would on a full 0-to-20 scale."

mean_df <- aggregate(infected ~ group, df, mean)
ggplot(mean_df, aes(x = group, y = infected)) +
  xlab("Group") +
  ylab("Mean infected bats") +
  geom_bar(stat = "identity") +
  coord_cartesian(ylim = c(6, 10)) +
  theme_classic(base_size = 20)

```

The bars are not the same height: Group A and Group B have different mean counts. But both groups were drawn from the very same population of sites, so this isn't a real difference in infection risk, it's sampling error.

This is the core inference problem: differences appear even when nothing real caused them. **How can we tell whether an observed difference is real or just the result of chance?**

### Differences due to chance can be simulated

Chapter 2 showed that chance alone can produce spurious correlations. The same is true for group means, and it can be pinned down: run the same arbitrary split many times and record what comes out.

The first step is to replicate the sampling scenario many times. @fig-4gumballsims shows bar graphs of the mean infected count for Group A and Group B across 10 replications of the same survey: all 10 sites sampled from the same population, split 5+5, and means computed.

```{r}
#| label: fig-4gumballsims
#| fig-cap: "Ten replications of surveying 10 caves and splitting them arbitrarily into two groups of 5. Every difference here is chance alone."
#| fig-alt: "Ten small bar-chart panels, each with two bars for Group A and Group B. The two bars are a similar height in most panels but noticeably uneven in others, with no consistent pattern of which group is higher from one panel to the next."
#| out-width: "100%"

group <- rep(rep(c("A", "B"), each = 5), 10)
experiment <- rep(1:10, each = 10)
infected <- rbinom(10 * 10, size = 20, prob = 0.40)
df <- data.frame(experiment, group, infected)
mean_df <- aggregate(infected ~ experiment * group, df, mean)

ggplot(mean_df, aes(x = group, y = infected)) +
  geom_bar(stat = "identity") +
  xlab("Group") +
  ylab("Mean infected bats") +
  facet_wrap(~experiment, nrow = 2)

```

These 10 replications show that chance produces different mean differences each time. Sometimes Group A is higher, sometimes Group B is higher, sometimes the means are nearly equal. All of this variation arises from sampling error alone: there is no real difference in the population.

## Chance makes some differences more likely than others

Ten replications show that chance produces *a* difference every time, but not how big those differences usually get. For that we need a summary we can pile up. We summarize each replication as a single **difference score**: the mean count for Group A minus the mean count for Group B. @fig-4gumballdiffs shows those difference scores for the 10 replications above.

```{r}
#| label: fig-4gumballdiffs
#| fig-cap: "Difference in mean infected count between Group A and Group B across 10 replications. The true difference is zero, but sampling error produces non-zero differences each time."
#| fig-alt: "Bar chart of the A-minus-B difference score for each of 10 replications, with a horizontal reference line at zero. Bars extend above and below the line by varying amounts, some replications favoring Group A, others favoring Group B, none of them exactly zero."

differences <- mean_df[mean_df$group == "A", ]$infected -
  mean_df[mean_df$group == "B", ]$infected

dif_df <- data.frame(experiment = c(1:10), differences)
dif_df$experiment <- as.factor(dif_df$experiment)

ggplot(dif_df, aes(y = differences, x = experiment)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0) +
  theme_classic(base_size = 18) +
  xlab("Replication") +
  ylab("Difference in mean count (A - B)")
```

A bar at zero means the two groups had the same mean count in that replication. Positive values indicate Group A was higher; negative values indicate Group B was higher.

The differences in these 10 replications mostly fall in a narrow range. To better characterize what chance can do, we run 100 replications. The results are shown in @fig-4manydiffs.

```{r}
#| label: fig-4manydiffs
#| fig-cap: "Difference in mean infected count between Group A and Group B across 100 replications. There are many possible differences that chance can produce, but the y-axis shows the range is bounded."
#| fig-alt: "Bar chart of the A-minus-B difference score for 100 replications, indexed along the x-axis. The bars vary in height and sign but stay within a bounded band around zero, with none reaching far beyond the rest."

group <- rep(rep(c("A", "B"), each = 5), 100)
experiment <- rep(1:100, each = 10)
infected <- rbinom(10 * 100, size = 20, prob = 0.40)
df <- data.frame(experiment, group, infected)
mean_df <- aggregate(infected ~ experiment * group, df, mean)

differences <- mean_df[mean_df$group == "A", ]$infected -
  mean_df[mean_df$group == "B", ]$infected

dif_df <- data.frame(experiment = c(1:100), differences)
dif_df$experiment <- as.factor(dif_df$experiment)

ggplot(dif_df, aes(y = differences, x = experiment)) +
  geom_bar(stat = "identity", color = "white", linewidth = 0.2) +
  scale_x_discrete(breaks = c(1, 25, 50, 75, 100)) +
  theme_classic(base_size = 16) +
  xlab("Replication") +
  ylab("Difference in mean count (A - B)")
```

The x-axis just indexes the 100 replications; the information is on the y-axis. Many different difference sizes occur, but the range is bounded: the bars stay within a limited band around zero, and differences several times larger never show up. This begins to define the window of what chance can produce.

To get an even clearer picture, we now run 10,000 replications and plot the distribution of difference scores as a histogram. This gives a precise view of which differences happen often and which are extremely rare.

```{r}
#| label: fig-4histdiffgumball
#| fig-cap: "Histogram of mean count differences between Group A and Group B over 10,000 replications. All differences arise from chance alone. The most frequent difference is near zero; larger differences occur less often. Chance cannot produce everything."
#| fig-alt: "Histogram of 10,000 A-minus-B difference scores, tallest at zero in the center and tapering off symmetrically toward both larger positive and larger negative differences, with almost nothing beyond roughly plus or minus 3."

group <- rep(rep(c("A", "B"), each = 5), 10000)
experiment <- rep(1:10000, each = 10)
infected <- rbinom(10 * 10000, size = 20, prob = 0.40)
df <- data.frame(experiment, group, infected)
mean_df <- aggregate(infected ~ experiment * group, df, mean)

differences <- mean_df[mean_df$group == "A", ]$infected -
  mean_df[mean_df$group == "B", ]$infected

ggplot(data.frame(differences), aes(x = differences)) +
  geom_histogram(bins = 50, color = "white", linewidth = 0.2, binwidth = 1) +
  labs(x = "Difference in mean count (A - B)", y = "Count")
```

This histogram is the **chance window** for this study design. It shows that chance most often produces a difference near zero, the tallest bar in the center. Larger differences in either direction occur less and less frequently. We can see that differences larger than +/- 3-ish are possible, but very rare.

We can use this window to evaluate any observed difference. If a study found a difference of 1 infected bat between two groups, that falls well inside the chance window: differences of that size occur frequently by random sampling alone. If the study found a difference of 6 bats, that would fall far outside the window; chance produced a difference that large 0 times out of 10,000. When an observed result almost never happens by chance, we have grounds to conclude that *something other than chance* produced it.

## Making a decision from the window

The window tells us what chance can do. It does not tell us where to draw the line between "chance could have" and "chance didn't." That judgment is next, and it is a judgment, not a calculation.

::: {.callout-caution title="Warning: Beyond ruling out chance"}
Every hypothesis test in this book asks one question: could chance alone have produced this result? It is not the only question available.

Other methods start from two or more candidate explanations, say a steady linear decline versus a sharp threshold response, and ask which one the data support better.

Ruling out chance tells you that *something* is going on, but not always *what*. It is worth knowing the difference, because "significant" only ever answers the first question.
:::

Every formal test in this book ends up drawing the same line. It just does it with a formula instead of ten thousand reshuffles.

### How rare is rare enough?

**Question**: How many times does something need to happen for it to happen "a lot"? How rare does something need to be before you stop worrying about it?

Environmental scientists already have a working vocabulary for this: the **100-year flood**. A "100-year flood" doesn't mean floods that size happen once a century on a schedule; it means that in any given year, there's about a 1-in-100 chance of a flood that severe. Would you build a house in that floodplain? Many people do, and often regret it, because a 1% annual chance still adds up over a 30-year mortgage (roughly a 26% chance of at least one such flood in that span). A location with a 1-in-10,000 annual flood risk is very different: over the same 30 years, the cumulative chance is under 0.3%.

That's the intuition we need going forward: **rare events still happen, but how rare something is changes what you're willing to conclude from seeing it.** Later in this chapter, we'll use 1-in-10,000 as our own working definition of "rare enough to call surprising."

### Drawing the line: chance vs. not chance

We already have everything we need to make this concrete: the bat simulation above gave us 10,000 mean differences that chance alone can produce when there is no real difference between groups. That histogram (@fig-4histdiffgumball) *is* our reference distribution. Now we use it to make a decision.

Let's draw some lines on that same histogram and label two kinds of regions:

1.  **Region of chance.** Chance did it.

2.  **Region of not chance.** Chance didn't do it.

The boundaries are the minimum and maximum values chance actually produced across the 10,000 replications. @fig-4simdecision shows this applied to our count differences.

```{r}
#| label: fig-4simdecision
#| fig-cap: "Decision boundaries applied to the histogram of mean count differences from @fig-4histdiffgumball. Grey marks the range chance did produce across 10,000 replications; gold marks the edges of that range, which chance reached only rarely; red marks the differences chance never produced at all."
#| fig-alt: "The same histogram of 10,000 difference scores, now overlaid with colored zones. A grey band across the middle, labeled CHANCE, covers the full range chance actually produced. Narrow gold slivers sit at the two edges of that grey band, and red zones beyond them on both sides, labeled NOT CHANCE, mark differences chance never produced in any of the 10,000 replications."

diff_min <- min(differences)
diff_max <- max(differences)
pad <- (diff_max - diff_min) * 0.6
grey_width <- (diff_max - diff_min) * 0.15

ggplot(data.frame(differences), aes(x = differences)) +
  annotate("rect", xmin = diff_min, xmax = diff_max, ymin = 0, ymax = Inf,
           alpha = 0.35, fill = "grey75") +
  annotate("rect", xmin = diff_min, xmax = diff_min + grey_width, ymin = 0, ymax = Inf,
           alpha = 0.5, fill = "#f5d547") +
  annotate("rect", xmin = diff_max - grey_width, xmax = diff_max, ymin = 0, ymax = Inf,
           alpha = 0.5, fill = "#f5d547") +
  annotate("rect", xmin = -Inf, xmax = diff_min, ymin = 0, ymax = Inf,
           alpha = 0.35, fill = "#b22222") +
  annotate("rect", xmin = diff_max, xmax = Inf, ymin = 0, ymax = Inf,
           alpha = 0.35, fill = "#b22222") +
  geom_histogram(binwidth = 0.2, center = 0, color = "white") +
  geom_vline(xintercept = c(diff_min, diff_max)) +
  coord_cartesian(xlim = c(diff_min - pad, diff_max + pad)) +
  annotate("text", x = 0, y = Inf, vjust = 2, fontface = "bold", label = "CHANCE") +
  annotate("text", x = diff_min - pad * 0.4, y = Inf, vjust = 2, fontface = "bold", label = "NOT\nCHANCE") +
  annotate("text", x = diff_max + pad * 0.4, y = Inf, vjust = 2, fontface = "bold", label = "NOT\nCHANCE") +
  xlab("Difference in mean count (Group A - Group B)") +
  ylab("Count")

```

Here's how the decision works. The grey band in the middle is the chance window: our 10,000 simulated replications produced differences from roughly `r round(diff_min,1)` to `r round(diff_max,1)` bats, and a result sitting well inside it, close to 0, is exactly what chance turns out routinely. No grounds for suspicion. The red zones on either side are the ones chance never reached, so a difference out there would give real grounds to suspect something beyond sampling error. This is the region where you would reject chance as the explanation.

The gold bands are the honest part: differences *near the edge* of what chance can do, but not quite past it. Chance *could* produce something in the gold zone, it is technically still inside the window, but it does so rarely. That's not a clean yes/no; it's a "probably not chance, but I can't rule it out," and how you handle that ambiguity is a judgment call. More conservative researchers would shrink the gold zone (fewer false alarms, more missed real effects); more permissive researchers would expand it (the reverse trade-off). All of science formalizes this trade-off with a standard convention ($\alpha = .05$), which we'll discuss more later in this chapter, so that researchers aren't each inventing their own boundary from scratch.

## From simulation to formal hypothesis tests

Everything so far was built by brute force: split the sites into groups at random ten thousand times, collect the difference each time, and look at where a real result falls against that pile. That works, and it makes the logic visible, but nobody wants to rebuild a chance window from scratch for every question they ask.

There is a shortcut, and you already have most of the pieces. In Chapter 3 we established that a sample comes from a population, that the sample mean is an estimate of that population's mean, and that sample means have a predictable distribution of their own with a known spread. Put those together and the simulated chance window can be replaced by a curve whose shape we already know. The logic does not change. Only the bookkeeping does.

## The four distributions

Hypothesis testing juggles four distributions at once, which is the main reason it feels harder than it is. Three of them you have already met. Only one is new.

**Already familiar, from Chapter 3:**

- **True population distribution.** The actual distribution the data came from. We never observe it. Chapter 3 called this the population distribution, and everything about estimation assumed it was out there.
- **Estimated population distribution.** What the sample suggests that population looks like. This is estimation from Chapter 3, viewed as a whole distribution rather than a single parameter. In a test it stands in for $H_a$.
- **Sampling distribution of the sample mean.** What you would get if you drew many samples and computed $\bar{x}$ each time. Chapter 3 built this one in detail and gave it a spread, the standard error, $\sigma/\sqrt{n}$.

**New here:**

- **Null distribution.** The population distribution we would have *if $H_0$ were true*. This is the formal version of the chance window we just simulated: instead of generating the differences chance can produce, we assume a distribution and read them off it.

That is the whole change. Testing does not add three new objects; it adds one, and then asks how the sample looks against it.

**Null value notation.** We write the hypothesized population mean as $\mu_0$. Hypotheses compare the population mean $\mu$ to this constant. The sample mean $\bar{x}$ is never part of $H_0$ or $H_a$; it is the evidence we bring to test them.

### Seeing all four at once

The four are much easier to hold onto as a picture than as a list. Let's build that picture one curve at a time, ending at @fig-4key.keyfig, which is the legend for every graph after this point.

**1. Null vs. true population.**

In an idealized world where $H_0$ is exactly true, the true population (orange) and the null distribution (blue) are identical. In reality, we never know the orange curve. It's shown here for illustration.

```{r}
#| label: fig-4key.null_identical_sample
#| fig-cap: 'True population distribution equals the null distribution ($H_0$ true): $\mu=\mu_0$.'
#| fig-alt: "A single bell-shaped curve, since the orange true population curve and the blue null distribution curve sit exactly on top of each other, both centered at the same point marked with a dashed vertical line. Text reads H sub 0 colon mu equals mu sub 0."
#| out-width: "100%"

# Define the data and functions
x_vals <- seq(-4.5, 4.5, by=0.01)
h0 <- dnorm(x_vals, mean=0, sd=1)

# Plot: Two identical distributions
p3 <- ggplot() + 
  geom_line(aes(x_vals, h0, color="True Population Distribution"), linetype="solid", size=4, alpha=1) +
  geom_line(aes(x_vals, h0, color="Null Distribution"), linetype="solid", size=1) +
  annotate("text", x = -2.5, y = 0.35, label = expression("H"[0] ~ ": " ~ mu == mu[0]), size = 5) +
  geom_vline(aes(xintercept=0), linetype="dashed", color="magenta", size=0.5) +
scale_color_manual(values=c("Null Distribution"="blue", "True Population Distribution"="orange"))+
  theme_classic() + 
  theme(legend.position="none", axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank())

print(p3)
```

**2. Adding the sampling distribution.** We never know the true parent distribution directly. What we do have is a sample, and from that we can construct the **sampling distribution of the mean** (dark-green dot-dash). This distribution underlies our ability to estimate a CI (dark green horizontal line).

```{r}
#| label: fig-4key.perfectly_representative_sample
#| fig-cap: 'Perfectly representative sample: estimated population, null, and true population distributions align; the CI for $\bar{x}$ includes $\mu_0$.'
#| fig-alt: "Four curves, all nearly stacked on top of one another near the same center: the blue null distribution, the orange true population, the light-green dashed estimated population, and the narrow dark-green sampling distribution of the mean. A thick green confidence interval bar sits directly under this shared center, comfortably covering the null mean."
#| out-width: "100%"

p4 <-  plot_distributions(null_mean=0.5, null_sd=1, sample_mean=0.52, sample_sd=0.3, inferred_mean=0.52, inferred_sd=1, true_mean=0.54, true_sd=1)
print(p4)

```

**3. Adding the estimated population distribution.**

From our sample we also infer an **estimated population distribution** (light-green dashed). In the best case, this lines up closely with the true population (orange). The CI bar shows that the sample mean is consistent with $\mu_0$.

```{r}
#| label: fig-4key.sample_imperfect_convenient
#| fig-cap: 'Imperfect but convenient sample: estimated population close to the null; the two-sided CI for $\bar{x}$ includes $\mu_0$ (fail to reject $H_0$).'
#| fig-alt: "Four curves: the blue null distribution sits somewhat left of the orange true population, light-green dashed estimated population, and dark-green sampling distribution, which are clustered together to its right. Despite that offset, the thick green confidence interval bar under the sampling distribution is wide enough to still reach back and include the null mean, marked with a dashed vertical line."
#| out-width: "100%"

p7 <- plot_distributions(null_mean=-0.2, null_sd=1, sample_mean=0.5, sample_sd=0.5, inferred_mean=0.5, inferred_sd=1, true_mean=0.6, true_sd=1)

print(p7)

```

All four together give the figure you will see variations on for the rest of the book.

Key reminder: $\mu_0$ (pink dashed line) comes from the null hypothesis. The unknown $\mu$ is what we're trying to learn about, and we approximate it from the sample through the estimated population distribution and the sampling distribution of $\bar{x}$.

```{r}
#| label: fig-4key.keyfig
#| fig-cap: 'Key elements of hypothesis testing graphs: null (blue), true population (orange), estimated population (light-green dashed), sampling distribution of $\bar{x}$ (dark-green dot-dash), null mean $\mu_0$ (magenta dashed), and CI for $\bar{x}$ (thick dark-green bar).'
#| fig-alt: "The reference legend figure: a blue null distribution on the left, a dashed vertical line marking its mean mu sub 0, and to the right a cluster of three curves close together, the orange true population, the light-green dashed estimated population, and the narrow dark-green dot-dash sampling distribution of the mean, with a thick dark-green confidence interval bar beneath them, clearly separated from the null distribution's mean."
#| out-width: "100%"

pkey <-  plot_distributions(null_mean=-0.5, null_sd=1, sample_mean=0.5, sample_sd=0.3, inferred_mean=0.5, inferred_sd=1, true_mean=0.75, true_sd=1)
print(pkey)
```

To summarize, here is what each element means in every figure from here on:

- **Blue solid line** = Null distribution ($H_0$).
- **Orange solid line** = True population distribution (what actually generated the data).
- **Light green dashed line** = Estimated population distribution (inferred from the sample; our stand-in for $H_a$).
- **Dark green dot-dashed line** = Sampling distribution of the sample mean, $\bar{x}$.
- **Magenta dashed line** = Hypothesized mean $\mu_0$ (under $H_0$).\
- **Thick dark green bar** = 95% confidence interval (CI) around the sample mean $\bar{x}$.

## Terms for Decision Rules

When running a hypothesis test, three terms guide our decision:

**Significance level (**$\alpha$). Chosen *before* the test. $\alpha$ is the long-run probability of a Type I error (rejecting a true $H_0$; more on error types next chapter). A common choice is $\alpha = 0.05$. Smaller $\alpha$ makes tests more cautious and produces wider confidence intervals.

$p$-value. Assuming $H_0$ is true, the $p$-value is the probability of seeing a test statistic at least as extreme as what we observed. Small $p$ means our data look unusual under $H_0$.

**Confidence level.** Confidence level is the long-run proportion of confidence intervals that contain the true parameter when we repeat the entire data collection and interval procedure.

$$\text{Confidence Level} = 1 - \alpha$$

With $\alpha=0.05$, about 95% of intervals constructed this way will capture the true value.

::: {.callout-caution title="Warning: Don't misread a confidence interval"}
Do **not** say:

> "There is a 95% probability this interval contains $\mu$."

Instead say:

> "We used a procedure that captures $\mu$ about 95% of the time in the long run."
:::

## Three equivalent ways to test $H_0$

All three routes below run on the same quantity: a **test statistic**. It is worth saying plainly what one is before using the term.

A test statistic is a ratio. On top goes the effect you observed, the distance between what you measured and what $H_0$ predicted. On the bottom goes an estimate of how much that distance could move around by chance alone:

$$\text{statistic}=\frac{\text{effect}}{\text{error}}$$

That is the whole idea, and every test in this book is a version of it. A difference of 2 units is not impressive or unimpressive on its own. It is impressive only relative to how much difference chance produces for a study of this size, which is exactly what the chance window measured earlier in this chapter. The test statistic is that comparison written as a single number: how many units of ordinary noise is this effect worth?

Large statistic means the effect is big compared to the noise. Small statistic means it is not. Everything after this is bookkeeping about where to draw the line.

1.  **Test statistic / critical values.** Choose $\alpha$, find the corresponding **critical value(s)** for your test statistic, and compute your observed statistic. Reject $H_0$ if the statistic falls in the **rejection region**.

2.  **Confidence interval (two-sided).** Build a $(1-\alpha)$ CI for the parameter. Reject $H_0$ if $\mu_0$ lies **outside** the CI; fail to reject if it lies **inside**.

3.  $p$-value (probability). Compute the $p$-value for your observed statistic under $H_0$ and compare to $\alpha$. Reject $H_0$ if $p<\alpha$.

**Note.** Methods 1--2 yield a yes/no at $\alpha$; method 3 yields a yes/no plus a graded measure of evidence (exact $p$).

### Test statistic/critical values

@fig-rejection-region shows how these ideas come together in a two-sided test. The blue curve is the sampling distribution of the test statistic if the null hypothesis $H_0$ is true. The dashed magenta line marks the expected center at $\mu_0$. The red vertical lines are the **critical values** for $\alpha = 0.05$. They cut off the most extreme 5% of the distribution: 2.5% in each tail. Those shaded tails are the **rejection regions**.

Here's the decision rule:

- If the observed test statistic falls inside the central region, we **fail to reject** $H_0$ because the result is consistent with chance variation.
- If it falls in either rejection region, we **reject** $H_0$ because the result is too extreme to attribute to chance alone at the 5% level.

This picture makes clear why $\alpha$ and the $p$-value are linked. $\alpha$ sets the cutoff for how extreme is "too extreme," while the $p$-value tells us how far into the tails our actual result lies.

```{r}
#| label: fig-rejection-region
#| fig-cap: 'Two-sided test: blue curve is the sampling distribution under $H_0$ (centered at $\mu_0$). Magenta dashed line marks $\mu_0$. Red vertical lines mark the critical values for $\alpha=0.05$; shaded tails are the rejection regions.'
#| fig-alt: "A blue bell curve centered at mu sub 0, marked with a dashed vertical line at its center. Two red vertical lines sit symmetrically in the tails, each labeled critical value, with the shaded area beyond each one labeled rejection region. The large middle portion of the curve between the two red lines is unshaded."
#| fig-width: 7
#| fig-height: 5

library(tidyverse)

# Parameters (edit alpha if needed)
mu0   <- 0
sigma <- 1
alpha <- 0.05
crit  <- qnorm(1 - alpha/2, mean = mu0, sd = sigma)
# Curve data
df <- tibble::tibble(
  x = seq(mu0 - 4*sigma, mu0 + 4*sigma, length.out = 2000),
  y = dnorm(x, mean = mu0, sd = sigma)
)

y_peak <- dnorm(mu0, mu0, sigma)

ggplot(df, aes(x, y)) +
  # Shade rejection regions (now mapping x and y explicitly)
  geom_area(
    data = df %>% filter(x <= -crit),
    aes(x = x, y = y),
    inherit.aes = FALSE, fill = "red", alpha = 0.15
  ) +
  geom_area(
    data = df %>% filter(x >=  crit),
    aes(x = x, y = y),
    inherit.aes = FALSE, fill = "red", alpha = 0.15
  ) +
  # Blue normal curve
  geom_line(color = "blue", linewidth = 1.2) +
  # Magenta dashed line at mu0 (use annotate to avoid recycling warning)
  annotate("segment",
           x = mu0, xend = mu0, y = 0, yend = y_peak,
           colour = "magenta", linetype = "dashed", linewidth = 1) +
  # Red critical value lines
  geom_vline(xintercept = c(-crit, crit), color = "red", linewidth = 0.9) +
  # Labels
  annotate("text", x =  crit, y = y_peak*0.95,
           label = "critical value", color = "black", hjust = -0.05, size = 3.6) +
  annotate("text", x = -crit, y = y_peak*0.95,
           label = "critical value", color = "black", hjust =  1.05, size = 3.6) +
  annotate("text", x = mu0 + 2.6*sigma, y = dnorm(crit, mu0, sigma)*0.55,
           label = "rejection region", color = "black", size = 3.6) +
  annotate("text", x = mu0 - 2.6*sigma, y = dnorm(crit, mu0, sigma)*0.55,
           label = "rejection region", color = "black", size = 3.6) +
  labs(x = "x", y = "Density") +
  coord_cartesian(ylim = c(0, y_peak*1.05)) +
  theme_classic(base_size = 12)

```

### Confidence intervals

@fig-ci-lens-illustrative shows another way to make the same decision. The **blue curve** is the null distribution ($H_0$ true), and the **dark green curve** is the sampling distribution of the mean based on our sample. The **dashed green curve** is the estimated population distribution we infer from the sample.

The **thick dark green bar** at the bottom is the 95% confidence interval around the sample mean $\bar{x}$. The **magenta dashed line** marks $\mu_0$, the hypothesized mean under $H_0$.

Decision rule:

- If $\mu_0$ lies **inside** the CI bar, we **fail to reject** $H_0$.
- If $\mu_0$ lies **outside** the CI bar, we **reject** $H_0$ at $\alpha=0.05$.

```{r}
#| label: fig-ci-lens-illustrative
#| fig-cap: 'Confidence-interval lens: null (blue), sampling distribution of $\bar{x}$ (dark green), estimated population (light-green dashed). The thick bar is the $(1-\alpha)$ CI around $\bar{x}$; the magenta dashed line marks $\mu_0$. Decision uses whether $\mu_0$ is inside or outside the CI.'
#| fig-alt: "A blue null distribution centered at mu sub 0, marked with a dashed vertical line, sitting just left of a dot-dash dark-green sampling distribution and a dashed light-green estimated population curve, both centered near the sample mean x-bar. A thick green confidence interval bar spans beneath the sampling distribution's peak and its left edge reaches back to include mu sub 0, so a text label reads 'Fail to reject at alpha = 0.05.'"
#| out-width: "100%"

library(ggplot2)
library(dplyr)
library(tibble)

# Parameters you can edit ---
null_mean    <- 0
null_sd      <- 1
sample_mean  <- 0.42   # \bar{x}
SE           <- 0.4   # standard error for sampling dist + CI
inferred_sd  <- 1      # for the light-green "estimated population"
alpha        <- 0.05

# Derived quantities
x_vals   <- seq(-6, 6, by = 0.01)
h0       <- dnorm(x_vals, mean = null_mean, sd = null_sd)
samp     <- dnorm(x_vals, mean = sample_mean, sd = SE)
inferred <- dnorm(x_vals, mean = sample_mean, sd = inferred_sd)

zcrit   <- qnorm(1 - alpha/2)
ci_lo   <- sample_mean - zcrit * SE
ci_hi   <- sample_mean + zcrit * SE
y_mu0   <- dnorm(null_mean, mean = null_mean, sd = null_sd)
inside  <- (null_mean >= ci_lo) && (null_mean <= ci_hi)

df <- tibble(x = x_vals, y_h0 = h0, y_samp = samp, y_inf = inferred)

# Helpful y-levels for placing labels
y_top   <- max(h0, samp, inferred)
y_mid   <- dnorm(sample_mean, sample_mean, SE)

ggplot() +
  # Curves
  geom_line(data = df, aes(x, y_h0),    color = "blue",      linewidth = 1.2) +
  geom_line(data = df, aes(x, y_samp),  color = "darkgreen", linewidth = 1.0, linetype = "dotdash") +
  geom_line(data = df, aes(x, y_inf),   color = "green3",    linewidth = 1.0, linetype = "dashed") +

  # mu0 reference (magenta dashed down from the peak of H0)
  annotate("segment",
           x = null_mean, xend = null_mean, y = 0, yend = y_mu0,
           colour = "magenta", linetype = "dashed", linewidth = 1) +

  # CI bar and ticks on baseline
  annotate("segment",
           x = ci_lo, xend = ci_hi, y = 0, yend = 0,
           colour = "darkgreen", linewidth = 2.8) +
  annotate("segment", x = ci_lo, xend = ci_lo, y = -0.01, yend = 0.03) +
  annotate("segment", x = ci_hi, xend = ci_hi, y = -0.01, yend = 0.03) +

  # xbar marker
  annotate("point", x = sample_mean, y = 0, size = 2) +
  annotate("text",  x = sample_mean, y = 0.06, label = expression(bar(X)), size = 4) +

  # Bracket and label for (1 - alpha) CI
  annotate("segment", x = ci_lo, xend = ci_lo, y = -0.06, yend = -0.01) +
  annotate("segment", x = ci_hi, xend = ci_hi, y = -0.06, yend = -0.01) +
  annotate("segment", x = ci_lo, xend = ci_hi, y = -0.06, yend = -0.06) +
  annotate("text",   x = (ci_lo + ci_hi)/2, y = -0.095,
           label = expression(paste("(1 - ", alpha, ") CI")), size = 3.8) +

  # Labels for the three curves (inline, color-matched)
  annotate("text", x = -3.2, y = dnorm(-3.2, null_mean, null_sd) + 0.015,
           label = "Null distribution", color = "blue", size = 3.6, hjust = 0, vjust=-0.4) +
  annotate("text", x = sample_mean + 1.6, y = dnorm(sample_mean + 1.6, sample_mean, SE) + 0.015,
           label = expression("Sampling dist. of " * bar(X)), color = "darkgreen", size = 3.6, hjust = 0, vjust=0.4) +
  annotate("text", x = sample_mean - 1.8, y = dnorm(sample_mean - 1.8, sample_mean, inferred_sd) + 0.015,
           label = "Estimated population", color = "green3", size = 3.6, hjust = 0) +

  # mu0 label
  annotate("text", x = null_mean, y = y_mu0 + 0.08, label = expression(mu[0]),
           colour = "magenta", size = 5) +

  # Decision banner
  annotate("label",
           x = sample_mean, y = y_top * 0.92,
           label = ifelse(inside,
                          paste0("Fail to reject at \u03B1 = ", alpha),
                          paste0("Reject at \u03B1 = ", alpha)),
           size = 3.7, fill = "white", label.size = 0.2) +

  # Axes and theme
  coord_cartesian(ylim = c(-0.12, y_top * 1.05)) +
  labs(x = NULL, y = NULL) +
  theme_classic(base_size = 12) +
  theme(
    axis.text  = element_blank(),
    axis.ticks = element_blank()
  )
```

For common models, a two-sided level-$\alpha$ test of $H_0:\ \mu=\mu_0$ is equivalent to checking whether $\mu_0$ lies inside the $100(1-\alpha)\%$ confidence interval. The **CI** is often the most intuitive way to see this: it shows both the decision (reject or not) *and* the range of parameter values still consistent with the data.

### p-values

The $p$-value is the probability, **under** $H_0$, of obtaining a test statistic **at least as extreme** as the one you observed. In practice, you compute it from your observed statistic using software or a table. Reject if $p<\alpha$.

```{r}
#| label: fig-pvalue-lens
#| fig-cap: 'p-value lens: compute the $p$-value from the observed statistic (vertical line) under the null.'
#| fig-alt: "A blue standard normal curve centered at zero, marked with a dashed vertical line at the center for the null value and a solid black vertical line further right labeled 'observed z' at about 1.85. Text in the corner reports a two-sided p-value of about 0.064, labeled fail to reject."
#| out-width: "100%"

library(tidyverse)

alpha <- 0.05
z_obs <- 1.85     # observed z (edit this)
p_two <- 2*pnorm(-abs(z_obs))   # two-sided p under N(0,1)

x <- tibble(
  x = seq(-4, 4, length.out = 2000),
  y = dnorm(x)
)

ggplot(x, aes(x, y)) +
  geom_line(color = "blue", linewidth = 1.2) +
  # observed statistic (black)
  geom_vline(xintercept = z_obs, color = "black", linewidth = 0.9) +
  # mu0 reference (magenta dashed at 0)
  annotate("segment", x = 0, xend = 0, y = 0, yend = dnorm(0),
           linetype = "dashed", colour = "magenta", linewidth = 1) +
  annotate("text", x = z_obs, y = dnorm(0)*0.95,
           label = expression(paste("observed ", z)),
           hjust = ifelse(z_obs > 0, -0.05, 1.05), size = 3.6) +
  annotate("text", x = 2.9, y = dnorm(0)*0.35,
           label = paste0("two-sided p ≈ ", sprintf("%.3f", p_two),
                          ifelse(p_two < alpha, "  (reject)", "  (fail to reject)")),
           size = 4.2) +
  labs(x = "z", y = "Density") +
  coord_cartesian(ylim = c(0, dnorm(0)*1.05)) +
  theme_classic(base_size = 12)
```

## One- and Two-Tailed Tests

Two-tailed tests allow for differences in *either* direction:

```{r}
#| label: fig-4key.twotailed
#| fig-cap: 'Two-sided test: $H_0:\ \mu=\mu_0$ vs $H_a:\ \mu\neq\mu_0$; estimated population distributions shifted left and right of the null.'
#| fig-alt: "Three bell curves: a solid blue null distribution in the center, and two dashed green estimated-population curves of equal height, one shifted well to its left and one shifted well to its right. Text reads H sub 0 colon mu equals mu sub 0, H sub a colon mu not equal mu sub 0."
#| out-width: "100%"

# Define the data and functions
x_vals <- seq(-6, 6, by=0.01)
ha_left <- dnorm(x_vals, mean=-3, sd=0.75)
h0 <- dnorm(x_vals, mean=0, sd=0.75)
ha_right <- dnorm(x_vals, mean=3, sd=0.75)

# Plot
p <- ggplot() + 
  geom_line(aes(x_vals, ha_left, color="Estimated Population Distribution"), linetype="dashed", size=1.2) +
  geom_line(aes(x_vals, h0, color="Null Distribution"), linetype="solid", size=1.2) +
  geom_line(aes(x_vals, ha_right, color="Estimated Population Distribution"), linetype="dashed", size=1.2) +
  annotate("text", x = 0, y = 0.62, label = expression(H[0] * ": " * mu == mu[0] ~ "   " ~ H[a] * ": " * mu != mu[0]),
  size = 5) +
  scale_color_manual(values=c("Null Distribution"="blue", "Estimated Population Distribution"="green3"))+
  theme_classic() + 
  ylim(0,0.8)+
  theme(legend.position="none", axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank()) 

print(p)

```

One-tailed tests only look in *one* direction (smaller or larger than $\mu_0$):

```{r}
#| label: fig-4key.onetailed
#| fig-cap: 'One-sided tests: $H_a:\ \mu<\mu_0$ (left) and $H_a:\ \mu>\mu_0$ (right), compared to the null.'
#| fig-alt: "Two side-by-side panels, each with a solid blue null curve and a dashed green estimated-population curve of the same height. In the left panel the green curve sits to the left of the blue one, labeled H sub a colon mu less than mu sub 0. In the right panel the green curve sits to the right instead, labeled H sub a colon mu greater than mu sub 0."
#| out-width: "100%"

# Define the data and functions for the one-tailed alternatives
x_vals <- seq(-10, 10, by=0.01)
ha_left <- dnorm(x_vals, mean=-5, sd=2)  # Adjusted standard deviation for wider curve
h0 <- dnorm(x_vals, mean=0, sd=2)       # Adjusted standard deviation for wider curve
ha_right <- dnorm(x_vals, mean=4, sd=2) # Adjusted standard deviation for wider curve

# First Plot: Ha: µ < X
p1 <- ggplot() + 
  geom_line(aes(x_vals, ha_left, color="Estimated Population Distribution"), linetype="dashed", size=1.2) +
  geom_line(aes(x_vals, h0, color="Null Distribution"), linetype="solid", size=1.2) +
    ylim(0, 0.25)+
  annotate("text", x = 0, y = 0.22, label = expression("H"[a] ~ ": " ~ mu < mu[0]), size = 5) + # adjusted y value
scale_color_manual(values=c("Null Distribution"="blue", "Estimated Population Distribution"="green3"))+
  theme_classic() + 
  theme(legend.position="none", axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank())

# Second Plot: Ha: µ > X
p2 <- ggplot() + 
  geom_line(aes(x_vals, h0, color="Null Distribution"), linetype="solid", size=1.2) +
  geom_line(aes(x_vals, ha_right, color="Estimated Population Distribution"), linetype="dashed", size=1.2) +
  annotate("text", x = 0, y = 0.22, label = expression("H"[a] ~ ": " ~ mu > mu[0]), size = 5) + # adjusted y value
scale_color_manual(values=c("Null Distribution"="blue", "Estimated Population Distribution"="green3"))+
    ylim(0, 0.25)+
  theme_classic() + 
  theme(legend.position="none", axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank())

# Plot them side by side
grid.arrange(p1, p2, ncol=2)
```

### One- vs. Two-Sided: which and why?

**Use this rule of thumb:**

- Two-sided (default): You care whether $\mu$ differs from $\mu_0$ **in either direction**.
- One-sided (only if justified): Differences in the opposite direction are **scientifically irrelevant** and **would not change your decision or action**.

Default to two-sided unless your theory genuinely rules out the other direction. There is a second reason beyond caution. A one-tailed critical value always sits closer to zero than a two-tailed one, so a one-tailed test rejects on weaker evidence. Run them habitually and you reject more often than your stated $\alpha$ implies, which is a cost the next chapter makes precise.

The practical upshot: commit to one or two tails *before* you see the data. Choosing afterward, because one of them gives the smaller $p$-value, is the version of this that gets people in trouble.

## Making the call

Test outcomes are **binary**: we either **reject** $H_0$ or **fail to reject** $H_0$. We never *accept* $H_0$, because sampling variability means we can't prove it's true, only that the data are consistent with it. Similarly, we don't formally *accept* $H_a$, but we can say the results provide **support for** $H_a$ when $H_0$ is rejected.

If the observed statistic falls in either tail beyond the critical value(s), reject $H_0$. Otherwise, fail to reject $H_0$. Report the exact $p$ and an effect size or CI.

Here are our options (for a two tailed test).

1.  **Reject the Null Hypothesis (Two-Tailed Test).** The sample mean falls beyond the critical values, in the rejection region. The confidence interval also excludes the null mean (magenta line). In this example, the mean is more than 3 standard errors from $\mu_0$, which is typically considered significant for common choices of $\alpha$.

```{r}
#| label: fig-4key.reject_2_two_tailed
#| fig-cap: 'Two-sided test, reject $H_0$: sample mean far from $\mu_0$ in either direction.'
#| fig-alt: "A blue null distribution centered at mu sub 0, marked with a dashed vertical line, well separated from a narrow dot-dash sampling distribution and dashed estimated population curve centered far to the left. A thick green confidence interval bar under those two curves sits entirely to the left of mu sub 0, not reaching it at all."
#| out-width: "100%"

plot_distributions_no_error(null_mean=0, null_sd=1, sample_mean=-1.5, sample_sd=0.3, inferred_mean=-1.5, inferred_sd=1)

```

2.  **Fail to Reject the Null Hypothesis (Two-Tailed Test).** The sample mean falls in the non-rejection region, and the confidence interval includes the null mean. In this case, the evidence is consistent with chance variation, so we fail to reject $H_0$.

```{r}
#| label: fig-4key.reject_1
#| fig-cap: 'Two-sided test, fail to reject $H_0$: sample mean not greater than critical value (within $3\,SE$ from $\mu_0$).'
#| fig-alt: "A blue null distribution centered at mu sub 0, marked with a dashed vertical line, overlapping closely with a wider dot-dash sampling distribution and dashed estimated population curve centered a short distance to its left. A thick green confidence interval bar under those two curves extends far enough to reach and include mu sub 0."
#| out-width: "100%"

plot_distributions_no_error(null_mean=0, null_sd=1, sample_mean=-0.4, sample_sd=0.7, inferred_mean=-0.4, inferred_sd=1)
```

## Chapter Summary

**Why it matters.** This chapter is the hinge between describing data and drawing conclusions from it. Everything after this point is a variation on one move: work out what chance alone could produce, then ask whether your result is something chance can comfortably explain.

**Core ideas**

- **Chance produces differences on its own.** Split one population into two arbitrary groups and their means will differ anyway. The question is never "is there a difference," it is "is this difference bigger than chance can account for."
- **The chance window is buildable.** Repeat a study thousands of times under conditions where you know nothing real is going on, collect the differences, and you have a picture of exactly what chance is capable of for a design your size.
- **Rarity is a judgment, and it needs a convention.** The 100-year flood shows the reasoning already exists outside statistics: a 1% annual chance is not zero, but it changes what you build. $\alpha = .05$ is that same judgment, standardized so researchers aren't each drawing their own line.
- **A hypothesis test is the practical shortcut.** Once you can assume a known reference distribution, the simulated window becomes a curve, and the decision becomes a calculation. The logic is unchanged.
- **Three routes, one decision.** A test statistic against a critical value, a confidence interval, and a $p$-value are three ways of asking the same question. For a given $\alpha$ they always agree.
- **Commit to one- or two-tailed in advance.** A one-tailed test is only justified when theory rules out the other direction. Choosing after seeing the data quietly raises your real error rate above the $\alpha$ you claimed.
