---
author: "Mallory L. Barnes"
---

```{r, include = FALSE}
source("global_stuff.R")
```

```{r, include=FALSE}
library(ggplot2)
library(dplyr)
library(gridExtra)
library(kableExtra)

plot_distributions <- function(null_mean, null_sd, sample_mean, sample_sd, inferred_mean, inferred_sd, true_mean, true_sd) {
  x_vals <- seq(-6, 6, by=0.01)
  
  h0 <- dnorm(x_vals, mean=null_mean, sd=null_sd)
  sampling_mean <- dnorm(x_vals, mean=sample_mean, sd=sample_sd)
  inferred_parent <- dnorm(x_vals, mean=inferred_mean, sd=inferred_sd)
  true_parent <- dnorm(x_vals, mean=true_mean, sd=true_sd)
  CI_left <- qnorm(0.025, mean=sample_mean, sd=sample_sd)
  CI_right <- qnorm(0.975, mean=sample_mean, sd=sample_sd)
  y_intersect = dnorm(null_mean, mean=null_mean, sd=null_sd)

  
  p <- ggplot() + 
    geom_line(aes(x=x_vals, y=h0, color="Null Distribution"), size=1.5) +
    geom_line(aes(x=x_vals, y=sampling_mean, color="Sampling Distribution of the Sample Mean"), linetype="dotdash", size=1) +
    geom_line(aes(x=x_vals, y=inferred_parent, color="Estimated Population Distribution"), linetype="dashed", size=1.5) +
    geom_line(aes(x=x_vals, y=true_parent, color="True Population Distribution"), size=1.5) +
    geom_segment(aes(x = CI_left, xend = CI_right, y = 0., yend = 0), color="darkgreen", size=1.5) +
    geom_segment(aes(x = null_mean, xend = null_mean, y = y_intersect, yend = 0), linetype="dashed", color="magenta", size=1) +
    theme_classic() + 
    theme(legend.position='bottom', 
          axis.text = element_blank(), 
          axis.title = element_blank(), 
          axis.ticks = element_blank(), 
          legend.text = element_text(size = 12)) + 
    guides(color=guide_legend(title=NULL, nrow=2)) +
    scale_color_manual(values=c("Null Distribution"="blue", 
                                "Sampling Distribution of the Sample Mean"="darkgreen", 
                                "Estimated Population Distribution"="green3", 
                                "True Population Distribution"="orange"))

  return(p)
}

plot_distributions_no_error <- function(
  null_mean, null_sd,
  sample_mean, sample_sd,     # sample_sd here is the SE used for the sampling dist + CI
  inferred_mean, inferred_sd,
  show_ci = TRUE
){
  x_vals <- seq(-6, 6, by = 0.01)

  h0              <- dnorm(x_vals, mean = null_mean,    sd = null_sd)
  sampling_mean   <- dnorm(x_vals, mean = sample_mean,  sd = sample_sd)
  inferred_parent <- dnorm(x_vals, mean = inferred_mean, sd = inferred_sd)

  # 95% CI for the sample mean (normal approx)
  CI_left  <- qnorm(0.025, mean = sample_mean, sd = sample_sd)
  CI_right <- qnorm(0.975, mean = sample_mean, sd = sample_sd)

  # height at mu0 for the magenta reference drop line
  y_intersect <- dnorm(null_mean, mean = null_mean, sd = null_sd)

  p <- ggplot() +
    geom_line(aes(x = x_vals, y = h0,             color = "Null Distribution"), size = 1.5) +
    geom_line(aes(x = x_vals, y = sampling_mean,  color = "Sampling Distribution of the Sample Mean"),
              linetype = "dotdash", size = 1) +
    geom_line(aes(x = x_vals, y = inferred_parent, color = "Estimated Population Distribution"),
              linetype = "dashed", size = 1.5) +
    # mu0 (magenta dashed)
    geom_segment(aes(x = null_mean, xend = null_mean, y = y_intersect, yend = 0),
                 linetype = "dashed", color = "magenta", size = 1) +
    # CI bar (thick dark green): keep this!
    { if (show_ci) geom_segment(aes(x = CI_left, xend = CI_right, y = 0, yend = 0),
                                color = "darkgreen", size = 1.8) } +
    theme_classic() +
    theme(
      legend.position = "bottom",
      axis.text = element_blank(),
      axis.title = element_blank(),
      axis.ticks = element_blank(),
      legend.text = element_text(size = 12)
    ) +
    guides(color = guide_legend(title = NULL, nrow = 2)) +
    scale_color_manual(values = c(
      "Null Distribution" = "blue",
      "Sampling Distribution of the Sample Mean" = "darkgreen",
      "Estimated Population Distribution" = "green3"
    ))

  p
}

```

# Errors, Power, and Effect Size

> *It is better to be roughly right than precisely wrong.*
>
> Attributed to John Maynard Keynes

The previous chapter built the machinery of a hypothesis test and left you with a yes or no: reject $H_0$, or fail to reject it. That decision is only part of the story.

Two questions remain. First, how often will that decision be wrong, and in which direction? Second, when we do reject, how *large* is the effect we found, and was the study ever capable of detecting it in the first place? This chapter answers both.

## Errors in Hypothesis Testing

Sometimes, even when we apply the test correctly, sampling variability can lead us to the wrong conclusion.

-   **Type I Error (**$\alpha$): Rejecting $H_0$ when it is actually true. This is like a smoke alarm going off when there's no fire: a false alarm. By convention, $\alpha$ is often set at 0.05 (5%).

-   **Type II Error (**$\beta$): Failing to reject $H_0$ when it is actually false. This is like a smoke alarm staying silent when there *is* a fire: a missed signal.

| Decision                 | If the null hypothesis is True | If the null hypothesis is False |
|---------------------|-------------------------|--------------------------|
| **Reject** $H_0$         | Type I error (prob = $\alpha$) | Correct (prob = 1 - $\beta$)    |
| **Fail to reject** $H_0$ | Correct (prob = 1 - $\alpha$)  | Type II error (prob = $\beta$)  |

> **Key Takeaway**: As $\alpha$ gets smaller, $\beta$ gets bigger, and vice-versa.

Errors happen because samples are imperfect.

-   In @fig-5key.type1_error, the true population (orange) is almost identical to the null (blue). By chance, however, the sample mean lands far away, and the CI excludes $\mu_0$. The test rejects $H_0$ even though it is true: a **Type I error**.

```{r}
#| label: fig-5key.type1_error
#| fig-cap: 'Type I error: $H_0$ is true ($\mu=\mu_0$), but the sample leads us to reject $H_0$.'
#| fig-alt: "Four overlaid normal curves: a blue null distribution and an orange true population distribution sit almost on top of each other, both centered near negative 0.5, while a dashed estimated-population curve and a dot-dash sampling-distribution curve are centered far to the right, near 2. A thick horizontal bar marking the 95 percent confidence interval sits around that same right-shifted location and does not reach back to the null mean, marked with a dashed vertical line, so the interval excludes the null value even though the true and null distributions are nearly identical."
#| out-width: "100%"

p8 <- plot_distributions(null_mean=-0.5, null_sd=1, sample_mean=2, sample_sd=0.5, inferred_mean=2, inferred_sd=1, true_mean=-0.52, true_sd=1)
print(p8)
```

-   In @fig-5key.type2_error, the true population mean (orange) is far from $\mu_0$, but the particular sample looks close to the null. The CI includes $\mu_0$, so the test fails to reject when it should reject the null: a **Type II error**.

```{r}
#| label: fig-5key.type2_error
#| fig-cap: 'Type II error: $H_0$ is false (true population mean differs from $\mu_0$), but the sample leads us to fail to reject $H_0$.'
#| fig-alt: "Four overlaid normal curves: a blue null distribution centered near negative 0.5, an orange true population distribution centered far to the right near 3, and a dashed estimated-population curve with a dot-dash sampling-distribution curve both centered near 0, much closer to the null than to the true population. A thick horizontal bar marking the 95 percent confidence interval sits near 0 and reaches back to include the null mean, marked with a dashed vertical line, so the interval fails to exclude the null value even though the true population is far away."
#| out-width: "100%"

p9 <- plot_distributions(null_mean=-0.5, null_sd=1, sample_mean=0, sample_sd=0.4, inferred_mean=0, inferred_sd=1, true_mean=3, true_sd=1)
print(p9)
```

### Effect of Sample Size on Errors

With fixed $\alpha$ and correct assumptions, the long-run Type I error rate is $\alpha$ for any $n$. **Sample size mainly affects Type II error:** with small $n$, the sampling distribution is wide, true effects are harder to detect, and $\beta$ increases (power decreases). As $n$ grows, the sampling distribution tightens, $\beta$ falls, and power rises. Apparent changes in the Type I rate at small $n$ usually reflect **assumption violations** (e.g., non-normality, poor variance estimates, multiple looks), not $n$ itself.

**Takeaway:** Increasing $n$ improves estimation precision and **power** (reduces $\beta$); **Type I error remains at** $\alpha$ when assumptions hold.

**The one-tailed cost, made precise.** Chapter 4 said that habitually running one-tailed tests makes you reject more often than your stated $\alpha$ implies. Here is what that means: a one-tailed test at $\alpha = .05$ puts the whole 5% in one tail, so if you pick the tail *after* seeing which way the data went, you have effectively run a two-tailed test at $\alpha = .10$ and doubled your Type I error rate. The test itself is not wrong. Choosing its direction from the data is.

### How to interpret $p$-values

A $p$-value is the probability, **assuming** $H_0$ is true, of obtaining a test statistic **at least as extreme** as the one we observed. We **reject** at level $\alpha$ when $p<\alpha$.

**Tails.** Match the tail(s) to $H_a$:

-   Two-sided ($H_a:\ \mu \neq \mu_0$): $p = 2\cdot P(\text{tail beyond }|z_{\text{obs}}|)$\
-   One-sided ($H_a:\ \mu > \mu_0$): $p = P(Z \ge z_{\text{obs}})$\
-   One-sided ($H_a:\ \mu < \mu_0$): $p = P(Z \le z_{\text{obs}})$

::: {.callout-caution title="Warning: Common mistakes to avoid"}
1.  $p$ is **not** the probability that $H_0$ is true.
2.  A smaller $p$ does **not** imply a larger effect; report effect size and a CI.
3.  With very **large** $n$, tiny effects can yield tiny $p$; assess practical importance.
4.  With small $n$, true effects can yield large $p$ (low power).
:::

**Takeaway.** Use $p$ for the decision, but interpret results alongside **effect size** and a **95% CI**.

### Reporting results

A results statement gives the plain-language conclusion and the key statistics in a fixed order. Aim for one clean sentence plus context.

**Template (two-sided test):** *Plain-language finding.* *(test, statistic, df or n, value of statistic, p-value, effect size/CI if relevant).*

**What to include (in order):**

1.  Test name (e.g., one-sample *z*, two-sample *t*, χ², regression).
2.  Test statistic and its degrees of freedom (or sample size for *z*).
3.  Exact $p$ (report $p<0.001$ when very small).
4.  An effect size and/or a 95% CI when available.

**Examples**

-   **One-sample *z*:** The average mercury level within 10 km of the smelter was higher than the legal limit (*z* = 2.85, *n* = 32, p = 0.004).

-   **One-sample *t* with CI:** Mean nitrate exceeded the target by 0.41 mg/L (*t*(31) = 2.6, p = 0.013, 95% CI \[0.09, 0.73\] mg/L).

-   **Two-sample *t* with effect size:** Downstream sites had higher turbidity than upstream sites (*t*(58) = 3.1, p = 0.003, Cohen's *d* = 0.80).

-   **Not significant (be explicit):** Mean canopy temperature did not differ from the baseline (*t*(47) = 1.2, p = 0.24); we fail to reject $H_0$.

**Rounding & style**

-   Round statistics to 2 decimals; $p$ to 3 decimals (use $p<0.001$ when needed).
-   Use "fail to reject $H_0$" (not "accept").
-   Match tail to hypothesis in prose (e.g., "higher than," "lower than," or "different from").

**Do / Don't**

-   **Do** pair the sentence with a short, substantive takeaway (what it means for the question).
-   **Do** include a CI or effect size when it aids interpretation.
-   **Don't** restate methods; **don't** claim proof; **don't** omit units.

> **Takeaway:** One sentence is enough: state the direction in words, then report the test, the statistic, df or *n*, the $p$-value, and an effect size or CI.

### Beyond the 0.05 cutoff: Effect-size and power

Why go beyond a yes/no at $\alpha=0.05$? Because scientists, managers, policymakers, and communities need to know **how large** an effect is and **how likely** a study is to detect it. That's what **effect size** and **power** provide.

**Working definitions**

-   **Effect size**: how big the difference or association is (in natural units or standardized units like Cohen's $d$).

-   **Power (**$1-\beta$): the probability your design will detect a real effect of a specified size at your chosen $\alpha$.

**Use them before you sample.** Decide what effect is *meaningful* for your context (e.g., a 0.3 mg/L nitrate reduction) and size your study to have adequate power to detect that effect.

### Effect size: practical vs. standardized

**What is effect size?** It's how big the difference or association is. You can express it in **original units** (practical meaning) or in **standardized units** (comparability across scales).

**Practical (original units).** Use the scale stakeholders recognize.

-   Exams: "Treatment improved scores by **5 percentage points**" (a letter-grade--sized change).
-   Time-on-task: "New training reduces practice time by **1,000 hours** out of 10,000" (a 10% reduction).

These statements are easy to interpret, but they don't travel well across contexts or measures.

**Standardized (unitless).** When units are unfamiliar or scales differ, standardize by typical variability. The most common standardized effect size metric is *Cohen's d* [@Cohen1988]. For two means with a common SD $\sigma$,

$$d = \frac{\mu_1 - \mu_2}{\sigma}$$

Now "how big?" is expressed in **SD units**.

**This is the same ratio as a test statistic, with one part swapped.** Chapter 4 gave the general shape, effect over error, and the $t$-test filled it in as

$$t = \frac{\bar{x}-\mu_0}{s/\sqrt{n}} \qquad \text{versus} \qquad d = \frac{\mu_1-\mu_2}{\sigma}$$

Same numerator: the effect. Different denominator, and that difference is the whole point. The $t$ statistic divides by the **standard error**, which shrinks as $n$ grows, so $t$ gets larger with more data even when the effect stays exactly the same. Cohen's $d$ divides by the **standard deviation**, which does not shrink with $n$, so $d$ reports the size of the effect itself.

That is why a large $t$ does not mean a large effect, and why the two numbers answer different questions. Collect enough data and you can drive $t$ as high as you like off a trivial difference. You cannot do that to $d$.

**Environmental example (biodiversity index).** Suppose an index increases from 3 to 4 after reforestation.

-   In **raw units**: +1 index point.

-   In **standardized terms**:

    -   If the SD across sites is 1.0, then $d = 1.0$: a large shift.
    -   If the SD is 10, then $d = 0.10$: a small shift.

> **Takeaway:** Report both when possible: original units for meaning (policy/action) and a standardized effect (e.g., Cohen's $d$) for comparison across studies.

**Notes for practice**

-   In real data, $\sigma$ is unknown; estimate it (pooled SD for two groups, with small-sample corrections as needed).
-   Heuristics (context matters): $d \approx 0.2$ (small), $d \approx 0.5$ (medium), $d \approx 0.8$ (large).
-   A "small" $d$ can still be important (e.g., small reductions in lead or $\mathrm{PM}_{2.5}$ can matter for health).
-   Pair effect sizes with **95% CIs** to show precision, not just magnitude.

Let's ground effect size in a concrete example. Suppose we're measuring **final exam performance** (percent correct). The class mean is 65% with a standard deviation of 5%. Group A is a control, while Group B receives some instructional treatment.

@fig-5.5effectdistsB shows four possible scenarios:

```{r}
#| label: fig-5.5effectdistsB
#| fig-cap: 'Same idea with concrete scores (mean $65$, $sd=5$): treatment shifts correspond to $d=0,\,0.5,\,1,\,2$.'
#| fig-alt: "Four panels, each showing two overlaid bell-shaped density curves for groups A and B. In the first panel the curves sit exactly on top of each other (no shift). In the next three panels group B's curve shifts progressively further right of group A's, first by a small amount, then a larger amount, and finally a shift so large the two curves barely overlap at all."
#| out-width: "100%"


X <- c(
  seq(25, 100, 1),
  seq(25, 100, 1),
  seq(25, 100, 1),
  seq(25, 100, 1),
  seq(25, 100, 1),
  seq(25, 100, 1),
  seq(25, 100, 1),
  seq(25, 100, 1)
)
Y <- c(
  dnorm(seq(25, 100, 1), 65, 5),
  dnorm(seq(25, 100, 1), 65, 5),
  dnorm(seq(25, 100, 1), 65, 5),
  dnorm(seq(25, 100, 1), 67.5, 5),
  dnorm(seq(25, 100, 1), 65, 5),
  dnorm(seq(25, 100, 1), 70, 5),
  dnorm(seq(25, 100, 1), 65, 5),
  dnorm(seq(25, 100, 1), 75, 5)
)
effect_size <-
  rep(c("65, d=0", "67.5,d=.5", "70, d=1", "75, d=2"), each = 76 * 2)
condition <- rep(rep(c("A", "B"), each = 76), 2)
df <- data.frame(effect_size,
                 condition,
                 X, Y)

ggplot(df, aes(
  x = X,
  y = Y,
  group = condition,
  color = condition
)) +
  geom_line() +
  theme_classic(base_size = 15) +
  facet_wrap( ~ effect_size) +
  xlab("values") +
  ylab("density")


```

-   **Panel 1 (**$d=0$): No effect. Both groups sample from the same distribution (mean = 65).
-   **Panel 2 (**$d=0.5$): Group B's mean shifts upward by 2.5 points (67.5).
-   **Panel 3 (**$d=1$): A 5-point shift (70).
-   **Panel 4 (**$d=2$): A large 10-point shift (75).

The takeaway is that effect size puts raw differences into a standard-deviation scale. A half--SD shift ($d=0.5$) is noticeable but modest; a two--SD shift ($d=2$) is very large. In practice, we rarely know beforehand how big a shift to expect, that's why we conduct the study.

## Power

When a real effect exists, your design needs to be sensitive enough to detect it, otherwise the test has little value. We've already seen that effects can vary in size. Now we add a second idea: study designs differ in their ability to reliably detect those effects. This sensitivity is called **statistical power**.

Power rises when:

1.  **Effect size is larger.**
2.  **Sample size (**$n$) is larger (smaller $SE$).
3.  **The test is less conservative** (larger $\alpha$).
4.  **Measurement noise is lower** (smaller $SD$).

> **Question:** "With $\alpha = 0.05$, how large must $n$ be to achieve $\ge 80\%$ power for the smallest effect that matters?"

### Simulation walkthrough

Let's see how design choices affect power with a simple simulation. Suppose we have two groups, each with $n=10$. Group A (control) is sampled from a normal distribution with mean = 10 and $SD = 5$. Group B (treatment) has a mean = 12.5, a shift of 2.5 points (Cohen's $d=0.5$), considered a medium effect.

If we simulate this experiment 1,000 times and run an independent-samples $t$-test each time, how often do we reject $H_0$ at $\alpha=0.05$?

```{r, echo=TRUE}
set.seed(2025)
p <- numeric(1000)
for(i in 1:1000){
  A <- rnorm(10, 10, 5)
  B <- rnorm(10, 12.5, 5)
  p[i] <- t.test(A,B,var.equal=TRUE)$p.value
}
```

```{r}
mean(p < 0.05)
```

This proportion is the **power** of the design: the probability of detecting the true effect in repeated experiments. With $n=10$, it's only about 20%. Even though we know a medium-sized effect exists, the test usually fails to pick it up with such a small sample.

### How power changes

1.  **Larger** $n$. Doubling to $n=20$ per group reduces sampling error and roughly doubles power:

```{r, echo=TRUE}
p <- numeric(1000)
for(i in 1:1000){
  A <- rnorm(20, 10, 5)
  B <- rnorm(20, 12.5, 5)
  p[i] <- t.test(A,B,var.equal=TRUE)$p.value
}
mean(p < 0.05)
```

2.  **Smaller** $\alpha$. Making the test more conservative (e.g., $\alpha=0.01$) lowers power, because fewer outcomes count as "significant". Using the same $n=20$ simulations, we just change the cutoff:

```{r, echo=TRUE}
mean(p < 0.01)
```

3.  **Bigger effects are easier to detect.** If the treatment produces a big shift (say 2 standard deviations), the same design has near-perfect power:

```{r, echo=TRUE}
p <- numeric(1000)
for(i in 1:1000){
  A <- rnorm(20, 10, 5)
  B <- rnorm(20, 20, 5)   # 2SD effect
  p[i] <- t.test(A,B,var.equal=TRUE)$p.value
}
```

```{r}
mean(p < 0.01)
```

### Power curves

It's best to think of power as a **profile**, not a single number. A power curve shows how sensitive a design is across a range of possible effects.

In @fig-5.5powercurve, we fix the design to a two-sample $t$-test with $n=10$ per group and $\alpha=0.05$. Power is plotted on the y-axis, and the true effect size (Cohen's $d$) is on the x-axis.

```{r}
#| label: fig-5.5powercurve
#| fig-cap: 'Power vs. effect size (Cohen’s $d$) for a two-sample $t$-test with $n=10$ per group and $\alpha=0.05$.'
#| fig-alt: "Line graph of power against effect size (Cohen's d) from 0 to 2. The curve starts near 0.05 at d = 0, rises slowly at first, then climbs steeply through the middle of the range, and flattens out near 1.0 as d approaches 2."
#| out-width: "100%"

# Exact power for a two-sample t-test (not the normal approximation):
# power.t.test uses the noncentral t, which is what the test actually uses.
effect_size <- seq(0, 2, .1)
power <- sapply(effect_size, function(d) {
  if (d == 0) return(0.05)                     # at d = 0, rejection rate is alpha
  power.t.test(n = 10, delta = d, sd = 1, sig.level = 0.05)$power
})

plot_df <- data.frame(power, effect_size)

ggplot(plot_df, aes(x=effect_size, y=power))+
  geom_line()+
  theme_classic()
```

**Walkthrough.**

-   With $n=10$, the design only detects a medium effect ($d=0.5$) about 18% of the time.
-   A larger effect ($d=0.8$) is caught about 40% of the time.
-   Very large effects ($d=2$) are detected almost every time (about 99%).

This shows that small samples are only sensitive to big, obvious effects.

Now suppose prior research suggests the effect size is small ($d=0.2$). Instead of varying the effect, we can ask: *how many subjects per group do we need to have decent power?*

@fig-5.5powercurveN plots power against sample size for detecting $d=0.2$. The green line marks 80% power, a common planning target.

```{r}
#| label: fig-5.5powercurveN
#| fig-cap: 'Power vs. sample size for detecting $d=0.2$ with a two-sample $t$-test at $\alpha=0.05$; green line marks $0.80$ power.'
#| fig-alt: "Line graph of power against sample size per group, from 10 to 800. Power starts near zero at small sample sizes and rises in a curve that flattens as it approaches 1.0. A horizontal reference line at 0.80 power crosses the curve around a sample size of roughly 394 per group."
#| out-width: "100%"

n_seq <- seq(10, 800, 10)
power <- sapply(n_seq, function(n)
  power.t.test(n = n, delta = 0.2, sd = 1, sig.level = 0.05)$power)

plot_df <- data.frame(power, n_seq)

ggplot(plot_df, aes(x=n_seq, y=power))+
  geom_line()+
  theme_classic()+
  geom_hline(yintercept=.8, color="green")+
  labs(x = "n per group", y = "Power")
```

**Walkthrough.**

-   At $n=10$ per group, power to detect $d=0.2$ is near zero.
-   To reach 80% power, we need about **394 per group**.
-   Even then, 20% of studies would still fail to reject $H_0$ just by chance.

**Takeaway:**

-   A power curve shows, for a fixed design, how power rises as the true effect grows.
-   A power-vs-$n$ curve shows how many samples you need to hit a target power for a specific effect.
-   Power analysis is mainly a **planning tool**, not a results-reporting tool. When you design a study, don't just aim for "power = 0.8." Always state the **assumed effect size** and the **standard deviation** you used for that calculation. Otherwise, "0.8" has no meaning: power depends entirely on those assumptions.

### Planning your design

Our discussion of effect size and power highlights the importance of understanding the statistical limitations of an experimental design. Planning a study means working through the following, in order:

-   **Define a meaningful effect** (environmentally or policy-relevant minimum).
-   **Choose** $\alpha$ (often 0.05; justify if different).
-   **Estimate variability** (pilot data, literature, or historical monitoring).
-   **Compute** $n$ for $\ge 80\%$ power (or explain tradeoffs if you can't reach it).
-   **Pre-specify** one- vs two-tailed before data collection.

As a general rule of thumb, small-$n$ designs can only reliably detect very large effects, whereas large-$n$ designs can reliably detect much smaller effects. As a researcher, it is your responsibility to plan your design accordingly so that it is capable of reliably detecting the kinds of effects it is intended to measure.

### Two cautions (common pitfalls)

-   **Low power inflates doubt, even when** $p<0.05$. Underpowered studies rarely replicate. A design with only 30% power to detect an expected effect might, by luck, produce a "significant" result, but most replications will not. Worse, false positives from small samples tend to look exaggerated: if a spurious finding reaches significance, the estimated effect size is often large enough to seem convincing.

-   **Huge** $n$ finds tiny effects. With very large samples, trivial differences can register as statistically significant. For example, a billion-person trial might show a drug reduces pain by 0.01%. That effect is "real" but practically meaningless. This is why effect size and confidence intervals should be reported alongside $p$-values: significance alone is not enough.

@fig-5.5effectsizeType1 shows the first pitfall graphically. Even when the null hypothesis is true, 5% of tests are expected to fall below $\alpha = 0.05$. In small samples, those false positives tend to be paired with large apparent effect sizes. This can give the illusion of a strong, replicable finding when in reality it is just sampling noise.

```{r}
#| label: fig-5.5effectsizeType1
#| fig-cap: 'Apparent effect size among Type I errors ($p<0.05$) under the null, by sample size, for a one-sample $t$-test. Effect sizes are shown as absolute values, since a false positive is equally likely in either direction and the sign carries no information here. Small $n$ inflates apparent effects.'
#| fig-alt: "Scatter plot of apparent effect size against p-value, restricted to the false positives with p below 0.05, colored by sample size (10, 50, 100, and 1000). Points from the smallest sample size, 10, reach the highest apparent effect sizes, well above 1. Points cluster progressively lower as sample size increases, with the sample size of 1000 group sitting close to zero across the full range of p-values."
#| out-width: "100%"

all_df<-data.frame()
for(i in 1:1000){
  for(n in c(10,50,100,1000)){
    some_data<-rnorm(n,0,1)
    p_value<-t.test(some_data)$p.value
    effect_size<-mean(some_data)/sd(some_data)
    t_df<-data.frame(sim=i,sample_size=n,p_value,effect_size)
    all_df<-rbind(all_df,t_df)
  }
}

type_I_error <- all_df[all_df$p_value<.05,]
type_I_error$sample_size<-factor(type_I_error$sample_size,
                                 levels = c(10, 50, 100, 1000))

ggplot(type_I_error, aes(x = p_value, y = abs(effect_size),
                         group = sample_size, color = sample_size)) +
  geom_point(size = 1.3) +
  theme_classic(base_size = 13) +
  xlab("p-value") +
  ylab("Apparent effect size (absolute)") +
  labs(color = "Sample size")
```

**Takeaway:** statistical significance does not guarantee scientific significance. Always interpret $p$ in the context of effect size, confidence intervals, and study design.

**Key point**

Power is not a fixed property of a test; it depends entirely on assumptions about $n$, effect size, $\alpha$, and variability. Think of it as a **family of numbers**, not a single value. That's why power analysis is most useful **before** you collect data: it tells you whether your design is capable of detecting the kind of effect you care about.

## Chapter Summary

**Why it matters.** A test result on its own is a yes or a no. This chapter is about everything that yes or no leaves out: how often the decision is wrong, how large the effect actually is, and whether the study was ever built to detect it.

**Core ideas**

-   **Two ways to be wrong.** A Type I error rejects a true $H_0$, at a rate you set with $\alpha$. A Type II error fails to reject a false one. Tightening one loosens the other, which is why $\alpha = .05$ is a balance rather than a fact about nature.
-   **What a** $p$-value is not. It is not the probability that $H_0$ is true, and on its own it says nothing about how large an effect is.
-   **Significance is not size.** With a large enough $n$, a trivially small difference clears any threshold. Report an effect size and a confidence interval next to every $p$-value so a reader can judge whether the result matters, not just whether it cleared a line.
-   **Effect size in two currencies.** Original units carry practical meaning for the people who act on your result; a standardized effect like Cohen's $d$ makes results comparable across studies and scales. Report both when you can.
-   **Power is a design decision.** Power depends on $n$, effect size, $\alpha$, and variability together. An underpowered study fails twice over: it usually misses real effects, and when it does reach significance the effect it reports tends to be inflated. That is why power analysis belongs *before* data collection.
