---
title: "Hypothesis Testing"
author: "Mallory Barnes"
---

```{r, include = FALSE}
source("global_stuff.R")
```

```{r, include=FALSE}
library(ggplot2)
library(dplyr)
library(gridExtra)
library(kableExtra)

plot_distributions <- function(null_mean, null_sd, sample_mean, sample_sd, inferred_mean, inferred_sd, true_mean, true_sd) {
  x_vals <- seq(-6, 6, by=0.01)
  
  h0 <- dnorm(x_vals, mean=null_mean, sd=null_sd)
  sampling_mean <- dnorm(x_vals, mean=sample_mean, sd=sample_sd)
  inferred_parent <- dnorm(x_vals, mean=inferred_mean, sd=inferred_sd)
  true_parent <- dnorm(x_vals, mean=true_mean, sd=true_sd)
  CI_left <- qnorm(0.025, mean=sample_mean, sd=sample_sd)
  CI_right <- qnorm(0.975, mean=sample_mean, sd=sample_sd)
  y_intersect = dnorm(null_mean, mean=null_mean, sd=null_sd)

  
  p <- ggplot() + 
    geom_line(aes(x=x_vals, y=h0, color="Null Distribution"), size=1.5) +
    geom_line(aes(x=x_vals, y=sampling_mean, color="Sampling Distribution of the Sample Mean"), linetype="dotdash", size=1) +
    geom_line(aes(x=x_vals, y=inferred_parent, color="Estimated Population Distribution"), linetype="dashed", size=1.5) +
    geom_line(aes(x=x_vals, y=true_parent, color="True Population Distribution"), size=1.5) +
    geom_segment(aes(x = CI_left, xend = CI_right, y = 0., yend = 0), color="darkgreen", size=1.5) +
    geom_segment(aes(x = null_mean, xend = null_mean, y = y_intersect, yend = 0), linetype="dashed", color="magenta", size=1) +
    theme_classic() + 
    theme(legend.position='bottom', 
          axis.text = element_blank(), 
          axis.title = element_blank(), 
          axis.ticks = element_blank(), 
          legend.text = element_text(size = 12)) + 
    guides(color=guide_legend(title=NULL, nrow=2)) +
    scale_color_manual(values=c("Null Distribution"="blue", 
                                "Sampling Distribution of the Sample Mean"="darkgreen", 
                                "Estimated Population Distribution"="green3", 
                                "True Population Distribution"="orange"))

  return(p)
}

plot_distributions_no_error <- function(
  null_mean, null_sd,
  sample_mean, sample_sd,     # sample_sd here is the SE used for the sampling dist + CI
  inferred_mean, inferred_sd,
  show_ci = TRUE
){
  x_vals <- seq(-6, 6, by = 0.01)

  h0              <- dnorm(x_vals, mean = null_mean,    sd = null_sd)
  sampling_mean   <- dnorm(x_vals, mean = sample_mean,  sd = sample_sd)
  inferred_parent <- dnorm(x_vals, mean = inferred_mean, sd = inferred_sd)

  # 95% CI for the sample mean (normal approx)
  CI_left  <- qnorm(0.025, mean = sample_mean, sd = sample_sd)
  CI_right <- qnorm(0.975, mean = sample_mean, sd = sample_sd)

  # height at mu0 for the magenta reference drop line
  y_intersect <- dnorm(null_mean, mean = null_mean, sd = null_sd)

  p <- ggplot() +
    geom_line(aes(x = x_vals, y = h0,             color = "Null Distribution"), size = 1.5) +
    geom_line(aes(x = x_vals, y = sampling_mean,  color = "Sampling Distribution of the Sample Mean"),
              linetype = "dotdash", size = 1) +
    geom_line(aes(x = x_vals, y = inferred_parent, color = "Estimated Population Distribution"),
              linetype = "dashed", size = 1.5) +
    # mu0 (magenta dashed)
    geom_segment(aes(x = null_mean, xend = null_mean, y = y_intersect, yend = 0),
                 linetype = "dashed", color = "magenta", size = 1) +
    # CI bar (thick dark green): keep this!
    { if (show_ci) geom_segment(aes(x = CI_left, xend = CI_right, y = 0, yend = 0),
                                color = "darkgreen", size = 1.8) } +
    theme_classic() +
    theme(
      legend.position = "bottom",
      axis.text = element_blank(),
      axis.title = element_blank(),
      axis.ticks = element_blank(),
      legend.text = element_text(size = 12)
    ) +
    guides(color = guide_legend(title = NULL, nrow = 2)) +
    scale_color_manual(values = c(
      "Null Distribution" = "blue",
      "Sampling Distribution of the Sample Mean" = "darkgreen",
      "Estimated Population Distribution" = "green3"
    ))

  p
}

```

## What you'll learn

Hypothesis testing is the backbone of statistical inference: it lets us decide whether the patterns we see in a sample are strong enough to draw conclusions about the population. By the end of this chapter, you should be able to:

-   **Define the key players** in hypothesis testing (null, alternative, true/estimated, and population/sampling distributions).
-   **Explain and use decision rules**: decide whether to reject $H_0$.
-   **Connect confidence intervals and hypothesis tests** for common models.
-   **Identify and interpret errors** (Type I and Type II) that can arise in testing.
-   **Recognize the role of effect size and power** in planning a study, even if your main focus is running tests.
-   **Write a clear results statement** that communicates both the statistical and substantive meaning of your test.

## Core Players (distributions)

To make sense of hypothesis testing, it helps to picture four kinds of distributions. Later in the chapter we'll put them together, but first we define them one at a time:

1.  **Null Distribution** - The hypothesized population distribution under the assumption that the null hypothesis $H_0$ is true.

2.  **Estimated Population Distribution** - The distribution we infer from our sample. This represents what the data suggest about $H_a$. (Sometimes called the *inferred parent distribution*.)

3.  **True Population Distribution** - The actual distribution from which our sample was drawn. We never know this perfectly, but it's the reality we're trying to approximate.

4.  **Sampling Distribution of the Sample Mean** - The distribution we'd get if we took many samples and calculated $\bar{X}$ each time. This is key for connecting sample evidence back to the population.

**Null value notation.** We'll write the hypothesized population mean as $\mu_0$. Hypotheses compare the population mean $\mu$ to this constant $\mu_0$. The sample mean $\bar{X}$ is never part of $H_0$ or $H_a$; it's the evidence we bring to test them.

*Note on terms.* In this chapter, you'll sometimes see the phrase **parent distribution** in the prose. This is just a metaphor: a sample is like a "child" drawn from a "parent" distribution. The standard statistical term is **population distribution**. Both terms mean the same thing here. For clarity, we'll use *population distribution* in figures and definitions, and you can think of *parent* as a memory aid.

## Terms for Decision Rules

When running a hypothesis test, three terms guide our decision:

**Significance level (**$\alpha$). Chosen *before* the test. $\alpha$ is the long-run probability of a Type I error (rejecting a true $H_0$). A common choice is $\alpha = 0.05$. Smaller $\alpha$ makes tests more cautious and produces wider confidence intervals.

$p$-value. Assuming $H_0$ is true, the $p$-value is the probability of seeing a test statistic at least as extreme as what we observed. Small $p$ means our data look unusual under $H_0$.

**Confidence level.** Confidence level is the long-run proportion of confidence intervals that contain the true parameter when we repeat the entire data collection and interval procedure.

$$
\text{Confidence Level} = 1 - \alpha
$$

With $\alpha=0.05$, about 95% of intervals constructed this way will capture the true value.

It's tempting to misinterpret confidence intervals. Do **not** say:

> "There is a 95% probability this interval contains $\mu$."

Instead say:

> "We used a procedure that captures $\mu$ about 95% of the time in the long run."

## Three equivalent ways to test $H_0$

1.  **Test statistic / critical values.** Choose $\alpha$, find the corresponding **critical value(s)** for your test statistic, and compute your observed statistic. Reject $H_0$ if the statistic falls in the **rejection region**.

2.  **Confidence interval (two-sided).** Build a $(1-\alpha)$ CI for the parameter. Reject $H_0$ if $\mu_0$ lies **outside** the CI; fail to reject if it lies **inside**.

3.  $p$-value (probability). Compute the $p$-value for your observed statistic under $H_0$ and compare to $\alpha$. Reject $H_0$ if $p<\alpha$.

**Note.** Methods 1--2 yield a yes/no at $\alpha$; method 3 yields a yes/no plus a graded measure of evidence (exact $p$).

### Test statistic/critical values

@fig-rejection-region shows how these ideas come together in a two-sided test. The blue curve is the sampling distribution of the test statistic if the null hypothesis $H_0$ is true. The dashed magenta line marks the expected center at $\mu_0$. The red vertical lines are the **critical values** for $\alpha = 0.05$. They cut off the most extreme 5% of the distribution: 2.5% in each tail. Those shaded tails are the **rejection regions**.

Here's the decision rule:

-   If the observed test statistic falls inside the central region, we **fail to reject** $H_0$ because the result is consistent with chance variation.
-   If it falls in either rejection region, we **reject** $H_0$ because the result is too extreme to attribute to chance alone at the 5% level.

This picture makes clear why $\alpha$ and the $p$-value are linked. $\alpha$ sets the cutoff for how extreme is "too extreme," while the $p$-value tells us how far into the tails our actual result lies.

```{r}
#| label: fig-rejection-region
#| fig-cap: 'Two-sided test: blue curve is the sampling distribution under $H_0$ (centered at $\mu_0$). Magenta dashed line marks $\mu_0$. Red vertical lines mark the critical values for $\alpha=0.05$; shaded tails are the rejection regions.'
#| fig-width: 7
#| fig-height: 5

library(tidyverse)

# Parameters (edit alpha if needed)
mu0   <- 0
sigma <- 1
alpha <- 0.05
crit  <- qnorm(1 - alpha/2, mean = mu0, sd = sigma)
# Curve data
df <- tibble::tibble(
  x = seq(mu0 - 4*sigma, mu0 + 4*sigma, length.out = 2000),
  y = dnorm(x, mean = mu0, sd = sigma)
)

y_peak <- dnorm(mu0, mu0, sigma)

ggplot(df, aes(x, y)) +
  # Shade rejection regions (now mapping x and y explicitly)
  geom_area(
    data = df %>% filter(x <= -crit),
    aes(x = x, y = y),
    inherit.aes = FALSE, fill = "red", alpha = 0.15
  ) +
  geom_area(
    data = df %>% filter(x >=  crit),
    aes(x = x, y = y),
    inherit.aes = FALSE, fill = "red", alpha = 0.15
  ) +
  # Blue normal curve
  geom_line(color = "blue", linewidth = 1.2) +
  # Magenta dashed line at mu0 (use annotate to avoid recycling warning)
  annotate("segment",
           x = mu0, xend = mu0, y = 0, yend = y_peak,
           colour = "magenta", linetype = "dashed", linewidth = 1) +
  # Red critical value lines
  geom_vline(xintercept = c(-crit, crit), color = "red", linewidth = 0.9) +
  # Labels
  annotate("text", x =  crit, y = y_peak*0.95,
           label = "critical value", color = "black", hjust = -0.05, size = 3.6) +
  annotate("text", x = -crit, y = y_peak*0.95,
           label = "critical value", color = "black", hjust =  1.05, size = 3.6) +
  annotate("text", x = mu0 + 2.6*sigma, y = dnorm(crit, mu0, sigma)*0.55,
           label = "rejection region", color = "black", size = 3.6) +
  annotate("text", x = mu0 - 2.6*sigma, y = dnorm(crit, mu0, sigma)*0.55,
           label = "rejection region", color = "black", size = 3.6) +
  labs(x = "x", y = "Density") +
  coord_cartesian(ylim = c(0, y_peak*1.05)) +
  theme_classic(base_size = 12)

```

### Confidence intervals

@fig-ci-lens-illustrative shows another way to make the same decision. The **blue curve** is the null distribution ($H_0$ true), and the **dark green curve** is the sampling distribution of the mean based on our sample. The **dashed green curve** is the estimated population distribution we infer from the sample.

The **thick dark green bar** at the bottom is the 95% confidence interval around the sample mean $\bar{X}$. The **magenta dashed line** marks $\mu_0$, the hypothesized mean under $H_0$.

Decision rule:

-   If $\mu_0$ lies **inside** the CI bar, we **fail to reject** $H_0$.
-   If $\mu_0$ lies **outside** the CI bar, we **reject** $H_0$ at $\alpha=0.05$.

```{r}
#| label: fig-ci-lens-illustrative
#| fig-cap: 'Confidence-interval lens: null (blue), sampling distribution of $\bar{X}$ (dark green), estimated population (light-green dashed). The thick bar is the $(1-\alpha)$ CI around $\bar{X}$; the magenta dashed line marks $\mu_0$. Decision uses whether $\mu_0$ is inside or outside the CI.'
#| out-width: "100%"

library(ggplot2)
library(dplyr)
library(tibble)

# Parameters you can edit ---
null_mean    <- 0
null_sd      <- 1
sample_mean  <- 0.42   # \bar{X}
SE           <- 0.4   # standard error for sampling dist + CI
inferred_sd  <- 1      # for the light-green "estimated population"
alpha        <- 0.05

# Derived quantities
x_vals   <- seq(-6, 6, by = 0.01)
h0       <- dnorm(x_vals, mean = null_mean, sd = null_sd)
samp     <- dnorm(x_vals, mean = sample_mean, sd = SE)
inferred <- dnorm(x_vals, mean = sample_mean, sd = inferred_sd)

zcrit   <- qnorm(1 - alpha/2)
ci_lo   <- sample_mean - zcrit * SE
ci_hi   <- sample_mean + zcrit * SE
y_mu0   <- dnorm(null_mean, mean = null_mean, sd = null_sd)
inside  <- (null_mean >= ci_lo) && (null_mean <= ci_hi)

df <- tibble(x = x_vals, y_h0 = h0, y_samp = samp, y_inf = inferred)

# Helpful y-levels for placing labels
y_top   <- max(h0, samp, inferred)
y_mid   <- dnorm(sample_mean, sample_mean, SE)

ggplot() +
  # Curves
  geom_line(data = df, aes(x, y_h0),    color = "blue",      linewidth = 1.2) +
  geom_line(data = df, aes(x, y_samp),  color = "darkgreen", linewidth = 1.0, linetype = "dotdash") +
  geom_line(data = df, aes(x, y_inf),   color = "green3",    linewidth = 1.0, linetype = "dashed") +

  # mu0 reference (magenta dashed down from the peak of H0)
  annotate("segment",
           x = null_mean, xend = null_mean, y = 0, yend = y_mu0,
           colour = "magenta", linetype = "dashed", linewidth = 1) +

  # CI bar and ticks on baseline
  annotate("segment",
           x = ci_lo, xend = ci_hi, y = 0, yend = 0,
           colour = "darkgreen", linewidth = 2.8) +
  annotate("segment", x = ci_lo, xend = ci_lo, y = -0.01, yend = 0.03) +
  annotate("segment", x = ci_hi, xend = ci_hi, y = -0.01, yend = 0.03) +

  # xbar marker
  annotate("point", x = sample_mean, y = 0, size = 2) +
  annotate("text",  x = sample_mean, y = 0.06, label = expression(bar(X)), size = 4) +

  # Bracket and label for (1 - alpha) CI
  annotate("segment", x = ci_lo, xend = ci_lo, y = -0.06, yend = -0.01) +
  annotate("segment", x = ci_hi, xend = ci_hi, y = -0.06, yend = -0.01) +
  annotate("segment", x = ci_lo, xend = ci_hi, y = -0.06, yend = -0.06) +
  annotate("text",   x = (ci_lo + ci_hi)/2, y = -0.095,
           label = expression(paste("(1 - ", alpha, ") CI")), size = 3.8) +

  # Labels for the three curves (inline, color-matched)
  annotate("text", x = -3.2, y = dnorm(-3.2, null_mean, null_sd) + 0.015,
           label = "Null distribution", color = "blue", size = 3.6, hjust = 0, vjust=-0.4) +
  annotate("text", x = sample_mean + 1.6, y = dnorm(sample_mean + 1.6, sample_mean, SE) + 0.015,
           label = expression("Sampling dist. of " * bar(X)), color = "darkgreen", size = 3.6, hjust = 0, vjust=0.4) +
  annotate("text", x = sample_mean - 1.8, y = dnorm(sample_mean - 1.8, sample_mean, inferred_sd) + 0.015,
           label = "Estimated population", color = "green3", size = 3.6, hjust = 0) +

  # mu0 label
  annotate("text", x = null_mean, y = y_mu0 + 0.08, label = expression(mu[0]),
           colour = "magenta", size = 5) +

  # Decision banner
  annotate("label",
           x = sample_mean, y = y_top * 0.92,
           label = ifelse(inside,
                          paste0("Fail to reject at \u03B1 = ", alpha),
                          paste0("Reject at \u03B1 = ", alpha)),
           size = 3.7, fill = "white", label.size = 0.2) +

  # Axes and theme
  coord_cartesian(ylim = c(-0.12, y_top * 1.05)) +
  labs(x = NULL, y = NULL) +
  theme_classic(base_size = 12) +
  theme(
    axis.text  = element_blank(),
    axis.ticks = element_blank()
  )
```

For common models, a two-sided level-$\alpha$ test of $H_0:\ \mu=\mu_0$ is equivalent to checking whether $\mu_0$ lies inside the $100(1-\alpha)\%$ confidence interval. The **CI** is often the most intuitive way to see this: it shows both the decision (reject or not) *and* the range of parameter values still consistent with the data.

### p-values

The $p$-value is the probability, **under** $H_0$, of obtaining a test statistic **at least as extreme** as the one you observed. In practice, you compute it from your observed statistic using software or a table. Reject if $p<\alpha$.

```{r}
#| label: fig-pvalue-lens
#| fig-cap: 'p-value lens: compute the $p$-value from the observed statistic (vertical line) under the null.'
#| out-width: "100%"

library(tidyverse)

alpha <- 0.05
z_obs <- 1.85     # observed z (edit this)
p_two <- 2*pnorm(-abs(z_obs))   # two-sided p under N(0,1)

x <- tibble(
  x = seq(-4, 4, length.out = 2000),
  y = dnorm(x)
)

ggplot(x, aes(x, y)) +
  geom_line(color = "blue", linewidth = 1.2) +
  # observed statistic (black)
  geom_vline(xintercept = z_obs, color = "black", linewidth = 0.9) +
  # mu0 reference (magenta dashed at 0)
  annotate("segment", x = 0, xend = 0, y = 0, yend = dnorm(0),
           linetype = "dashed", colour = "magenta", linewidth = 1) +
  annotate("text", x = z_obs, y = dnorm(0)*0.95,
           label = expression(paste("observed ", z)),
           hjust = ifelse(z_obs > 0, -0.05, 1.05), size = 3.6) +
  annotate("text", x = 2.9, y = dnorm(0)*0.35,
           label = paste0("two-sided p ≈ ", sprintf("%.3f", p_two),
                          ifelse(p_two < alpha, "  (reject)", "  (fail to reject)")),
           size = 4.2) +
  labs(x = "z", y = "Density") +
  coord_cartesian(ylim = c(0, dnorm(0)*1.05)) +
  theme_classic(base_size = 12)
```

## One- and Two-Tailed Tests

Two-tailed tests allow for differences in *either* direction:

```{r}
#| label: fig-5-5.twotailed
#| fig-cap: 'Two-sided test: $H_0:\ \mu=\mu_0$ vs $H_a:\ \mu\neq\mu_0$; estimated population distributions shifted left and right of the null.'
#| out-width: "100%"

# Define the data and functions
x_vals <- seq(-6, 6, by=0.01)
ha_left <- dnorm(x_vals, mean=-3, sd=0.75)
h0 <- dnorm(x_vals, mean=0, sd=0.75)
ha_right <- dnorm(x_vals, mean=3, sd=0.75)

# Plot
p <- ggplot() + 
  geom_line(aes(x_vals, ha_left, color="Estimated Population Distribution"), linetype="dashed", size=1.2) +
  geom_line(aes(x_vals, h0, color="Null Distribution"), linetype="solid", size=1.2) +
  geom_line(aes(x_vals, ha_right, color="Estimated Population Distribution"), linetype="dashed", size=1.2) +
  annotate("text", x = 0, y = 0.62, label = expression(H[0] * ": " * mu == mu[0] ~ "   " ~ H[a] * ": " * mu != mu[0]),
  size = 5) +
  scale_color_manual(values=c("Null Distribution"="blue", "Estimated Population Distribution"="green3"))+
  theme_classic() + 
  ylim(0,0.8)+
  theme(legend.position="none", axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank()) 

print(p)

```

One-tailed tests only look in *one* direction (smaller or larger than $\mu_0$):

```{r}
#| label: fig-5-5.onetailed
#| fig-cap: 'One-sided tests: $H_a:\ \mu<\mu_0$ (left) and $H_a:\ \mu>\mu_0$ (right), compared to the null.'
#| out-width: "100%"

# Define the data and functions for the one-tailed alternatives
x_vals <- seq(-10, 10, by=0.01)
ha_left <- dnorm(x_vals, mean=-5, sd=2)  # Adjusted standard deviation for wider curve
h0 <- dnorm(x_vals, mean=0, sd=2)       # Adjusted standard deviation for wider curve
ha_right <- dnorm(x_vals, mean=4, sd=2) # Adjusted standard deviation for wider curve

# First Plot: Ha: µ < X
p1 <- ggplot() + 
  geom_line(aes(x_vals, ha_left, color="Estimated Population Distribution"), linetype="dashed", size=1.2) +
  geom_line(aes(x_vals, h0, color="Null Distribution"), linetype="solid", size=1.2) +
    ylim(0, 0.25)+
  annotate("text", x = 0, y = 0.22, label = expression("H"[a] ~ ": " ~ mu < mu[0]), size = 5) + # adjusted y value
scale_color_manual(values=c("Null Distribution"="blue", "Estimated Population Distribution"="green3"))+
  theme_classic() + 
  theme(legend.position="none", axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank())

# Second Plot: Ha: µ > X
p2 <- ggplot() + 
  geom_line(aes(x_vals, h0, color="Null Distribution"), linetype="solid", size=1.2) +
  geom_line(aes(x_vals, ha_right, color="Estimated Population Distribution"), linetype="dashed", size=1.2) +
  annotate("text", x = 0, y = 0.22, label = expression("H"[a] ~ ": " ~ mu > mu[0]), size = 5) + # adjusted y value
scale_color_manual(values=c("Null Distribution"="blue", "Estimated Population Distribution"="green3"))+
    ylim(0, 0.25)+
  theme_classic() + 
  theme(legend.position="none", axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank())

# Plot them side by side
grid.arrange(p1, p2, ncol=2)
```

### One- vs. Two-Sided: which and why?

**Use this rule of thumb:**

-   Two-sided (default): You care whether $\mu$ differs from $\mu_0$ **in either direction**.
-   One-sided (only if justified): Differences in the opposite direction are **scientifically irrelevant** and **would not change your decision or action**.

## Putting it all together

Let's bring the key players together. These are the elements we'll use throughout the chapter:

-   **Blue solid line** = Null distribution ($H_0$).
-   **Orange solid line** = True population distribution (what actually generated the data).
-   **Light green dashed line** = Estimated population distribution (inferred from the sample; our stand-in for $H_a$).
-   **Dark green dot-dashed line** = Sampling distribution of the sample mean, $\bar{X}$.
-   **Magenta dashed line** = Hypothesized mean $\mu_0$ (under $H_0$).\
-   **Thick dark green bar** = 95% confidence interval (CI) around the sample mean $\bar{X}$.

Up to now we've introduced each distribution separately. Let's layer them one at a time to see how they connect, building up to @fig-5-5.keyfig, which serves as the legend for every graph after it.

**1. Null vs. true population.**

In an idealized world where $H_0$ is exactly true, the true population (orange) and the null distribution (blue) are identical. In reality, we never know the orange curve. It's shown here for illustration.

```{r}
#| label: fig-5-5.null_identical_sample
#| fig-cap: 'True population distribution equals the null distribution ($H_0$ true): $\mu=\mu_0$.'
#| out-width: "100%"

# Define the data and functions
x_vals <- seq(-4.5, 4.5, by=0.01)
h0 <- dnorm(x_vals, mean=0, sd=1)

# Plot: Two identical distributions
p3 <- ggplot() + 
  geom_line(aes(x_vals, h0, color="True Population Distribution"), linetype="solid", size=4, alpha=1) +
  geom_line(aes(x_vals, h0, color="Null Distribution"), linetype="solid", size=1) +
  annotate("text", x = -2.5, y = 0.35, label = expression("H"[0] ~ ": " ~ mu == mu[0]), size = 5) +
  geom_vline(aes(xintercept=0), linetype="dashed", color="magenta", size=0.5) +
scale_color_manual(values=c("Null Distribution"="blue", "True Population Distribution"="orange"))+
  theme_classic() + 
  theme(legend.position="none", axis.text = element_blank(), axis.title = element_blank(), axis.ticks = element_blank())

print(p3)
```

**2. Adding the sampling distribution.** We never know the true parent distribution directly. What we do have is a sample, and from that we can construct the **sampling distribution of the mean** (dark-green dot-dash). This distribution underlies our ability to estimate a CI (dark green horizontal line).

```{r}
#| label: fig-5-5.perfectly_representative_sample
#| fig-cap: 'Perfectly representative sample: estimated population, null, and true population distributions align; the CI for $\bar{X}$ includes $\mu_0$.'
#| out-width: "100%"

p4 <-  plot_distributions(null_mean=0.5, null_sd=1, sample_mean=0.52, sample_sd=0.3, inferred_mean=0.52, inferred_sd=1, true_mean=0.54, true_sd=1)
print(p4)

```

**3. Adding the estimated population distribution.**

From our sample we also infer an **estimated population distribution** (light-green dashed). In the best case, this lines up closely with the true population (orange). The CI bar shows that the sample mean is consistent with $\mu_0$.

```{r}
#| label: fig-5-5.sample_imperfect_convenient
#| fig-cap: 'Imperfect but convenient sample: estimated population close to the null; the two-sided CI for $\bar{X}$ includes $\mu_0$ (fail to reject $H_0$).'
#| out-width: "100%"

p7 <- plot_distributions(null_mean=-0.2, null_sd=1, sample_mean=0.5, sample_sd=0.5, inferred_mean=0.5, inferred_sd=1, true_mean=0.6, true_sd=1)

print(p7)

```

Putting everything together, you'll see graphs moving forward that are all variations on the figure below. By now you should recognize each element.

Key reminder: $\mu_0$ (pink dashed line) comes from the null hypothesis. The unknown $\mu$ is what we're trying to learn about, and we approximate it from the sample through the estimated population distribution and the sampling distribution of $\bar{X}$.

```{r}
#| label: fig-5-5.keyfig
#| fig-cap: 'Key elements of hypothesis testing graphs: null (blue), true population (orange), estimated population (light-green dashed), sampling distribution of $\bar{X}$ (dark-green dot-dash), null mean $\mu_0$ (magenta dashed), and CI for $\bar{X}$ (thick dark-green bar).'
#| out-width: "100%"

pkey <-  plot_distributions(null_mean=-0.5, null_sd=1, sample_mean=0.5, sample_sd=0.3, inferred_mean=0.5, inferred_sd=1, true_mean=0.75, true_sd=1)
print(pkey)
```

### Decision Outcomes

Test outcomes are **binary**: we either **reject** $H_0$ or **fail to reject** $H_0$. We never *accept* $H_0$, because sampling variability means we can't prove it's true, only that the data are consistent with it. Similarly, we don't formally *accept* $H_a$, but we can say the results provide **support for** $H_a$ when $H_0$ is rejected.

If the observed statistic falls in either tail beyond the critical value(s), reject $H_0$. Otherwise, fail to reject $H_0$. Report the exact $p$ and an effect size or CI.

Here are our options (for a two tailed test).

1.  **Reject the Null Hypothesis (Two-Tailed Test).** The sample mean falls beyond the critical values, in the rejection region. The confidence interval also excludes the null mean (magenta line). In this example, the mean is more than 3 standard errors from $\mu_0$, which is typically considered significant for common choices of $\alpha$.

```{r}
#| label: fig-5-5.reject_2_two_tailed
#| fig-cap: 'Two-sided test, reject $H_0$: sample mean far from $\mu_0$ in either direction.'
#| out-width: "100%"

plot_distributions_no_error(null_mean=0, null_sd=1, sample_mean=-1.5, sample_sd=0.3, inferred_mean=-1.5, inferred_sd=1)

```

2.  **Fail to Reject the Null Hypothesis (Two-Tailed Test).** The sample mean falls in the non-rejection region, and the confidence interval includes the null mean. In this case, the evidence is consistent with chance variation, so we fail to reject $H_0$.

```{r}
#| label: fig-5-5.reject_1
#| fig-cap: 'Two-sided test, fail to reject $H_0$: sample mean not greater than critical value (within $3\,SE$ from $\mu_0$).'
#| out-width: "100%"

plot_distributions_no_error(null_mean=0, null_sd=1, sample_mean=-0.4, sample_sd=0.7, inferred_mean=-0.4, inferred_sd=1)
```

## Errors in Hypothesis Testing

Sometimes, even when we apply the test correctly, sampling variability can lead us to the wrong conclusion.

-   **Type I Error (**$\alpha$): Rejecting $H_0$ when it is actually true. This is like a smoke alarm going off when there's no fire: a false alarm. By convention, $\alpha$ is often set at 0.05 (5%).

-   **Type II Error (**$\beta$): Failing to reject $H_0$ when it is actually false. This is like a smoke alarm staying silent when there *is* a fire: a missed signal.

| Decision                 | If the null hypothesis is True | If the null hypothesis is False |
|--------------------------|--------------------------------|---------------------------------|
| **Reject** $H_0$         | Type I error (prob = $\alpha$) | Correct (prob = 1 - $\beta$)    |
| **Fail to reject** $H_0$ | Correct (prob = 1 - $\alpha$)  | Type II error (prob = $\beta$)  |

> **Key Takeaway**: As $\alpha$ gets smaller, $\beta$ gets bigger, and vice-versa.

Errors happen because samples are imperfect.

-   In @fig-5-5.type1_error, the true population (orange) is almost identical to the null (blue). By chance, however, the sample mean lands far away, and the CI excludes $\mu_0$. The test rejects $H_0$ even though it is true: a **Type I error**.

```{r}
#| label: fig-5-5.type1_error
#| fig-cap: 'Type I error: $H_0$ is true ($\mu=\mu_0$), but the sample leads us to reject $H_0$.'
#| out-width: "100%"

p8 <- plot_distributions(null_mean=-0.5, null_sd=1, sample_mean=2, sample_sd=0.5, inferred_mean=2, inferred_sd=1, true_mean=-0.52, true_sd=1)
print(p8)
```

-   In @fig-5-5.type2_error, the true population mean (orange) is far from $\mu_0$, but the particular sample looks close to the null. The CI includes $\mu_0$, so the test fails to reject when it should reject the null: a **Type II error**.

```{r}
#| label: fig-5-5.type2_error
#| fig-cap: 'Type II error: $H_0$ is false (true population mean differs from $\mu_0$), but the sample leads us to fail to reject $H_0$.'
#| out-width: "100%"

p9 <- plot_distributions(null_mean=-0.5, null_sd=1, sample_mean=0, sample_sd=0.4, inferred_mean=0, inferred_sd=1, true_mean=3, true_sd=1)
print(p9)
```

### Effect of Sample Size on Errors

With fixed $\alpha$ and correct assumptions, the long-run Type I error rate is $\alpha$ for any $n$. **Sample size mainly affects Type II error:** with small $n$, the sampling distribution is wide, true effects are harder to detect, and $\beta$ increases (power decreases). As $n$ grows, the sampling distribution tightens, $\beta$ falls, and power rises. Apparent changes in the Type I rate at small $n$ usually reflect **assumption violations** (e.g., non-normality, poor variance estimates, multiple looks), not $n$ itself.

**Takeaway:** Increasing $n$ improves estimation precision and **power** (reduces $\beta$); **Type I error remains at** $\alpha$ when assumptions hold.

### How to interpret $p$-values

A $p$-value is the probability, **assuming** $H_0$ is true, of obtaining a test statistic **at least as extreme** as the one we observed. We **reject** at level $\alpha$ when $p<\alpha$.

**Tails.** Match the tail(s) to $H_a$:

-   Two-sided ($H_a:\ \mu \neq \mu_0$): $p = 2\cdot P(\text{tail beyond }|z_{\text{obs}}|)$\
-   One-sided ($H_a:\ \mu > \mu_0$): $p = P(Z \ge z_{\text{obs}})$\
-   One-sided ($H_a:\ \mu < \mu_0$): $p = P(Z \le z_{\text{obs}})$

**Common mistakes to avoid.**

1.  $p$ is **not** the probability that $H_0$ is true.
2.  A smaller $p$ does **not** imply a larger effect; report effect size and a CI.
3.  With very **large** $n$, tiny effects can yield tiny $p$; assess practical importance.
4.  With small $n$, true effects can yield large $p$ (low power).

**Takeaway.** Use $p$ for the decision, but interpret results alongside **effect size** and a **95% CI**.

### Reporting results

A results statement gives the plain-language conclusion and the key statistics in a fixed order. Aim for one clean sentence plus context.

**Template (two-sided test):** *Plain-language finding.* *(test, statistic, df or n, value of statistic, p-value, effect size/CI if relevant).*

**What to include (in order):**

1.  Test name (e.g., one-sample *z*, two-sample *t*, χ², regression).
2.  Test statistic and its degrees of freedom (or sample size for *z*).
3.  Exact $p$ (report $p<0.001$ when very small).
4.  An effect size and/or a 95% CI when available.

**Examples**

-   **One-sample *z*:** The average mercury level within 10 km of the smelter was higher than the legal limit (*z* = 2.85, *n* = 32, p = 0.004).

-   **One-sample *t* with CI:** Mean nitrate exceeded the target by 0.41 mg/L (*t*(31) = 2.6, p = 0.013, 95% CI \[0.09, 0.73\] mg/L).

-   **Two-sample *t* with effect size:** Downstream sites had higher turbidity than upstream sites (*t*(58) = 3.1, p = 0.003, Cohen's *d* = 0.80).

-   **Not significant (be explicit):** Mean canopy temperature did not differ from the baseline (*t*(47) = 1.2, p = 0.24); we fail to reject $H_0$.

**Rounding & style**

-   Round statistics to 2 decimals; $p$ to 3 decimals (use $p<0.001$ when needed).
-   Use "fail to reject $H_0$" (not "accept").
-   Match tail to hypothesis in prose (e.g., "higher than," "lower than," or "different from").

**Do / Don't**

-   **Do** pair the sentence with a short, substantive takeaway (what it means for the question).
-   **Do** include a CI or effect size when it aids interpretation.
-   **Don't** restate methods; **don't** claim proof; **don't** omit units.

> **Takeaway:** One sentence is enough: state the direction in words, then report the test, the statistic, df or *n*, the $p$-value, and an effect size or CI.

### Beyond the 0.05 cutoff: Effect-size and power

Why go beyond a yes/no at $\alpha=0.05$? Because scientists, managers, policymakers, and communities need to know **how large** an effect is and **how likely** a study is to detect it. That's what **effect size** and **power** provide.

**Working definitions**

-   **Effect size**: how big the difference or association is (in natural units or standardized units like Cohen's $d$).

-   **Power (**$1-\beta$): the probability your design will detect a real effect of a specified size at your chosen $\alpha$.

**Use them before you sample.** Decide what effect is *meaningful* for your context (e.g., a 0.3 mg/L nitrate reduction) and size your study to have adequate power to detect that effect.

### Effect size: practical vs. standardized

**What is effect size?** It's how big the difference or association is. You can express it in **original units** (practical meaning) or in **standardized units** (comparability across scales).

**Practical (original units).** Use the scale stakeholders recognize.

-   Exams: "Treatment improved scores by **5 percentage points**" (a letter-grade--sized change).
-   Time-on-task: "New training reduces practice time by **1,000 hours** out of 10,000" (a 10% reduction).

These statements are easy to interpret, but they don't travel well across contexts or measures.

**Standardized (unitless).** When units are unfamiliar or scales differ, standardize by typical variability. The most common standardized effect size metric is *Cohen's d*. For two means with a common SD $\sigma$,

$$
d = \frac{\mu_1 - \mu_2}{\sigma}
$$

Now "how big?" is expressed in **SD units**. If you notice, this is just a kind of z-score. It is a way to standardize the mean difference in terms of the population standard deviation.

**Environmental example (biodiversity index).** Suppose an index increases from 3 to 4 after reforestation.

-   In **raw units**: +1 index point.

-   In **standardized terms**:

    -   If the SD across sites is 1.0, then $d = 1.0$: a large shift.
    -   If the SD is 10, then $d = 0.10$: a small shift.

> **Takeaway:** Report both when possible: original units for meaning (policy/action) and a standardized effect (e.g., Cohen's $d$) for comparison across studies.

**Notes for practice**

-   In real data, $\sigma$ is unknown; estimate it (pooled SD for two groups, with small-sample corrections as needed).
-   Heuristics (context matters): $d \approx 0.2$ (small), $d \approx 0.5$ (medium), $d \approx 0.8$ (large).
-   A "small" $d$ can still be important (e.g., small reductions in lead or $\mathrm{PM}_{2.5}$ can matter for health).
-   Pair effect sizes with **95% CIs** to show precision, not just magnitude.

Let's ground effect size in a concrete example. Suppose we're measuring **final exam performance** (percent correct). The class mean is 65% with a standard deviation of 5%. Group A is a control, while Group B receives some instructional treatment.

@fig-5.5effectdistsB shows four possible scenarios:

```{r}
#| label: fig-5.5effectdistsB
#| fig-cap: 'Same idea with concrete scores (mean $65$, $sd=5$): treatment shifts correspond to $d=0,\,0.5,\,1,\,2$.'
#| out-width: "100%"


X <- c(
  seq(25, 100, 1),
  seq(25, 100, 1),
  seq(25, 100, 1),
  seq(25, 100, 1),
  seq(25, 100, 1),
  seq(25, 100, 1),
  seq(25, 100, 1),
  seq(25, 100, 1)
)
Y <- c(
  dnorm(seq(25, 100, 1), 65, 5),
  dnorm(seq(25, 100, 1), 65, 5),
  dnorm(seq(25, 100, 1), 65, 5),
  dnorm(seq(25, 100, 1), 67.5, 5),
  dnorm(seq(25, 100, 1), 65, 5),
  dnorm(seq(25, 100, 1), 70, 5),
  dnorm(seq(25, 100, 1), 65, 5),
  dnorm(seq(25, 100, 1), 75, 5)
)
effect_size <-
  rep(c("65, d=0", "67.5,d=.5", "70, d=1", "75, d=2"), each = 76 * 2)
condition <- rep(rep(c("A", "B"), each = 76), 2)
df <- data.frame(effect_size,
                 condition,
                 X, Y)

ggplot(df, aes(
  x = X,
  y = Y,
  group = condition,
  color = condition
)) +
  geom_line() +
  theme_classic(base_size = 15) +
  facet_wrap( ~ effect_size) +
  xlab("values") +
  ylab("density")


```

-   **Panel 1 (**$d=0$): No effect. Both groups sample from the same distribution (mean = 65).
-   **Panel 2 (**$d=0.5$): Group B's mean shifts upward by 2.5 points (67.5).
-   **Panel 3 (**$d=1$): A 5-point shift (70).
-   **Panel 4 (**$d=2$): A large 10-point shift (75).

The takeaway is that effect size puts raw differences into a standard-deviation scale. A half--SD shift ($d=0.5$) is noticeable but modest; a two--SD shift ($d=2$) is very large. In practice, we rarely know beforehand how big a shift to expect, that's why we conduct the study.

## Power

When a real effect exists, your design needs to be sensitive enough to detect it, otherwise the test has little value. We've already seen that effects can vary in size. Now we add a second idea: study designs differ in their ability to reliably detect those effects. This sensitivity is called **statistical power**.

Power rises when:

1.  **Effect size is larger.**
2.  **Sample size (**$n$) is larger (smaller $SE$).
3.  **The test is less conservative** (larger $\alpha$).
4.  **Measurement noise is lower** (smaller $SD$).

> **Question:** "With $\alpha = 0.05$, how large must $n$ be to achieve $\ge 80\%$ power for the smallest effect that matters?"

### Simulation walkthrough

Let's see how design choices affect power with a simple simulation. Suppose we have two groups, each with $n=10$. Group A (control) is sampled from a normal distribution with mean = 10 and $SD = 5$. Group B (treatment) has a mean = 12.5, a shift of 2.5 points (Cohen's $d=0.5$), considered a medium effect.

If we simulate this experiment 1,000 times and run an independent-samples $t$-test each time, how often do we reject $H_0$ at $\alpha=0.05$?

```{r, echo=TRUE}
set.seed(2025)
p <- numeric(1000)
for(i in 1:1000){
  A <- rnorm(10, 10, 5)
  B <- rnorm(10, 12.5, 5)
  p[i] <- t.test(A,B,var.equal=TRUE)$p.value
}
```

```{r}
mean(p < 0.05)
```

This proportion is the **power** of the design: the probability of detecting the true effect in repeated experiments. With $n=10$, it's only about 20%. Even though we know a medium-sized effect exists, the test usually fails to pick it up with such a small sample.

### How power changes

1.  **Larger** $n$. Doubling to $n=20$ per group reduces sampling error and roughly doubles power:

```{r}
p <- numeric(1000)
for(i in 1:1000){
  A <- rnorm(20, 10, 5)
  B <- rnorm(20, 12.5, 5)
  p[i] <- t.test(A,B,var.equal=TRUE)$p.value
}
mean(p < 0.05)
```

2.  **Smaller** $\alpha$. Making the test more conservative (e.g., $\alpha=0.01$) lowers power, because fewer outcomes count as "significant":

```{r}
mean(p < 0.01)
```

3.  **Bigger effects are easier to detect.** If the treatment produces a big shift (say 2 standard deviations), the same design has near-perfect power:

```{r, echo=TRUE}
p <- numeric(1000)
for(i in 1:1000){
  A <- rnorm(20, 10, 5)
  B <- rnorm(20, 20, 5)   # 2SD effect
  p[i] <- t.test(A,B,var.equal=TRUE)$p.value
}
```

```{r}
mean(p < 0.01)
```

### Power curves

It's best to think of power as a **profile**, not a single number. A power curve shows how sensitive a design is across a range of possible effects.

In @fig-5.5powercurve, we fix the design to a two-sample $t$-test with $n=10$ per group and $\alpha=0.05$. Power is plotted on the y-axis, and the true effect size (Cohen's $d$) is on the x-axis.

```{r}
#| label: fig-5.5powercurve
#| fig-cap: 'Power vs. effect size (Cohen’s $d$) for a two-sample $t$-test with $n=10$ per group and $\alpha=0.05$.'
#| out-width: "100%"

power<-c()
for(i in seq(0,2,.1)){
  sd_AB <- 1
  n<-10
  C <- qnorm(0.975)
  se <- sqrt(sd_AB/n + sd_AB/n)
  delta<-i
  power <- c(power,1-pnorm(C-delta/se) + pnorm(-C-delta/se))
}

plot_df<-data.frame(power,
                    effect_size = seq(0,2,.1))

ggplot(plot_df, aes(x=effect_size, y=power))+
  geom_line()+
  theme_classic()
```

**Walkthrough.**

-   With $n=10$, the design only detects a medium effect ($d=0.5$) about 20% of the time.
-   A larger effect ($d=0.8$) is caught about half the time.
-   Very large effects ($d=2$) are detected almost every time.

This shows that small samples are only sensitive to big, obvious effects.

Now suppose prior research suggests the effect size is small ($d=0.2$). Instead of varying the effect, we can ask: *how many subjects per group do we need to have decent power?*

@fig-5.5powercurveN plots power against sample size for detecting $d=0.2$. The green line marks 80% power, a common planning target.

```{r}
#| label: fig-5.5powercurveN
#| fig-cap: 'Power vs. sample size for detecting $d=0.2$ with a two-sample $t$-test at $\alpha=0.05$; green line marks $0.80$ power.'
#| out-width: "100%"

power<-c()
for(i in seq(10,800,10)){
  sd_AB <- 1
  n<-i
  C <- qnorm(0.975)
  se <- sqrt(sd_AB/n + sd_AB/n)
  delta<-0.2
  power <- c(power,1-pnorm(C-delta/se) + pnorm(-C-delta/se))
}

plot_df<-data.frame(power,
                    N = seq(10,800,10))

ggplot(plot_df, aes(x=N, y=power))+
  geom_line()+
  theme_classic()+
  geom_hline(yintercept=.8, color="green")+
  labs(x = "n per group", y = "Power")
```

**Walkthrough.**

-   At $n=10$ per group, power to detect $d=0.2$ is near zero.
-   To reach 80% power, we need about **380 per group**.
-   Even then, 20% of studies would still fail to reject $H_0$ just by chance.

**Takeaway:**

-   A power curve shows, for a fixed design, how power rises as the true effect grows.
-   A power-vs-$n$ curve shows how many samples you need to hit a target power for a specific effect.
-   Power analysis is mainly a **planning tool**, not a results-reporting tool. When you design a study, don't just aim for "power = 0.8." Always state the **assumed effect size** and the **standard deviation** you used for that calculation. Otherwise, "0.8" has no meaning: power depends entirely on those assumptions.

### Planning your design

Our discussion of effect size and power highlights the importance of understanding the statistical limitations of an experimental design. Planning a study means working through the following, in order:

-   **Define a meaningful effect** (environmentally or policy-relevant minimum).
-   **Choose** $\alpha$ (often 0.05; justify if different).
-   **Estimate variability** (pilot data, literature, or historical monitoring).
-   **Compute** $n$ for $\ge 80\%$ power (or explain tradeoffs if you can't reach it).
-   **Pre-specify** one- vs two-tailed before data collection.

As a general rule of thumb, small-$n$ designs can only reliably detect very large effects, whereas large-$n$ designs can reliably detect much smaller effects. As a researcher, it is your responsibility to plan your design accordingly so that it is capable of reliably detecting the kinds of effects it is intended to measure.

### Two cautions (common pitfalls)

-   **Low power inflates doubt, even when** $p<0.05$. Underpowered studies rarely replicate. A design with only 30% power to detect an expected effect might, by luck, produce a "significant" result, but most replications will not. Worse, false positives from small samples tend to look exaggerated: if a spurious finding reaches significance, the estimated effect size is often large enough to seem convincing.

-   **Huge** $n$ finds tiny effects. With very large samples, trivial differences can register as statistically significant. For example, a billion-person trial might show a drug reduces pain by 0.01%. That effect is "real" but practically meaningless. This is why effect size and confidence intervals should be reported alongside $p$-values: significance alone is not enough.

@fig-5.5effectsizeType1 shows the first pitfall graphically. Even when the null hypothesis is true, 5% of tests are expected to fall below $\alpha = 0.05$. In small samples, those false positives tend to be paired with large apparent effect sizes. This can give the illusion of a strong, replicable finding when in reality it is just sampling noise.

```{r}
#| label: fig-5.5effectsizeType1
#| fig-cap: 'Effect size estimates among Type I errors ($p<0.05$) under the null, by sample size, for a one-sample $t$-test. Small $n$ inflates apparent effects.'
#| out-width: "100%"

all_df<-data.frame()
for(i in 1:1000){
  for(n in c(10,20,50,100,1000)){
    some_data<-rnorm(n,0,1)
    p_value<-t.test(some_data)$p.value
    effect_size<-mean(some_data)/sd(some_data)
    t_df<-data.frame(sim=i,sample_size=n,p_value,effect_size)
    all_df<-rbind(all_df,t_df)
  }
}

type_I_error <- all_df[all_df$p_value<.05,]
type_I_error$sample_size<-as.factor(type_I_error$sample_size)

ggplot(type_I_error,aes(x=p_value,y=effect_size, group=sample_size,color=sample_size))+
  geom_point()+
  theme_classic()+
  ggtitle("Effect sizes for Type I errors")
```

**Takeaway:** statistical significance does not guarantee scientific significance. Always interpret $p$ in the context of effect size, confidence intervals, and study design.

**Key point**

Power is not a fixed property of a test; it depends entirely on assumptions about $n$, effect size, $\alpha$, and variability. Think of it as a **family of numbers**, not a single value. That's why power analysis is most useful **before** you collect data: it tells you whether your design is capable of detecting the kind of effect you care about.

## Chapter Summary

**Why it matters.** Hypothesis testing is the machinery behind most quantitative claims you will read, make, or have to defend. It gives you a disciplined way to decide whether a pattern in your sample says something real about the population, and, just as importantly, an honest account of how often that decision will be wrong.

**Core ideas**

-   **Three routes, one decision.** A test statistic against a critical value, a confidence interval, and a $p$-value are three ways of asking the same question, and for a given $\alpha$ they always agree. Which you report is a matter of audience and convention, not correctness.
-   **What a $p$-value is.** The probability of data at least this extreme *if* $H_0$ were true. It is not the probability that $H_0$ is true, and on its own it tells you nothing about how large the effect is.
-   **Two ways to be wrong.** A Type I error rejects a true $H_0$, at a rate you choose by setting $\alpha$. A Type II error fails to reject a false one. Tightening one loosens the other, which is why $\alpha = .05$ is a convention balancing the two rather than a fact about nature.
-   **Significance is not size.** With a large enough $n$, a trivially small difference will clear any threshold. Report an effect size and a confidence interval next to every $p$-value so a reader can judge whether the result matters, not just whether it cleared a line.
-   **Power is a design decision.** Power is your chance of detecting a real effect when one exists, and it depends on $n$, effect size, $\alpha$, and variability together. An underpowered study fails in both directions: it usually misses real effects, and on the occasions it does reach significance, the effect it reports tends to be badly inflated. That is why power analysis belongs *before* data collection.
-   **Commit to one- or two-tailed in advance.** A one-tailed test is only justified when theory genuinely rules out the other direction. Deciding after seeing the data quietly raises your real Type I error rate above the $\alpha$ you claimed.
