---
editor: source
author: 
  - Mallory Barnes
aliases: [redirects/t-tests.html]
---

```{r, include = FALSE}
source("global_stuff.R")
set.seed(538)
```

```{r, echo=FALSE}
library(ggplot2)
library(dplyr)
library(broom)

fmt_p <- function(p) ifelse(p < .001, "< .001", formatC(p, digits = 3, format = "f"))

# Histogram helpers for the simulation section, so those figures match the
# rest of the book instead of falling back to base graphics.
hist_t <- function(ts, title = NULL) {
  ggplot(data.frame(t = ts), aes(x = t)) +
    geom_histogram(bins = 30) +
    labs(x = "Simulated t", y = "Count", title = title)
}

hist_p <- function(ps, title = NULL) {
  ggplot(data.frame(p = ps), aes(x = p)) +
    geom_histogram(breaks = seq(0, 1, .05)) +
    scale_x_continuous(breaks = seq(0, 1, .1)) +
    labs(x = "Simulated p-value", y = "Count", title = title)
}
```

# t-tests

## Intro

In the early 1900s, William Sealy Gosset joined Guinness, then the world's largest brewery. Guinness hired scientifically trained brewers and gave them latitude to pursue research that improved the product. As production scaled, quality control mattered: a pint should taste the same in Dublin as in Detroit. Until then, hop quality was judged mostly by appearance and smell, which was unreliable. Chemical assays could measure ingredients, but the core problem was statistical. Inference relied on the normal ($z$) distribution, which works well when samples are large enough for a normal approximation. In practice, Guinness could not destroy large volumes of beer or ingredients just to test them.

Gosset set out to quantify the extra uncertainty introduced by small samples. How much wider is the error around an estimate when $n=10$ instead of $n=1000$? He derived a new sampling distribution that accounts for small-sample variability, now known as the **Student's $t$ distribution**. Because Guinness restricted external publications, he published the 1908 paper under the pseudonym "Student" [@Student1908]. That is why we still say Student's $t$-test. This chapter uses the same idea: measure a signal, scale it by its standard error, and compare the resulting $t$ to the $t$ distribution for your $n-1$ degrees of freedom.

**Why this matters.** We often compare a sample mean to a reference (drinking-water standard, historical average) or compare two conditions (before/after, site A vs site B). A $t$-test asks whether the observed difference is larger than you would expect from sampling variation alone, given your $n$ and variability.

**Three common $t$-tests you will use.**

1.  **One-sample $t$-test.** Compare a sample mean to a target ($\mu_0$).
2.  **Paired $t$-test.** Compare before/after (or matched) measurements on the same units by running a one-sample $t$-test on the *differences*.
3.  **Independent-samples $t$-test (also called two-sample $t$-test).** Compare means from two groups measured on different units. The pooled version assumes the two groups have similar spread; Welch's version [@Welch1947] does not, and it is what R's `t.test()` runs unless you set `var.equal = TRUE`.

::: {.callout-tip title="Tip: What you will report"}
Always report:

-   the effect size (mean difference)
-   a 95% CI
-   the test statistic with its df
-   the $p$-value

Example: "Mean change was -0.42 units (95% CI \[-0.73, -0.11\]; *t*(11) = -2.92; *p* = .014)."
:::

## Signal over noise, for a mean

Chapter 4 introduced the general shape of a test statistic: the effect you observed, divided by an estimate of how much it could move by chance.

$$\text{statistic}=\frac{\text{effect}}{\text{error}}$$

The $t$-test is the first place we fill that in with real quantities. The effect is the distance between the sample mean and the value $H_0$ names. The error is not the raw standard deviation, though: it is the standard error, $s/\sqrt{n}$, because what varies from study to study is the *mean*, not an individual observation.

$$t = \frac{\text{effect}}{\text{standard error}} = \frac{\bar{x} - \mu_0}{s / \sqrt{n}}$$

## One-sample t-test

Use a one-sample $t$ when you want to test whether a sample mean differs from a reference value ($\mu_0$). The question: is the observed difference bigger than you'd expect from sampling variability, given $n$ and your sample SD $s$?

$$t = \frac{\bar{x}-\mu_0}{s/\sqrt{n}}, \qquad df = n-1$$

After calculating $t$ from our data, we compare the observed $t$ to a $t$ distribution with $n-1$ degrees of freedom to get a $p$-value and a decision at $\alpha$.

### Comparing $z$ and $t$

Under a known population SD ($\sigma$), the standardized mean uses the $z$ statistic: $$z = \frac{\bar{x}-\mu_0}{\sigma/\sqrt{n}}$$

When $\sigma$ is *unknown* (the usual case), we estimate it with the sample SD ($s$) and use the $t$ statistic: $$t = \frac{\bar{x}-\mu_0}{s/\sqrt{n}}, \qquad df = n-1$$

Because $s$ varies from sample to sample, the $t$ distribution has heavier tails than the standard normal. As $n$ grows, $s$ stabilizes and $t$ converges to $z$.

```{r}
#| label: fig-6-z-vs-t-linear-log
#| fig-cap: 'Normal ($z$) vs $t$ distributions: on a linear scale (top), tail differences look small; on a log scale (bottom), the heavy tails of small-$n$ $t$ are clear.'
#| fig-alt: "Two stacked panels showing the same three overlaid curves: the standard normal, t with 10 degrees of freedom, and t with 2 degrees of freedom. On the linear-scale top panel, all three curves look almost identical, bell-shaped and close together. On the log-scale bottom panel, the same three curves separate clearly in the tails, with the t distribution at 2 degrees of freedom sitting well above the other two far from the center, revealing much heavier tails than the linear panel suggested."
#| out-width: "100%"

library(tidyverse)
library(patchwork)

xs <- seq(-4, 4, length.out = 2000)

df_vals <- tibble(
  dist = c("z (standard normal)", "t (df=10)", "t (df=2)"),
  df    = c(NA, 10, 2)   # NA for z
)

# Build a long data frame: every x crossed with each distribution
plot_df <- tidyr::crossing(x = xs, df_vals) |>
  mutate(density = case_when(
    dist == "z (standard normal)" ~ dnorm(x),
    dist == "t (df=10)"           ~ dt(x, df = 10),
    dist == "t (df=2)"            ~ dt(x, df = 2)
  ))

base_cols <- c("z (standard normal)" = "#0072B2",
               "t (df=10)"           = "#E69F00",
               "t (df=2)"            = "#009E73")

p_lin <- ggplot(plot_df, aes(x, density, color = dist)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = base_cols, guide = guide_legend(title = NULL)) +
  coord_cartesian(xlim = c(-4, 4), ylim = c(0, 0.45)) +
  labs(x = NULL, y = "Density (linear scale)") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

p_log <- ggplot(plot_df, aes(x, density, color = dist)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = base_cols, guide = "none") +
  scale_y_log10() +
  coord_cartesian(xlim = c(-4, 4)) +
  labs(x = "value", y = "Density (log scale)") +
  theme_classic(base_size = 12)

p_lin / p_log

```

::: {.callout-tip title="Tip: When is $z$ appropriate?"}
If the population SD $\sigma$ is known *and* the sampling distribution of the sample mean is (approximately) normal, use $z$. In practice, $\sigma$ is rarely known, so we use $t$ with $s$ and $df=n-1$.
:::

**Rule of thumb.** For two-tailed tests at $\alpha=0.05$:

-   $z$ uses critical values $\pm 1.96$.
-   $t$ is a bit wider for small $n$ (e.g., $df=9 \Rightarrow \pm 2.262$; $df=29 \Rightarrow \pm 2.045$), and approaches $\pm 1.96$ as $n$ increases.

### Degrees of freedom

Every $t$ formula so far has carried $df = n-1$ alongside it, and @fig-6-z-vs-t-linear-log drew a different $t$ curve for each $df$. That number is the **degrees of freedom**. It is also the number in parentheses whenever a $t$-test is reported, as in $t(9) = 1.94$.

For one-sample and paired $t$-tests, $df = n-1$. A sample of 10 has $df=9$.

Degrees of freedom counts how many values are still free to vary once something has been estimated from the data. Consider three numbers with a mean of 2: pick the first two freely, say 51 and $-3$, and the third number is no longer free. It has to be $-42$ for the mean to come out to 2. Two of the three numbers had freedom; the third didn't. That's $df=n-1=2$ for three numbers.

In a $t$-test, the mean is the thing estimated, and the $df$ tells you how much information is left for estimating the spread, $s$. With few degrees of freedom, $s$ can land well above or below $\sigma$, so large $t$ values turn up more often and the $t$ curve has heavier tails. As $df$ grows, $s$ settles close to $\sigma$ and the $t$ curve approaches the normal. That is why each $df$ gets its own curve.

**Simulating how degrees of freedom affect the $t$ distribution.**

We can see the effect of degrees of freedom directly by simulating the $t$ distribution at two different sample sizes. @fig-6dft compares $df=9$ (a sample of 10) to $df=100$ (a much larger sample).

```{r}
#| label: fig-6dft
#| fig-cap: "The width of the t distribution shrinks as sample size and degrees of freedom (from 9 to 100) increase."
#| fig-alt: "Two side-by-side histograms of 10,000 simulated t values each, one for 9 degrees of freedom and one for 100 degrees of freedom. The 9-degrees-of-freedom histogram is visibly shorter and wider, with more values spread into the tails, while the 100-degrees-of-freedom histogram is taller and narrower, concentrated closer to zero."
#| fig-asp: 0.5
#| out-width: "100%"

ts <- c(rt(10000, 9), rt(10000, 100))
dfs <- as.factor(rep(c(9, 100), each = 10000))

t_df <- data.frame(dfs, ts)
t_df <- t_df[abs(t_df$ts) < 5, ]

ggplot(t_df, aes(x = ts)) +
  geom_histogram(fill = "#008cff", color = "white") +
  facet_wrap( ~ dfs) +
  xlab("t value") +
  ylab("Count")


```

As degrees of freedom increase, the $t$ distribution gets taller in the middle and thinner in the tails. With more data the sample SD is a more reliable stand-in for $\sigma$, so $t$ behaves more like $z$.

### What t represents

$t$ expresses how far a sample mean is from the hypothesized mean, in standard error units (signal over noise):

$$t = \frac{\bar{x}-\mu_0}{s/\sqrt{n}}$$

$t$ can be positive or negative depending on direction. Under $H_0$, $t \sim t_{(df=n-1)}$.\
**Rule of thumb.** Under $H_0$, $|t|$ beyond about 2 is rare. The exact two-tailed cutoff at $\alpha = .05$ is 2.09 at $df = 20$, and higher for smaller $df$ (2.26 at $df = 9$).

::: {.callout-note title="Note: Standard error, again"}
Remember that $s/\sqrt{n}$ is the standard error (SE), the standard deviation of the sampling distribution of the sample mean.
:::

Let's see how $t$ values behave when samples really come from the null hypothesis. Under the null hypothesis, $t$ follows a $t$ distribution with $df = n-1$.

```{r}
#| label: fig-6t-sim
#| fig-cap: 'Histogram of 1,000 simulated $t$ values ($n=10$) from a population whose mean equals $\mu_0 = 0$. Most $t$s cluster near 0; large positive or negative values are rare.'
#| fig-alt: "Histogram of 1000 simulated t values, tallest near zero and tapering off symmetrically toward both larger positive and larger negative values, with very few beyond about plus or minus 3."
t_vals <- replicate(1000, {
  s <- rnorm(10, 0, 1)
  (mean(s) - 0) / (sd(s) / sqrt(length(s)))
})
ggplot(data.frame(t_vals), aes(x = t_vals)) +
  geom_histogram(bins = 30, color = "white", linewidth = 0.2) +
  labs(x = "t values", y = "Count")
```

@fig-6t-sim shows 1,000 $t$ values from samples of 10 drawn with $H_0$ true. Most fall between $-2$ and $2$, and very few are beyond $\pm 3$.

**Takeaway.** $t$ measures how far the sample mean is from the null expectation, scaled by its uncertainty. Under $H_0$ it follows a known $t$ distribution, so we can check how often chance alone would produce a $t$ as large as ours.

### Calculating t from data

Let's calculate a one-sample $t$ statistic step by step using a short example. Ten students each took a 10-question true/false quiz (50% expected by chance). Their scores (percent correct) are:

**Scores:** 50, 70, 60, 40, 80, 30, 90, 60, 70, 60

```{r}
#| label: ex-onesample-t
#| echo: FALSE

scores <- c(50, 70, 60, 40, 80, 30, 90, 60, 70, 60)
n  <- length(scores)
x  <- mean(scores)
s  <- sd(scores)
SE <- s / sqrt(n)

tt <- t.test(scores, mu = 50)
ci_lo <- round(tt$conf.int[[1]], 2)
ci_hi <- round(tt$conf.int[[2]], 2)
```

By hand, the standard error of the mean is the sample SD divided by the square root of $n$:

$$\text{SE} = \frac{s}{\sqrt{n}} = \frac{`r round(s,2)`}{\sqrt{`r n`}} = `r round(SE,2)`$$

And $t$ is the difference between the sample mean (`r round(x,2)`) and the reference value (50, chance level), divided by that SE:

$$t = \frac{\bar{x}-\mu_0}{SE} = \frac{`r round(x,2)`-50}{`r round(SE,2)`} = `r round(tt$statistic,2)`$$

We can confirm this with a single R command, `t.test(scores, mu = 50)`:

**Results:** $t(9)=`r round(tt$statistic,2)`$, $p=`r signif(tt$p.value,3)`$; mean $=`r round(tt$estimate,2)`$ (95% CI \[`r ci_lo`, `r ci_hi`\]).

**Interpretation.** The mean (`r round(x,0)`%) is above chance (50%), but the difference is not statistically significant at $\alpha=.05$.

**Assumptions check.** The 10 scores spread evenly around 60 with no outliers, so approximate normality is reasonable. @sec-6assumptions covers how to check this.

### How does t behave?

Extreme sample means are unlikely if $H_0$ is true, which is why large $|t|$ values are rare. That is the basis of the decision rule for $t$-tests: if your observed $t$ falls beyond the critical values, you reject $H_0$.

### Interpreting $t$

We're asking whether our $t$-value could plausibly have come from the null distribution: the distribution of $t$-values we'd see across many samples of size 10, if the true mean were exactly 50 (chance level). @fig-6tnull shows what that distribution looks like, at $df=9$.

```{r}
#| label: fig-6tnull
#| fig-cap: 'The distribution of $t$-values that chance alone produces when the population mean equals $\mu_0$ ($df = 9$).'
#| fig-alt: "A symmetric t-distribution curve with 9 degrees of freedom, centered at zero with a grey vertical line marking the center. Labels on the left and right halves each read 50 percent, showing the curve splits evenly into negative and positive t-values."


range <- seq(-3, 3, .01)
null_distribution <- dt(range, 9, log = FALSE)
plot_df <- data.frame(range, null_distribution)

ggplot(plot_df, aes(x = range, y = null_distribution)) +
  geom_line(colour = "blue", linetype = "dotdash", linewidth = 1.1) +
  annotate("segment", x = 0, xend = 0, y = 0, yend = dt(0, 9), colour = "grey40", linewidth = 0.8) +
  annotate("text", x = c(-.7, .7), y = .1, label = c("50%\n(-)", "50%\n(+)"), size = 4.5, colour = "grey20") +
  scale_x_continuous(breaks = (seq(-3, 3, .5))) +
  labs(x = "t", y = "Density")


```

$t$ is most likely to land near zero, which makes sense: this is the distribution of *no* differences, so most of the time chance alone won't push it far from 0. Because the distribution is symmetric, a $t$-value from the null distribution is just as likely to be positive as negative.

Because the distribution is symmetric, half of all $t$-values are at or above zero and half are at or below.

We can get more precise than a 50/50 split. @fig-6percentregions divides the same distribution into ten regions, each containing 10% of the distribution. Notice the regions get wider out in the tails: since most $t$-values cluster near 0, it takes a wider interval out there to capture the same 10%.

```{r}
#| label: fig-6percentregions
#| fig-cap: "Splitting the t distribution up into regions each containing 10% of the $t$-values. The width between the bars narrows as they approach the center of the distribution, where there are more $t$-values."
#| fig-alt: "The same t-distribution curve with 9 degrees of freedom, divided by nine grey vertical lines into ten regions, each labeled 10 percent. The regions near the two outer edges are noticeably wider than the regions clustered near the center of the curve."
#| out-width: "100%"

range <- seq(-3, 3, .01)
null_distribution <- dt(range, 9, log = FALSE)
plot_df <- data.frame(range, null_distribution)

t_ps <- qt(seq(.1, .9, .1), 9)

# Decile boundaries are guides, not critical values, so grey and stopping at the curve.
ggplot(plot_df, aes(x = range, y = null_distribution)) +
  geom_line(colour = "blue", linetype = "dotdash", linewidth = 1.1) +
  annotate("segment", x = t_ps, xend = t_ps, y = 0, yend = dt(t_ps, 9), colour = "grey40", linewidth = 0.7) +
  scale_x_continuous(breaks = seq(-3, 3, 1)) +
  # one label per region, turned sideways so the narrow central regions fit
  annotate("text", x = c(-1.75, (t_ps[-9] + t_ps[-1]) / 2, 1.75), y = 0.04, angle = 90,
           label = "10%", size = 3.6, colour = "grey20") +
  labs(x = "t", y = "Density")


```

For example:

1.  $t \leq -1.38$: $p = 10\%$
2.  $-1.38 \leq t \leq -0.88$: $p = 10\%$
3.  $-0.88 \leq t \leq -0.54$: $p = 10\%$
4.  $t \geq 1.38$: $p = 10\%$

Each range captures the same 10% of the distribution, even though the ranges themselves are different widths.

### Getting the p-values for $t$-values

Older textbooks printed $p$-values for $t$ in a table at the back of the book. R, Excel, and [online calculators](http://www.socscistatistics.com/pvalues/tdistribution.aspx) all work; you'll get hands-on practice with this in lab, both with software and by simulation.

Software needs two things from you to compute a $p$-value: the degrees of freedom, which you already have, and whether the test is one-tailed or two-tailed, which comes next.

### One-tailed vs. two-tailed, applied

Chapter 4 introduced the concept: a two-tailed test asks whether $\mu$ differs from $\mu_0$ in either direction, while a one-tailed test commits to one direction in advance, and is only justified when a result in the other direction would lead to the same action as no result at all. For a $t$-test, that choice changes the critical value, and therefore the $p$-value, for the exact same data.

For our quiz-score example ($df=9$), a one-tailed test only needs to clear 5% of the distribution on one side. A two-tailed test has to clear 2.5% on each side, which pushes the critical value farther out:

```{r}
#| label: fig-6critT
#| fig-cap: "One-tailed critical value at df=9: t-values of 1.83 or greater occur 5% of the time."
#| fig-alt: "A t-distribution curve with 9 degrees of freedom, with a single red vertical line at t equal to 1.83 labeled plus crit equals 1.83. The tail to the right of that line is shaded red and labeled 5 percent; everything to the left is unshaded. A green dotted line marks the observed t of 1.94, just inside the shaded tail."

range <- seq(-3.5, 3.5, .01)
plot_df <- data.frame(range, null_distribution = dt(range, 9))
crit1 <- qt(.95, 9)
t_obs <- 1.94   # the quiz example, t(9) = 1.94

ggplot(plot_df, aes(x = range, y = null_distribution)) +
  geom_area(data = subset(plot_df, range >= crit1), fill = "red", alpha = 0.18) +
  geom_line(colour = "blue", linetype = "dotdash", linewidth = 1.1) +
  annotate("segment", x = crit1, xend = crit1, y = 0, yend = 0.41, colour = "red", linewidth = 0.8) +
  annotate("text", x = crit1, y = 0.425, label = paste0("+crit = ", round(crit1, 2)), colour = "red", size = 4.2) +
  annotate("segment", x = t_obs, xend = t_obs, y = 0, yend = 0.33, colour = "#1E9E1E", linetype = "dotted", linewidth = 1.4) +
  annotate("text", x = t_obs + 0.08, y = 0.36, label = "observed\nt = 1.94", colour = "#1E9E1E", hjust = 0, size = 3.8, lineheight = 0.9) +
  annotate("text", x = 2.9, y = 0.06, label = "5%", size = 4.2, colour = "grey20") +
  labs(x = "t", y = "Density")
```

```{r}
#| label: fig-6twotailedt
#| fig-cap: "Two-tailed critical values at df=9: t-values beyond ±2.26 occur 5% of the time total, 2.5% in each tail."
#| fig-alt: "A t-distribution curve with 9 degrees of freedom, with two red vertical lines labeled minus crit and plus crit at negative 2.26 and positive 2.26. The tails beyond each line are shaded red and labeled 2.5 percent, one on each side, leaving the large central region between them unshaded. A green dotted line marks the observed t of 1.94, just short of the right critical value."

range <- seq(-4, 4, .01)
plot_df <- data.frame(range, null_distribution = dt(range, 9))
crit2 <- qt(.975, 9)
t_obs <- 1.94   # the quiz example, t(9) = 1.94

ggplot(plot_df, aes(x = range, y = null_distribution)) +
  geom_area(data = subset(plot_df, range <= -crit2), fill = "red", alpha = 0.18) +
  geom_area(data = subset(plot_df, range >=  crit2), fill = "red", alpha = 0.18) +
  geom_line(colour = "blue", linetype = "dotdash", linewidth = 1.1) +
  annotate("segment", x = c(-crit2, crit2), xend = c(-crit2, crit2), y = 0, yend = 0.41, colour = "red", linewidth = 0.8) +
  annotate("text", x = c(-crit2, crit2), y = 0.425,
           label = paste0(c("-crit = -", "+crit = "), round(crit2, 2)), colour = "red", size = 4.2) +
  annotate("segment", x = t_obs, xend = t_obs, y = 0, yend = 0.33, colour = "#1E9E1E", linetype = "dotted", linewidth = 1.4) +
  annotate("text", x = t_obs - 0.08, y = 0.36, label = "observed\nt = 1.94", colour = "#1E9E1E", hjust = 1, size = 3.8, lineheight = 0.9) +
  annotate("text", x = c(-3.4, 3.4), y = 0.06, label = "2.5%", size = 4.2, colour = "grey20") +
  labs(x = "t", y = "Density")
```

Applied to our actual result, $t(9) = 1.94$: the one-tailed $p$ is .042, the two-tailed $p$ is .084, exactly double, since a two-tailed test counts the matching extreme region on both sides instead of just one. Here the choice actually changes the decision: one-tailed, this result is significant at $\alpha=.05$; two-tailed, it isn't. That's why a one-tailed test has to be justified by your theory *before* you see the data, not chosen afterward because it gives the answer you wanted.

### Which one should you use?

The rule is Chapter 4's, unchanged: default to two-tailed, and go one-tailed only when a result in the opposite direction would lead you to do the same thing as no result. With a $t$-test you can see the difference in the critical value: at the same $\alpha$, the one-tailed $t_{crit}$ is closer to 0 than the two-tailed one.

### Assumptions, and what to do when they don't hold {#sec-6assumptions}

Every test in this book has assumptions. Assumptions are conditions that need to be roughly true for the math behind the test to be trustworthy. There are two different kinds of assumptions.

**The universal assumption.** Every test in this book, including the nonparametric tests in Chapter 7, assumes your observations are independent (in paired designs, the pairs) and came from a sound study design (a fair sample, not one that systematically favors certain values). This assumption is so constant that from here on, we won't re-list it test by test. Keep it in the back of your mind, but if an exam or homework question asks you to "list the assumptions of this test," this isn't the answer they're looking for.

**Test-specific assumptions.** These come from how a particular test's math actually works, and they differ test to test. For the one-sample and paired $t$-tests, the specific assumption is approximate normality of the outcome itself (one-sample) or of the difference scores (paired). The $t$-distribution we've been using to get $p$-values is *derived* assuming the population is normal, which makes the sample mean normally distributed too. If your data are badly skewed and your sample is small, that assumption fails, and the $p$-value may not be the true probability of a result this extreme under $H_0$.

**How worried should you be?** It depends on $n$. With a large sample, the Central Limit Theorem does a lot of the work for you. The sampling distribution of the sample mean tends toward normal even when individual observations are skewed, so the $t$-test is fairly **robust** to non-normality once $n$ is reasonably large. A lot of environmental data, particularly field data, instead comes from small samples that are also skewed. Examples include a handful of expensive water samples, a short list of monitoring wells, or the five plots you had time to visit before the storm rolled in. A $t$-test's answer becomes questionable when the sample is small and the data are also non-normal.

**What to do about it.** First, look at your data. A histogram or a Q-Q plot will usually tell you whether "approximately normal" is a reasonable description or a stretch. You can also run a formal normality test, the **Shapiro-Wilk test** (`shapiro.test()` in R), which tests $H_0$: the data came from a normal distribution. Treat it as one more piece of evidence, though, rather than a strict gate. With a small sample, exactly when a violation matters most, it has low power and will often say "no problem" even when there is one. If your data look far from normal and your sample is small, you have another option: a **nonparametric test**, which does not assume normality. You'll meet these nonparametric alternatives in the next chapter, once we've covered all three $t$-test forms below.

## Paired t: before/after or matched

When the same units are measured twice (before/after, matched pairs), the question is whether they changed. The natural data unit is the **within-unit difference**.

**Design in this worked example.** Ecologists measured soil moisture at 20 field plots before wetland restoration and again one year later, to see whether restoration increased moisture enough to support plant reestablishment and carbon sequestration. @fig-6paireddesign shows the design.

```{r}
#| label: fig-6paireddesign
#| fig-cap: "A paired design. The dashed boxes and the note underneath mark the feature that defines it: there is one set of 20 plots, measured on two occasions, not two separate samples."
#| fig-alt: "Study design schematic. IV: restoration, before versus after. Two dashed boxes labeled Before and After, each n = 20 plots, with a note that these are the same 20 plots measured on both occasions. DV: soil moisture, percent."
#| fig-asp: 0.34
#| out-width: "100%"

design_groups("Restoration (before vs. after)", c("Before", "After"), 20, "plots",
              "Soil moisture (%)", "Did moisture change within each plot?",
              kind = "Paired", dashed = TRUE,
              note = "the same 20 plots, measured on both occasions")
```

The dashed boxes mark what makes this design paired. In every design so far the boxes held *different* units, and the test compared one set against another. Here both boxes hold the same 20 plots, so each plot's Before value can be subtracted from its own After value.

**Define differences and their meaning.** Choose an order and stick to it:

-   $d_i = \text{After}_i - \text{Before}_i$
-   $d_i>0$: plot had more soil moisture after restoration (increase)
-   $d_i<0$: plot had less soil moisture after restoration (decrease)

This subtraction cancels between-plot variability (some plots are naturally wetter than others) and focuses on within-plot change.

**Test on differences (paired $t$).** Compute one difference per plot, then run a one-sample $t$ against 0:

$$\bar d=\frac{1}{n}\sum_{i=1}^n d_i, \quad
s_d=\text{SD}(d_1,\ldots,d_n), \quad
t = \frac{\bar d - 0}{s_d/\sqrt{n}}, \quad df=n-1.$$

The sign of $t$ matches $\bar d$ and your difference direction.

**What to report.** $\bar d$ (the mean change), a 95% CI, $t(df)$, and $p$. Optionally add a standardized effect size, $\bar d / s_d$, the paired version of Cohen's $d$.

**R pointers (you'll do this in lab).**

-   Compute differences: `d <- after - before`
-   Paired $t$: `t.test(d, mu = 0)` or `t.test(after, before, paired = TRUE)`
-   CI for $\bar d$: from the `t.test` output

### Calculate t

To calculate $t$ for a paired-samples $t$-test, we use the one-sample $t$-test formula from the previous section, applied to the difference scores instead of raw scores. The paired-samples $t$-test is a one-sample $t$-test on a single group of numbers, the differences.

Let's find $t$ for the mean difference scores:

```{r, echo=FALSE}
# Soil moisture (%) at 5 paired restoration plots: before and after restoration
before <- c(18, 22, 16, 25, 21)
after  <- c(21, 26, 22, 29, 18)

paired_sample_df <- data.frame(
  plot   = 1:5,
  Before = before,
  After  = after
)

differences     <- paired_sample_df$After - paired_sample_df$Before
diff_from_mean  <- differences - mean(differences)
Squared_differences <- diff_from_mean^2

paired_sample_df <- cbind(paired_sample_df,
                          differences,
                          diff_from_mean,
                          Squared_differences)

paired_sample_df <- paired_sample_df %>%
  rbind(c("Sums",  colSums(paired_sample_df[1:5, 2:6]))) %>%
  rbind(c("Means", colMeans(paired_sample_df[1:5, 2:6]))) %>%
  rbind(c(" ", " ", " ", " ", "sd",  round(sd(paired_sample_df[1:5, 4]), digits=3))) %>%
  rbind(c(" ", " ", " ", " ", "SE",  round(sd(paired_sample_df[1:5, 4])/sqrt(5), digits=3))) %>%
  rbind(c(" ", " ", " ", " ", "t",   mean(differences)/round(sd(paired_sample_df[1:5, 4])/sqrt(5), digits=3)))
paired_sample_df[6, 5] <- 0
paired_sample_df[7, 5] <- 0
knitr::kable(paired_sample_df,
             col.names = c("Plot", "Before", "After", "Difference",
                           "Deviation from mean difference", "Squared deviation"))
```

The five differences average 2.8 percentage points, with $s_d = 3.42$, so $SE = 3.42/\sqrt{5} = 1.53$ and $t = 2.8/1.53 = 1.83$.

If we did this test using R, we would obtain almost the same numbers (there is a little bit of rounding in the table).

```{r}
paired5_out <- t.test(differences, mu = 0)
paired5_tidy <- broom::tidy(paired5_out) |>
  mutate(statistic = round(statistic, 2), p.value = fmt_p(p.value))

knitr::kable(
  paired5_tidy[, c("estimate", "statistic", "parameter", "p.value")],
  col.names = c("Mean difference", "t", "df", "p-value"),
  caption = "Paired t-test on 5 restoration plots"
)
```

Our $t$-test results are: *t*(4) = 1.83, *p* = .141.

The sign of $t$ matches the sign of the mean difference (ours was $+2.8$ percentage points), so soil moisture increased on average, as $\bar d$ showed. Getting a $p$-value from this $t$, and choosing one-tailed vs. two-tailed, works exactly like it did for the one-sample case (see the One-sample $t$-test section above). Two-tailed, $p=.141$, which is not significant. With only five plots, a mean increase this large would not be unusual even if restoration had no effect on soil moisture.

### Scaling to the full sample

The five plots above are a subset of the full restoration study, which measured all 20 paired plots.

```{r, echo=FALSE}
before_full <- c(18, 22, 16, 25, 21, 27, 26, 23, 28, 23, 19, 19, 20, 25, 15, 24, 23, 16, 18, 21)
after_full  <- c(21, 26, 22, 29, 18, 31, 31, 29, 29, 28, 16, 19, 20, 19, 18, 28, 25, 21, 18, 19)

full20_out <- t.test(after_full, before_full, paired = TRUE)
full20_tidy <- broom::tidy(full20_out) |>
  mutate(statistic = round(statistic, 2), p.value = fmt_p(p.value))

knitr::kable(
  full20_tidy[, c("estimate", "statistic", "parameter", "p.value")],
  col.names = c("Mean difference", "t", "df", "p-value"),
  caption = "Paired t-test on all 20 restoration plots"
)
```

At $n=20$, $t(19) = `r round(full20_out$statistic, 2)`$, $p =`r fmt_p(full20_out$p.value)`$: significant. The mean difference (`r round(full20_out$estimate, 1)` percentage points) is smaller than the 5-plot subset's `r round(mean(differences), 1)`. The larger $t$ comes from the standard error, which fell from `r round(sd(differences)/sqrt(5), 2)` to `r round(sd(after_full - before_full)/sqrt(20), 2)` with four times as many plots. A mean increase of `r round(full20_out$estimate, 1)` points would be unlikely if restoration had no effect, so the extra plots gave more power to detect a real effect.

## Two independent groups

The independent-samples $t$-test is the third and last design in this chapter. The logic is the same as the one-sample and paired cases: a mean difference divided by an estimate of its variability. Only the research design is different.

Use an independent-samples $t$-test for a **between-subjects** design: different people or units in each condition, each measured once. With two groups of 5, that's 10 units in total, none of them paired. Without repeated measures on the same unit, there's no meaningful way to subtract scores between conditions the way the paired $t$-test does.

The numerator works the same way it always has: the mean difference between groups. The denominator differs from the paired case, because now we need a single estimate of variability built from two separate samples instead of one.

A natural first guess is to average the two groups' standard errors:

$$t = \frac{\bar{A}-\bar{B}}{\left(\dfrac{SE_A+SE_B}{2}\right)}$$

That average understates how much a difference between two means varies. The standard error of a difference between two independent means is $\sqrt{SE_A^2 + SE_B^2}$, which is always larger than the average of the two SEs. Instead of averaging, the independent-samples $t$-test pools the two groups' variances, weighted by their degrees of freedom, into a single estimate for the denominator:

$$t = \frac{\bar{x}_A-\bar{x}_B}{s_p\sqrt{\dfrac{1}{n_A}+\dfrac{1}{n_B}}}$$

where the pooled standard deviation $s_p$ is (note $s_A^2$ and $s_B^2$ here are variances, not standard deviations):

$$s_p = \sqrt{\frac{(n_A-1)s_A^2 + (n_B-1)s_B^2}{n_A+n_B-2}}$$

The pooled test has $df = n_A + n_B - 2$, one degree of freedom used for each group's mean. With two groups of 5, that is $df = 8$, the 8 in $t(8)$ below.

The pooled test adds two test-specific assumptions: the outcome is approximately normal within each group, and the two groups have similar variances. When the spreads clearly differ, use Welch's version instead.

The example below computes $t$ for two small groups by following that formula step by step in R, then checks it against `t.test()`.

Here we deliberately print R's raw console output instead of a formatted table, because this is the block you will see in lab, and you can check that the number you built by hand matches the `t = ` value in it.

```{r, echo=TRUE }

## By "hand" using R
a <- c(1,2,3,4,5)
b <- c(3,5,4,7,9)

mean_difference <- mean(a)-mean(b) # compute mean difference

variance_a <- var(a) # compute variance for A
variance_b <- var(b) # compute variance for B

# Compute top part and bottom part of sp formula

sp_numerator <- (4*variance_a + 4* variance_b) 
sp_denominator <- 5+5-2
sp <- sqrt(sp_numerator/sp_denominator) # compute sp


# compute t following the formula

t <- mean_difference / ( sp * sqrt( (1/5) +(1/5) ) )

t # our hand-computed t


# the same test via R's built-in function: compare the "t = " value
t.test(a,b, paired=FALSE, var.equal = TRUE)


```

@tbl-6three-tests sets the three forms side by side. Each one divides an effect by its standard error. The design decides what counts as the effect and how its standard error is estimated.

| | One-sample | Paired | Independent-samples |
|---|---|---|---|
| Design | One sample against a reference value | The same units measured twice | Two separate groups |
| Effect (numerator) | $\bar{x} - \mu_0$ | $\bar{d}$ | $\bar{x}_A - \bar{x}_B$ |
| Standard error (denominator) | $s/\sqrt{n}$ | $s_d/\sqrt{n}$ | $s_p\sqrt{1/n_A + 1/n_B}$ |
| $df$ | $n - 1$ | $n - 1$ ($n$ = number of pairs) | $n_A + n_B - 2$ |
| R call | `t.test(x, mu = 50)` | `t.test(after, before, paired = TRUE)` | `t.test(a, b, var.equal = TRUE)` |

: The three $t$-tests compared. {#tbl-6three-tests}

::: {.callout-important title="Results reporting: t-tests"}
All three forms report the same four things: the direction in plain language, the effect size, the test statistic with its df, and $p$. The three forms differ only in what counts as the effect.

One-sample, where the effect is the sample mean against the reference value:

> Mean quiz score was 61%, which did not differ significantly from the 50% expected by chance, *t*(9) = 1.94, *p* = .084, 95% CI \[48.18, 73.82\].

Paired, where the effect is the mean *difference* score:

> Soil moisture increased by 1.90 percentage points after restoration, *t*(19) = 2.50, *p* = .022, 95% CI \[0.31, 3.49\].

Independent-samples, where the effect is the difference between two group means:

> Group A scored lower than Group B, a mean difference of -2.6, *t*(8) = -2.02, *p* = .078, 95% CI \[-5.57, 0.37\].

Note the paired write-up never mentions the before and after means separately. The difference score *is* the unit of analysis, so that single number plus its CI is the whole result.
:::

## Simulating data for t-tests

Chapter 4 simulated *group mean differences* to build a chance window. Here we simulate the test statistic itself, which shows two things the earlier simulation could not: what the null distribution of $t$ actually looks like, and what happens to $p$-values when $H_0$ is true.

Simulating under conditions you choose builds intuition for how sampling error affects your results given your design: sample size, the size of the difference you expect, and the amount of variation. Formal power analyses use the same methods.

### Simulating a one-sample t-test

A simulation has two steps.

1.  Decide what the population looks like. Here each sample is 30 sites, with values drawn from a normal distribution with mean 50 and SD 25.
2.  Draw one sample from that population, run the $t$-test, and save the statistics you care about ($t$ and $p$). Repeat 1,000 times and look at how they are distributed.

Every test compares the sample mean to $\mu_0 = 50$, which is also the true population mean, so $H_0$ is true in every repetition. @fig-6nullt is the null distribution of $t$ that this produces.

```{r}
# steps to create fake data from a distribution
# and conduct t-tests on the simulated data
save_ps <- numeric(1000)
save_ts <- numeric(1000)
for (i in 1:1000) {
  my_sample <- rnorm(n = 30, mean = 50, sd = 25)
  t_test <- t.test (my_sample, mu = 50)
  save_ps[i] <- t_test$p.value
  save_ts[i] <- t_test$statistic
}
```

```{r}
#| label: fig-6nullt
#| fig-cap: "The distribution of $t$-values under the null. These are the $t$ values that are produced by chance alone."
#| fig-alt: "Histogram of 1000 simulated t statistics from samples with no true effect, symmetric and bell-shaped, centered at zero, with fewer and fewer values as t moves further from zero in either direction."

hist_t(save_ts)
```

```{r}
#| label: fig-6flatp
#| fig-cap: "The distribution of $p$-values that are observed is flat under the null."
#| fig-alt: "Histogram of 1000 simulated p-values from samples with no true effect, roughly flat across the full range from 0 to 1, with bars of similar height in every bin including the bin closest to zero."

hist_p(save_ps)
```

The simulated $t$ distribution looks like a $t$ distribution should, and it shows us how often we get $t$ values of particular sizes. The $p$-distribution is flat under the null, which we are simulating here. This means you have the same chance of getting a $t$ with a p-value between 0 and 0.05 as you would of getting one between .90 and .95. Those ranges are both ranges of 5%, so they hold an equal share of the $t$ values by definition.

You can write the same simulation more compactly with `replicate()` instead of a `for` loop; it produces the same two distributions above:

```{r, echo=TRUE, eval=FALSE}
simulated_ts <- replicate(1000, t.test(rnorm(30, 50, 25), mu = 50)$statistic)
simulated_ps <- replicate(1000, t.test(rnorm(30, 50, 25), mu = 50)$p.value)
```

### Simulating paired and independent-samples tests

The paired and independent-samples cases behave the same way, because they use the same $t$ calculation on different inputs. A paired test runs a one-sample $t$ on the difference scores; an independent-samples test swaps in the pooled standard error. Simulate either one under $H_0$ and you get the same two pictures: a $t$ distribution centered on zero, and a flat $p$ distribution.

To adapt the code above, change only the line that generates the data:

```{r, echo=TRUE, eval=FALSE}
# paired: one sample of difference scores
d <- rnorm(10, 0, 5) - rnorm(10, 0, 5)
t.test(d, mu = 0)$p.value

# independent samples: two separate groups
t.test(rnorm(10, 10, 5), rnorm(10, 10, 5), var.equal = TRUE)$p.value
```

Next, give one condition a larger mean than the other, so the simulated data contain a real difference:

```{r, echo=TRUE}

save_ps <- numeric(1000)
save_ts <- numeric(1000)
for ( i in 1:1000 ){
  condition_A <- rnorm(10,10,5)
  condition_B <- rnorm(10,13,5)
  t_test <- t.test(condition_A, condition_B, var.equal = TRUE)
  save_ps[i] <- t_test$p.value
  save_ts[i] <- t_test$statistic
}

```

```{r}
#| label: fig-6simtrueps
#| fig-cap: "1,000 simulated $p$-values ($n = 10$) when there is a true difference."
#| fig-alt: "Histogram of 1000 simulated p-values from samples of size 10 with a real underlying difference. Unlike the flat null histogram, this one is right-skewed, with the tallest bars near zero and progressively shorter bars moving toward 1."
hist_p(save_ps)
```

Now you can see that the $p$-value distribution is right-skewed: most of the values pile up near 0, with a long tail toward 1. A true effect produces larger $t$ values than chance alone would, so $p$-values below .05 turn up more often.

Here $H_0$ is false, so rejecting it when $p < .05$ is a correct decision, not a Type I error. The code below counts how often that happens in the 1,000 experiments.

```{r, echo=TRUE}
length(save_ps[save_ps<.05])
length(save_ps[save_ps<.05])/1000
```

We got `r length(save_ps[save_ps<.05])` simulations where p was less than .05, or `r 100*length(save_ps[save_ps<.05])/1000`% of experiments. That percentage is the design's power. A study with power this low will usually miss a difference that is really there, so running it as designed spends the time and money of data collection on a result that is most likely to come back not significant. Improve the design before you run it.

The most direct lever is $n$, the number of subjects. Let's increase $n$ from 10 to 100 and see what happens to the number of "significant" simulated experiments.

```{r, echo=TRUE}

save_ps <- numeric(1000)
save_ts <- numeric(1000)
for ( i in 1:1000 ){
  condition_A <- rnorm(100,10,5)
  condition_B <- rnorm(100,13,5)
  t_test <- t.test(condition_A, condition_B, var.equal = TRUE)
  save_ps[i] <- t_test$p.value
  save_ts[i] <- t_test$statistic
}

```

```{r}
#| label: fig-6simpsB
#| fig-cap: "1,000 simulated $p$-values ($n = 100$) when there is a true difference."
#| fig-alt: "Histogram of 1000 simulated p-values from samples of size 100 with the same real underlying difference as the previous figure. Almost every simulation now produces a p-value near zero, with a tall spike in the leftmost bin and almost nothing anywhere else."
hist_p(save_ps)
```

Now almost all of the experiments (`r sum(save_ps < .05)` of 1000) show a $p$-value of less than .05 (using a two-tailed test, the default in R). You can use this simulation process to determine how many subjects you need to reliably detect your effect.

## Chapter Summary

**Why it matters.** The $t$-test measures a signal against the noise around it. ANOVA, regression, and the rest of the course use the same logic.

**Core ideas**

-   **Signal over noise.** A difference is only impressive relative to the variability around it. Every $t$ statistic is a ratio: the effect you observed on top, an estimate of how much that effect could bounce around by chance on the bottom.
-   **Three forms, one idea.** The one-sample, paired, and independent-samples $t$-tests differ in how they compute the effect and its standard error, but all three ask whether an observed difference is large compared to its own standard error. The paired test is just the one-sample test applied to difference scores.
-   **Why $t$ and not $z$.** When the population SD $\sigma$ is unknown, we estimate it with $s$, which adds uncertainty. The $t$ distribution accounts for that by being fatter in the tails than the normal, and it tightens toward the normal as degrees of freedom grow.
-   **Assumptions come in two kinds.** *Independence and a sound study design* apply to every test in this book and are assumed from here on. The *test-specific* assumption is approximate normality: of the outcome (one-sample), of the difference scores (paired), or of each group (independent samples). The pooled independent-samples test also assumes similar variances.
-   **How worried to be depends on $n$.** Thanks to the Central Limit Theorem, $t$-tests are fairly robust to non-normality once samples are reasonably large. $t$-tests are least trustworthy when samples are small *and* skewed, which describes a lot of environmental field data. Plot your data first, and treat a Shapiro-Wilk result as evidence rather than a gate, since it has the least power with small samples, when a violation matters most.
-   **Choose your tails before you look.** Default to two-tailed. Reserve one-tailed tests for cases where a difference in the opposite direction would lead to the same action as no difference, and decide before seeing the result.
