---
author: 
  - Mallory Barnes
aliases: [redirects/t-tests.html]
---

```{r, include = FALSE}
source("global_stuff.R")
set.seed(538)
```

```{r, echo=FALSE}
library(ggplot2)
library(dplyr)
library(broom)

fmt_p <- function(p) ifelse(p < .001, "< .001", formatC(p, digits = 3, format = "f"))

# Histogram helpers for the simulation section, so those figures match the
# rest of the book instead of falling back to base graphics.
hist_t <- function(ts, title = NULL) {
  ggplot(data.frame(t = ts), aes(x = t)) +
    geom_histogram(bins = 30) +
    labs(x = "Simulated t", y = "Count", title = title) +
    theme_classic(base_size = 13)
}

hist_p <- function(ps, title = NULL) {
  ggplot(data.frame(p = ps), aes(x = p)) +
    geom_histogram(breaks = seq(0, 1, .05)) +
    scale_x_continuous(breaks = seq(0, 1, .1)) +
    labs(x = "Simulated p-value", y = "Count", title = title) +
    theme_classic(base_size = 13)
}
```

# t-tests [note: chapter title rendering incorrectly in sidebar table of contents]

## Intro

In the early 1900s, William Sealy Gosset joined **Guinness**, then the world's largest brewery. Guinness hired scientifically trained brewers and gave them latitude to pursue research that improved the product. As production scaled, quality control mattered: a pint should taste the same in Dublin as in Detroit. Until then, hop quality was judged mostly by appearance and smell, which was unreliable. Chemical assays could measure ingredients, but the core problem was statistical. Inference relied on the normal (Z) distribution, which works well when samples are large enough for a normal approximation. In practice, Guinness could not destroy large volumes of beer or ingredients just to test them.

Gosset set out to quantify the extra uncertainty introduced by **small** samples. How much wider is the error around an estimate when $n=10$ instead of $n=1000$? He derived a new sampling distribution that accounts for small-sample variability, now known as the **Student's** $t$ distribution. Because Guinness restricted external publications, he published the 1908 paper under the pseudonym **"Student"** [@Student1908]. That is why we still say **Student's** $t$-test. The core idea you will use today is the same: measure a signal, scale it by its standard error, and compare the resulting $t$ to the appropriate $t$ distribution for your $n-1$ degrees of freedom.

**Why this matters.** We often compare a sample mean to a reference (drinking-water standard, historical average) or compare two conditions (before/after, site A vs site B). A t-test asks whether the observed difference is larger than you would expect from sampling variation alone, given your $n$ and variability.

**Three common t-tests you will use.**

1.  **One-sample t-test.** Compare a sample mean to a target ($\mu_0$).
2.  **Paired t-test.** Compare before/after (or matched) measurements on the same units by running a one-sample t on the *differences*.
3.  **Independent-samples t-test (also called two-sample t-test)** Compare means from two groups measured on different units. Use Welch's test by default when variances differ.

::: {.callout-tip title="Tip: What you will report"}
Always report:

- the effect size (mean difference)
- a 95% CI
- the test statistic with its df
- the $p$-value

Example: "Mean change was -0.42 units (95% CI \[-0.73, -0.11\]; *t*(11) = -2.92; p = 0.014)."
:::

## From mean & SD to "signal over noise"

A mean tells you where the center is; the SD tells you how spread out the data are. A mean of 50 could be very representative (if SD is tiny) or nearly meaningless (if SD is huge). You need both.

A simple way to combine them is a ratio:

$$\frac{\text{mean}}{\text{SD}}$$

- Big numerator, small denominator → big ratio (more confidence that the mean reflects the data).

- Small numerator or big denominator → small ratio (less confidence).

For instance, mean = 50 with SD = 1 gives 50; mean = 50 with SD = 100 gives 0.5. Same mean, very different stories.

This "signal over noise" pattern shows up everywhere in inference. In general,

$$\text{statistic}=\frac{\text{effect}}{\text{error}}.$$

For $t$-tests, we won't actually use SD in the denominator, we use the **standard error** $SE$, which is the SD of the sample mean. But the logic is the same: measure a signal, scale it by its uncertainty, and see whether the resulting ratio is unusually large or small under the right $t$ distribution.

$$t = \frac{\text{effect}}{\text{standard error}} = \frac{\bar{X} - \mu_0}{s / \sqrt{n}}$$

We now turn that ratio into a test statistic for a *single mean*.

## One-sample t-test

Use a one-sample $t$ when you want to test whether a sample mean differs from a reference value ($\mu_0$). The question: is the observed difference bigger than you'd expect from sampling variability, given $n$ and your sample SD $s$?

$$t = \frac{\bar{X}-\mu_0}{s/\sqrt{n}}, \qquad df = n-1$$

After calculating $t$ from our data, we compare the observed $t$ to a $t$ distribution with $n-1$ degrees of freedom to get a $p$-value and a decision at $\alpha$.

### Comparing $z$ and $t$

When the population SD is known (or we have a large enough sample to assume convergence), the ratio is called $z$; when it's estimated from the data, it's $t$.

Under a known population SD ($\sigma$), the standardized mean uses the $z$ statistic:
$$z = \frac{\bar{X}-\mu_0}{\sigma/\sqrt{n}}$$

When $\sigma$ is **unknown** (the usual case), we estimate it with the sample SD ($s$) and use the $t$ statistic:
$$t = \frac{\bar{X}-\mu_0}{s/\sqrt{n}}, \qquad df = n-1$$

Because $s$ varies from sample to sample, the $t$ distribution has **heavier tails** than the standard normal. As $n$ grows, $s$ stabilizes and $t$ **converges** to $z$.

```{r}
#| label: fig-7-z-vs-t-linear-log
#| fig-cap: 'Normal ($z$) vs $t$ distributions: on a linear scale (top), tail differences look small; on a log scale (bottom), the heavy tails of small-$n$ $t$ are clear.'
#| out-width: "100%"

library(tidyverse)
library(patchwork)

xs <- seq(-4, 4, length.out = 2000)

df_vals <- tibble(
  dist = c("z (standard normal)", "t (df=10)", "t (df=2)"),
  df    = c(NA, 10, 2)   # NA for z
)

# Build a long data frame: every x crossed with each distribution
plot_df <- tidyr::crossing(x = xs, df_vals) |>
  mutate(density = case_when(
    dist == "z (standard normal)" ~ dnorm(x),
    dist == "t (df=10)"           ~ dt(x, df = 10),
    dist == "t (df=2)"            ~ dt(x, df = 2)
  ))

base_cols <- c("z (standard normal)" = "#0072B2",
               "t (df=10)"           = "#E69F00",
               "t (df=2)"            = "#009E73")

p_lin <- ggplot(plot_df, aes(x, density, color = dist)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = base_cols, guide = guide_legend(title = NULL)) +
  coord_cartesian(xlim = c(-4, 4), ylim = c(0, 0.45)) +
  labs(x = NULL, y = "Density (linear scale)") +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

p_log <- ggplot(plot_df, aes(x, density, color = dist)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = base_cols, guide = "none") +
  scale_y_log10() +
  coord_cartesian(xlim = c(-4, 4)) +
  labs(x = "value", y = "Density (log scale)") +
  theme_classic(base_size = 12)

p_lin / p_log

```

::: {.callout-tip title="Tip: When is $z$ appropriate?"}
If the population SD $\sigma$ is known *and* the sampling distribution of the mean is (approximately) normal, use $z$. In practice, $\sigma$ is rarely known, so we use $t$ with $s$ and $df=n-1$.
:::

**Rule of thumb.** For two-tailed tests at $\alpha=0.05$:

- $z$ uses critical values $\pm 1.96$.
- $t$ is a bit wider for small $n$ (e.g., $df=9 \Rightarrow \pm 2.262$; $df=29 \Rightarrow \pm 2.045$), and approaches $\pm 1.96$ as $n$ increases.

### What t represents

Either way, the ratio asks the same question: how far is the mean from the reference in SE units?

$t$ expresses how far a sample mean is from the hypothesized mean, in **standard error units** (signal over noise):

$$t = \frac{\bar{X}-\mu_0}{s/\sqrt{n}}$$

$t$ can be positive or negative depending on direction. Under $H_0$, $t \sim t_{(df=n-1)}$.\
**Rule of thumb:** at $\alpha=.05$ two-tailed, $|t|\gtrsim 2$ is uncommon when $df\approx 20$.[Is it really a "rule of thumb" if it's just for n = 20? Seems ewird to say that]

::: {.callout-note title="Note: Standard error, again"}
Remember that $s/\sqrt{n}$ is the **standard error (SE)**, the standard deviation of the sampling distribution of the mean.
:::

Let's see how $t$ values behave when samples really come from the null hypothesis. Under the null hypothesis, $t$ follows a $t$ distribution with $df = n-1$.

```{r}
#| label: fig-7t-sim
#| fig-cap: 'Histogram of 1,000 simulated $t$ values ($n=10$) drawn from a population with $\mu_0=0$. Most $t$s cluster near 0; large positive or negative values are rare.'
t_vals <- replicate(1000, {
  s <- rnorm(10, 0, 1)
  (mean(s) - 0) / (sd(s) / sqrt(length(s)))
})
ggplot(data.frame(t_vals), aes(x = t_vals)) +
  geom_histogram(bins = 30, color = "white", linewidth = 0.2) +
  labs(x = "t values", y = "Count")
```

**Takeaway:** $t$ measures how far the sample mean is from the null expectation, scaled by its uncertainty. It behaves predictably under $H_0$, allowing us to decide whether a difference is likely due to chance.

### Calculating t from data

Let's calculate a one-sample $t$ statistic step by step using a short example. Ten students each took a 5-question true/false quiz (50% expected by chance). Their scores (percent correct) are:

**Scores**: 50, 70, 60, 40, 80, 30, 90, 60, 70, 60

```{r}
#| label: ex-onesample-t
#| echo: FALSE

scores <- c(50, 70, 60, 40, 80, 30, 90, 60, 70, 60)
n  <- length(scores)
x  <- mean(scores)
s  <- sd(scores)
SE <- s / sqrt(n)

tt <- t.test(scores, mu = 50)
ci_lo <- round(tt$conf.int[[1]], 2)
ci_hi <- round(tt$conf.int[[2]], 2)
```

By hand, the standard error of the mean is the sample SD divided by the square root of $n$:

$$\text{SEM} = \frac{s}{\sqrt{n}} = \frac{`r round(s,2)`}{\sqrt{`r n`}} = `r round(SE,2)`$$

And $t$ is the difference between the sample mean (`r round(x,2)`) and the reference value (50, chance level), divided by that SEM:

$$t = \frac{\bar{X}-\mu_0}{SEM} = \frac{`r round(x,2)`-50}{`r round(SE,2)`} = `r round(tt$statistic,2)`$$

We can confirm this with a single R command, `t.test(scores, mu = 50)`:

**Results:** $t(9)=`r round(tt$statistic,2)`$, $p=`r signif(tt$p.value,3)`$; mean $=`r round(tt$estimate,2)`$ (95% CI \[`r ci_lo`,`r ci_hi\`\]).

**Interpretation:** The mean (`r round(x,0)`%) is above chance (50%), but the difference is not statistically significant at $\alpha=.05$.

**Assumptions check:** Observations are independent, and with $n=10$ the data are approximately normal, sufficient for a one-sample $t$-test.

### Assumptions, and what to do when they don't hold

Every test in this book rests on assumptions. Assumptions are conditions that need to be roughly true for the math behind the test to be trustworthy. There are two different kinds of assumptions.

**The universal assumption.** Every single test we cover in this class (one-sample $t$, paired $t$, independent $t$, ANOVA, regression) assumes your observations are **independent** and came from a **sound study design** (a fair sample, not one that systematically favors certain values). This assumption is so constant that from here on, we won't re-list it test by test. Keep it in the back of your mind, but if an exam or homework question asks you to "list the assumptions of this test," this isn't the answer they're looking for.

**Test-specific assumptions.** These come from how a particular test's math actually works, and they differ test to test. For the one-sample and paired $t$-tests, the specific assumption is **approximate normality** of the outcome itself (one-sample) or of the difference scores (paired). Why does this matter? The $t$-distribution we've been using to get $p$-values is *derived* assuming the sampling distribution of the mean is normal. If your data are badly skewed and your sample is small, that derivation is on shaky ground, and the $p$-value it hands you may not mean what it claims to mean.

**How worried should you be?** It depends enormously on $n$. With a large sample, the Central Limit Theorem does a lot of the work for you. The sampling distribution of the mean tends toward normal even when individual observations are skewed, so the $t$-test is fairly **robust** to non-normality once $n$ is reasonably large. A lot of environmental data, particularly field data, instead comes from **small samples that are also skewed**. Examples include a handful of expensive water samples, a short list of monitoring wells, five plots you had time to visit before the storm rolled in. Small $n$ and non-normality reinforcing each other is when a $t$-test's answer becomes questionable.

**What to do about it.** First, look at your data. A histogram or a QQ-plot will usually tell you whether "approximately normal" is a reasonable description or a stretch. You can also run a formal normality test, the **Shapiro-Wilk test** (`shapiro.test()` in R), which tests $H_0$: the data came from a normal distribution. Treat it as one more piece of evidence rather than a strict gate though: with a small sample, exactly when a violation matters most, it has low power and will often say "no problem" even when there is one. If your data look like a stretch and your sample is small, you have another option: a **nonparametric test** that doesn't lean on the normality assumption at all. You'll meet these nonparametric alternatives in the next chapter, once we've covered all three t-test forms below.

### How does t behave?

Extreme sample means are unlikely if $H_0$ is true, which is why large $|t|$ values are rare. Returning to the rule of thumb above [rule of thumb seems weird here], values of $t$ beyond roughly $-2$ or $+2$ occur less than 5% of the time under the null when $df \approx 20$. That's the foundation of the decision rule for $t$ tests: if your observed $t$ lies in those rare tails, you reject $H_0$.

Remember, if we obtained a single $t$ from one sample we collected, we could consult the chance window in @fig-7-t-rejection below to find out whether the $t$ we obtained from the sample was likely or unlikely to occur by chance.[This renders weird, all of blue says "critical values", and the shading is inconsistent on the two sides of the graph]

```{r}
#| label: fig-7-t-rejection
#| fig-cap: 'Two-tailed rejection regions at $\alpha=.05$ for $t_{(df=9)}$. Shaded tails mark unlikely values under $H_0$.'
#| out-width: "80%"


library(tidyverse)

df   <- 9
xs   <- seq(-4.5, 4.5, length.out = 2000)
dens <- dt(xs, df)

crit <- qt(c(.025, .975), df)

t_tibble <- tibble(x = xs, y = dens,
                   tail = case_when(x <= crit[1] ~ "Left tail",
                                    x >= crit[2] ~ "Right tail",
                                    TRUE ~ "Center"))

ggplot(t_tibble, aes(x, y, fill = tail)) +
  geom_area(alpha = 0.4, color = NA) +
  geom_line(aes(y = y), linewidth = 0.9, color = "black") +
  geom_vline(xintercept = crit, linetype = 2) +
  annotate("text", x = mean(crit), y = 0.02, label = "critical values", vjust = -0.7) +
  labs(x = "t", y = "Density") +
  theme_classic(base_size = 12) +
  theme(legend.position = "none")

```

### Interpreting $t$

We're asking whether our sample could plausibly have come from the null distribution: the distribution of $t$-values we'd see across many samples of size 10, if the true mean were exactly 50 (chance level). @fig-7tnull shows what that distribution looks like, at $df=9$.

```{r}
#| label: fig-7tnull
#| fig-cap: "A distribution of $t$-values that can occur by chance alone, when there is no difference between the sample and a population"


range <- seq(-3, 3, .1)
null_distribution <- dt(range, 9, log = FALSE)
plot_df <- data.frame(range, null_distribution)

ggplot(plot_df, aes(x = range, y = null_distribution)) +
  geom_line() +
  xlab("t-values") +
  ylab("Probability") +
  theme(axis.text.y = element_blank(), axis.ticks = element_blank()) +
  scale_x_continuous(breaks = (seq(-3, 3, .5))) +
  geom_label(data = data.frame(x = -.7, y = .1, label = "50% \n (-)"),
             aes(x = x, y = y, label = label)) +
  geom_label(data = data.frame(x = .7, y = .1, label = "50% \n (+)"), aes(x = x, y = y, label = label)) +
  geom_vline(xintercept = 0)+
  theme_classic(base_size = 20)


```

$t$ is most likely to land near zero, which makes sense: this is the distribution of *no* differences, so most of the time chance alone won't push it far from 0. Because the distribution is symmetric, a $t$-value from the null distribution is just as likely to be positive as negative.

That symmetry gives us a few basic probabilities for free:

1.  $t$ is always zero, positive, or negative, so $p(t=0 \text{ or } t>0 \text{ or } t<0) = 1$.
2.  $p(t \geq 0) = .5$: half of all $t$-values are zero or greater.
3.  $p(t \leq 0) = .5$: half are zero or smaller.

We can get more precise than a 50/50 split. @fig-7percentregions divides the same distribution into ten regions, each containing 10% of the distribution [labels not matching up to regions and x-axis crowded label-wise]. Notice the regions get wider out in the tails: since most $t$-values cluster near 0, it takes a wider interval out there to capture the same 10%.

```{r}
#| label: fig-7percentregions
#| fig-cap: "Splitting the t distribution up into regions each containing 10% of the $t$-values. The width between the bars narrows as they approach the center of the distribution, where there are more $t$-values."
#| out-width: "100%"

range <- seq(-3, 3, .1)
null_distribution <- dt(range, 9, log = FALSE)
plot_df <- data.frame(range, null_distribution)

t_ps <- qt(seq(.1, .9, .1), 9)

ggplot(plot_df, aes(x = range, y = null_distribution)) +
  geom_line() +
  xlab("t-values") +
  ylab("Probability") +
  geom_vline(xintercept = t_ps) +
  theme_classic(base_size = 15) +
  theme(axis.text.y = element_blank(), axis.ticks = element_blank()) +
  scale_x_continuous(breaks = round(t_ps, digits = 1)) +
  geom_label(data = data.frame(x = -2, y = .2, label = "10%"), aes(x = x, y = y, label = label)) +
  geom_label(data = data.frame(x = -1.13, y = .2, label = "10%"), aes(x = x, y = y, label = label)) +
  geom_label(data = data.frame(x = 2, y = .2, label = "10%"), aes(x = x, y = y, label = label)) +
  geom_label(data = data.frame(x = 1.13, y = .2, label = "10%"), aes(x = x, y = y, label = label))


```

For example:

1.  $t \leq -1.38$: $p = 10\%$
2.  $-1.38 \leq t \leq -0.88$: $p = 10\%$
3.  $-0.88 \leq t \leq -0.54$: $p = 10\%$
4.  $t \geq 1.38$: $p = 10\%$

Each range captures the same 10% of the distribution, even though the ranges themselves are different widths.

### Getting the p-values for $t$-values

Where do values like these actually come from? Older textbooks put a table in the back of the book; this one doesn't need one, because software gives you exact $t$ and $p$ values directly. R, Excel, and [online calculators](http://www.socscistatistics.com/pvalues/tdistribution.aspx) all work; you'll get hands-on practice with this in lab, both with software and by simulation.

Software does need a couple of things from you to compute a $p$-value correctly: the degrees of freedom, and whether the test is one-tailed or two-tailed. We haven't covered either yet, so that's next, starting with one-tailed tests, then two-tailed, then degrees of freedom.

### One-tailed vs. two-tailed, applied

Chapter 4 introduced the concept: a two-tailed test asks whether $\mu$ differs from $\mu_0$ in either direction, while a one-tailed test commits to one direction in advance, and is only justified when your theory already rules out the other direction. Here's how that choice plays out mechanically for a $t$-test: it changes the critical value, and therefore the $p$-value, for the exact same data.

For our quiz-score example ($df=9$), a one-tailed test only needs to clear 5% of the distribution on one side. A two-tailed test has to clear 2.5% on each side, which pushes the critical value farther out:

```{r}
#| label: fig-7critT
#| fig-cap: "One-tailed critical value at df=9: t-values of 1.83 or greater occur 5% of the time."

range <- seq(-3, 3, .1)
null_distribution <- dt(range, 9, log = FALSE)
plot_df <- data.frame(range, null_distribution)

ggplot(plot_df, aes(x = range, y = null_distribution)) +
  geom_line() +
  xlab("t-values") +
  ylab("Probability") +
  geom_vline(xintercept = qt(.95, 9, lower.tail = TRUE)) +
  theme_classic(base_size = 12) +
  theme(axis.text.y = element_blank(), axis.ticks = element_blank()) +
  annotate("rect", xmin = qt(.95, 9, lower.tail = TRUE), xmax = 3, ymin = 0, ymax = Inf, alpha = 0.15, fill = "#b22222") +
  geom_label(data = data.frame(x = 2.4, y = .2, label = "5%"), aes(x = x, y = y, label = label)) +
  geom_label(data = data.frame(x = qt(.95, 9, lower.tail = TRUE), y = .3,
                               label = paste0("Critical t: ", round(qt(.95, 9, lower.tail = TRUE), 2))),
             aes(x = x, y = y, label = label))
```

```{r}
#| label: fig-7twotailedt
#| fig-cap: "Two-tailed critical values at df=9: t-values beyond ±2.26 occur 5% of the time total, 2.5% in each tail."

range <- seq(-4, 4, .1)
null_distribution <- dt(range, 9, log = FALSE)
plot_df <- data.frame(range, null_distribution)

ggplot(plot_df, aes(x = range, y = null_distribution)) +
  geom_line() +
  xlab("t-values") +
  ylab("Probability") +
  geom_vline(xintercept = c(qt(.975, 9, lower.tail = TRUE), qt(.025, 9, lower.tail = TRUE))) +
  theme_classic(base_size = 12) +
  theme(axis.text.y = element_blank(), axis.ticks = element_blank()) +
  annotate("rect", xmin = qt(.975, 9, lower.tail = TRUE), xmax = 4, ymin = 0, ymax = Inf, alpha = 0.15, fill = "#b22222") +
  annotate("rect", xmin = -4, xmax = qt(.025, 9, lower.tail = TRUE), ymin = 0, ymax = Inf, alpha = 0.15, fill = "#b22222") +
  geom_label(data = data.frame(x = 3.4, y = .2, label = "2.5%"), aes(x = x, y = y, label = label)) +
  geom_label(data = data.frame(x = -3.4, y = .2, label = "2.5%"), aes(x = x, y = y, label = label)) +
  geom_label(data = data.frame(x = qt(.975, 9, lower.tail = TRUE), y = .3, label = "2.26"), aes(x = x, y = y, label = label)) +
  geom_label(data = data.frame(x = qt(.025, 9, lower.tail = TRUE), y = .3, label = "-2.26"), aes(x = x, y = y, label = label))
```

Applied to our actual result, $t(9) = 1.94$: the one-tailed $p$ is .042, the two-tailed $p$ is .084, exactly double, since a two-tailed test counts the matching extreme region on both sides instead of just one. Here the choice actually changes the decision: one-tailed, this result is significant at $\alpha=.05$; two-tailed, it isn't. That's exactly why a one-tailed test has to be justified by your theory *before* you see the data, not chosen afterward because it gives the answer you wanted.

### Which one should you use? [This is nice and succinct - consider for Chapter 4? But otherwise feels like too much of a reminder, like too much space? Maybe it's fine]

Use a one-tailed test only when your theory already rules out the opposite direction; default to two-tailed otherwise (Chapter 4's rule). There's a second reason two-tailed is the safer default: a one-tailed critical value is always closer to 0 than a two-tailed one, so routinely running one-tailed tests inflates your long-run Type I error rate above your stated $\alpha$ (recall from Chapter 5: a Type I error means rejecting $H_0$ when it's actually true). That's why most researchers default to two-tailed tests even when they do have a directional prediction, and why it's worth being able to justify your choice rather than picking whichever gives the smaller $p$-value.

### Degrees of freedom

You've probably noticed a number in parentheses every time we report a $t$-test, like $t(9) = 1.94$. That's the degrees of freedom, and it's worth understanding what it means, not just where to find it.

For one-sample and paired $t$-tests, $df = n-1$. In our quiz example, $n=10$ students, so $df=9$.

Degrees of freedom is both a concept and a correction [don't think prose follows up explicitly on how this is a correction]. The concept: once you've estimated something from your data, like the mean, that estimate constrains how many of the numbers are still free to vary. Consider three numbers with a mean of 2: pick the first two freely, say 51 and $-3$, and the third number is no longer free. It has to be $-42$ for the mean to come out to 2. Two of the three numbers had freedom; the third didn't. That's $df=n-1=2$ for three numbers.

The same logic applies to the sample standard deviation. Computing it requires the sample mean first, which uses up one degree of freedom, exactly why the sample SD divides by $n-1$ rather than $n$ (Chapter 1).

**Simulating how degrees of freedom affects the** $t$ distribution.

We can see the effect of degrees of freedom directly by simulating the $t$ distribution at two different sample sizes. @fig-7dft compares $df=9$ (our quiz-score example) to $df=100$ (a much larger sample).

```{r}
#| label: fig-7dft
#| fig-cap: "The width of the t distribution shrinks as sample size and degrees of freedom (from 9 to 100) increases."
#| fig-asp: 0.5
#| out-width: "100%"

ts <- c(rt(10000, 9), rt(10000, 100))
dfs <- as.factor(rep(c(9, 100), each = 10000))

t_df <- data.frame(dfs, ts)
t_df <- t_df[abs(t_df$ts) < 5, ]

ggplot(t_df, aes(x = ts)) +
  geom_histogram(fill = "#63b8ff", color = "white") +
  theme_classic(base_size = 15) +
  facet_wrap( ~ dfs) +
  xlab("t value") +
  ylab("Count")


```

As degrees of freedom increase, the $t$ distribution gets taller in the middle and narrower in the tails, converging toward the standard normal. That's the same pattern from "Comparing $z$ and $t$" earlier: with more data, the sample SD becomes a more reliable estimate of the population SD, so $t$ behaves more like $z$.

Why $n-1$ and not $n$? Because $t$ uses the sample SD to estimate the standard error, and that SD calculation already spent one degree of freedom estimating the mean, so the $t$-distribution is built assuming $n-1$ from the start.

## Paired t: before/after or matched

When the **same units** are measured twice (before/after, matched pairs), the question is whether they **changed**. The natural data unit is the **within-unit difference**.

**Design in this worked example.** Ecologists measured soil moisture at 20 field plots before wetland restoration and again one year later, to see whether restoration **increased** moisture enough to support plant reestablishment and carbon sequestration. @fig-7paireddesign shows the design.

```{r}
#| label: fig-7paireddesign
#| fig-cap: "A paired design. The dashed boxes and the note underneath mark the feature that defines it: there is one set of 20 plots, measured on two occasions, not two separate samples."
#| fig-alt: "Study design schematic. IV: restoration, before versus after. Two dashed boxes labelled Before and After, each n = 20 plots, with a note that these are the same 20 plots measured on both occasions. DV: soil moisture, percent."
#| fig-asp: 0.34
#| out-width: "100%"

design_groups("Restoration (before vs. after)", c("Before", "After"), 20, "plots",
              "Soil moisture (%)", "Did moisture change within each plot?",
              kind = "Paired", dashed = TRUE,
              note = "the same 20 plots, measured on both occasions")
```

The dashed boxes are doing real work there. In every design so far the boxes held *different* units, and the test compared one set against another. Here both boxes hold the same 20 plots, which is what makes the design paired and what lets us subtract.

**Define differences and their meaning.** Choose an order and stick to it:

- $d_i = \text{After}_i - \text{Before}_i$
- $d_i>0$: plot had **more** soil moisture after restoration (increase)
- $d_i<0$: plot had **less** soil moisture after restoration (decrease)

This subtraction **cancels between-plot variability** (some plots are naturally wetter than others) and focuses on **within-plot change**.

**Test on differences (paired** $t$). Compute one difference per plot, then run a **one-sample** $t$ against 0:

$$\bar d=\frac{1}{n}\sum_{i=1}^n d_i, \quad
s_d=\text{SD}(d_1,\ldots,d_n), \quad
t = \frac{\bar d - 0}{s_d/\sqrt{n}}, \quad df=n-1.$$

The **sign of** $t$ matches $\bar d$ and your difference direction.

**Mini walk-through (first 5 plots).** Using the first five plots, $\bar d = 2.8$ and $s_d = 3.42$. Then
$$t=\frac{2.8}{3.42/\sqrt{5}} \approx 1.83,\qquad df=4,$$
which (two-tailed) is **not** significant. That's expected with $n=5$. [This seems.a little far away from the "calculate t" table at this point? Or repetitive with it? Why are we saying it twice?]

**Full sample effect.** With all 20 plots, $\bar d$ is similar but $s_d/\sqrt{n}$ gets smaller, so $t$ increases and the result becomes significant. This is the payoff of pairing: more power from the same number of plots.

**What to report.** $\bar d$ (the mean change), a 95% CI, $t(df)$, and $p$. Optionally add a standardized effect (Cohen's $d=\bar d/s_d$).

**R pointers (you'll do this in lab).**

- Compute differences: `d <- after - before`
- Paired $t$: `t.test(d, mu = 0)` or `t.test(after, before, paired = TRUE)`
- CI for $\bar d$: from the `t.test` output

### Calculate t

How do we calculate $t$ for a paired-samples $t$-test? We use the exact same one-sample $t$-test formula from the previous section, just applied to the difference scores instead of raw scores. The paired-samples $t$-test is really just a variation on the one-sample $t$-test: there's still only one group of numbers involved, the differences.

Let's find $t$ for the mean difference scores:

```{r, echo=FALSE}
# Soil moisture (%) at 5 paired restoration plots: before and after restoration
before <- c(18, 22, 16, 25, 21)
after  <- c(21, 26, 22, 29, 18)

paired_sample_df <- data.frame(
  plot   = 1:5,
  Before = before,
  After  = after
)

differences     <- paired_sample_df$After - paired_sample_df$Before
diff_from_mean  <- differences - mean(differences)
Squared_differences <- diff_from_mean^2

paired_sample_df <- cbind(paired_sample_df,
                          differences,
                          diff_from_mean,
                          Squared_differences)

paired_sample_df <- paired_sample_df %>%
  rbind(c("Sums",  colSums(paired_sample_df[1:5, 2:6]))) %>%
  rbind(c("Means", colMeans(paired_sample_df[1:5, 2:6]))) %>%
  rbind(c(" ", " ", " ", " ", "sd",  round(sd(paired_sample_df[1:5, 4]), digits=3))) %>%
  rbind(c(" ", " ", " ", " ", "SEM", round(sd(paired_sample_df[1:5, 4])/sqrt(5), digits=3))) %>%
  rbind(c(" ", " ", " ", " ", "t",   mean(differences)/round(sd(paired_sample_df[1:5, 4])/sqrt(5), digits=3)))
paired_sample_df[6, 5] <- 0
paired_sample_df[7, 5] <- 0
knitr::kable(paired_sample_df)
```

If we did this test using R, we would obtain almost the same numbers (there is a little bit of rounding in the table).

```{r}
paired5_out <- t.test(differences, mu = 0)
paired5_tidy <- broom::tidy(paired5_out) |>
  mutate(statistic = round(statistic, 2), p.value = fmt_p(p.value))

knitr::kable(
  paired5_tidy[, c("estimate", "statistic", "parameter", "p.value")],
  col.names = c("Mean difference", "t", "df", "p-value"),
  caption = "Paired t-test on 5 restoration plots"
)
```

Our t-test results are: t(4) = 1.83, p = .141.

The sign of $t$ matches the sign of the mean difference (ours was $+2.8$ percentage points), so soil moisture increased on average, exactly what $\bar d$ already told us directly. Getting a $p$-value from this $t$, and choosing one-tailed vs. two-tailed, works exactly like it did for the one-sample case (see the One-sample $t$-test section above). Two-tailed, $p=.141$: not significant, this five-plot sample is too small to rule out chance.

### Scaling to the full sample

The five plots above are a subset of the full restoration study, which measured all **20** paired plots. With only $n=5$, $t(4)=1.83$ wasn't enough to distinguish the observed increase from chance ($p=.141$). Here's what changes with the full sample.

```{r, echo=FALSE}
before_full <- c(18, 22, 16, 25, 21, 27, 26, 23, 28, 23, 19, 19, 20, 25, 15, 24, 23, 16, 18, 21)
after_full  <- c(21, 26, 22, 29, 18, 31, 31, 29, 29, 28, 16, 19, 20, 19, 18, 28, 25, 21, 18, 19)

full20_out <- t.test(after_full, before_full, paired = TRUE)
full20_tidy <- broom::tidy(full20_out) |>
  mutate(statistic = round(statistic, 2), p.value = fmt_p(p.value))

knitr::kable(
  full20_tidy[, c("estimate", "statistic", "parameter", "p.value")],
  col.names = c("Mean difference", "t", "df", "p-value"),
  caption = "Paired t-test on all 20 restoration plots"
)
```

At $n=20$, $t(19) = `r round(full20_out$statistic, 2)`$, $p =`r fmt_p(full20_out$p.value)`$: significant. The mean difference (`r round(full20_out$estimate, 1)` percentage points) is in the same range as what the 5-plot subset already suggested; what changed is the standard error, which shrank enough with the larger sample to rule out chance. This is the payoff of pairing: same design, more plots, more power to detect a real effect.

## Two independent groups

The independent-samples $t$-test is the third and last design in this chapter. The logic is the same as the one-sample and paired cases: a mean difference divided by an estimate of its variability. What changes is the research design.

Use an independent-samples $t$-test for a **between-subjects** design: different people or units in each condition, each measured once. With two groups of 10, that's 20 people total, none of them paired. Without repeated measures on the same unit, there's no meaningful way to subtract scores between conditions the way the paired $t$-test does.

The numerator works the same way it always has: the mean difference between groups. The denominator is where independent groups differ from the paired case, since now we need a single estimate of variability built from two separate samples instead of one.

A natural first guess is to average the two groups' standard errors: [Why are we saying "SEM" notation in instead of "SE" like in the rest of the book?]

$$t = \frac{\bar{A}-\bar{B}}{\left(\dfrac{SEM_A+SEM_B}{2}\right)}$$

That average, however, is a biased estimator of the variability under the null. Instead, the independent-samples $t$-test **pools** the two groups' variances, weighted by their degrees of freedom, into a single estimate for the denominator:

$$t = \frac{\bar{X}_A-\bar{X}_B}{s_p\sqrt{\dfrac{1}{n_A}+\dfrac{1}{n_B}}}$$

where the pooled standard deviation $s_p$ is (note $s_A^2$ and $s_B^2$ here are variances, not standard deviations):

$$s_p = \sqrt{\frac{(n_A-1)s_A^2 + (n_B-1)s_B^2}{n_A+n_B-2}}$$

The example below computes $t$ for two small groups by following that formula step by step in R, then checks it against `t.test()`.

Everywhere else this book formats test results into a clean table. [I don't need this meta-narration for the studnets] Here we deliberately print R's raw console output instead, because this is the block you will actually be looking at in lab, and it is worth seeing that the number you built by hand is the same number sitting in it.

```{r, echo=TRUE }

## By "hand" using R
a <- c(1,2,3,4,5)
b <- c(3,5,4,7,9)

mean_difference <- mean(a)-mean(b) # compute mean difference

variance_a <- var(a) # compute variance for A
variance_b <- var(b) # compute variance for B

# Compute top part and bottom part of sp formula

sp_numerator <- (4*variance_a + 4* variance_b) 
sp_denominator <- 5+5-2
sp <- sqrt(sp_numerator/sp_denominator) # compute sp


# compute t following the formula

t <- mean_difference / ( sp * sqrt( (1/5) +(1/5) ) )

t # our hand-computed t


# the same test via R's built-in function: compare the "t = " value
t.test(a,b, paired=FALSE, var.equal = TRUE)


```

::: {.callout-important title="Results reporting: t-tests"}
All three forms report the same four things: the direction in plain language, the effect size, the test statistic with its df, and $p$. What changes is what counts as the effect.

One-sample, where the effect is the sample mean against the reference value:

> Mean quiz score was 61%, which did not differ significantly from the 50% expected by chance, *t*(9) = 1.94, *p* = .084, 95% CI \[48.18, 73.82\].

Paired, where the effect is the mean *difference* score:

> Soil moisture increased by 1.90 percentage points after restoration, *t*(19) = 2.50, *p* = .022, 95% CI \[0.31, 3.49\].

Independent-samples, where the effect is the difference between two group means:

> Group A scored lower than Group B, a mean difference of -2.6, *t*(8) = -2.02, *p* = .078.

Note the paired write-up never mentions the before and after means separately. The difference score *is* the unit of analysis, so that single number plus its CI is the whole result.
:::

## Simulating data for t-tests [How much of htis is helpful and/or necessary given Chatper 4's simulation points? Or at least do we need all of them?]

We can also use R to simulate data for $t$-tests, which is worth knowing how to do.

Recall that $t$ is a property of a sample: we calculate it from the sample we collected. If we took thousands of samples and calculated $t$ for each one, the distribution of those values would be the sampling distribution of $t$.

Simulating data under conditions you choose builds intuition for how sampling error influences your results, given the specific parameters of your design: sample size, the size of the mean difference you expect, and the amount of variation. The same methods can be used formally to conduct power analyses, or informally for data sense.

### Simulating a one-sample t-test

Here are the steps you might follow to simulate data for a one sample $t$-test.

1.  Make some assumptions about what your sample (that you might be planning to collect) might look like. For example, you might be planning to collect 30 subjects worth of data. The scores of those data points might come from a normal distribution (mean = 50, sd = 10).

2.  sample simulated numbers from the distribution, then conduct a $t$-test on the simulated numbers. Save the statistics you want (such as $t$s and $p$s), and then see how things behave.

Let's do this a couple different times. First, let's simulate samples with N = 30, taken from a normal (mean= 50, sd = 25). We'll do a simulation with 1000 simulations. For each simulation, we will compare the sample mean with a population mean of 50. There should be no difference on average here. @fig-7nullt is the null distribution that we are simulating.

```{r}
# steps to create fake data from a distribution
# and conduct t-tests on the simulated data
save_ps <- length(1000)
save_ts <- length(1000)
for (i in 1:1000) {
  my_sample <- rnorm(n = 30, mean = 50, sd = 25)
  t_test <- t.test (my_sample, mu = 50)
  save_ps[i] <- t_test$p.value
  save_ts[i] <- t_test$statistic
}
```

```{r}
#| label: fig-7nullt
#| fig-cap: "The distribution of $t$-values under the null. These are the $t$ values that are produced by chance alone."

hist_t(save_ts)
```

```{r}
#| label: fig-7flatp
#| fig-cap: "The distribution of $p$-values that are observed is flat under the null."

hist_p(save_ps)
```

The simulated $t$ distribution looks like a $t$ distribution should, and it shows us how often we get $t$ values of particular sizes. The $p$-distribution is flat under the null, which we are simulating here. This means you have the same chance of getting a $t$ with a p-value between 0 and 0.05 as you would of getting one between .90 and .95. Those ranges are both ranges of 5%, so there are an equal amount of $t$ values in them by definition.

You can write the same simulation more compactly with `replicate()` instead of a `for` loop; it produces the same two distributions above:

```{r, echo=TRUE, eval=FALSE}
simulated_ts <- replicate(1000, t.test(rnorm(30, 50, 25), mu = 50)$statistic)
simulated_ps <- replicate(1000, t.test(rnorm(30, 50, 25), mu = 50)$p.value)
```

### Simulating a paired samples t-test

The code below is set up to sample 10 scores for condition A and B from the same normal distribution. The simulation is conducted 1000 times, and the $t$s and $p$s are saved and plotted for each.

```{r, echo=TRUE}

save_ps <- length(1000)
save_ts <- length(1000)
for ( i in 1:1000 ){
  condition_A <- rnorm(10,10,5)
  condition_B <- rnorm(10,10,5)
  differences <- condition_A - condition_B
  t_test <- t.test(differences, mu=0)
  save_ps[i] <- t_test$p.value
  save_ts[i] <- t_test$statistic
}

```

```{r}
#| label: fig-7simts1000
#| fig-cap: "1000 simulated ts from the null distribution"
hist_t(save_ts)

```

```{r}
#| label: fig-7simps1000
#| fig-cap: "1000 simulated ps from the null distribution"
hist_p(save_ps)

```

According to the simulation, when there are no differences between the conditions, and the samples are being pulled from the very same distribution, you get these two distributions for $t$ and $p$. These again show how the null distribution of no differences behaves.

For any of these simulations, if you rejected the null-hypothesis (that your difference was only due to chance), you would be making a type I error. If you set your alpha criteria to $\alpha = .05$, we can ask how many type I errors were made in these 1000 simulations. The answer is:

```{r, echo=TRUE}
length(save_ps[save_ps<.05])
length(save_ps[save_ps<.05])/1000
```

We happened to make `r length(save_ps[save_ps<.05])` type I errors. The expectation over the long run is a 5% type I error rate (if your alpha is .05).

What happens if there actually is a difference in the simulated data, let's set one condition to have a larger mean than the other:

```{r, echo=TRUE}

save_ps <- length(1000)
save_ts <- length(1000)
for ( i in 1:1000 ){
  condition_A <- rnorm(10,10,5)
  condition_B <- rnorm(10,13,5)
  differences <- condition_A - condition_B
  t_test <- t.test(differences, mu=0)
  save_ps[i] <- t_test$p.value
  save_ts[i] <- t_test$statistic
}

```

```{r}
#| label: fig-7simtrueps
#| fig-cap: "1000 ps when there is a true difference"
hist_p(save_ps)
```

Now you can see that the $p$-value distribution is skewed to the left. This is because when there is a true effect, you will get p-values that are less than .05 more often. Or, rather, you get larger $t$ values than you normally would if there were no differences.

In this case, we wouldn't be making a type I error if we rejected the null when p was smaller than .05. How many times would we do that out of our 1000 experiments?

```{r, echo=TRUE}
length(save_ps[save_ps<.05])
length(save_ps[save_ps<.05])/1000
```

We got `r length(save_ps[save_ps<.05])` simulations where p was less than .05, or `r 100*length(save_ps[save_ps<.05])/1000`% of experiments. A design that detects a real effect that rarely is not one you want to run, which is the argument for designing a better experiment.

The most direct lever is $n$, the number of subjects. Let's increase $n$ from 10 to 100 and see what happens to the number of "significant" simulated experiments.

```{r, echo=TRUE}

save_ps <- length(1000)
save_ts <- length(1000)
for ( i in 1:1000 ){
  condition_A <- rnorm(100,10,5)
  condition_B <- rnorm(100,13,5)
  differences <- condition_A - condition_B
  t_test <- t.test(differences, mu=0)
  save_ps[i] <- t_test$p.value
  save_ts[i] <- t_test$statistic
}

```

```{r}
#| label: fig-7simpsB
#| fig-cap: "1000 ps for n =100, when there is a true effect"
hist_p(save_ps)


length(save_ps[save_ps<.05])
length(save_ps[save_ps<.05])/1000
```

Now almost all of the experiments show a $p$-value of less than .05 (using a two-tailed test, the default in R). You can use this simulation process to determine how many subjects you need to reliably detect your effect.

### Simulating an independent samples t-test

The setup is the same, you just change what goes into `t.test()`. Here's the null case, assuming no difference between two independent groups of $n=10$:

```{r, echo=TRUE}

save_ps <- length(1000)
save_ts <- length(1000)
for ( i in 1:1000 ){
  group_A <- rnorm(10,10,5)
  group_B <- rnorm(10,10,5)
  t_test <- t.test(group_A, group_B, paired=FALSE, var.equal=TRUE)
  save_ps[i] <- t_test$p.value
  save_ts[i] <- t_test$statistic
}

```

```{r}
#| label: fig-7simpsC
#| fig-cap: "1000 ps for the independent-samples null, n=10 per group"
hist_p(save_ps)


length(save_ps[save_ps<.05])
length(save_ps[save_ps<.05])/1000
```

Same story as before: about 5% of these simulated experiments come out "significant" by chance alone. And the same two levers we saw for the paired test still apply here, giving `group_A` and `group_B` different means will pull power up, and so will increasing $n$. Try editing the code above to see it for yourself.

## Chapter Summary

**Why it matters.** The $t$-test is the first real inferential test in this course, and the logic it runs on, a signal measured against the noise around it, is the same logic behind ANOVA, regression, and everything else that follows. Getting comfortable here pays off for the rest of the semester.

**Core ideas**

- **Signal over noise.** A difference is only impressive relative to the variability around it. Every $t$ statistic is a ratio: the effect you observed on top, an estimate of how much that effect could bounce around by chance on the bottom.
- **Three forms, one idea.** The one-sample, paired, and independent-samples $t$-tests differ in what goes into the numerator, but all three ask whether an observed difference is large compared to its own standard error. The paired test is just the one-sample test applied to difference scores.
- **Why** $t$ and not $z$. When the population SD $\sigma$ is unknown, we estimate it with $s$, which adds uncertainty. The $t$ distribution accounts for that by being fatter in the tails than the normal, and it tightens toward the normal as degrees of freedom grow.
- **Assumptions come in two kinds.** *Independence and a sound study design* apply to every test in this book and are assumed from here on. The *test-specific* assumption for these tests is approximate normality, of the outcome for a one-sample test or of the difference scores for a paired test.
- **How worried to be depends on** $n$. Thanks to the Central Limit Theorem, $t$-tests are fairly robust to non-normality once samples are reasonably large. The danger zone is small *and* skewed, which describes a lot of environmental field data. Plot your data first; treat a Shapiro-Wilk result as evidence rather than a gate, since it has least power exactly when you need it most.
- **Choose your tails before you look.** Default to two-tailed. Reserve one-tailed tests for cases where theory rules out the opposite direction, and decide before seeing the result.
