---
author:
  - Mallory L. Barnes
aliases: [redirects/simple-linear-regression.html]
---

```{r, include = FALSE}
source("global_stuff.R")
fmt_p <- function(p) ifelse(p < .001, "< 0.001", formatC(p, digits = 3, format = "f"))

# ANOVA table in the book's style (as in Chapters 8 and 10)
anova_kable <- function(fit, caption = NULL) {
  broom::tidy(anova(fit)) |>
    dplyr::mutate(
      sumsq     = round(sumsq, 2),
      meansq    = round(meansq, 2),
      statistic = ifelse(is.na(statistic), "", formatC(statistic, digits = 2, format = "f")),
      p.value   = ifelse(is.na(p.value), "", fmt_p(p.value))
    ) |>
    knitr::kable(col.names = c("Source", "df", "SS", "MS", "F", "p"),
                 align = "lrrrrr", caption = caption)
}
```

# Simple Linear Regression

> *Essentially, all models are wrong, but some are useful.*
>
> George E. P. Box

```{r}
library(tidyverse)
library(broom)
library(knitr)
```

Back in Chapter 2, we used the covariance between two variables to quantify the strength and direction of a relationship (Pearson's $r$), and fit a best-fit line through a scatterplot. We could not yet ask the inference question every other test in this book has asked:

*Could chance alone plausibly have produced our result?*

Now that we've learned hypothesis testing, sampling distributions, and the logic of comparing an observed statistic to a null distribution, we can apply that same logic to the regression line and test whether a given slope is evidence of a real relationship.

## Regression basics

The equation for a linear regression will be familiar, and it's one of the most heavily used equations in applied science:

$$y = \beta_0 + \beta_1 x + \varepsilon$$

Here $\beta_1$ is the slope and $\beta_0$ is the intercept.

A linear equation, fit to real data, is a statistical **model**. We never observe $\beta_0$ or $\beta_1$ directly. Real data also do not fall exactly on a line, so the model needs one more term, the **error** $\varepsilon$: the variability in the DV that the population line doesn't explain.

### Computing the best-fit line by hand

In practice, we let software identify the line with the smallest total squared error, but working through one example by hand shows where that line actually comes from. Here are the formulas for the slope and intercept, calculated straight from the data:

$$\text{intercept} = b_0 = \frac{\sum{y}\sum{x^2}-\sum{x}\sum{xy}}{n\sum{x^2}-(\sum{x})^2}$$

$$\text{slope} = b_1 = \frac{n\sum{xy}-\sum{x}\sum{y}}{n\sum{x^2}-(\sum{x})^2}$$

Here $x$ and $y$ are the individual scores. This is the same line from @fig-2corwithLine in Chapter 2. To find its slope and intercept, we start with this table:

```{r}
x  <- c(1, 4, 3, 6, 5, 7, 8)
y  <- c(2, 5, 1, 8, 6, 8, 9)
tab <- data.frame(Pair = as.character(1:7), x = x, y = y, x2 = x^2, xy = x * y)
tab <- rbind(tab, data.frame(Pair = "Sums", x = sum(x), y = sum(y), x2 = sum(x^2), xy = sum(x * y)))

knitr::kable(tab, col.names = c("Pair", "x", "y", "x²", "xy"),
             caption = "Sums for the slope and intercept formulas")

```

The table has 7 pairs of $x$ and $y$ scores, plus columns for $x^2$ and $xy$ (each $x$ score times its matching $y$ score), with column sums at the bottom. These sums are everything the formulas need:

$$\text{intercept} = b_0 = \frac{\sum{y}\sum{x^2}-\sum{x}\sum{xy}}{n\sum{x^2}-(\sum{x})^2} = \frac{39 \times 200 - 34 \times 231}{7 \times 200-34^2} = -0.221$$

$$\text{slope} = b_1 = \frac{n\sum{xy}-\sum{x}\sum{y}}{n\sum{x^2}-(\sum{x})^2} = \frac{7 \times 231-34 \times 39}{7 \times 200-34^2} = 1.19$$

With $b_0$ and $b_1$ in hand, we can compute a residual for any observed point. A residual is the difference between what we observed and what the line predicts, $e = y - \hat y$, where $\hat y = b_0 + b_1x$. For the first row above ($x=1, y=2$): $\hat y = -0.221 + 1.19(1) = 0.969$, so $e = 2 - 0.969 = 1.031$. That point is above the line. Repeating this for every point, then squaring and summing the residuals, gives the quantity that $b_0$ and $b_1$ were chosen to minimize.

The residual $e$ is the sample's version of the error $\varepsilon$. The error $\varepsilon$ belongs to the population line, which we never see, and the residual $e$ belongs to the line we fit.

### Parameters and sample statistics

The beta coefficients in the model are population parameters. They describe the population line, which we infer from a sample but never know for certain.

| Symbol             | What is it?                         | Do we know it?                                  |
|--------------------|-------------------------------------|-------------------------------------------------|
| $\beta_0, \beta_1$ | True population intercept and slope | Almost never known for sure                     |
| $b_0, b_1$         | Sample intercept and slope          | Yes, calculated by fitting the line to our data |

## Worked regression example: stream temperature and dissolved oxygen

Warmer water holds less dissolved gas, so ecologists expect dissolved oxygen (DO) to decline as stream temperature rises. This matters for stream health, because low DO can stress or kill fish and invertebrates. @fig-9streamdesign lays out the design.

```{r}
#| label: fig-9streamdesign
#| fig-cap: "A regression design. Instead of a few groups to sort sites into, every site sits somewhere along a continuous range, and every temperature value is treated as its own group. A slope describes how the mean changes across that range."
#| fig-alt: "Study design schematic. IV: stream temperature, continuous, an arrow running from 10 to 24 degrees Celsius across 30 sites. DV: dissolved oxygen in mg per liter."
#| fig-asp: 0.34
#| out-width: "100%"

design_continuous("Stream temperature (continuous)", "10 \u00b0C", "24 \u00b0C",
                  "30 sites", "Dissolved oxygen (mg/L)",
                  "Does DO change as temperature rises?")
```

Up to now every design put units into groups and asked whether the groups differed. Now, with a continuous predictor, every temperature value is treated as its own group. Each of the 30 sites falls somewhere along the range, and the line describes how mean DO changes from one temperature to the next. The question becomes how much DO changes per degree rather than whether two groups differ.

The 30 sites in this example are invented. We built them from a population line with $\beta_0 = 13.5$ mg/L and $\beta_1 = -0.28$ mg/L per °C, plus random scatter, so here, unlike in a real study, we know the true slope that $b_1$ estimates.

```{r}
#| label: fig-9regscatter
#| fig-cap: "Stream temperature and dissolved oxygen at 30 sample sites, with the fitted regression line."
#| fig-alt: "Scatter plot of dissolved oxygen against stream temperature for 30 sites, with a fitted line sloping down from about 11.7 mg/L at 10 degrees C to about 7 mg/L at 23 degrees C. Points scatter around the line by up to about 4 mg/L."

set.seed(538)
n <- 30
stream_temp <- round(runif(n, 10, 24), 1)
DO <- round(13.5 - 0.28 * stream_temp + rnorm(n, 0, 1.8), 2)
stream_df <- data.frame(stream_temp, DO)

ggplot(stream_df, aes(x = stream_temp, y = DO)) +
  geom_point(size=1.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#008cff") +
  xlab(expression("Stream temperature ("*degree*"C)")) +
  ylab("Dissolved oxygen (mg/L)")

```

The scatterplot in @fig-9regscatter shows a clear downward trend. Just from the graph, we can guess a couple of things:

-   Pearson's $r$ should be negative, since DO falls as temperature rises.

-   It should be well short of $-1$, since the points scatter noticeably around the line.

Regression tests whether a relationship like this one is more than chance by writing it as a model.

### The regression model

We model dissolved oxygen as a linear function of stream temperature, plus error:

$$DO = \beta_0 + \beta_1(\text{Temperature}) + \varepsilon$$

$\beta_0$ is the population intercept, $\beta_1$ is the population slope, and $\varepsilon$ is the error, the part of DO that temperature alone doesn't explain.

We don't know $\beta_0$ and $\beta_1$. We estimate them from the sample as $b_0$ and $b_1$, using the same least-squares logic from Chapter 2 (the line that minimizes the sum of squared residuals).

In R, this is a single line of code:

```{r 10fitshow, echo = TRUE, eval = FALSE}
fit <- lm(DO ~ stream_temp, data = stream_df)
```

This fitted line (@fig-9regresidviz) is the best-fit line that minimizes the total squared distance between the data and the line.

```{r 10fit}
fit <- lm(DO ~ stream_temp, data = stream_df)
fit_tidy <- tidy(fit, conf.int = TRUE)
fit_glance <- glance(fit)
```

```{r}
#| label: fig-9regresidviz
#| fig-cap: "Stream temperature and dissolved oxygen with the fitted line. Vertical red segments are residuals."
#| fig-alt: "Scatter plot of dissolved oxygen against stream temperature with a fitted downward-sloping line. Open circles mark each point's predicted value on the line, and a red vertical segment connects each observed point to its prediction."

stream_df$fitted_line <- fitted(fit)

ggplot(stream_df, aes(x = stream_temp, y = DO)) +
  geom_smooth(method = "lm", se = FALSE, color = "#008cff") +
  geom_segment(aes(xend = stream_temp, yend = fitted_line), color = "#b22222", alpha = 0.5) +
  geom_point(size=1.3) +
  geom_point(aes(y = fitted_line), shape = 1) +
  xlab(expression("Stream temperature ("*degree*"C)")) +
  ylab("Dissolved oxygen (mg/L)")
```

Here's the table of model coefficients R gives us.

```{r}
#| echo: true
#| eval: false
summary(fit)
```

```{r fitKable}
kable(
  fit_tidy %>%
    mutate(across(c(estimate, std.error, statistic, conf.low, conf.high), ~round(., 2)),
           p.value = fmt_p(p.value)) %>%
    rename(Term = term, Estimate = estimate, `Std. Error` = std.error,
           `t` = statistic, `p` = p.value, `CI low` = conf.low, `CI high` = conf.high),
  caption = "Coefficient table for DO ~ stream temperature"
)
```

With the results of the regression, we can get an equation for the fitted line:

$\widehat{DO} = `r sprintf("%.2f", coef(fit)[1])` `r ifelse(coef(fit)[2] >= 0, "+", "-")` `r sprintf("%.2f", abs(coef(fit)[2]))` \times \text{Temperature}$.

The hat over $DO$ marks it as a *predicted* value, what the model outputs for a given temperature, as distinct from an actually observed $DO$ measurement.

In the coefficient table, the `Estimate` values for the intercept and slope are $b_0$ and $b_1$, not $\beta_0$ and $\beta_1$.

The intercept ($b_0 = `r sprintf("%.2f", coef(fit)[1])`$) is the model's predicted DO at 0°C. That temperature is outside the range we sampled (`r sprintf("%.1f", min(stream_df$stream_temp))` to `r sprintf("%.1f", max(stream_df$stream_temp))`°C), so we won't interpret the intercept.

The slope ($b_1 = `r sprintf("%.2f", coef(fit)[2])`$) is the part we care about most. For every 1°C increase in stream temperature, predicted dissolved oxygen drops by about `r sprintf("%.2f", abs(coef(fit)[2]))` mg/L.

## Testing the slope

The slope $b_1$ is a sample estimate. Recall the hypothesis test asks whether the population slope $\beta_1$ could plausibly be zero:

-   $H_0: \beta_1 = 0$ (no linear relationship between stream temperature and dissolved oxygen)

-   $H_a: \beta_1 \neq 0$ (there is a linear relationship)

Like any other statistic, $b_1$ varies from sample to sample. If we drew a new sample of streams and fit a new line, we'd get a slightly different $b_1$, just by chance. The test asks whether our $b_1$ is large relative to that sampling variability, the same signal-over-noise pattern as every test so far:

$$\text{statistic}=\frac{\text{effect}}{\text{error}}.$$

R reports the $t$-test for the slope in the coefficient table from `summary(fit)`, so there is no separate test to run. You get the `Std. Error`, $t$, and $p$ for each coefficient at once.

The test statistic divides the estimated slope by its standard error, the `Std. Error` column in the coefficient table above. The standard error measures how much $b_1$ would vary from sample to sample if we repeated this study many times:

$$t = \frac{b_1}{SE(b_1)} = \frac{`r sprintf("%.3f", coef(fit)[2])`}{`r sprintf("%.3f", fit_tidy$std.error[2])`} = `r sprintf("%.2f", fit_tidy$statistic[2])`$$

with $df = n - 2 = `r nrow(stream_df) - 2`$ (one degree of freedom spent estimating the intercept, one for the slope).

**The $p$-value comes from a $t$-distribution.** If temperature and DO were truly unrelated in the population, repeating this study many times would still produce a nonzero $b_1$ just by chance, and a $t$-statistic that follows a $t$-distribution with $df = `r nrow(stream_df) - 2`$. @fig-9regslopenull shows that distribution, with the critical values marking the boundary of the most extreme 5% under $H_0$, and our observed $t$ marked against it.

```{r}
#| label: fig-9regslopenull
#| fig-cap: "The null distribution of t for the slope test at df = 28 (blue). The shaded tails beyond the red critical values are the rejection region at alpha = 0.05; the green dotted line marks our observed t, far into the left tail."
#| fig-alt: "A blue t-distribution curve centered at zero with 28 degrees of freedom. Red vertical lines labeled minus crit and plus crit mark the critical values near plus and minus 2.05, and the tails beyond them are shaded red. A green dotted vertical line marks the observed t of about negative 5.3, well to the left of the lower critical value."

df_slope <- nrow(stream_df) - 2
obs_t <- fit_tidy$statistic[2]
xs <- seq(-8, 8, by = 0.01)
t_null <- data.frame(x = xs, y = dt(xs, df_slope))
crit <- qt(.975, df_slope)

ggplot(t_null, aes(x, y)) +
  geom_area(data = subset(t_null, x <= -crit), fill = "red", alpha = 0.18) +
  geom_area(data = subset(t_null, x >=  crit), fill = "red", alpha = 0.18) +
  geom_line(colour = "blue", linetype = "dotdash", linewidth = 1.1) +
  annotate("segment", x = c(-crit, crit), xend = c(-crit, crit), y = 0,
           yend = 0.41, colour = "red", linewidth = 0.8) +
  annotate("text", x = c(-crit, crit), y = 0.425, label = c("-crit", "+crit"),
           colour = "red", size = 4.2) +
  annotate("segment", x = obs_t, xend = obs_t, y = 0, yend = 0.36,
           colour = "#1E9E1E", linetype = "dotted", linewidth = 1.4) +
  annotate("text", x = obs_t - 0.2, y = 0.37, label = paste0("observed\nt = ", sprintf("%.2f", obs_t)),
           colour = "#1E9E1E", hjust = 1, size = 4, lineheight = 0.9) +
  annotate("text", x = 4.5, y = 0.3, label = "'Under '*H[0]", parse = TRUE,
           colour = "blue", hjust = 0, size = 4.5) +
  labs(x = "t", y = "Density")
```

Our observed $t = `r sprintf("%.2f", fit_tidy$statistic[2])`$ is far past the critical values of about $\pm`r sprintf("%.2f", qt(.975, nrow(stream_df)-2))`$, well outside where 95% of $t$-values would fall if temperature and DO were unrelated. The $p$-value is the probability of a $t$ that far out (or farther) in either tail under $H_0$. Here $p`r fmt_p(fit_tidy$p.value[2])`$, which crosses the conventional $\alpha = 0.05$ threshold.

We reject $H_0$ and conclude that stream temperature is linearly related to dissolved oxygen.

The 95% confidence interval for the slope, also in the coefficient table above, is 95% CI [`r sprintf("%.3f", fit_tidy$conf.low[2])`, `r sprintf("%.3f", fit_tidy$conf.high[2])`] mg/L per °C. Because this interval does not contain 0, it agrees with the hypothesis test, as it always will. The interval also contains $\beta_1 = -0.28$, the true slope we built into the data. A 95% interval catches the true value in about 95% of samples, and this sample is one of them.

## Partitioning variation: regression as ANOVA

Besides testing the slope, we want to know how much of the variation in DO the line explains. We answer that the same way as one-way ANOVA in Chapter 8, by splitting the total variation into a piece the model explains and a piece it leaves over.

@fig-9regpartition shows the three pieces for the stream data. Every site's distance from mean DO (left) splits into the distance from the mean to the line (middle), the part explained by the line, and the distance from the line to the point (right), the residual.

```{r}
#| label: fig-9regpartition
#| fig-cap: "Partitioning variation in dissolved oxygen. Left: each site's distance from mean DO (dashed line). Middle: the part of that distance explained by the line. Right: what is left over, the residuals. Squaring and summing each set of segments gives the three sums of squares."
#| fig-alt: "Three panels of the same scatter plot of dissolved oxygen against stream temperature, each with the fitted line and a dashed horizontal line at mean DO. The left panel draws grey segments from every point to the mean line. The middle panel draws blue segments from the fitted line to the mean line, the part explained by the line. The right panel draws red segments from every point to the fitted line, the residuals. Each panel is labeled with its sum of squares, SS Total, SS Regression and SS Residual, and the middle and right values add up to the left."
#| fig-width: 10
#| fig-height: 4
#| out-width: "100%"

ybar <- mean(stream_df$DO)
part_levels <- c("Total variation: point to mean", "Explained by the line: line to mean",
                 "Residual: point to line")
part_df <- rbind(
  data.frame(stream_df[, c("stream_temp", "DO")], fit_y = fitted(fit), panel = part_levels[1], y0 = ybar,         y1 = stream_df$DO),
  data.frame(stream_df[, c("stream_temp", "DO")], fit_y = fitted(fit), panel = part_levels[2], y0 = ybar,         y1 = fitted(fit)),
  data.frame(stream_df[, c("stream_temp", "DO")], fit_y = fitted(fit), panel = part_levels[3], y0 = fitted(fit), y1 = stream_df$DO)
)
part_df$panel <- factor(part_df$panel, levels = part_levels)

ss_total <- sum((stream_df$DO - ybar)^2)
ss_reg   <- sum((fitted(fit) - ybar)^2)
ss_res   <- sum(resid(fit)^2)
ss_labels <- data.frame(panel = factor(part_levels, levels = part_levels),
                        label = sprintf("SS[%s] == '%.1f'", c("Total", "Regression", "Residual"),
                                        c(ss_total, ss_reg, ss_res)))
mean_labels <- data.frame(panel = factor(part_levels, levels = part_levels))

ggplot(part_df, aes(stream_temp)) +
  geom_hline(yintercept = ybar, linetype = "dashed", colour = "grey40") +
  geom_line(aes(y = fit_y), colour = "#008cff", linewidth = 0.8) +
  geom_segment(aes(xend = stream_temp, y = y0, yend = y1, colour = panel), alpha = 0.8) +
  geom_point(aes(y = DO), size = 1.2, colour = "black") +
  geom_text(data = ss_labels, aes(x = 18.5, y = max(stream_df$DO) + 0.6, label = label),
            inherit.aes = FALSE, parse = TRUE, size = 3.6) +
  geom_text(data = mean_labels, aes(x = 23.5, y = ybar + 0.15, label = "mean DO"),
            inherit.aes = FALSE, hjust = 1, vjust = 0, size = 3.2, colour = "grey40") +
  scale_colour_manual(values = setNames(c("grey45", "#10377f", "#b22222"), part_levels), guide = "none") +
  facet_wrap(~panel) +
  labs(x = expression("Stream temperature (" * degree * "C)"), y = "Dissolved oxygen (mg/L)") +
  theme(strip.text = element_text(size = 9))
```

Squaring and summing each set of segments gives the partition:

$$SS_\text{Total} = SS_\text{Regression} + SS_\text{Residual}, \qquad `r sprintf("%.2f", ss_total)` = `r sprintf("%.2f", ss_reg)` + `r sprintf("%.2f", ss_res)`$$

The pieces line up with Chapter 8's. $SS_\text{Regression}$ is the variation the predictor explains, the role $SS_\text{Between}$ played for group means. $SS_\text{Residual}$ is what's left over around the line, the role $SS_\text{Within}$ played around each group's mean. Treating every temperature value as its own group means that the line supplies a predicted mean at each value of $x$, in place of one mean per group.

**R-squared** is the proportion of the total variation explained by the line:

$$R^2 = \frac{SS_\text{Regression}}{SS_\text{Total}} = \frac{`r sprintf("%.2f", ss_reg)`}{`r sprintf("%.2f", ss_total)`} = `r sprintf("%.2f", ss_reg / ss_total)`$$

It runs from 0 to 1, and it is the same quantity as $\eta^2$ from Chapter 8. For simple regression with one predictor, $R^2$ is also Pearson's $r$ from Chapter 2, squared: $(`r sprintf("%.3f", cor(stream_df$stream_temp, stream_df$DO))`)^2 = `r sprintf("%.3f", cor(stream_df$stream_temp, stream_df$DO)^2)`$.

**The regression ANOVA table.** Dividing each sum of squares by its degrees of freedom gives mean squares, and their ratio is $F$, exactly as in Chapter 8. The regression row uses 1 degree of freedom, one for the slope. The residual row uses $n-2$, since the line estimated two things, an intercept and a slope.

```{r}
#| echo: true
#| eval: false
anova(fit)
```

```{r}
#| label: tbl-9anova
#| tbl-cap: "Regression ANOVA table for DO ~ stream temperature."

anova_kable(fit)
```

$$F = \frac{MS_\text{Regression}}{MS_\text{Residual}} = \frac{`r sprintf("%.2f", ss_reg)`}{`r sprintf("%.3f", ss_res / (nrow(stream_df) - 2))`} = `r sprintf("%.2f", fit_glance$statistic)`$$

With one predictor, $F$ and the slope's $t$ test the same thing: $F = t^2 = (`r sprintf("%.2f", fit_tidy$statistic[2])`)^2 = `r sprintf("%.2f", fit_tidy$statistic[2]^2)`$, with the same $p$. $F$ is the form that generalizes to models with more than one predictor, which is why software reports both.

Here $R^2 = `r sprintf("%.2f", fit_glance$r.squared)`$, so stream temperature explains about `r round(fit_glance$r.squared*100)`% of the variation we observed in dissolved oxygen across sites. The remaining `r 100 - round(fit_glance$r.squared*100)`% comes from everything else the model doesn't capture (discharge, shading, groundwater input, and so on), and it shows up as the residuals.

::: {.callout-important title="Results reporting: regression"}
We report a regression result the same way as any other test: the slope with its confidence interval, the test statistic, degrees of freedom, and $p$, plus $R^2$ as the effect size.

> Dissolved oxygen declined significantly with stream temperature, $b_1 = `r sprintf("%.2f", coef(fit)[2])`$ mg/L per °C, 95% CI [`r sprintf("%.2f", fit_tidy$conf.low[2])`, `r sprintf("%.2f", fit_tidy$conf.high[2])`], $t(`r nrow(stream_df)-2`) = `r sprintf("%.2f", fit_tidy$statistic[2])`$, $p < 0.001$, $R^2 = `r sprintf("%.2f", fit_glance$r.squared)`$.
:::

## Consider both significance and effect size

As with every test in this book, the significance test is not the whole picture. Testing the slope asks whether $\beta_1$ differs from 0. $R^2$, regression's effect size, says how much of the variation the line explains. The two can disagree in either direction.

A very large sample, the kind common in remote sensing and other big-data environmental work, can make a trivial effect significant, because more observations shrink the standard error until almost any nonzero slope clears $p<0.05$. A small sample can fail the other way and miss a real, sizeable effect because its standard error is too large. Because $p$ and $R^2$ answer different questions, we report both.

@fig-9regeffectpanels shows three of these combinations: a strong relationship with a non-significant slope (underpowered), a weak relationship with a significant slope (a large sample), and a relationship that is both strong and significant. In the left panel, six points give $R^2 = 0.63$ but $p = 0.058$. In the middle panel, 5,000 points give $p = 0.004$ for a line that explains 0.2% of the variation.

```{r}
#| label: fig-9regeffectpanels
#| fig-cap: "R-squared and the slope's p-value can disagree. Left: a strong relationship (high R-squared) that isn't statistically significant. Middle: a weak relationship (low R-squared) that is significant anyway. Right: strong and significant together."
#| fig-alt: "Three scatter plots with fitted lines. The left panel shows a steep, tight relationship with few points and a non-significant p-value. The middle panel shows a nearly flat, scattered relationship with many points and a significant p-value. The right panel shows a steep, tight relationship with many points and a significant p-value."

mk_panel <- function(seed, n, slope, noise, label) {
  set.seed(seed)
  x <- runif(n, 0, 10)
  y <- 5 + slope * x + rnorm(n, 0, noise)
  d <- data.frame(x, y)
  f <- lm(y ~ x, data = d)
  g <- glance(f)
  p_txt <- fmt_p(g$p.value)
  p_txt <- ifelse(startsWith(p_txt, "<"), p_txt, paste("=", p_txt))
  d$panel <- sprintf("%s\nR² = %.3f, p %s", label, g$r.squared, p_txt)
  d
}

panel_under  <- mk_panel(seed = 15,  n = 6,    slope = 1.4,  noise = 3.2, label = "Strong & underpowered")
panel_huge_n <- mk_panel(seed = 538, n = 5000, slope = 0.05, noise = 3.0, label = "Weak & huge sample")
panel_both   <- mk_panel(seed = 538, n = 60,   slope = 1.4,  noise = 1.6, label = "Strong & significant")

all_panels <- rbind(panel_under, panel_huge_n, panel_both)
all_panels$panel <- factor(all_panels$panel, levels = unique(all_panels$panel))

ggplot(all_panels, aes(x, y)) +
  geom_point(alpha = 0.5, size = 1.4, color = "black") +
  geom_smooth(method = "lm", se = FALSE, color = "#008cff") +
  facet_wrap(~panel, scales = "free") +
  theme(strip.text = element_text(size = 12))
```

::: {.callout-caution title="Warning: How big is a \"good\" $R^2$?"}
It depends on the field. In studies of human behavior, where many unmeasured factors shape any outcome, 0.20 is a strong result. Environmental systems tend toward more direct physical relationships, so 0.5 to 0.8 for a well-chosen single predictor isn't unusual. An $R^2$ that looks *too* good, say 0.98 from a single predictor in a noisy field system, more often points to a problem in how the analysis was set up than to a real result.
:::

## Checking assumptions

Simple linear regression makes four assumptions. @tbl-9assumptions lists each with the check this chapter uses.

| Assumption | What it means | How to check | What a violation looks like |
|:-----------|:--------------|:-------------|:----------------------------|
| Linearity | The true relationship between $X$ and $Y$ is a straight line. | Residuals-versus-fitted plot | A curve, such as a U-shape |
| Independence | Each observation's error is unrelated to the others'. This is the universal assumption from Chapter 6. | The study design | Sites along the same stream, or repeated measurements over time |
| Constant variance (homoscedasticity) | The spread of residuals is about the same across the whole range of $X$. | Residuals-versus-fitted plot | A funnel that widens or narrows |
| Normal residuals | The residuals are approximately normally distributed. | Q-Q plot and Shapiro-Wilk | Points bending away from the Q-Q diagonal |

: Assumptions of simple linear regression and how to check them. {#tbl-9assumptions}

### Sometimes the truth isn't a line at all

> You can do linear regression without thinking about whether the phenomenon you're modeling is actually close to linear. But you shouldn't.\
> Jordan Ellenberg, *How Not to Be Wrong*

A straight line is not guaranteed to be the right shape for your data. A residuals-versus-fitted plot checks two of the assumptions above at once. A curve means linearity is violated, and a funnel shape means constant variance is violated.

```{r}
#| label: fig-9regresid
#| fig-cap: "Residuals versus fitted values for the stream temperature model. A random scatter around zero, with no funnel shape or curve, supports linearity and constant variance."
#| fig-alt: "Residuals plotted against fitted DO values for the stream model, scattered above and below a dashed line at zero with no curve and no widening from left to right."

stream_df$fitted <- fitted(fit)
stream_df$resid <- resid(fit)

ggplot(stream_df, aes(x = fitted, y = resid)) +
  geom_point(color = "black", size = 1.8) +
  # residual 0 = exactly on the fitted line, so the reference takes the line's color
  geom_hline(yintercept = 0, linetype = "dashed", color = "#008cff", linewidth = 0.8) +
  # label sits in the right margin, clear of the points
  annotate("text", x = Inf, y = 0, label = "0 = on the\nregression line", colour = "#008cff",
           hjust = -0.05, size = 3.6, lineheight = 0.9) +
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(5.5, 95, 5.5, 5.5)) +
  xlab("Fitted values (predicted DO)") +
  ylab("Residuals")
```

@fig-9regresid, for our stream temperature data, shows residuals scattered evenly above and below zero, with no curve (which would suggest the true relationship isn't linear) and no funnel shape (which would suggest variance changes systematically across the range of $X$).

Next, check the residuals for a second data set, 14 days of invented avian influenza case counts. We built the counts from an exponential curve, $6e^{0.28 \times \text{day}}$, plus random scatter, so we know the true shape that a straight line will miss. The design is the same shape as the stream example, with time as the continuous predictor.

```{r}
#| label: fig-9outbreakdesign
#| fig-cap: "The outbreak data as a design. It has the same structure as the stream example."
#| fig-alt: "Study design schematic. IV: day of outbreak, continuous, an arrow running from day 1 to day 14. DV: new cases detected per day."
#| fig-asp: 0.34
#| out-width: "100%"

design_continuous("Day of outbreak (continuous)", "day 1", "day 14", "14 days",
                  "New cases detected per day",
                  "Do new cases change over time?")
```

```{r}
#| label: fig-9outbreakfit
#| fig-cap: "New avian influenza cases per day, days 1-14 of an outbreak, with a fitted straight line."
#| fig-alt: "Scatter plot of new cases per day against day of outbreak, with a fitted straight regression line. Points curve gently above and below the line."

set.seed(538)
days <- 1:14
true_rate <- 0.28
cases <- round(6 * exp(true_rate * days) + rnorm(14, 0, 5))
cases <- pmax(cases, 1)
outbreak_df <- data.frame(days, cases)

fit_outbreak <- lm(cases ~ days, data = outbreak_df)
outbreak_tidy <- glance(fit_outbreak)

ggplot(outbreak_df, aes(days, cases)) +
  geom_point(color = "black", size = 1.8) +
  geom_smooth(method = "lm", se = FALSE, color = "#008cff") +
  xlab("Day of outbreak") +
  ylab("New cases detected")
```

A straight line fit to these 14 days gives $R^2 = `r sprintf("%.2f", outbreak_tidy$r.squared)`$ and $p`r fmt_p(outbreak_tidy$p.value)`$, which by the benchmark in the "How big is a good $R^2$?" warning above looks like a solid, significant fit for a field system. Nothing about those numbers alone raises a flag. The plot does raise one, though. The fitted line starts at about -40 cases on day 1, and a count of cases can't be negative.

**Check the residuals.** The residuals-versus-fitted plot shows a problem that the $R^2$ and slope test did not.

```{r}
#| label: fig-9outbreakresid
#| fig-cap: "Residuals versus fitted values for the linear outbreak model. A clear U-shape, not random scatter, is the signature of a nonlinear relationship forced into a straight line."
#| fig-alt: "Residuals plotted against fitted values. The residuals form a clear U-shaped curve: negative in the middle, positive at both low and high fitted values, rather than scattering randomly around zero."

outbreak_df$fitted <- fitted(fit_outbreak)
outbreak_df$resid <- resid(fit_outbreak)

ggplot(outbreak_df, aes(fitted, resid)) +
  geom_point(color = "black", size = 1.8) +
  # residual 0 = exactly on the fitted line, so the reference takes the line's color
  geom_hline(yintercept = 0, linetype = "dashed", color = "#008cff", linewidth = 0.8) +
  # label sits in the right margin, clear of the points
  annotate("text", x = Inf, y = 0, label = "0 = on the\nregression line", colour = "#008cff",
           hjust = -0.05, size = 3.6, lineheight = 0.9) +
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(5.5, 95, 5.5, 5.5)) +
  xlab("Fitted values (predicted cases)") +
  ylab("Residuals")
```

The residuals form a U. The line underpredicts early and late in the outbreak and overpredicts in the middle, which is the pattern a straight line fit to a curve produces.

If a residual plot curves instead of scattering randomly, a straight line is the wrong model for the data.

Fixes typically include:

-   a different model (often adding a polynomial term)
-   a transformation, re-expressing a variable on a different scale so a straight line fits it (log-transforming a variable that grows multiplicatively, like the outbreak data above, is the most common case)
-   or a nonlinear model entirely, beyond what this course covers

### Checking normality

For normality, we check the residuals two ways: visually, with a **Q-Q plot**, and formally, with the Shapiro-Wilk test.

A Q-Q plot sorts the residuals and compares them to where a perfectly normal distribution would place values at the same percentiles. If the residuals are normal, the points fall close to the diagonal line. A systematic curve or S-shape means they don't.

```{r}
#| label: fig-9regqq
#| fig-cap: "Normal Q-Q plot of the stream temperature model's residuals. Points close to the diagonal support normality."
#| fig-alt: "A normal quantile-quantile plot of the residuals. The points fall close to the diagonal reference line, showing no strong departure from normality."

ggplot(stream_df, aes(sample = resid)) +
  stat_qq(color = "black", size = 1.8) +
  stat_qq_line(color = "#008cff") +
  xlab("Theoretical quantiles") +
  ylab("Sample quantiles")
```

@fig-9regqq looks close to straight, consistent with normal residuals. Shapiro-Wilk asks the same question formally. Its $H_0$ is that the residuals come from a normal distribution, so a small $p$-value is evidence against normality.

```{r}
#| echo: true
#| eval: false
shapiro.test(resid(fit))
```

```{r}
sh_out <- shapiro.test(resid(fit))
kable(
  data.frame(Statistic = round(unname(sh_out$statistic), 2),
             `p-value` = fmt_p(sh_out$p.value), check.names = FALSE),
  caption = "Shapiro-Wilk test on the regression residuals"
)
```

The residuals don't depart from normality here ($p = `r fmt_p(sh_out$p.value)`$). As with any formal normality test, treat this as one piece of evidence alongside the Q-Q plot and residual plot, not a strict gate. Shapiro-Wilk has the least power to detect real departures when the sample is small, which is also when the visual checks matter most.

Now apply the same two normality checks to the avian influenza model, the one we already know is badly nonlinear.

```{r}
#| label: fig-9outbreakqq
#| fig-cap: "Normal Q-Q plot of the outbreak model's residuals. The lowest residuals bend away from the diagonal, but not by enough for Shapiro-Wilk to reject normality."
#| fig-alt: "A normal quantile-quantile plot of the 14 residuals. The middle and upper points fall near the diagonal reference line, but the lowest five level off near minus 40, well above where the line would put them."

ggplot(outbreak_df, aes(sample = resid)) +
  stat_qq(color = "black", size = 1.8) +
  stat_qq_line(color = "#008cff") +
  xlab("Theoretical quantiles") +
  ylab("Sample quantiles")
```

```{r}
sh_outbreak <- shapiro.test(resid(fit_outbreak))
```

The Q-Q plot bends away from the diagonal at the low end, but Shapiro-Wilk does not reject normality ($p = `r fmt_p(sh_outbreak$p.value)`$). On a model we already know is the wrong shape, the normality checks give at most a borderline signal. They check *normality of the residuals*, which is a different assumption from linearity, and a model can badly violate linearity while its residuals look only mildly non-normal.

## Using the model: prediction

Once we've confirmed a real linear relationship, the fitted equation lets us predict DO at any stream temperature. Plugging temperature into the fitted line directly gives the prediction: at 20°C,

$$\widehat{DO} = b_0 + b_1(20) = `r sprintf("%.2f", coef(fit)[1])` + (`r sprintf("%.2f", coef(fit)[2])`)(20) = `r sprintf("%.2f", coef(fit)[1] + coef(fit)[2]*20)`$$

R gives the same number, along with a confidence interval:

```{r}
#| echo: true
#| eval: false
predict(fit, newdata = data.frame(stream_temp = 20), interval = "confidence")
```

```{r}
newd <- data.frame(stream_temp = 20)
pred_out <- predict(fit, newd, interval = "confidence")
kable(
  data.frame(`Stream temperature` = 20, Predicted = round(pred_out[1],2),
             `CI low` = round(pred_out[2],2), `CI high` = round(pred_out[3],2),
             check.names = FALSE),
  caption = "Predicted dissolved oxygen at 20°C, with 95% confidence interval"
)
```

At 20°C, the model predicts DO of `r sprintf("%.2f", pred_out[1])` mg/L, 95% CI [`r sprintf("%.2f", pred_out[2])`, `r sprintf("%.2f", pred_out[3])`]. Because 20°C falls inside the observed range of `r sprintf("%.1f", min(stream_df$stream_temp))` to `r sprintf("%.1f", max(stream_df$stream_temp))`°C, this prediction is interpolation, which is what regression is good for.

Predicting DO at, say, 35°C would be extrapolation. The fitted line gives `r sprintf("%.1f", predict(fit, data.frame(stream_temp = 35)))` mg/L at 35°C, a number with no data behind it. Nothing in the data tells us the relationship stays linear (or even stays a relationship at all) beyond the range we actually observed, and the further outside that range, the less trustworthy the prediction. @fig-9regextrap shows both zones on the same line.

```{r}
#| label: fig-9regextrap
#| fig-cap: "The fitted line within the observed temperature range (solid blue, interpolation) and extended beyond it in both directions (dashed red, extrapolation). Dotted grey lines mark the lowest and highest observed temperatures, 10.0 and 23.3°C."
#| fig-alt: "Scatter plot of dissolved oxygen against stream temperature. A solid blue line covers the observed range from 10 to about 23 degrees C, and a dashed red line extends the same fitted trend below and above that range."

obs_lim <- range(stream_df$stream_temp)
obs_range <- data.frame(stream_temp = seq(obs_lim[1], obs_lim[2], length.out = 50))
obs_range$DO <- predict(fit, obs_range)
obs_range$part <- "Observed range (interpolation)"

extrap_range <- data.frame(stream_temp = seq(0, 35, length.out = 100))
extrap_range$DO <- predict(fit, extrap_range)
extrap_range$part <- "Beyond observed range (extrapolation)"
extrap_range <- extrap_range[extrap_range$stream_temp < obs_lim[1] | extrap_range$stream_temp > obs_lim[2], ]

extrap_range$part <- factor(extrap_range$part, levels = c("Observed range (interpolation)", "Beyond observed range (extrapolation)"))
obs_range$part <- factor(obs_range$part, levels = levels(extrap_range$part))

ggplot() +
  geom_line(data = extrap_range, aes(stream_temp, DO, color = part, linetype = part), linewidth = 1.5) +
  geom_line(data = obs_range, aes(stream_temp, DO, color = part, linetype = part), linewidth = 1) +
  geom_point(data = stream_df, aes(stream_temp, DO), size = 1.5) +
  geom_vline(xintercept = obs_lim, linetype = "dotted", color = "grey60", linewidth=1.2) +
  scale_color_manual(values = c("Observed range (interpolation)" = "#008cff", "Beyond observed range (extrapolation)" = "#b22222")) +
  scale_linetype_manual(values = c("Observed range (interpolation)" = "solid", "Beyond observed range (extrapolation)" = "dashed")) +
  labs(x = expression("Stream temperature ("*degree*"C)"), y = "Dissolved oxygen (mg/L)", color = NULL, linetype = NULL) +
  theme(legend.position = "top")
```

Regression is used for extrapolation all the time, and often reasonably. A well-understood physical process can justify projecting a line beyond the data you happened to collect. Extrapolating a straight line is only safe if the true relationship actually stays straight past the range you observed.

The problem is that a narrow window of data can't tell you which kind of curve you're actually on. @fig-9curveshapes shows four very different relationships, exponential growth, a sigmoid, a hump-shaped curve, and a truly straight line, each with the same short observed window highlighted in blue. In each panel a straight line (dashed red) is fit to that window and extended. Once you see the full curve (grey), you can see how badly linear extrapolation beyond the range of observed data would perform in each case.

```{r}
#| label: fig-9curveshapes
#| fig-cap: "Four different true relationships that all look locally linear over a short observed window (blue). A line fit to each window (dashed red) matches it closely inside the window, but only the linear case stays correct once the full curve (grey) is revealed."
#| fig-alt: "Four panels: exponential, sigmoid, hump-shaped, and linear curves. In each, a short initial segment is highlighted in blue and looks nearly straight. A dashed red line extends that local trend across the full range. The full grey curve shows the true shape, which diverges sharply from the dashed line except in the linear panel, where the two coincide."

curve_x <- seq(0, 10, length.out = 300)
mk_curve <- function(x, y, label) data.frame(x = x, y = y, curve = label)

curve_df <- rbind(
  mk_curve(curve_x, exp(0.5 * curve_x), "Exponential"),
  mk_curve(curve_x, 10 / (1 + exp(-(curve_x - 5))), "Sigmoid"),
  mk_curve(curve_x, -(curve_x - 5)^2 + 25, "Hump-shaped"),
  mk_curve(curve_x, 2 * curve_x, "Linear")
)
curve_df$curve <- factor(curve_df$curve, levels = c("Exponential", "Sigmoid", "Hump-shaped", "Linear"))

curve_window <- curve_df[curve_df$x <= 3, ]

curve_fits <- do.call(rbind, lapply(split(curve_window, curve_window$curve), function(w) {
  f <- lm(y ~ x, data = w)
  data.frame(x = curve_x, pred = predict(f, data.frame(x = curve_x)), curve = w$curve[1])
}))

ggplot() +
  geom_line(data = curve_df, aes(x, y), color = "grey50", linewidth = 0.9) +
  geom_line(data = curve_fits, aes(x, pred), color = "#b22222", linetype = "dashed", linewidth = 0.9) +
  geom_line(data = curve_window, aes(x, y), color = "#008cff", linewidth = 2.2) +
  facet_wrap(~curve, scales = "free_y") +
  theme_classic(base_size = 13) +
  labs(x = "x", y = "y")
```

Suppose officials use the linear fitted line from the outbreak example above to project avian influenza cases at day 30, well past the 14 days observed.

```{r}
#| label: fig-9outbreakextrap
#| fig-cap: "The linear fit extended (dashed) past the observed range, against the true exponential trajectory (dotted grey). The straight line badly underestimates how large the outbreak actually becomes."
#| fig-alt: "Line graph of cases against day of outbreak from day 1 to day 30. A solid blue line and observed points cover days 1 through 14. A dashed red line continues the same linear trend from day 14 to day 30, staying low. A dotted grey curve shows the true exponential trajectory, which rises steeply and dramatically outpaces the dashed line by day 30."

obs_days <- 1:14
extrap_days <- 14:30
full_days <- 1:30

lin_obs <- data.frame(days = obs_days,
                       cases = predict(fit_outbreak, data.frame(days = obs_days)),
                       part = "Observed range (interpolation)")
lin_extrap_df <- data.frame(days = extrap_days,
                             cases = predict(fit_outbreak, data.frame(days = extrap_days)),
                             part = "Beyond observed range (extrapolation)")
lin_obs$part <- factor(lin_obs$part, levels = c("Observed range (interpolation)", "Beyond observed range (extrapolation)"))
lin_extrap_df$part <- factor(lin_extrap_df$part, levels = levels(lin_obs$part))
true_df <- data.frame(days = full_days, cases = 6 * exp(true_rate * full_days))

ggplot() +
  geom_line(data = true_df, aes(days, cases), color = "grey40", linetype = "dotted", linewidth = 1.1) +
  geom_line(data = lin_obs, aes(days, cases, linetype = part, color = part), linewidth = 1.3) +
  geom_line(data = lin_extrap_df, aes(days, cases, linetype = part, color = part), linewidth = 1.3) +
  geom_point(data = outbreak_df, aes(days, cases)) +
  geom_vline(xintercept = 14, linetype = "dashed", color = "grey60") +
  annotate("text", x = 14.3, y = 26500, label = "last observed day", hjust = 0, size = 5, color = "grey40") +
  annotate("text", x = 26, y = 14000, label = "true trajectory", hjust = 1, color = "grey40", size = 5) +
  scale_color_manual(values = c("Observed range (interpolation)" = "#008cff", "Beyond observed range (extrapolation)" = "#b22222")) +
  scale_linetype_manual(values = c("Observed range (interpolation)" = "solid", "Beyond observed range (extrapolation)" = "dashed")) +
  labs(x = "Day of outbreak", y = "New cases detected", color = NULL, linetype = NULL) +
  theme(legend.position = "top")
```

```{r}
lin_pred30 <- predict(fit_outbreak, data.frame(days = 30))
true_val30 <- 6 * exp(true_rate * 30)
```

In @fig-9outbreakextrap, the dashed line keeps rising slowly past day 14, while the true exponential curve rises far above it.

The linear model predicts about `r round(lin_pred30)` cases on day 30. On the exponential curve the data were built from, the count is closer to `r formatC(round(true_val30), big.mark=",", format="d")`, roughly `r round(true_val30/lin_pred30)` times higher.

Planning surveillance and containment staffing off the linear forecast would leave the response badly under-resourced for the outbreak that actually arrives. A straight line can look fine over a short early window and still be very wrong about where that window leads.

This problem calls for a different model. More data or a better-fitting straight line would not fix it. Fitting curves and transformations is beyond this course, but you can recognize when you need one by plotting the data and the residuals.

## Chapter Summary

**Why it matters.** Chapter 2 drew a best-fit line through a scatterplot. This chapter turns that line into something you can test, quantify, and make predictions from.

**Core ideas**

-   **The slope can be tested.** Chapter 2's $r$ described a relationship but came with no test of whether it was more than chance. Fitting a line turns that relationship into a model whose slope we can test: $H_0: \beta_1 = 0$ says there is no linear relationship. Testing it uses a $t$-test on the slope divided by its standard error, with $df = n-2$.
-   **Regression partitions variation the way ANOVA does.** $SS_\text{Total}$ splits into $SS_\text{Regression}$ (Chapter 8's between) and $SS_\text{Residual}$ (Chapter 8's within), and $F = MS_\text{Regression}/MS_\text{Residual}$, which equals $t^2$ with one predictor. $R^2$, the proportion of variation explained by the line, is the effect size. A relationship can be real ($p$ small) and still explain very little, which is why you report both.
-   **Look at your data, then look at your residuals.** The assumptions that must be met are linearity, independence, constant variance, and roughly normal residuals. A residuals-versus-fitted plot catches violations a formal test can miss entirely, as the outbreak example showed.
-   **Predict inside the observed range, and extrapolate beyond it only with good reason.** Interpolating within the observed range of $x$ is what regression is for. Extrapolating beyond it is sometimes justified, but only when you have good reason to believe the relationship stays the same shape past the range you observed.
-   **A good $R^2$ can still be the wrong model.** The outbreak example fit a straight line to exponential growth and got $R^2=`r sprintf("%.2f", outbreak_tidy$r.squared)`$, a residuals-versus-fitted plot that clearly curved, and an extrapolated forecast about 26,000 cases short of the exponential curve the data came from. The residual plot catches this. A normality check on the same residuals gives only a borderline signal, because it tests a different assumption.
