---
author: 
  - Mallory L. Barnes
aliases: [redirects/ancova.html]
---

```{r, include = FALSE}
source("global_stuff.R")
knitr::opts_chunk$set(cache=FALSE)
```

# Analysis of Covariance (ANCOVA)

```{r}
# Libraries
library(tidyverse)
library(ggpubr)

```

\vspace{1cm}

**ANCOVA** uses the same linear model ideas from regression, with one continuous covariate and one categorical factor. We first test whether groups have different **slopes** with respect to the covariate, then, only if slopes are parallel, we test whether groups differ in **adjusted means**.

**A quick refresher, since regression was a few weeks ago.** A simple linear model is

$$y = \beta_0 + \beta_1 x + \varepsilon$$

one intercept, one slope, one line drawn through the data. ANOVA has the same shape, with a categorical predictor standing in for the continuous one. ANCOVA puts both kinds of predictor in the same equation, and that raises the question this whole chapter turns on: do our groups share that single line, or does each group need its own?

::: callout-note
**Going deeper.** Every test in this book, $t$-tests, ANOVA, regression, and ANCOVA, is the same underlying model with different kinds of predictors. Chasing that unification all the way, through reference coding across many categories, planned contrasts, and several continuous predictors at once, is what a second course in regression is for. Here we need exactly one piece of it: the two-category indicator variable.
:::

## Moving Beyond Regression and ANOVA to ANCOVA

As we've explored statistical modeling, we've understood the power of both regression and ANOVA. Now, let's merge them. ANCOVA allows us to examine relationships across different categories.

Consider these questions that may arise in environmental research:

-   Are dbh (Diameter at Breast Height) and height related similarly for tulip poplars and oaks?

-   Are biomass and BTUs (British Thermal Units) related similarly for corn stover and Miscanthus?

-   Does the exposure to PFAS correlate with the lifetime incidence of cancer uniformly across low- and high-income American populations?

This specific subsection of general linear models is known as **Analysis of Covariance -- ANCOVA**. ANCOVA is a linear model with one continuous covariate and one categorical factor; it tests group differences **after** adjusting for the covariate and, first, whether slopes differ across groups.

### Conceptual Background for ANCOVA

At its core, ANCOVA checks whether groups share the same line, or whether each group needs its own slope and intercept. That's exactly what the interaction test is doing. @fig-11onevstwo shows the three possibilities, and it's worth keeping this picture in mind for the rest of the chapter.

```{r}
#| label: fig-11onevstwo
#| fig-cap: "The question ANCOVA answers. Left: one line describes both groups. Middle: the groups need different intercepts but share a slope, so the lines are parallel. Right: the groups need different slopes, so the lines are not parallel and the gap between them depends on where you look."
#| fig-alt: "Three scatter plots, each with two colored groups of points. In the first, a single line runs through both groups. In the second, two parallel lines sit at different heights. In the third, two lines with visibly different slopes diverge from left to right."
#| out-width: "100%"
#| fig-asp: 0.45

set.seed(11)
n  <- 30
x  <- runif(n * 2, 0, 10)
g  <- rep(c("Group A", "Group B"), each = n)
B  <- g == "Group B"

# Three scenarios, deliberately kept on a similar y range so the shared axis
# doesn't squash the first two panels.
y_same     <- 8 + 0.5 * x + rnorm(n * 2, 0, 0.9)
y_parallel <- 5 + 0.5 * x + ifelse(B, 5, 0) + rnorm(n * 2, 0, 0.9)
y_slopes   <- ifelse(B, 5 + 0.9 * x, 12 - 0.2 * x) + rnorm(n * 2, 0, 0.9)

lines_df <- rbind(
  data.frame(x, g, y = y_same,     panel = "One shared line"),
  data.frame(x, g, y = y_parallel, panel = "Parallel lines\n(same slope, different intercept)"),
  data.frame(x, g, y = y_slopes,   panel = "Different slopes\n(the interaction)")
)
lines_df$panel <- factor(lines_df$panel, levels = unique(lines_df$panel))

ggplot(lines_df, aes(x, y, color = g)) +
  geom_point(alpha = 0.55, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  facet_wrap(~panel) +
  labs(x = "Covariate", y = "Response", color = NULL) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top", strip.text = element_text(size = 9.5))
```

ANCOVA works right to left across that figure: first it asks whether the third panel is what we have (do the slopes differ?), and only if the answer is no does it ask whether we have the second panel or the first (do the intercepts differ?).

### Terminology in LMs

A few key terms help clarify how variables play different roles in linear models:

-   **Fixed factor:** A categorical variable with specific, intentional levels chosen by the researcher.

    *Example:* comparing plant growth across seasons (spring, summer, autumn, winter). Each season is a fixed, predefined category of interest.

-   **Random factor:** A categorical variable whose levels represent a random sample from a larger population.

    *Example:* if tree plots are randomly selected from many possible sites, "plot" is a random factor because it reflects sampling variation rather than a treatment.

-   **Covariate:** A continuous variable that changes alongside the response variable and helps explain part of its variation.

    In environmental studies, common covariates include temperature, rainfall, soil moisture, or pollution concentration, all factors that may influence the outcome but are not the main treatment of interest.

    The purpose of including a covariate is to **adjust for its influence**, allowing a clearer test of the categorical effects once that continuous background variation is accounted for.

## What ANCOVA Is

**Analysis of Covariance (ANCOVA)** extends ANOVA by adding a continuous predictor. It fits a linear model with both a categorical factor and a covariate, allowing us to compare groups **and** account for how the response changes with the continuous variable.

The questions that opened this chapter all ask the same kind of thing: does a *relationship* hold up across groups? That framing fits environmental work particularly well, because you usually can't experimentally strip out a covariate like temperature, rainfall, or year anyway. Often you wouldn't want to. It may be the reason you designed the study in the first place. **The covariate is the point, and the interaction is the finding.**

::: callout-tip
**The other way you'll see ANCOVA.** Look this test up elsewhere and the covariate is usually a **nuisance**: something quietly influencing the outcome that you add to the model to control for, so a group comparison comes out fairer. There the interaction is a box to check rather than a result. Most "which test should I use" flowcharts assume that version, which is why ANCOVA rarely gets its own branch on them. Both are the same test; the difference is only which term you came to interpret, so decide that before you fit anything.
:::

For ANCOVA, you need:

1)  A continuous dependent variable (the main effect you're studying)
2)  At least one continuous independent variable (covariate)
3)  At least one categorical independent variable (which can be either a fixed or random factor)

### Indicator (dummy) variables and how they work

Categorical predictors get into a linear model by being coded as **indicator variables**. For a factor with $k$ groups you include $k-1$ indicators; the omitted group becomes the **reference**, and every coefficient is interpreted relative to it. With two groups, one indicator does the whole job:

$$x_c = \begin{cases} 0 & \text{group 1 (the reference group)} \\ 1 & \text{group 2} \end{cases}$$

In R, `lm(y ~ x + Group)` creates these indicators for you automatically.

**Indicator variables in action.** When an indicator variable is turned on (set to 1), it activates the terms multiplied by that variable. Depending on where the indicator appears in the model, that can change the intercept, the slope, or both.

Consider two predictors: a continuous variable $x_l$ and an indicator $x_c$ (0 or 1). Use this model:

$$y = 2x_l + 1(x_l x_c) - 7x_c + 3$$

**When** $x_c = 0$ (indicator off), only the baseline terms are active:

$$y = 2x_l + 3 \qquad \text{slope} = 2, \ \text{intercept} = 3$$

**When** $x_c = 1$ (indicator on), the new terms activate:

$$y = (2x_l + 3) + (1x_l - 7) = 3x_l - 4 \qquad \text{slope} = 3, \ \text{intercept} = -4$$

Turning on the indicator adds **+1 to the slope** and **−7 to the intercept**. One equation, two different lines, with the indicator deciding which one applies.

This is how linear models represent different groups: the indicator ($x_c$) and its interaction with $x_l$ let a single equation shift the intercept and the slope for each group. ANCOVA uses exactly this logic to test, in order, whether the **slopes differ** between groups (the $x_l x_c$ term) and, if they don't, whether the **adjusted means** differ (the $x_c$ term).

**Naming the pieces.** Write that same equation in the general form and each coefficient now has a job:

$$y = \underbrace{\beta_0}_{\text{reference intercept}} + \underbrace{\beta_1 x_l}_{\text{reference slope}} + \underbrace{\beta_2 x_c}_{\text{intercept shift}} + \underbrace{\beta_3 x_l x_c}_{\text{slope shift}} + \varepsilon$$

-   $\beta_0$ and $\beta_1$ describe the line for the **reference group** ($x_c=0$).
-   $\beta_2$ is how much the **intercept** moves for the other group.
-   $\beta_3$ is how much the **slope** changes for the other group. This is the interaction term.

If we only include $\beta_2$ and leave out $\beta_3$, we get a restricted version: the intercept can shift but the slope cannot, so the two lines are forced to be **parallel**. That is the parallel-lines model, and it is the one we fall back to when the slope difference turns out not to be significant.

::: callout-note
**Why this is worth the detour.** Every coefficient you will read out of an ANCOVA table is one of these four. Without the indicator, $\beta_2$ and $\beta_3$ are just numbers R hands you. With it, $\beta_3$ has a plain-language meaning you can check against a graph: it is the amount by which one group's line tilts away from the other's.
:::

### The Null Hypothesis in ANCOVA

**Why this matters.** ANCOVA asks two questions in order. First, do groups have the **same slope** for the covariate--response relation. Only if slopes are the same do we ask whether groups differ in **adjusted means** after controlling for the covariate.

**Model.** For one covariate (x) and a two-level group indicator $g\in{0,1}$,

$y=\beta_0+\beta_1 x+\beta_2 g+\beta_3(x\times g)+\varepsilon .$

**Step 1: Slopes equal across groups?** This tests whether the covariate effect is the same in each group.

-   Null: $H_{0,\text{int}}:\beta_3=0$ (parallel lines, same slope)
-   Alternative: $H_{A,\text{int}}:\beta_3\neq 0$ (non-parallel lines, different slopes)

**Interpretation if rejected.** Report the two fitted lines and describe how the slope differs by group. Do not test adjusted means, because the idea of a single adjusted mean per group assumes parallel slopes.

**Step 2: Adjusted means equal?** *(only if Step 1 retained parallel slopes)* Fit the model without the interaction,

$y=\beta_0+\beta_1 x+\beta_2 g+\varepsilon .$

-   Null: $H_{0,\text{group}}:\beta_2=0$ (no difference in adjusted means)
-   Alternative: $H_{A,\text{group}}:\beta_2\neq 0$

**What the parameters mean.**

-   $\beta_1$: common slope of $y$ on $x$ when slopes are parallel.
-   $\beta_2$: difference in group means **after** adjusting for $x$ (tested only when $\beta_3=0$).
-   $\beta_3$: difference in slopes between groups.

**Intercepts and centering.** Intercepts are evaluated at $x=0$. Compare intercepts only if you center the covariate, for example $\tilde x=x-\bar x$, so that the intercepts represent group means at the average covariate value.

**What to report.**

-   If $x\times g$ is significant: "The effect of $x$ differs by group," then provide the two line equations and a figure with separate fitted lines.
-   If $x\times g$ is not significant: report the common slope $\beta_1$ and the adjusted group difference $\beta_2$, with a figure showing parallel lines.

**Common pitfalls.**

-   Interpreting main effects when the interaction is significant.
-   Comparing raw intercepts when $x$ is not centered.
-   Treating adjusted means as meaningful when slopes are not parallel.

### Conducting the Test

When we run an ANCOVA, the output includes an **omnibus F-test**, a single test of whether there are any differences in the adjusted group means after accounting for the covariate.

If this test is **significant**, we reject the overall null hypothesis that all adjusted means are equal.

**What happens next:** A significant omnibus test tells us that at least one group differs, but it doesn't tell us *which* groups differ or *why*.

To interpret the result, we must **decompose the omnibus effect** into its components:

1.  **Interaction (slope difference):** Are the slopes of the covariate--response relationship the same across groups?

2.  **Adjusted means (intercept difference):** If the slopes are equal, do the groups still differ in their adjusted means?

This two-step logic, testing slope equality first and then adjusted means, is central to ANCOVA and parallels how we follow a significant ANOVA with post-hoc comparisons.

**Interpreting the Findings:**

Rejecting the omnibus null indicates that the categorical factor affects the dependent variable *after controlling for the covariate*.

However, as always, statistical significance does not guarantee ecological or practical importance. The results should be interpreted in light of the study's design, measurement precision, and the size of the observed effects.

::: callout-note
**ANCOVA ASSUMPTIONS**

-   Linearity within each group (covariate--response relation is linear).
-   **Homogeneity of slopes** (test via interaction).
-   Independent observations.
-   Approximately normal residuals and similar variances across groups.
-   Covariate measured without substantial error and not affected by treatment
:::

## Full ANCOVA Example

### Question

Imagine we're environmental scientists studying the impact of **air pollution** (a continuous variable) on **bird species density** (our dependent variable).

We want to know whether this relationship differs between **urban** and **rural** areas (our categorical variable).

For this purpose, we've created a synthetic dataset to model the scenario.

### Model

We'll fit an ANCOVA model to test whether location type influences bird density **after accounting for** air pollution.

The categorical variable (urban vs. rural) is represented as a dummy variable, and air pollution enters as a covariate.

-   **Dataset Overview:** 200 observations of bird density across urban and rural sites, each with a measured level of air pollution.

### Why We're Running the ANCOVA

-   **Research Question:** Does the effect of air pollution on bird density differ between urban and rural areas?

-   **Expectation:** Air pollution will have a stronger negative effect in urban areas than in rural areas.

### Null and Alternative Hypotheses in our example ANCOVA

0.  **Omnibus (adjusted means) hypothesis** After adjusting for air pollution, do urban and rural sites differ in mean bird density?

    $H_{0}: \mu_{\text{Urban, adj}} = \mu_{\text{Rural, adj}}$

    $H_{A}: \mu_{\text{Urban, adj}} \neq \mu_{\text{Rural, adj}}$

    Here, $\mu_{\text{Urban, adj}}$ and $\mu_{\text{Rural, adj}}$ are the *adjusted* group means once the covariate (air pollution) is accounted for.

1.  **Slope (interaction) hypothesis**

    Does air pollution affect bird density in the same way across urban and rural sites?

    $H_{0,1}: \beta_{\text{interaction}} = 0$

    $H_{A,1}: \beta_{\text{interaction}} \neq 0$

    A significant interaction means the slopes differ, and air pollution's relationship with bird density depends on area type.

2.  **Intercept (baseline difference) hypothesis**

    If slopes are parallel (interaction not significant), do the adjusted means differ?

    $H_{0,2}: \beta_{0,\text{Urban}} = \beta_{0,\text{Rural}}$

    $H_{A,2}: \beta_{0,\text{Urban}} \neq \beta_{0,\text{Rural}}$

    Rejecting $H_{0,2}$ indicates an inherent group difference in bird density that is independent of pollution.

### Analysis

```{r}
# Set seed for reproducibility
set.seed(123)

n <- 200                                                     # number of sites
area_type <- sample(c("Urban", "Rural"), n, replace = TRUE)  # randomly assign area type
air_pollution <- round(runif(n, min = 25, max = 150), 1)     # continuous covariate
urban <- ifelse(area_type == "Urban", 1, 0)                  # 0/1 indicator

# Bird density declines with pollution everywhere, but more steeply in urban
# sites. That difference in slopes is the interaction we want the model to find.
bird_density <- 40 -
  0.05 * air_pollution -                # rural slope
  3    * urban -                        # urban intercept offset
  0.10 * air_pollution * urban +        # extra urban steepness (the interaction)
  rnorm(n, mean = 0, sd = 5)            # noise

bird_density <- round(pmax(bird_density, 0), 0)

data <- data.frame(AreaType = area_type, AirPollution = air_pollution, BirdDensity = bird_density)
```

Before we look at the model results, it's always good practice to **look at the data first**.

Visual inspection helps us check for broad patterns and see whether the model output later "makes sense."

Let's start with the response variable, **bird density**, in @fig-11birddensity. This gives us a quick visual sense of whether urban and rural areas differ in their average values.

```{r}
#| label: fig-11birddensity
#| fig-cap: "Bird density by area type. Rural sites sit higher than urban ones."
#| fig-alt: "Boxplot of bird density for rural and urban area types. The rural box is centered higher than the urban box, with overlapping ranges."
#| echo: true

ggplot(data, aes(AreaType, BirdDensity)) +
  geom_boxplot() + theme_classic(base_size = 12) +
  labs(x = "Area Type", y = "Bird Density")
```

At first glance, bird density appears **higher in rural areas** than in urban ones, consistent with what we might expect ecologically.

Next, check the **covariate**, air pollution.

If the covariate differs sharply between groups, ANCOVA will "adjust" for that difference.

If it doesn't, group comparisons will look more like a simple ANOVA.

```{r}
#| label: fig-11airpollution
#| fig-cap: "Air pollution by area type. The two groups look broadly similar, so there is no large covariate imbalance for the model to adjust away."
#| fig-alt: "Boxplot of air pollution for rural and urban area types. The two boxes overlap substantially with similar centers."
#| echo: true

ggplot(data, aes(AreaType, AirPollution)) +
  geom_boxplot() + theme_classic(base_size = 12) +
  labs(x = "Area Type", y = "Air Pollution")
```

Air pollution looks fairly similar between urban and rural sites, with no obvious systematic difference.

This isn't diagnostic, but it helps us **anticipate what the model might show**: we may find that group differences in bird density aren't explained by pollution alone.

### Results of the ANCOVA Model

Now that we've explored the data visually, we can test our hypotheses formally using ANCOVA.

**Step 1: Fit the ANCOVA model**.

`ancova <- aov(BirdDensity ~ AirPollution * AreaType, data = data)`

```{r}
library(broom)

ancova <- aov(BirdDensity ~ AirPollution * AreaType, data = data)
ancova_tidy <- broom::tidy(ancova)

knitr::kable(
ancova_tidy[, c("term", "df", "sumsq", "meansq", "statistic", "p.value")],
digits = c(0, 0, 2, 2, 2, 4),
caption = "ANCOVA Table"
)
```

**Step 2: Interpret each term in light of the hypotheses**

**0. Omnibus (adjusted means)** :

$H_{0}: \mu_{\text{Urban, adj}} = \mu_{\text{Rural, adj}}$

The overall ANCOVA test examines whether adjusted means differ between groups after accounting for the covariate.

Because we included an interaction term, this omnibus test is **decomposed** into two parts:

-   the **interaction** (slope difference), and

-   the **main effects** (covariate and group).

So we interpret the omnibus test through these individual terms rather than as a separate $p$-value.

**1. Interaction (slope difference)**:

$H_{0,1}: \beta_{\text{interaction}} = 0$

The interaction between air pollution and area type is **significant** ($p = .001$).

Reject $H_{0,1}$ → our slopes are not parallel. Air pollution's relationship with bird density depends on area type.

**2. Covariate (main effect of Air Pollution)**

Although not a formal ANCOVA hypothesis, this term tests whether air pollution predicts bird density *on average*, across both groups.

It is **significant** ($p < .001$): bird density falls as pollution rises.

**3. Group (main effect of Area Type)** -- $H_{0,2}: \mu_{\text{Urban, adj}} = \mu_{\text{Rural, adj}}$

Area type is **significant** ($p < .001$), suggesting lower average bird density in urban areas.

However, because the interaction is significant, this group difference changes with air pollution level, so it should be interpreted conditionally, not as a single overall difference.

Reject $H_{0,2}$ with caution.

**Step 3: Examine regression coefficients**

We can also fit the same model using `lm()` to view the **regression coefficients** directly.

This shows how each term contributes to predicting bird density and corresponds to the hypothesis tests we just discussed.

`ancova_lm <- lm(BirdDensity ~ AirPollution * AreaType, data = data)`

```{r}
ancova_lm <- lm(BirdDensity ~ AirPollution * AreaType, data = data)
lm_tidy <- broom::tidy(ancova_lm)

knitr::kable(
lm_tidy[, c("term", "estimate", "std.error", "statistic", "p.value")],
digits = c(0, 3, 3, 2, 4),
caption = "Linear Model Coefficients"
)
```

R automatically converts the categorical `AreaType` variable into a 0/1 indicator behind the scenes. It picks one level (here, "Rural," alphabetically first) as the **reference**, and every coefficient involving `AreaType` is interpreted relative to that reference. That's why you see a coefficient called `AreaTypeUrban` but no separate `AreaTypeRural` term.

These coefficients give the *direction* and *magnitude* of each effect:

-   **(Intercept):** estimated mean bird density for rural sites (the reference group) when air pollution is zero.
-   **AirPollution:** slope of the relationship between pollution and bird density in rural sites.
-   **AreaTypeUrban:** difference in intercept between urban and rural areas.
-   **AirPollution:AreaTypeUrban:** difference in slope between urban and rural areas (the interaction).

**Step 4: Visualize the interaction**

As always, plotting helps confirm what the model tells us.

A scatter plot with fitted regression lines (@fig-11ancovalines) lets us see how air pollution relates to bird density for each area type.

```{r}
#| label: fig-11ancovalines
#| fig-cap: "Bird density against air pollution, with a separate fitted line for each area type. Both lines slope down, but the urban line falls about twice as steeply, which is what the interaction term tests."
#| fig-alt: "Scatter plot of bird density against air pollution, colored by area type, with separate straight fitted lines for rural and urban sites. Both slope downward, and the urban line is noticeably steeper, so the two lines diverge as pollution increases."

ggplot(data, aes(AirPollution, BirdDensity, color = AreaType)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic(base_size=14) +
  labs(title = "Bird Density vs Air Pollution by Area Type",
       x = "Air Pollution", y = "Bird Density")
```

Here, the slopes differ by area type, illustrating the **significant interaction** we detected statistically. Both lines slope downward, so pollution is associated with fewer birds in both settings, but the urban line drops roughly twice as fast. The two lines start close together at low pollution and pull apart as pollution rises, which is exactly what a slope difference looks like.

You'll also notice real scatter around each line, which makes sense: our overall $R^2$ is **0.65**.

Remember, $R^2$ represents the **proportion of variance in bird density explained by the model**.

So about **65% of the variation** in bird density is explained by air pollution, area type, and their interaction.

The remaining 35% reflects other factors, measurement noise, or natural ecological variability not captured by this simple model. That is a strong fit by the standards of field data, and it is strong here partly because this is a synthetic dataset built to illustrate the method. Real environmental data are usually messier.

**Step 5: Centering the covariate**

Interpreting intercepts at $x = 0$ can be misleading when "zero pollution" isn't meaningful.

By centering the covariate (subtracting its mean), we make the intercept represent **average bird density at mean air pollution** instead.

```{r}
data <- data |> mutate(AirPollution_c = AirPollution - mean(AirPollution))
fit_centered <- lm(BirdDensity ~ AirPollution_c * AreaType, data = data)

library(broom)
tidy_centered <- broom::tidy(fit_centered)[, c("term", "estimate", "p.value")]

knitr::kable(
  tidy_centered,
  digits = c(0, 3, 4),
  caption = "Centered ANCOVA Model Coefficients"
)
```

This re-centering doesn't change the slopes or $p$-values. It just shifts the intercepts, making them easier to interpret.

**Interpreting the Centered Model**

Centering does **not** change the slopes or their $p$-values.

Notice that the coefficients for `AirPollution_c` (--0.069) and `AirPollution_c:AreaTypeUrban` (--0.064) are **identical** to the slopes we estimated before centering.

Only the **intercept** and the **AreaTypeUrban** (group difference) term have changed. They are now interpreted at the *mean pollution level* (about 86 units) instead of pollution = 0.

-   **Intercept (35.8):** Predicted bird density for **rural sites** at *average air pollution*.

-   **AreaTypeUrban (--11.4):** Urban sites have, on average, **about 11 fewer birds** than rural sites *at the same mean pollution level*.

-   **AirPollution_c (--0.069):** Slope for rural sites: bird density falls by about 0.07 birds for each additional unit of pollution.

-   **Interaction (--0.064):** Difference in slopes between groups. The urban slope = $-0.069 + (-0.064) \approx$ **--0.13**, roughly twice as steep a decline as in rural sites.

Note how much the group difference changed when we centered: --5.9 before, --11.4 after. That is not a contradiction. Uncentered, the term compares groups at pollution = 0, a value we never actually observed. Centered, it compares them at average pollution, which is a real place in the data. When slopes differ, "the group difference" is not one number; it depends on where along the covariate you look.

**Step 6: Summarize and Interpret**

The significant interaction shows that **air pollution's effect on bird density depends on area type**: bird density declines with pollution in both settings, but the decline is about twice as steep in urban sites.

Because the interaction is significant, we focus on these **different slopes**, not on adjusted mean differences.

If the interaction had **not** been significant, we would instead drop the interaction term and interpret the **parallel slopes model**, comparing adjusted means between groups.

Our results sentences might look something like this:

Results from the ANCOVA indicated that air pollution's effect on bird density depended on area type, *F*(1, 196) = 10.91, *p* = .001. Bird density declined with increasing pollution in both settings, but the decline was roughly twice as steep at urban sites (slope --0.13) as at rural ones (slope --0.07). These findings suggest that **urbanization intensifies the pollution--bird density relationship** and may point toward the need for area-specific conservation strategies.

**Key takeaways**

1.  Always **check for interaction first**
2.  If interaction is **significant**, interpret slope differences.
3.  If not, **fit the reduced model** (no interaction) and compare adjusted means.
4.  Center covariates when the intercept needs to represent a meaningful point.

## Chapter Summary

In this chapter, we introduced **ANCOVA** as an extension of ANOVA that includes continuous covariates. We began by framing ANOVA within the linear modeling framework, showing how ANCOVA builds on it by combining categorical and continuous predictors. Using examples such as bird density across urban and rural environments, we developed and tested null and alternative hypotheses step by step. The main advantage of ANCOVA is its ability to adjust for continuous variables, allowing clearer interpretation of group effects.
