---
author: 
  - Mallory L. Barnes
aliases: [redirects/ancova.html]
---

```{r, include = FALSE}
source("global_stuff.R")
knitr::opts_chunk$set(cache=FALSE)
```

# Analysis of Covariance (ANCOVA)

```{r}
# Libraries
library(tidyverse)
library(broom)

set.seed(538)

fmt_p <- function(p) ifelse(p < .001, "< 0.001", formatC(p, digits = 3, format = "f"))
# For inline math: gives "< 0.001" or "= 0.035", so write $p `r fmt_p_eq(x)`$
fmt_p_eq <- function(p) ifelse(p < .001, "< 0.001", paste("=", formatC(p, digits = 3, format = "f")))

# ANOVA table in the book's style: R's row labels, blank F and p on the Residuals row
anova_kable <- function(fit, caption = NULL) {
  broom::tidy(fit) |>
    mutate(
      sumsq     = round(sumsq, 2),
      meansq    = round(meansq, 2),
      statistic = ifelse(is.na(statistic), "", formatC(statistic, digits = 2, format = "f")),
      p.value   = ifelse(is.na(p.value), "", fmt_p(p.value))
    ) |>
    knitr::kable(col.names = c("Source", "df", "SS", "MS", "F", "p"),
                 align = "lrrrrr", caption = caption)
}
```

We've now covered both regression and ANOVA. This chapter merges them:

$$\text{Regression} + \text{ANOVA} = \text{ANCOVA}$$

The last chapter compared macroinvertebrate species count across two discrete Pollution categories, present or absent, and two Flow categories, low or high. But suppose you measured each channel's pollution concentration instead of labeling it present or absent.

-   Does Flow's effect on species count still depend on pollution once pollution is treated as a continuous measurement instead of two groups? *(covariate: pollution concentration; factor: flow rate)*

-   Does exposure to PFAS correlate with lifetime cancer incidence uniformly across low- and high-income American populations? *(covariate: PFAS exposure; factor: income level)*

Each question pairs a continuous relationship, a continuous variable's effect on the outcome, with a categorical grouping. That pairing, one continuous variable and one categorical factor, is what makes a design an ANCOVA rather than a plain regression or a plain ANOVA.

The first question points to a limitation of the last chapter. "Pollution: present or absent" simplifies a continuous measurement, a concentration in parts per billion. Sorting continuous measurements into "high" and "low" buckets throws away the information a covariate needs, which is how much more polluted a channel is, not just whether it is polluted. Whenever you have the measured values, ANCOVA (or, if there's no categorical factor left at all, plain regression) is almost always the better tool than forcing them into groups first.

ANCOVA is a linear model built directly on regression's own equation. A simple linear model is

$$y = \beta_0 + \beta_1 x + \varepsilon$$

Here $\beta_0$ is the intercept, $\beta_1$ is the slope, and $\varepsilon$ is the error. The model describes a single line.

ANCOVA adds a categorical factor to that equation, and the chapter asks whether every group follows that same line or each group needs its own slope and intercept.

We answer it in order: first check that the model explains anything at all, then test whether the groups have different slopes, and only if the slopes are parallel, test whether the groups differ in adjusted means.

## Why adjust for anything?

The cleanest illustration is a result called **Simpson's paradox**, where a pattern in combined data reverses once you split the data into meaningful groups.

During the COVID-19 pandemic, early data showed that Italy's overall case fatality rate was higher than China's. But once researchers *disaggregated by age group,* the trend flipped. In every age group except the youngest, where neither country recorded a death, the fatality rate was higher in China. The apparent contradiction was explained by differences in the age distribution of cases [@vonkugelgen2021].

The numbers below are the ones that study compared, China's cases and deaths by age as reported on February 17, 2020, and Italy's as reported on March 9, 2020. The case fatality rate (CFR) is the percentage of confirmed cases that ended in death.

```{r}
#| label: fig-11simpson-overall
#| fig-cap: "Crude case fatality rates. Italy appears to have a higher fatality rate than China when all cases are combined."
#| fig-alt: "Bar chart of the crude case fatality rate with all ages combined. Italy's bar, labeled 4.4 percent, is nearly twice as tall as China's, labeled 2.3 percent."
age_levels <- c("0–9","10–19","20–29","30–39","40–49","50–59","60–69","70–79","80+")

dat <- tibble::tribble(
  ~Country, ~Age,    ~Cases, ~Deaths,
  "Italy",  "0–9",       43,       0,
  "Italy",  "10–19",     85,       0,
  "Italy",  "20–29",    296,       0,
  "Italy",  "30–39",    470,       0,
  "Italy",  "40–49",    891,       1,
  "Italy",  "50–59",   1453,       3,
  "Italy",  "60–69",   1471,      37,
  "Italy",  "70–79",   1785,     114,
  "Italy",  "80+",     1532,     202,

  "China",  "0–9",      416,       0,
  "China",  "10–19",    549,       1,
  "China",  "20–29",   3619,       7,
  "China",  "30–39",   7600,      18,
  "China",  "40–49",   8571,      38,
  "China",  "50–59",  10008,     130,
  "China",  "60–69",   8583,     309,
  "China",  "70–79",   3918,     312,
  "China",  "80+",     1408,     208
) %>%
  mutate(
    Age = factor(Age, levels = age_levels),
    CFR_band_pct = round(100 * Deaths / Cases, 1)
  )

fig1_df <- dat %>%
  group_by(Country) %>%
  summarise(
    Cases = sum(Cases),
    Deaths = sum(Deaths),
    CFR_crude_pct = 100 * Deaths / Cases
  )

ggplot(fig1_df, aes(x = Country, y = CFR_crude_pct, fill = Country)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", CFR_crude_pct)),
            vjust = -0.4, size = 4.2) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(y = "Fatality rate (%)", x = NULL) +
  theme(legend.position = "none")
```

At face value, Italy looks worse. Older people are much more likely to die from COVID than younger people. In these numbers, `r dat$CFR_band_pct[dat$Country == "Italy" & dat$Age == "80+"]`% of Italy's cases aged 80 and over died, against none of the `r dat$Cases[dat$Country == "Italy" & dat$Age == "0–9"]` cases under 10. So we split cases into 10-year age bands. Within every band where anyone died, China's fatality rate is higher.

```{r}
#| label: fig-11simpson-by-age
#| fig-cap: "Fatality rates by age group. In every age band where anyone died, China's CFR exceeds Italy's."
#| fig-alt: "Grouped bar chart of fatality rate by 10-year age band for China and Italy. Rates climb steeply with age in both countries, from zero in the youngest band to 14.8 percent for China and 13.2 percent for Italy at 80 and over, and China's bar is taller than Italy's in every band from 10 to 19 upward."

fig2_df <- dat %>%
  select(Country, Age, CFR_band_pct)

ggplot(fig2_df, aes(x = Age, y = CFR_band_pct, fill = Country)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  labs(x = "Age band", y = "Fatality rate (%)", fill = NULL) +
  theme(legend.position = "top")
```

```{r}
#| label: tbl-11simpsonrates
#| tbl-cap: "Fatality rate (%) by age band. China's rate is higher in every band except 0–9, where neither country recorded a death. Several bands under 50 are too small to see in the figure."

rate_tbl <- tidyr::pivot_wider(dat[, c("Age", "Country", "CFR_band_pct")],
                               names_from = Country, values_from = CFR_band_pct)
knitr::kable(rate_tbl[, c("Age", "Italy", "China")], col.names = c("Age band", "Italy", "China"))
```

@tbl-11simpsonrates lists the rates themselves. The reversal happens because Italy's cases skew older while China's skew younger. Italy therefore had a greater proportion of older people getting sick, which pulls Italy's overall average up. Similarly, China's average is pulled down because a greater proportion of younger people are getting sick. Aggregating without adjusting for age hid the within-group pattern entirely. Age is what Chapter 2 called a confounding variable, related both to which country a case came from and to whether the case was fatal.

```{r}
#| label: fig-11age-distributions
#| fig-cap: "Case age distribution by country. Italy’s cases are older on average; China’s are younger."
#| fig-alt: "Two bar charts side by side showing each country's share of cases in each 10-year age band. China's cases are concentrated between 30 and 69, peaking at 50 to 59, while Italy's are concentrated from 50 upward, peaking at 70 to 79, with almost one in five cases at 80 and over."

fig3_df <- dat %>%
  group_by(Country, Age) %>%
  summarise(Cases = sum(Cases), .groups = "drop") %>%
  group_by(Country) %>%
  mutate(Prop = Cases / sum(Cases))


ggplot(fig3_df, aes(x = Age, y = Prop, fill = Country, group=Country)) +
  geom_col(width = 0.7) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~Country)+
  labs(x = "Age band", y = "Share of cases", fill = NULL) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

This kind of reversal comes from a difference between the groups in the *composition* of a third variable, and the combined number averages over that difference. Once you condition on the third variable, the within-group pattern shows up again.

ANCOVA handles this situation when the third variable is continuous, instead of grouped into categories like the age bands above. It compares groups while holding a covariate fixed, so a difference between the groups on that covariate does not distort the comparison.

## Covariates and fixed factors

**Covariate:** a continuous variable that changes alongside the outcome and helps explain part of its variation. In environmental studies, common covariates include temperature, rainfall, soil moisture, or pollution concentration, which may influence the outcome but aren't the main treatment of interest. Including one lets us adjust for its influence, giving a clearer test of the categorical effect once that continuous background variation is accounted for.

The categorical predictor itself, in every example this chapter uses, is a **fixed factor**, with specific levels the researcher chose (urban vs. rural, shaded vs. open canopy) rather than a random sample of possible groups. That distinction matters more once you're building mixed-effects models, which mix fixed and random factors and are beyond where this course goes. They are a different idea from Chapter 10's mixed design, which mixes between-subjects and within-subjects factors. A random factor, by contrast, would be something like the counties your streams are in. You aren't interested in those ten counties for their own sake. They represent a larger population of counties, and you want your conclusions to apply beyond the ones you sampled.

## What ANCOVA is

ANCOVA asks whether a *relationship* holds up across groups. That framing fits environmental work particularly well, because you usually can't experimentally strip out a covariate like temperature, rainfall, or season anyway. Often you wouldn't want to. It may be the reason you designed the study in the first place. In this chapter's main example, the covariate is what we care about and the interaction is the result we're testing for.

::: {.callout-note title="Note: ANCOVA with a nuisance covariate"}
Elsewhere the covariate is a **nuisance** instead, something you control for so a group comparison is fair. In that version, you test the interaction hoping it is not significant.

A classic version from psychology compares anxiety scores across two therapy groups after treatment, controlling for pre-treatment scores.

Both versions use the same test and the same math. They differ only in which term you came to interpret, so decide that before you fit anything. The second example at the end of this chapter works through this version.
:::

For ANCOVA, you need:

1)  A continuous dependent variable (the outcome)
2)  At least one continuous independent variable (covariate)
3)  At least one categorical independent variable (a fixed factor, in this course)

### Indicator variables and how they work

Before reading coefficients off an ANCOVA table, it helps to see how one equation can represent two groups at once.

Categorical predictors get into a linear model by being coded as **indicator variables** (you'll also see them called dummy variables). For a factor with $k$ groups you include $k-1$ indicators. The omitted group becomes the **reference**, and every coefficient is interpreted relative to it. With two groups, one indicator does the whole job:

$$G = \begin{cases} 0 & \text{group 1 (the reference group)} \\ 1 & \text{group 2} \end{cases}$$

In R, `lm(y ~ x + Group)` creates these indicators for you automatically.

**Indicator variables in action.** When an indicator variable is turned on (set to 1), it activates the terms multiplied by that variable. Depending on where the indicator appears in the model, that can change the intercept, the slope, or both.

Consider two predictors: a continuous variable $x$ and an indicator $G$ (0 or 1). The numbers below are arbitrary, chosen to make the arithmetic clean. The worked example later in this chapter plugs in fitted coefficients for air pollution and area type. Use this model:

$$y = 2x + 1(x G) - 7G + 3$$

When $G = 0$ (indicator off), only the baseline terms are active:

$$y = 2x + 3 \qquad \text{slope} = 2, \ \text{intercept} = 3$$

When $G = 1$ (indicator on), the new terms activate:

$$y = (2x + 3) + (1x - 7) = 3x - 4 \qquad \text{slope} = 3, \ \text{intercept} = -4$$

Turning on the indicator adds +1 to the slope and −7 to the intercept. The one equation gives two different lines, and the value of the indicator decides which one applies.

Linear models represent different groups this way. The indicator ($G$) and its interaction with $x$ let a single equation shift the intercept and the slope for each group. ANCOVA uses exactly this logic to test, in order, whether the slopes differ between groups (the $x G$ term) and, if they don't, whether the adjusted means differ (the $G$ term).

**Naming the pieces.** Written in the general form, the same equation gives each coefficient a name and a meaning:

$$y = \underbrace{\beta_0}_{\text{reference intercept}} + \underbrace{\beta_1 x}_{\text{reference slope}} + \underbrace{\beta_2 G}_{\text{intercept shift}} + \underbrace{\beta_3 x G}_{\text{slope shift}} + \varepsilon$$

-   $\beta_0$ and $\beta_1$ describe the line for the reference group ($G=0$).
-   $\beta_2$ is how much the intercept moves for the other group.
-   $\beta_3$ is how much the slope changes for the other group. This is the interaction term.

If we only include $\beta_2$ and leave out $\beta_3$, we get a restricted version in which the intercept can shift but the slope cannot, so the two lines are forced to be parallel. That is the **parallel-lines model**, and it is the one we fall back to when the slope difference turns out not to be significant.

::: {.callout-note title="Note: How R names the four coefficients"}
These are the four coefficients R reports for this model, and each one can be checked against a graph of the two group lines. For `lm(y ~ x * Group)`, where `Group` has levels A (the reference) and B, R names the rows `(Intercept)` for $\beta_0$, `x` for $\beta_1$, `GroupB` for $\beta_2$, and `x:GroupB` for $\beta_3$. The worked example below follows the same pattern: `(Intercept)`, `AirPollution`, `AreaTypeUrban`, and `AirPollution:AreaTypeUrban`.
:::

### The steps of an ANCOVA

**Model.** For one covariate ($x$) and a two-level group indicator $G\in\{0,1\}$,

$$y=\beta_0+\beta_1 x+\beta_2 G+\beta_3(x\times G)+\varepsilon$$

@fig-11onevstwo shows the three possibilities this model can produce.

```{r}
#| label: fig-11onevstwo
#| fig-cap: "The question ANCOVA answers. Left: one line describes both groups. Middle: the groups need different intercepts but share a slope, so the lines are parallel. Right: the groups need different slopes, so the lines are not parallel and the gap between them depends on where you look."
#| fig-alt: "Three scatter plots, each with two colored groups of points. In the first, a single black line runs through both groups. In the second, two parallel lines sit at different heights. In the third, two lines with visibly different slopes diverge from left to right."
#| out-width: "100%"
#| fig-asp: 0.45

# Fixed seed: the caption and alt text describe this particular draw
set.seed(11)
n  <- 30
x  <- runif(n * 2, 0, 10)
g  <- rep(c("Group A", "Group B"), each = n)
B  <- g == "Group B"

# Three scenarios, deliberately kept on a similar y range so the shared axis
# doesn't squash the first two panels.
y_same     <- 8 + 0.5 * x + rnorm(n * 2, 0, 0.9)
y_parallel <- 5 + 0.5 * x + ifelse(B, 5, 0) + rnorm(n * 2, 0, 0.9)
y_slopes   <- ifelse(B, 5 + 0.9 * x, 12 - 0.2 * x) + rnorm(n * 2, 0, 0.9)

lines_df <- rbind(
  data.frame(x, g, y = y_same,     panel = "One shared line"),
  data.frame(x, g, y = y_parallel, panel = "Parallel lines\n(same slope, different intercept)"),
  data.frame(x, g, y = y_slopes,   panel = "Different slopes\n(an interaction)")
)
lines_df$panel <- factor(lines_df$panel, levels = unique(lines_df$panel))

one_panel <- levels(lines_df$panel)[1]

ggplot(lines_df, aes(x, y, color = g)) +
  geom_point(alpha = 0.55, size = 1.5) +
  # first panel: a single line fit to both groups together
  geom_smooth(data = subset(lines_df, panel == one_panel), aes(group = 1),
              method = "lm", se = FALSE, linewidth = 1, colour = "black") +
  # other panels: one line per group
  geom_smooth(data = subset(lines_df, panel != one_panel),
              method = "lm", se = FALSE, linewidth = 1) +
  facet_wrap(~panel) +
  scale_x_continuous(breaks = c(0, 5, 10)) +
  labs(x = "Covariate", y = "Outcome", color = NULL) +
  theme(legend.position = "top", strip.text = element_text(size = 9),
        panel.spacing = unit(1.5, "lines"))
```

ANCOVA works right to left across that figure. It first asks whether the rightmost panel is what we have (do the slopes differ?), and only if the answer is no does it ask whether we have the middle panel or the leftmost (do the intercepts differ?).

**Step 0: Does the model explain anything?** Before looking at any single term, check the whole-model $F$-test, which asks whether the covariate, the group, and their interaction together explain more of $y$ than chance would: $H_0:\beta_1=\beta_2=\beta_3=0$. R reports it on the last line of `summary(lm(...))`. If it is not significant, stop. The data give no evidence that the covariate, the group, or their interaction explains $y$, so no single term is worth interpreting.

**Step 1: Are slopes equal across groups?** This tests whether the covariate effect is the same in each group.

-   Null: $H_{0,\text{int}}:\beta_3=0$ (parallel lines, same slope)
-   Alternative: $H_{a,\text{int}}:\beta_3\neq 0$ (non-parallel lines, different slopes)

**If Step 1 rejects.** The slopes differ, so stop testing the group term. There is no single group difference to test, because the gap between the lines changes along the covariate. Describe each group's line instead. The model gives both slopes (the reference group's is $\beta_1$, the other group's is $\beta_1+\beta_3$), but it does not test whether each one differs from zero. A significant $\beta_3$ says the slopes differ from each other, not that each one differs from zero. To test each slope, fit a separate simple regression within each group.

**Step 2: Are adjusted means equal?** *(only if Step 1 failed to reject parallel slopes)* Drop the interaction and refit the parallel-lines model,

$$y=\beta_0+\beta_1 x+\beta_2 G+\varepsilon$$

-   Null: $H_{0,\text{group}}:\beta_2=0$
-   Alternative: $H_{a,\text{group}}:\beta_2\neq 0$

With parallel lines, $\beta_2$ is the vertical gap between the two lines, and it is the same at every value of $x$. That gap is the **adjusted difference**, how much the groups differ when they are compared at the same covariate value. Evaluating each group's line at the overall mean of $x$ gives the **adjusted means**, the group means you would expect if both groups had the same average covariate. The second example at the end of this chapter works through one.

In the parallel-lines model, $\beta_1$ is the common slope and $\beta_2$ is the adjusted difference between groups.

**Intercepts and centering.** Intercepts are evaluated at $x=0$. In the model with an interaction, the lines are not parallel, so $\beta_2$ compares the groups at $x=0$ only, which may be far outside the data. Centering the covariate, $\tilde x=x-\bar x$, moves that comparison to the average covariate value. In the parallel-lines model the gap is the same everywhere, so centering changes the intercept but not $\beta_2$.

**What to report.**

-   If $x\times G$ is significant: "The effect of $x$ differs by group," then each group's line with its own slope test, and a figure with separate fitted lines.
-   If $x\times G$ is not significant: the common slope $\beta_1$ and the adjusted group difference $\beta_2$ (or the adjusted means), with a figure showing parallel lines.

As always, statistical significance doesn't guarantee ecological or practical importance. Interpret a rejected null in light of the study's design, measurement precision, and the actual size of the effect, not just its $p$-value.

::: {.callout-caution title="Warning: Three ways to misread an ANCOVA"}
-   Interpreting the group term (the intercept shift) when the interaction is significant.
-   Comparing raw intercepts when $x$ is not centered.
-   Treating adjusted means as meaningful when slopes are not parallel.
:::

### Checking ANCOVA's assumptions

Most of these are ones ANOVA already needed. Three are new because a covariate has entered the model (@tbl-11assumptions).

| Assumption | New with ANCOVA? | How to check |
|:-----------|:-----------------|:-------------|
| Independent observations | No (universal, Chapter 6) | The study design |
| Normal residuals and similar variances across groups | No | Q-Q plot, residuals-versus-fitted plot, Levene's test |
| Linearity within each group | Yes | A scatter plot with a fitted line for each group |
| Homogeneity of slopes | Yes | The interaction test (Step 1), not a check run beforehand |
| The covariate is measured with little error and is not affected by the treatment | Yes | The study design |

: Assumptions of ANCOVA and how to check them. {#tbl-11assumptions}

If the covariate is measured with error, ANCOVA only partly adjusts for group differences in the covariate, and the leftover difference can look like a treatment effect.

## Full ANCOVA example

### Birds, air pollution and area type

Suppose we are studying the effect of air pollution on bird species density, and we want to know whether that relationship differs between urban and rural areas. @fig-11design lays out the design.

```{r}
#| label: fig-11design
#| fig-cap: "An ANCOVA design: one continuous predictor and one categorical factor. Compare it with the factorial design in Chapter 10, which has the same shape with a second set of levels in place of the gradient."
#| fig-alt: "Study design schematic. Two predictors: area type with levels Rural and Urban, crossed with air pollution as a continuous gradient from low to high, across 200 observations. DV: bird species density."
#| fig-asp: 0.36
#| out-width: "100%"

design_mixed("Air pollution (continuous) × Area type (categorical)",
             "low", "high", "200 observations", c("Rural", "Urban"),
             "Bird species density",
             "Does the pollution–density relationship differ between rural and urban?",
             note = "full pollution range measured in BOTH groups")
```

This design has the same structure as the factorial design in Chapter 10. Both cross two predictors and ask whether one changes the effect of the other. The only difference is that factorial's second predictor has levels, while ANCOVA's is a continuous range. Both groups span the full pollution range, so rural sites are not simply the clean ones. An adjusted comparison compares the groups at the same pollution level, and that only works where both groups have sites at that level. Without the overlap, each group's line would have to be extended past its own data, the extrapolation problem from Chapter 9.

We might expect pollution to have a stronger negative effect in urban areas. Urban habitat is already fragmented and offers birds fewer clean patches to retreat to, while rural areas have more intact habitat nearby.

We'll fit an ANCOVA model to test whether air pollution's relationship with bird density differs between urban and rural sites. Area type enters as an indicator variable, and air pollution enters as the covariate.

### Null and alternative hypotheses in our example ANCOVA

First, we state the hypotheses for each step:

0.  **Whole model.** Do air pollution, area type, and their interaction together explain bird density?

    $H_0: \beta_1=\beta_2=\beta_3=0$

1.  **Slopes (interaction).** Does air pollution affect bird density the same way at urban and rural sites?

    $H_{0,\text{int}}: \beta_3 = 0$ (parallel lines)

    $H_{a,\text{int}}: \beta_3 \neq 0$

    A significant interaction means the slopes differ, and air pollution's relationship with bird density depends on area type.

2.  **Adjusted means (only if the slopes are parallel).** At the same level of air pollution, do urban and rural sites differ in bird density?

    $H_{0,\text{group}}: \beta_2 = 0$

    $H_{a,\text{group}}: \beta_2 \neq 0$

### The data

The data in this example are invented. There are 200 sites, each assigned at random to urban or rural, with air pollution between 25 and 150. The true slope is $-0.05$ birds per unit of pollution at rural sites and $-0.15$ at urban sites, plus random scatter, so the fitted slopes below come out near these values but not on them.

```{r}
# Fixed seed: the invented bird data, and every description of them below, depend on this draw
set.seed(123)

n <- 200                                                     # number of sites
area_type <- sample(c("Urban", "Rural"), n, replace = TRUE)  # randomly assign area type
air_pollution <- round(runif(n, min = 25, max = 150), 1)     # continuous covariate
urban <- ifelse(area_type == "Urban", 1, 0)                  # 0/1 indicator

# Bird density declines with pollution everywhere, but more steeply in urban
# sites. That difference in slopes is the interaction we want the model to find.
bird_density <- 40 -
  0.05 * air_pollution -                # rural slope
  3    * urban -                        # urban intercept offset
  0.10 * air_pollution * urban +        # extra urban steepness (the interaction)
  rnorm(n, mean = 0, sd = 5)            # noise

bird_density <- round(pmax(bird_density, 0), 0)

data <- data.frame(AreaType = area_type, AirPollution = air_pollution, BirdDensity = bird_density)
```

Before fitting the model, look at the data. @fig-11birddensity shows the outcome, bird density, by area type, which gives a quick sense of whether urban and rural areas differ on average.

```{r}
#| label: fig-11birddensity
#| fig-cap: "Bird density by area type. Rural sites sit higher than urban ones."
#| fig-alt: "Boxplot of bird density for rural and urban area types. The rural box is centered higher than the urban box, with overlapping ranges."

ggplot(data, aes(AreaType, BirdDensity)) +
  geom_boxplot(aes(fill = AreaType), show.legend = FALSE) +
  labs(x = "Area type", y = "Bird density")
```

Bird density is higher at rural sites, with a median of `r median(data$BirdDensity[data$AreaType == "Rural"])` birds against `r median(data$BirdDensity[data$AreaType == "Urban"])` at urban sites, consistent with what we might expect ecologically.

Next, check the covariate, air pollution. If the covariate differs sharply between groups, ANCOVA adjusts for that difference. If it doesn't, group comparisons will look more like a simple ANOVA.

```{r}
#| label: fig-11airpollution
#| fig-cap: "Air pollution by area type. The two groups look broadly similar, so there is no large covariate imbalance for the model to adjust away."
#| fig-alt: "Boxplot of air pollution for rural and urban area types. The two boxes overlap substantially with similar centers."

ggplot(data, aes(AreaType, AirPollution)) +
  geom_boxplot(aes(fill = AreaType), show.legend = FALSE) +
  labs(x = "Area type", y = "Air pollution")
```

Air pollution is similar at urban and rural sites. The medians are `r median(data$AirPollution[data$AreaType == "Rural"])` at rural sites and `r median(data$AirPollution[data$AreaType == "Urban"])` at urban sites, a small gap compared with the range of about 25 to 150. This plot doesn't test anything, but it tells us what to expect. Because pollution is similar in the two groups, a group difference in bird density is unlikely to come from a difference in pollution.

### Results of the ANCOVA model

**Fit the model.** This is the same `aov()` call we used for one-way and factorial ANOVA. Because `AirPollution` is numeric instead of a factor, `aov()` treats it as a continuous covariate, so R needs no separate ANCOVA function.

```{r}
#| echo: true
#| eval: false
ancova <- aov(BirdDensity ~ AirPollution * AreaType, data = data)
```

```{r}
ancova <- aov(BirdDensity ~ AirPollution * AreaType, data = data)
anova_kable(ancova, caption = "ANCOVA table for bird density")
```

```{r}
ancova_lm <- lm(BirdDensity ~ AirPollution * AreaType, data = data)
fstat <- summary(ancova_lm)$fstatistic
whole_p <- pf(fstat[1], fstat[2], fstat[3], lower.tail = FALSE)
int_row <- broom::tidy(ancova) |> filter(term == "AirPollution:AreaType")
```

**Step 0: the whole model.** `summary(lm())` reports $F(`r fstat[2]`, `r fstat[3]`) = `r sprintf("%.2f", fstat[1])`$, $p `r fmt_p_eq(whole_p)`$, with $R^2 = `r sprintf("%.2f", summary(ancova_lm)$r.squared)`$. The model explains bird density, so we continue.

**Step 1: the interaction.** The `AirPollution:AreaType` row tests $H_{0,\text{int}}$: $F(1, `r fstat[3]`) = `r sprintf("%.2f", int_row$statistic)`$, $p `r fmt_p_eq(int_row$p.value)`$. We reject $H_{0,\text{int}}$, so the slopes are not parallel. Air pollution's relationship with bird density depends on area type, so we stop here and do not test adjusted means.

**What the other two rows are.** The table still has rows for `AirPollution` and `AreaType`, both with tiny $p$-values, so it is tempting to read them, but they don't answer this chapter's question. R builds these rows in sequence. `AirPollution` tests one average slope for all 200 sites, ignoring area type, and `AreaType` tests a group difference after that single slope is fitted. Once the interaction says the slopes differ, neither row describes the data. There is no single slope, and the group difference depends on where along the pollution gradient you look. Because the rows are built in sequence and the groups are unequal in size (`r sum(data$AreaType == "Rural")` rural, `r sum(data$AreaType == "Urban")` urban), listing the terms in a different order would also change their sums of squares.

**Reading the coefficients.**

Fitting the same model with `lm()` shows the regression coefficients directly, so we can see how each term contributes to predicting bird density.

```{r}
#| echo: true
#| eval: false
ancova_lm <- lm(BirdDensity ~ AirPollution * AreaType, data = data)
```

```{r}
lm_tidy <- broom::tidy(ancova_lm)

knitr::kable(
  lm_tidy |>
    transmute(term, estimate = round(estimate, 3), std.error = round(std.error, 3),
              statistic = round(statistic, 2), p.value = fmt_p(p.value)),
  col.names = c("Term", "Estimate", "SE", "t", "p"),
  caption = "Linear model coefficients"
)
```

R automatically converts the categorical `AreaType` variable into a 0/1 indicator, the same $G$ from the indicator-variable section earlier in this chapter. It picks one level (here, "Rural," alphabetically first) as the reference, and every coefficient involving `AreaType` is interpreted relative to that reference. That's why you see a coefficient called `AreaTypeUrban` but no separate `AreaTypeRural` term.

These coefficients give the *direction* and *magnitude* of each effect:

-   **`(Intercept)`.** Estimated mean bird density for rural sites (the reference group) when air pollution is zero.
-   **`AirPollution`.** Slope of the relationship between pollution and bird density in rural sites. This is not the same quantity as the `AirPollution` row of the ANOVA table above, which was one average slope for both groups.
-   **`AreaTypeUrban`.** Difference in intercept between urban and rural areas.
-   **`AirPollution:AreaTypeUrban`.** Difference in slope between urban and rural areas (the interaction).

**Visualize the interaction.**

A scatter plot with fitted regression lines (@fig-11ancovalines) lets us see how air pollution relates to bird density for each area type.

```{r}
#| label: fig-11ancovalines
#| fig-cap: "Bird density against air pollution, with a separate fitted line for each area type. Both lines slope down, but the urban line falls about twice as steeply, which is what the interaction term tests."
#| fig-alt: "Scatter plot of bird density against air pollution, colored by area type, with separate straight fitted lines for rural and urban sites. Both slope downward, and the urban line is noticeably steeper, so the two lines diverge as pollution increases."

ggplot(data, aes(AirPollution, BirdDensity, color = AreaType)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Air pollution", y = "Bird density", color = "Area type")
```

Here, the slopes differ by area type, illustrating the significant interaction we detected statistically. Both lines slope downward, so pollution is associated with fewer birds in both settings, but the urban line drops roughly twice as fast. The lines are about `r sprintf("%.1f", abs(coef(ancova_lm)[3] + coef(ancova_lm)[4]*min(data$AirPollution)))` birds apart at the lowest pollution levels and about `r sprintf("%.1f", abs(coef(ancova_lm)[3] + coef(ancova_lm)[4]*max(data$AirPollution)))` apart at the highest.

**Testing each group's slope.** The interaction told us the two slopes differ from each other. It did not tell us whether each one differs from zero, so we fit a separate simple regression within each group.

```{r}
#| echo: true
#| eval: false
lm(BirdDensity ~ AirPollution, data = subset(data, AreaType == "Rural"))
lm(BirdDensity ~ AirPollution, data = subset(data, AreaType == "Urban"))
```

```{r}
#| label: tbl-11slopes
#| tbl-cap: "Separate regressions of bird density on air pollution within each area type."

slopes_raw <- data |>
  group_by(AreaType) |>
  group_modify(~ {
    fit <- lm(BirdDensity ~ AirPollution, data = .x)
    broom::tidy(fit) |> mutate(df = fit$df.residual)
  }) |>
  ungroup() |>
  filter(term == "AirPollution")
sl_rur <- slopes_raw[slopes_raw$AreaType == "Rural", ]
sl_urb <- slopes_raw[slopes_raw$AreaType == "Urban", ]

slopes_tbl <- slopes_raw |>
  transmute(`Area type` = AreaType, Slope = round(estimate, 3), SE = round(std.error, 3),
            t = round(statistic, 2), df, p = fmt_p(p.value))
knitr::kable(slopes_tbl)
```

Both slopes are significant. Bird density falls with pollution at rural sites and falls about twice as steeply at urban sites. The slope estimates match the ANCOVA's, $`r sprintf('%.3f', coef(ancova_lm)[2])`$ for rural and $`r sprintf('%.3f', coef(ancova_lm)[2])` + (`r sprintf('%.3f', coef(ancova_lm)[4])`) \approx `r sprintf('%.3f', coef(ancova_lm)[2] + coef(ancova_lm)[4])`$ for urban, because a model with an interaction fits each group its own line. Only the tests are new, and their standard errors differ slightly because each regression estimates its own error variance.

The points scatter around each line. The model's $R^2$ is `r sprintf("%.2f", summary(ancova_lm)$r.squared)`, so air pollution, area type, and their interaction explain about `r round(100*summary(ancova_lm)$r.squared)`% of the variation in bird density. The remaining `r 100 - round(100*summary(ancova_lm)$r.squared)`% reflects other factors, measurement noise, or natural ecological variability not captured by this simple model. That is a strong fit by the standards of field data, and it is strong here partly because this is a synthetic dataset built to illustrate the method. Field data from actual sites are usually messier.

**Centering the covariate.**

Interpreting intercepts at $x = 0$ can be misleading when "zero pollution" isn't meaningful.

By centering the covariate (subtracting its mean), we make the intercept represent average bird density at mean air pollution instead.

```{r}
data <- data |> mutate(AirPollution_c = AirPollution - mean(AirPollution))
fit_centered <- lm(BirdDensity ~ AirPollution_c * AreaType, data = data)

tidy_centered <- broom::tidy(fit_centered) |>
  transmute(term, estimate = round(estimate, 3), p.value = fmt_p(p.value))
cc <- coef(fit_centered)

knitr::kable(
  tidy_centered,
  col.names = c("Term", "Estimate", "p"),
  caption = "Coefficients with the covariate centered"
)
```

Centering doesn't change the slopes, the interaction, or their $p$-values. It moves the intercept and the group term, and the group term's test now compares the groups at average pollution instead of at zero, so its $p$-value changes too.

**Interpreting the centered model.** The intercept and the `AreaTypeUrban` term are now interpreted at the *mean pollution level*, about `r round(mean(data$AirPollution))` units, instead of pollution = 0.

-   **Intercept (`r sprintf("%.1f", cc["(Intercept)"])`).** Predicted bird density for rural sites at average air pollution.

-   **`AreaTypeUrban` ($`r sprintf("%.1f", cc["AreaTypeUrban"])`$).** Urban sites have, on average, about `r round(abs(cc["AreaTypeUrban"]))` fewer birds than rural sites at the same mean pollution level.

-   **`AirPollution_c` ($`r sprintf("%.3f", cc["AirPollution_c"])`$).** Slope for rural sites: bird density falls by about `r sprintf("%.2f", abs(cc["AirPollution_c"]))` birds for each additional unit of pollution.

-   **Interaction ($`r sprintf("%.3f", cc["AirPollution_c:AreaTypeUrban"])`$).** Difference in slopes between groups. The urban slope = $`r sprintf("%.3f", cc["AirPollution_c"])` + (`r sprintf("%.3f", cc["AirPollution_c:AreaTypeUrban"])`) \approx `r sprintf("%.2f", cc["AirPollution_c"] + cc["AirPollution_c:AreaTypeUrban"])`$, roughly twice as steep a decline as in rural sites.

The group difference changed when we centered, from $`r sprintf("%.1f", coef(ancova_lm)["AreaTypeUrban"])`$ to $`r sprintf("%.1f", cc["AreaTypeUrban"])`$. Uncentered, the term compares groups at pollution = 0, a value outside the observed data. Centered, it compares them at average pollution, a value inside the observed range. When slopes differ, "the group difference" is not one number. It depends on where along the covariate you look.

**Summarize and interpret.**

The significant interaction shows that air pollution's effect on bird density depends on area type. Bird density declines with pollution in both settings, but the decline is about twice as steep in urban sites. Because the interaction is significant, we focus on these different slopes, not on adjusted mean differences.

If the interaction had *not* been significant, we would instead drop the interaction term and interpret the parallel-lines model, comparing adjusted means between groups. The next example shows that case.

::: {.callout-important title="Results reporting: ANCOVA"}
> Results from the ANCOVA indicated that air pollution's effect on bird density depended on area type, $F(1, `r fstat[3]`) = `r sprintf("%.2f", int_row$statistic)`$, $p `r fmt_p_eq(int_row$p.value)`$, model $R^2 = `r sprintf("%.2f", summary(ancova_lm)$r.squared)`$. Separate regressions showed that bird density declined with increasing pollution at both rural sites, $b_1 = `r sprintf("%.2f", sl_rur$estimate)`$ birds per unit of pollution, $t(`r sl_rur$df`) = `r sprintf("%.2f", sl_rur$statistic)`$, $p `r fmt_p_eq(sl_rur$p.value)`$, and urban sites, $b_1 = `r sprintf("%.2f", sl_urb$estimate)`$, $t(`r sl_urb$df`) = `r sprintf("%.2f", sl_urb$statistic)`$, $p `r fmt_p_eq(sl_urb$p.value)`$, with the urban decline roughly twice as steep. These findings suggest that urbanization intensifies the pollution-bird density relationship and may point toward the need for area-specific conservation strategies.
:::

## Second example: when the covariate is a nuisance

The bird example used ANCOVA the way this chapter leads with it. The covariate relationship was the question, and the interaction was the finding. The other way you'll see ANCOVA reverses those roles. The group comparison is the question, and the covariate is something to adjust for so that the comparison is fair. This is also the continuous version of the Simpson's paradox problem the chapter opened with.

### Retrofits and electricity use

A city facilities office wants to know whether energy retrofits (new insulation, lighting, and controls) reduced electricity use. It compares 30 retrofitted buildings with 30 standard ones. Electricity use depends heavily on how many people are in a building, and the retrofitted buildings happen to be the busier ones. The covariate is occupancy, and the question is about the group term. These buildings are invented. In the invented data, a retrofit cuts electricity use by 40 kWh per day at any occupancy, both kinds of building use 2 more kWh per day for each extra occupant, and retrofitted buildings hold 60 to 120 people against 30 to 90 for standard ones.

Occupancy also passes the last check on the assumptions list. A retrofit changes a building's equipment, not how many people use it, so adjusting for occupancy doesn't remove any of the retrofit's own effect.

```{r}
# Fixed seed: the invented building data and the figure's alt text depend on this draw
set.seed(2)
n_b <- 30
occ_standard    <- runif(n_b, 30, 90)
occ_retrofitted <- runif(n_b, 60, 120)   # retrofitted buildings are busier

energy <- data.frame(
  Type      = factor(rep(c("Standard", "Retrofitted"), each = n_b),
                     levels = c("Standard", "Retrofitted")),
  Occupancy = round(c(occ_standard, occ_retrofitted))
)
# Same slope in both groups; retrofits lower use by 40 kWh/day at any occupancy
energy$Energy <- round(200 + 2 * energy$Occupancy -
                         40 * (energy$Type == "Retrofitted") +
                         rnorm(2 * n_b, 0, 20))

occ_means <- tapply(energy$Occupancy, energy$Type, mean)
raw_means <- tapply(energy$Energy, energy$Type, mean)
raw_t     <- t.test(Energy ~ Type, data = energy, var.equal = TRUE)
```

### The raw comparison

Average daily occupancy is `r round(occ_means["Retrofitted"], 1)` people in the retrofitted buildings and `r round(occ_means["Standard"], 1)` in the standard ones. @fig-11energyraw compares electricity use without taking that into account. Retrofitted buildings use more, `r sprintf("%.1f", raw_means["Retrofitted"])` vs. `r sprintf("%.1f", raw_means["Standard"])` kWh per day, and a two-sample $t$-test calls that difference significant, $t(58) = `r sprintf("%.2f", abs(raw_t$statistic))`$, $p `r fmt_p_eq(raw_t$p.value)`$. Taken at face value, the retrofits made things worse.

```{r}
#| label: fig-11energyraw
#| fig-cap: "Daily electricity use by building type, ignoring occupancy. Diamonds mark the group means."
#| fig-alt: "Boxplots of daily electricity use for standard and retrofitted buildings. The retrofitted box sits higher, with a mean of about 342 kWh per day against about 318 for standard buildings."
#| out-width: "60%"
#| fig-asp: 0.8

ggplot(energy, aes(Type, Energy, fill = Type)) +
  geom_boxplot(width = 0.5, show.legend = FALSE) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "white") +
  labs(x = NULL, y = "Daily electricity use (kWh)")
```

### The ANCOVA

```{r}
energy_full <- lm(Energy ~ Occupancy * Type, data = energy)
e_f   <- summary(energy_full)$fstatistic
e_fp  <- pf(e_f[1], e_f[2], e_f[3], lower.tail = FALSE)
e_int <- broom::tidy(aov(Energy ~ Occupancy * Type, data = energy)) |>
  filter(term == "Occupancy:Type")

energy_par <- lm(Energy ~ Occupancy + Type, data = energy)
e_par <- broom::tidy(energy_par, conf.int = TRUE)
b1  <- e_par$estimate[e_par$term == "Occupancy"]
b2  <- e_par$estimate[e_par$term == "TypeRetrofitted"]
b2r <- e_par[e_par$term == "TypeRetrofitted", ]

occ_bar  <- mean(energy$Occupancy)
adj_std  <- coef(energy_par)[1] + b1 * occ_bar
adj_retr <- adj_std + b2
```

**Step 0: the whole model.** $F(`r e_f[2]`, `r e_f[3]`) = `r sprintf("%.2f", e_f[1])`$, $p `r fmt_p_eq(e_fp)`$. The model explains electricity use, so we continue.

**Step 1: the interaction.** The `Occupancy:Type` row gives $F(1, `r e_f[3]`) = `r sprintf("%.2f", e_int$statistic)`$, $p `r fmt_p_eq(e_int$p.value)`$. We fail to reject $H_{0,\text{int}}$, so we treat the slopes as parallel. Each additional occupant adds about the same amount of electricity in either kind of building. We drop the interaction and refit.

**Step 2: the adjusted difference.** @tbl-11energypar shows the parallel-lines model.

```{r}
#| echo: true
#| eval: false
lm(Energy ~ Occupancy + Type, data = energy)
```

```{r}
#| label: tbl-11energypar
#| tbl-cap: "Parallel-lines model: electricity use on occupancy and building type."

knitr::kable(
  e_par |>
    transmute(term, estimate = round(estimate, 2), std.error = round(std.error, 2),
              statistic = round(statistic, 2), p.value = fmt_p(p.value)),
  col.names = c("Term", "Estimate", "SE", "t", "p")
)
```

The `Occupancy` coefficient is the common slope, about `r sprintf("%.2f", b1)` kWh per day for each additional occupant. The `TypeRetrofitted` coefficient is $b_2$, the estimate of $\beta_2$, the adjusted difference. At the same occupancy, retrofitted buildings use `r sprintf("%.1f", abs(b2))` kWh per day less than standard ones, $t(57) = `r sprintf("%.2f", b2r$statistic)`$, $p `r fmt_p_eq(b2r$p.value)`$. We reject $H_{0,\text{group}}$.

@fig-11energyadj shows what that coefficient measures. The adjusted means turn it into two numbers. Evaluating each line at the overall mean occupancy, `r round(occ_bar, 1)` people, gives `r sprintf("%.1f", adj_std)` kWh per day for standard buildings and `r sprintf("%.1f", adj_retr)` for retrofitted ones. Their difference is $b_2$.

```{r}
#| label: fig-11energyadj
#| fig-cap: "Electricity use against occupancy, with the parallel lines from the model. Open circles mark the adjusted means, each line's value at the overall mean occupancy (dashed line); the arrow is the adjusted difference, the same at every occupancy."
#| fig-alt: "Scatter plot of daily electricity use against average daily occupancy for standard and retrofitted buildings, with two parallel fitted lines. The retrofitted points sit further right, at higher occupancy, but their line sits about 42 kWh per day below the standard line. A dashed vertical line at the mean occupancy of about 76 people marks two open circles, one on each line, joined by a double-headed arrow."
#| out-width: "90%"
#| fig-asp: 0.6

lines_df <- data.frame(
  Type = factor(c("Standard", "Retrofitted"), levels = levels(energy$Type)),
  b0   = c(coef(energy_par)[1], coef(energy_par)[1] + b2),
  b1   = b1
)
adj_df <- data.frame(Type = lines_df$Type, x = occ_bar, y = c(adj_std, adj_retr))

ggplot(energy, aes(Occupancy, Energy, colour = Type)) +
  geom_point(alpha = 0.6) +
  geom_abline(data = lines_df, aes(intercept = b0, slope = b1, colour = Type), linewidth = 1) +
  geom_vline(xintercept = occ_bar, linetype = "dashed", colour = "grey40") +
  geom_point(data = adj_df, aes(x, y), size = 3.5, shape = 21, fill = "white", stroke = 1.5) +
  annotate("segment", x = occ_bar - 3, xend = occ_bar - 3, y = adj_std, yend = adj_retr,
           arrow = arrow(ends = "both", length = unit(0.08, "in"))) +
  annotate("label", x = occ_bar - 5, y = (adj_std + adj_retr) / 2, hjust = 1, size = 3.3,
           label = sprintf("adjusted difference\n%.1f kWh/day", b2)) +
  labs(x = "Average daily occupancy (people)", y = "Daily electricity use (kWh)", colour = NULL) +
  theme(legend.position = "top")
```

The raw comparison and the adjusted one point in opposite directions, for the same reason as in the COVID example. The groups differed in a third variable. The retrofitted buildings were busier, and busier buildings use more electricity. Compared at the same occupancy, retrofitted buildings use less electricity.

```{r}
mse_raw   <- anova(lm(Energy ~ Type, data = energy))["Residuals", "Mean Sq"]
mse_ancova <- anova(energy_par)["Residuals", "Mean Sq"]
```

Adjusting for occupancy also shrank the error term. In the raw comparison the error mean square is `r round(mse_raw)`, and in the parallel-lines model it is `r round(mse_ancova)`. Much of the building-to-building variation in electricity use comes from occupancy, and once the model accounts for it, that variation no longer counts as error. Repeated-measures ANOVA gains power the same way in Chapter 8. There the known source of variation was the units themselves (the monitoring stations), and here it is a measured covariate. Either way, removing it from $MS_\text{Error}$ makes the denominator of the test smaller, so a real group difference is easier to detect.

::: {.callout-important title="Results reporting: ANCOVA with parallel slopes"}
> After adjusting for occupancy, retrofitted buildings used less electricity than standard buildings, $b_2 = `r sprintf("%.1f", b2)`$ kWh per day, $t(57) = `r sprintf("%.2f", b2r$statistic)`$, $p `r fmt_p_eq(b2r$p.value)`$, 95% CI [`r sprintf("%.1f", b2r$conf.low)`, `r sprintf("%.1f", b2r$conf.high)`] kWh per day. Adjusted means at the average occupancy of `r round(occ_bar, 1)` people were `r sprintf("%.1f", adj_retr)` kWh per day for retrofitted buildings and `r sprintf("%.1f", adj_std)` for standard ones. The occupancy × building type interaction was not significant, $F(1, `r e_f[3]`) = `r sprintf("%.2f", e_int$statistic)`$, $p `r fmt_p_eq(e_int$p.value)`$, so a common slope was used (`r sprintf("%.2f", b1)` kWh per day per occupant).
:::

## Chapter Summary

**Why it matters.** Environmental relationships rarely play out identically everywhere. ANCOVA answers a specific, recurring question, whether a continuous relationship, like pollution's effect on some outcome, holds the same way across groups or changes from one group to the next.

**Core ideas**

-   **ANCOVA is regression plus a categorical factor.** It takes the familiar $y=\beta_0+\beta_1 x+\varepsilon$ line and lets a categorical grouping shift the intercept, the slope, or both, using the same indicator-variable machinery that underlies every linear model in this book.
-   **Check the whole model, then the interaction.** If the whole-model $F$ is not significant, stop. Otherwise test whether the groups' slopes differ ($H_{0,\text{int}}:\beta_3=0$). That answer decides how everything else gets interpreted.
-   **A significant interaction means report the two lines, not one number.** If slopes differ, describe how the relationship changes by group instead of reducing it to a single adjusted-mean comparison, which assumes parallel lines. Test each group's slope with its own regression, since the interaction only says the slopes differ from each other.
-   **If slopes are parallel, compare adjusted means.** Drop the interaction term and test $\beta_2$, the gap between the parallel lines. When the groups differ on the covariate, the adjusted comparison can point the opposite way from the raw one, as it did for the retrofitted buildings.
-   **A covariate also shrinks the error term.** Variation the covariate explains leaves the error mean square, which is the same way repeated-measures ANOVA gains power by removing unit-to-unit variation.
-   **Center the covariate when the intercept should mean something.** Left uncentered, the intercept describes the groups at covariate = 0, a value the data may never actually contain.
