---
author: 
  - Mallory L. Barnes
aliases: [redirects/simulating-data.html]
---

```{r, include = FALSE}
source("global_stuff.R")
knitr::opts_chunk$set(cache=FALSE)
```

# Analysis of Covariance (ANCOVA)

```{r}
# Libraries (top)
library(tidyverse)
library(ggpubr)

# --- Data generation (replace your whole block) ---
set.seed(123)
n <- 200
data <- tibble(
  AreaType     = sample(c("Urban", "Rural"), n, replace = TRUE),
  AirPollution = round(runif(n, min = 25, max = 150), 1)
) |>
  mutate(
    # simple data-generating process: lower baseline in Urban, mild negative slope
    BirdDensity = 10 + 0.2*AirPollution - if_else(AreaType=="Urban", 5, 0) + rnorm(n, 0, 4),
    BirdDensity = pmax(0, round(BirdDensity, 0))
  )

```

ANCOVA uses the same linear model ideas from regression, with one continuous covariate and one categorical factor. We first test whether groups have different **slopes** with respect to the covariate, then, only if slopes are parallel, we test whether groups differ in **adjusted means**.

## Linear Models (LMs)

Linear Models are a class of models that encompass various types of statistical analyses, including ANOVA, and regression. They're flexible enough to handle different types of data and relationships. These assume the response variable is **continuous and normally distributed**. They include *t-tests, ANOVA, regression,* and *ANCOVA* as special cases.

A genera **linear model** is the common framework that includes ANOVA, regression, and ANCOVA. Each uses the same residual-based logic: a continuous response $Y$ is modeled as a combination of one or more predictors plus error. We've seen a few different cases so far:

-   **No Groups and No Relationships (H0):**
    -   This scenario often emerges when our ANOVA or regression analysis yields non-significant results.
    -   **What We Report:** The grand mean and overall variance or standard deviation. Alternatively, we can use a confidence interval to encapsulate this information.
-   **Two or More Categories with Significantly Different Means (t-test, ANOVA):**
    -   Here, we delve into data where group means are distinct and significant.
    -   **What We Report:** Group means. In classical ANOVA, we might report a common variance or standard deviation, or opt for group-specific measures (potentially through confidence intervals), ensuring to note that variances are not significantly different. With Welch's t-test, we focus on group-specific variance or standard deviation.
-   **Two Continuous Variables with a Significant Linear or Monotonic Relationship (Regression):**
    -   This model applies when there's a significant linear relationship between two continuous variables.
    -   **What We Report:** The equation of the regression line, along with its confidence limits. We can also select specific x-values and provide confidence intervals for the predicted y-values at those points.

### Model Statements:

-   **No Groups/Relationships:** $y \sim N(\mu, \sigma^2)$
-   **Categories with Different Means:** $y_{i,j} = \mu + \tau_i + \epsilon$ (t-test allows for unequal variance: $y_{i,j} = \mu + \tau_i + \epsilon_i$)
-   **Linear Relationship:** $y = \beta_0 + \beta_1x + \epsilon$

Linear models handle continuous $y$; for counts or non-Gaussian $y$, we use generalized linear models (not covered here).

### Flexibility and Applicability of LMs

LMs enable us to model more complex relationships beyond the basic categorical X with continuous Y (as in t-tests and ANOVA) or continuous X with continuous Y (as in regression). They can handle diverse data types and model relationships that aren't strictly linear, making them crucial for studying environmental systems.

So far, in our exploration of parametric tests, we have primarily focused on two types of relationships:

-   **A categorical X and a continuous Y:** This includes tests like the t-test and ANOVA.
-   **A continuous X and a continuous Y:** This is typically analyzed using linear regression.

### Understanding ANOVA as a linear model:

ANOVA, or Analysis of Variance, is traditionally viewed as a technique to compare means across multiple groups. However, at its core, ANOVA is a linear model. It's a special case where the predictors are categorical, not continuous. This distinction is important but subtle. 

Example: Imagine a researcher is exploring the effects of metal contamination on the species richness of sessile marine invertebrates. They're particularly interested in the impact of copper and the orientation of the substrate on which these organisms live. To investigate this, they conduct a factorial experiment, measuring species richness across different levels of copper enrichment (None, Low, High) and substrate orientation (Vertical, Horizontal). This setup allows us to not only consider each factor separately but also examine their potential interaction.

The ANOVA framework provides three null hypotheses to test:

1.  There are no differences in species richness across the copper levels.

2.  There are no differences in species richness between substrate orientations.

3.  There is no interaction effect between copper levels and substrate orientation.

These hypotheses can be represented in a linear model as follows:

$y_{ijk} = \beta_0 + \beta_{Copper_i} + \beta_{Orientation_j} + \beta_{Copper \times Orientation_{ij}} + \varepsilon_{ijk}$

Here, $\beta_0$ is the grand mean (intercept), $\beta_{Copper_i}$ and $\beta_{Orientation_j}$ are the main effects of the copper levels and orientation, respectively, and $\beta_{Copper \times Orientation_{ij}}$ represents the interaction between these factors. The $\varepsilon_{ijk}$ term captures the residual variance, or the random deviation of each observation from the model prediction. The significance of each beta coefficient is tested to determine if it makes a meaningful contribution to the model.

When conducting an ANOVA, we're essentially fitting a linear model with categorical predictors. These predictors are represented by beta coefficients in our model, which show the unique contribution of each factor (like copper levels or substrate orientation) to the outcome variable (such as species richness). The critical question we ask is whether each beta coefficient is significant---does it make a meaningful difference to our model?

We test each term with an $F$-test. The F-statistic is a ratio that compares the amount of variance explained by a particular factor to the variance not explained by the model (within-group variance or error). A higher F-statistic indicates that the factor explains a significant portion of the variability in the outcome variable, while a lower F-statistic suggests that the factor does not contribute much to our understanding of the outcome variable.

$F = \frac{MS_{Treatment}}{MS_{Error}}$.

In an ANOVA with two factors, we calculate the F-statistic for each factor and their interaction. Significance is tested with $F$; interpretation focuses on planned factors and interactions. If $F$ is significant, we keep the beta coefficient in our model. If not, we may consider omitting it for a simpler model.

It turns out that ANOVAs are just a type of linear model in which the predictor variable is categorical. In practice, we can perform ANOVA using the `lm()` function in R, treating the categorical predictors as factors. This allows us to use the familiar beta notation and interpret ANOVA as a linear regression model with categorical predictors.

## Moving Beyond Regression and ANOVA to ANCOVA

As we've explored statistical modeling, we've understood the power of both regression and ANOVA. Now, let's merge them. ANCOVA allows us to examine relationships across different categories. 

Consider these questions that may arise in environmental research:

-   Are dbh (Diameter at Breast Height) and height related similarly for tulip poplars and oaks?

-   Are biomass and BTUs (British Thermal Units) related similarly for corn stover and Miscanthus?

-   Does the exposure to PFAS correlate with the lifetime incidence of cancer uniformly across low- and high-income American populations?

This specific subsection of general linear models is known as **Analysis of Covariance -- ANCOVA**. ANCOVA is a linear model with one continuous covariate and one categorical factor; it tests group differences **after** adjusting for the covariate and, first, whether slopes differ across groups.

------------------------------------------------------------------------

### Understanding ANCOVA

With ANCOVA, if we identify a significant difference between two linear relationships, our model will represent two distinct lines.

#### Conceptualizing the Transformation

Consider the following equations:

-   `y = 2x + 3`
-   `y = 3x - 4`

The question we pose is: What modifications are required to transform the first equation into the second? This involves determining the adjustments needed in terms of x (the slope) and the constant term. Understanding this transformation is key to grasping how ANCOVA allows us to compare different linear relationships within a single model framework.

## Dummy (Indicator) Variables and How They Work

Categorical predictors can be included in a linear model by coding them as **indicator variables**.

For a factor with $k$ groups, include $k-1$ indicators; the omitted group is the **reference**, and coefficients are interpreted relative to it.

**Example:** modeling **algal biomass** across three nutrient treatments—Control, Low N, and High N.

Use two indicators, `LowN` and `HighN`:

| Treatment | LowN | HighN |
| --------- | ---: | ----: |
| Control   |    0 |     0 |
| Low N     |    1 |     0 |
| High N    |    0 |     1 |


The model is $y=\beta_0+\beta_1,\text{LowN}+\beta_2,\text{HighN}+\varepsilon$,
where:

* $\beta_0$ is the mean for the Control group (reference),
* $\beta_1$ is Low N minus Control,
* $\beta_2$ is High N minus Control.

In R, `lm(y ~ Treatment)` creates these indicators automatically.

### Indicator Variables in Action

When an indicator variable is turned on (set to 1), it activates the terms multiplied by that variable. This can change the intercept, the slope, or both, depending on where the indicator appears in the model.

Consider two predictors: a continuous variable $x_l$ and an indicator $x_c$ (0 or 1).
Use this model:

$y = 2x_l + 1(x_l x_c) - 7x_c + 3$

**When $x_c = 0$** (indicator off), only baseline terms are active:
$y = 2x_l + 3$
Slope = 2, intercept = 3.

**When $x_c = 1$** (indicator on), new terms activate:
$y = (2x_l + 3) + (1x_l - 7) = 3x_l - 4$
Slope = 3, intercept = −4.


Turning on the indicator adds **+1 to the slope** and **–7 to the intercept**. 

This is how linear models represent different groups: the indicator ($x_c$) and its interaction with $x_l$ let the equation shift the intercept and slope for each group.

Later, ANCOVA will use this same logic to formally test:

* whether the **slopes differ** between groups (the $x_lx_c$ term), and
* if not, whether the **adjusted means** differ (the $x_c$ term).

#### Visualizing the Effect

The R plots below demonstrate the change in the regression line when the indicator variable is toggled between being active (`xc = 1`) and inactive (`xc = 0`).

**When `xc` = 0:**

*Equation Simplified:* `y = 2xl + 3`

**When `xc` = 1:**

*Equation Modified:* `y = (2xl + 3) + (1xl - 7)`

The plot on the left shows the regression line when the indicator variable is inactive. The equation simplifies, reflecting a scenario where the categorical variable does not influence the outcome.

The plot on the right illustrates the regression line when the indicator variable is active. The equation now includes additional terms, showing how the presence of the categorical variable changes the relationship between `xl` and `y`.

```{r}
# Graph for when xc = 0
# Equation becomes: y = 2xl + 3

# Example data frame
data_xc0 <- data.frame(
  xl = c(1, 2, 3, 4, 5))
data_xc0$y= (2 * data_xc0$xl + 3) 

# Plot for xc = 0
p1 <- ggplot(data_xc0, aes(x = xl, y = y)) +
  geom_line() +
  labs(title = "Regression Line when xc = 0",
       x = "Continuous Variable (xl)",
       y = "Dependent Variable (y)")+
  theme_classic()+
  ylim(-1,15)+
  ggtitle("Plot for xc=0")

# Graph for when xc = 1
# Equation becomes: y = (2xl + 3) + (1xl - 7)
data_xc1 <- data.frame(
  xl = c(1, 2, 3, 4, 5))

data_xc1$y <- ((2 * data_xc1$xl + 3) + (1 * data_xc1$xl - 7)) 


# Plot for xc = 1
p2 <- ggplot(data_xc1, aes(x = xl, y = y)) +
  geom_line() +
  labs(title = "Regression Line when xc = 1",
       x = "Continuous Variable (xl)",
       y = "Dependent Variable (y)")+
  theme_classic()+
  ylim(-1,15)+
  ggtitle("Plot for xc=1")

ggpubr::ggarrange(p1,p2, ncol=2)
```

**Key idea:** to change only intercepts, include $x_c$; to change slopes, include $x_l x_c$; to change both, include both. The corresponding parameters in a general model $y=\beta_0+\beta_1 x_l+\beta_2 x_c+\beta_3(x_l x_c)+\varepsilon$ are $\beta_2$ (intercept shift) and $\beta_3$ (slope difference).

### Interaction Term in Linear Models

**Interaction terms** tell us whether the relationship between one predictor and the response changes across levels of another predictor.

In other words: does the slope of $x_l$ depend on $x_c$?"

**Model Equation Explained:**

Let's break down a typical equation:

$$
y = \beta_0 + \beta_1x_l + \beta_2x_c + \beta_3x_lx_c + \epsilon
$$

the term $x_l x_c$ is the **interaction**, and $\beta_3$ is its coefficient.

| Term              | Meaning                                                                    |
| ----------------- | -------------------------------------------------------------------------- |
| $y$               | response variable (e.g., plant growth)                                     |
| $\beta_0$         | intercept: expected value when $x_l$ and $x_c$ are zero                   |
| $\beta_1 x_l$     | effect of the continuous predictor (main effect of $x_l$)                  |
| $\beta_2 x_c$     | effect of the categorical predictor (main effect of $x_c$)                 |
| $\beta_3 x_l x_c$ | **interaction:** how the slope of $x_l$ changes across categories of $x_c$ |
| $\varepsilon$     | residual variation not explained by the model                              |

If $\beta_3$ is significant, the slope of $x_l$ differs across groups.

If $\beta_3$ is not significant, the relationship between $x_l$ and $y$ is consistent across groups, and the lines are parallel.

### Five Versions of the Linear Model

This is our most complex version of the linear model so far. Remember what we just learned in case #1, and how each simpler version (#2–5) nests within it. 

1. **Full model with interaction ($\beta_0, \beta_1, \beta_2, \beta_3$)**

   * When both the categorical ($x_c$) and continuous ($x_l$) predictors are included, and the interaction term ($\beta_3$) is significant, the model produces **different slopes and intercepts** for each group.
   * Equations:

     * For $x_c = 0$: $y = \beta_0 + \beta_1x_l$
     * For $x_c = 1$: $y = (\beta_0 + \beta_2) + (\beta_1 + \beta_3)x_l$

2. **Model without interaction ($\beta_0, \beta_1, \beta_2$)**

   * If both predictors are significant but the interaction is not, the model simplifies to **parallel lines**—the slope of $x_l$ is the same across categories of $x_c$, but intercepts differ.
   * Equations:

     * For $x_c = 0$: $y = \beta_0 + \beta_1x_l$
     * For $x_c = 1$: $y = (\beta_0 + \beta_2) + \beta_1x_l$

3. **Simple linear regression ($\beta_0, \beta_1$)**

   * If only the continuous variable ($x_l$) matters, the model reduces to a single regression line.
   * Equation: $y = \beta_0 + \beta_1x_l$

4. **Two different means (t-test equivalent) ($\beta_0, \beta_2$)**

   * If only the categorical variable ($x_c$) matters, the model compares two means—just like a t-test.
   * Equations:

     * For $x_c = 0$: $y = \beta_0$
     * For $x_c = 1$: $y = \beta_0 + \beta_2$

5. **Single mean (one-sample model) ($\beta_0$)**

   * If neither variable contributes, the model collapses to a single overall mean.
   * Equation: $y = \beta_0$

### Terminology in LMs

Here are some key terms:

-   **Fixed Factor:** This refers to a categorical variable with specific, predefined categories. For example, if you're studying the effect of different seasons (spring, summer, autumn, winter) on plant growth, 'season' is a fixed factor because it represents specific, distinct categories of interest.

-   **Random Factor:** This is a categorical variable where the categories represent a random sample from a larger population. For instance, if you're examining the output quality from different machines in a factory, and these machines are randomly selected from a larger set, then 'machine' is a random factor.

-   **Covariate:** This is your continuous variable that varies along with the dependent variable (Y). In environmental studies, this could be something like temperature, rainfall, or pollution levels, which you suspect might influence your outcome of interest (like species distribution or plant growth).

------------------------------------------------------------------------

## So, What Exactly is ANCOVA?

Analysis of Covariance (ANCOVA) is a statistical method that combines the principles of ANOVA and regression. It's designed to compare categorical independent variables---such as different treatments or groups---while controlling for the influence of continuous variables. These continuous variables, known as covariates, are typically not the primary focus but could affect the outcome.

For ANCOVA, you need:

1)  A continuous dependent variable (the main effect you're studying)
2)  At least one continuous independent variable (covariate)
3)  At least one categorical independent variable (which can be either a fixed or random factor)

### The Null Hypothesis in ANCOVA

**Why this matters.** ANCOVA asks two questions in order. First, do groups have the **same slope** for the covariate–response relation. Only if slopes are the same do we ask whether groups differ in **adjusted means** after controlling for the covariate.

**Model.** For one covariate (x) and a two-level group indicator $g\in{0,1}$,

$y=\beta_0+\beta_1 x+\beta_2 g+\beta_3(x\times g)+\varepsilon .$

**Step 1: Slopes equal across groups?**
This tests whether the covariate effect is the same in each group.

* Null: $H_{0,\text{int}}:\beta_3=0$  (parallel lines, same slope)
* Alternative: $H_{A,\text{int}}:\beta_3\neq 0$  (non-parallel lines, different slopes)

**Interpretation if rejected.** Report the two fitted lines and describe how the slope differs by group. Do not test adjusted means, because the idea of a single adjusted mean per group assumes parallel slopes.

**Step 2: Adjusted means equal?** *(only if Step 1 retained parallel slopes)*
Fit the model without the interaction,

$y=\beta_0+\beta_1 x+\beta_2 g+\varepsilon .$

* Null: $H_{0,\text{group}}:\beta_2=0$  (no difference in adjusted means)
* Alternative: $H_{A,\text{group}}:\beta_2\neq 0$

**What the parameters mean.**

* $\beta_1): common slope of (y) on (x) when slopes are parallel.
* $\beta_2$: difference in group means **after** adjusting for $x$ (tested only when $\beta_3=0$).
* $\beta_3$: difference in slopes between groups.

**Intercepts and centering.** Intercepts are evaluated at (x=0). Compare intercepts only if you center the covariate, for example (\tilde x=x-\bar x), so that the intercepts represent group means at the average covariate value.

**What to report.**

* If $x\times g$ is significant: “The effect of $x$ differs by group,” then provide the two line equations and a figure with separate fitted lines.
* If $x\times g$ is not significant: report the common slope $\beta_1$ and the adjusted group difference $\beta_2$, with a figure showing parallel lines.

**Common pitfalls.**

* Interpreting main effects when the interaction is significant.
* Comparing raw intercepts when $x$ is not centered.
* Treating adjusted means as meaningful when slopes are not parallel.


### Testing the Null Hypothesis

**Conducting the Test**: When we perform ANCOVA, we examine the null hypothesis through an omnibus F-test. If we find sufficient evidence to reject the null hypothesis, we infer that significant differences exist among the adjusted means of the dependent variable across the categorical groups, even after considering the covariate.

**Interpreting the Findings**: Rejecting the null hypothesis indicates that the categorical independent variable has a significant impact on the dependent variable, aside from the covariate's effect. It's crucial to remember, though, that statistical significance doesn't automatically translate to practical significance. We must always consider the results within the specific context and aims of the study.\

::: callout-note
**ANCOVA ASSUMPTIONS**

-   Linearity within each group (covariate–response relation is linear).
-   **Homogeneity of slopes** (test via interaction).
-   Independent observations.
-   Approximately normal residuals and similar variances across groups.
-   Covariate measured without substantial error and not affected by treatment
:::

## Example

### Question

Imagine we're environmental scientists studying the impact of air pollution---a continuous variable---on bird species distribution, our dependent variable. We want to see if this relationship differs between urban and rural areas---our categorical variable. For this purpose, we've created a synthetic dataset to model the scenario.

### Model

With our data in hand, we'd construct an ANCOVA model to discern whether location type influences bird distribution, beyond what can be explained by pollution levels alone. This would involve coding our categorical variable (urban vs. rural) into a dummy variable and including air pollution as a covariate in the model. I've done this for you already by creating the synthetic dataset.

-   **Dataset Overview:** Our dataset comprises 200 observations, capturing bird species density in various urban and rural locations, along with corresponding levels of air pollution.

### Why We're Running the ANCOVA

-   **Research Question:** Does the effect of air pollution on bird species density differ between urban and rural areas?

-   **Expectation:** We predict that air pollution will have a more pronounced negative effect on bird species density in urban areas than in rural areas.

### Null and Alternative Hypotheses in our example ANCOVA

1.  **Omnibus Null Hypothesis** ($H_0$): The null hypothesis posits that once we adjust for air pollution, the mean bird species density does not differ between urban and rural areas.

$H_0: \mu^*_{\text{Urban}} = \mu^*_{\text{Rural}}$

$H_A: \mu^*_{\text{Urban}} \neq \mu^*_{\text{Rural}}$

Where $\mu^*_{\text{Urban}}$ and $\mu^*_{\text{Rural}}$ represent the adjusted means of bird species density for urban and rural areas, respectively..

2.  **Null Hypothesis for Slope** ($H_{01}$): This hypothesis examines if the effect of air pollution on bird species density is consistent between urban and rural areas.

    -   $H_{01}$: There is no interaction between air pollution and area type; the slopes are parallel.
    -   $H_{A1}$: There is an interaction; the effect of air pollution on bird species density differs between urban and rural areas.

$H_{01}: \beta_{\text{interaction}} = 0$

$H_{A1}: \beta_{\text{interaction}} \neq 0$

3.  **Null Hypothesis for Intercept** ($H_{02}$): If we find no significant interaction, we then consider the intercepts. This hypothesis tests whether the baseline bird species density, at zero air pollution, differs between urban and rural areas.

    -   $H_{02}$: The intercepts are the same, indicating no difference in bird species density between urban and rural areas at zero air pollution.

    -   $H_{A2}$: The intercepts differ, suggesting an inherent difference in bird species density between urban and rural areas, independent of air pollution.

$H_{02}: \beta_{0\text{,Urban}} = \beta_{0\text{,Rural}}$

$H_{A2}: \beta_{0\text{,Urban}} \neq \beta_{0\text{,Rural}}$

Rejecting $H_{02}$ would imply that there is an inherent difference in bird species density due to area type, even before considering the air pollution levels.

These hypotheses guide our ANCOVA model and analysis, aiming to isolate the effects of urbanization from the influence of air pollution on avian populations.

### Analysis

#### Explore the data

```{r}
# Set seed for reproducibility
set.seed(123)

# Generate data
n <- 200  # Number of observations
area_type <- sample(c("Urban", "Rural"), n, replace = TRUE)  # Randomly assign area type
air_pollution <- runif(n, min = 25, max = 150)  # Generate continuous air pollution levels

# Generate bird density data
# Assuming a simple linear relationship with air pollution and different intercepts for urban and rural
# Generate bird density data
# Adjusting the formula to avoid negative values
bird_density <- 10 + 0.2 * air_pollution - ifelse(area_type == "Urban", 5, 0) + rnorm(n, mean = 0, sd = 4)

# Ensure all bird densities are positive
bird_density <-(round(ifelse(bird_density < 0, 0, bird_density), 0))

#Reverse air pollution for negative relationship
air_pollution <- rev(round(air_pollution, 1))

# Combine into a data frame
data <- data.frame(AreaType = area_type, AirPollution = air_pollution, BirdDensity = bird_density)

# View the first few rows of the dataset
head(data)
```

Let's create a summary of our current dataset

```{r, echo=TRUE}
summary(data)
```

This will give us a quick statistical summary of our variables.

Now, let's visualize the distribution of bird density in different areas.

```{r, echo=TRUE}
# Boxplot for bird density by area type
ggplot(data, aes(AreaType, BirdDensity)) +
  geom_boxplot() + theme_classic() +
  labs(title = "Bird Density by Area Type", x = "Area Type", y = "Bird Density")

```

And the same for air pollution:

```{r, echo=TRUE}
ggplot(data, aes(AreaType, AirPollution)) +
  geom_boxplot() + theme_classic() +
  labs(title = "Air Pollution by Area Type", x = "Area Type", y = "Air Pollution")
```

### Results of the ANCOVA Model

The anova results

```{r, echo=TRUE}
ancova <- aov(BirdDensity ~ AirPollution*AreaType, data=data)
summary(ancova)
```

The linear model results

```{r, echo=TRUE}
ancova_lm <- lm(BirdDensity ~ AirPollution*AreaType, data=data)
summary(ancova_lm)
```

```{r, eval=FALSE, echo=TRUE}
fit_full <- lm(BirdDensity ~ AirPollution * AreaType, data = data)
anova(fit_full)         # tests interaction and main effects in the full model
summary(fit_full)

# If interaction is NOT significant:
fit_parallel <- lm(BirdDensity ~ AirPollution + AreaType, data = data)
anova(fit_parallel)
summary(fit_parallel)
```
#### Main Effects:

-   **Air Pollution**: The ANCOVA model does not find a significant main effect of air pollution on bird density (p = 0.7690 from ANOVA), indicating no consistent impact across the areas. However, the linear model's marginal p-value (p = 0.081641) suggests a trend that might have been significant with a larger sample size or reduced variability.

-   **Area Type**: There is a significant main effect of AreaType on bird density (p \< 0.001), indicating that there are considerable differences in bird density between rural and urban areas, with urban areas showing lower bird density.

#### Interpreting the Beta Coefficients:

-   **Intercept**: Reflecting the expected bird density in rural areas at zero air pollution, the intercept is quite high, indicating a relatively healthy bird population in the absence of pollution.

-   **Area Type**: The beta coefficient for our area type factor is significantly different from zero (p \< 0.001), suggesting a distinct difference in bird density between urban and rural areas. However, this effect is captured more precisely in the interaction term.

-   **Interaction Term**: The significant beta for the interaction term confirms our hypothesis that air pollution's effect on bird density is not uniform across urban and rural areas.

#### Visual Representation:

The scatter plot with regression lines will demonstrate these relationships visually, showing the differing trends for urban and rural areas suggested by the significant interaction term.

```{r}
# Scatter plot with regression lines
ggplot(data, aes(AirPollution, BirdDensity, color = AreaType)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic() +
  labs(title = "Bird Density vs Air Pollution by Area Type",
       x = "Air Pollution", y = "Bird Density")
```

```{r}
data <- data |> mutate(AirPollution_c = AirPollution - mean(AirPollution))
fit_centered <- lm(BirdDensity ~ AirPollution_c * AreaType, data = data)
summary(fit_centered)   # intercepts now compare groups at mean AirPollution
```

#### Interpretation:

The significant interaction means that air pollution’s effect on bird density depends on area type. The relationship differs between urban and rural sites. In urban areas, the slope is slightly positive (or less negative), suggesting that other urban factors may influence bird populations differently than in rural settings.

Because the main effect of air pollution is not significant on its own, this interaction points to a more complex ecological pattern: pollution alone doesn’t predict bird density without considering land context. Recognizing these interactions helps refine management strategies and better target conservation efforts for urban versus rural habitats.

## Chapter Summary

In this chapter, we introduced **Analysis of Covariance (ANCOVA)** as an extension of ANOVA that incorporates continuous covariates. We began by framing ANOVA as a linear model, setting the groundwork for understanding how ANCOVA adds both continuous and categorical predictors. Using examples such as bird species density across environments, we practiced formulating and testing null and alternative hypotheses. ANCOVA’s key strength is its ability to adjust for additional variables, helping us isolate the effect of each factor more accurately. Through real-world examples and clear, step-by-step reasoning, we built intuition for how hypothesis testing in ANCOVA connects to the broader family of linear models in environmental science.

