---
author:
  - Mallory L. Barnes
aliases: [redirects/glm-dummy-variables.html]
---

```{r, include = FALSE}
source("global_stuff.R")
knitr::opts_chunk$set(cache=FALSE)
```

# The Linear Model Framework: Dummy Variables and Interactions

This chapter is not part of E-538. It was originally written as part of the ANCOVA chapter, but it goes a level deeper than a first course in statistics needs to go --- it's the kind of material that belongs in a second course, one built around the **general linear model** as the unifying idea behind everything else (t-tests, ANOVA, regression, and ANCOVA are all the same model wearing different hats).

It's preserved here, out of the book's numbered sequence, because the material took real work to build and is useful the moment you're ready for it --- whether that's a future course, a research project that calls for it, or your own curiosity. If you've just finished the ANCOVA chapter and want to understand *why* an interaction term does what it does, mechanically, this is where that explanation lives.

```{r}
library(tidyverse)
library(ggpubr)
```

## Linear Models (LMs)

Linear Models are a class of models that encompass various types of statistical analyses, including ANOVA, and regression. They're flexible enough to handle different types of data and relationships. These assume the response variable is **continuous and normally distributed**. They include *t-tests, ANOVA, regression,* and *ANCOVA* as special cases.

A general **linear model** is the common framework that includes ANOVA, regression, and ANCOVA. Each uses the same residual-based logic: a continuous response $Y$ is modeled as a combination of one or more predictors plus error. We've seen a few different cases so far:

-   **No Groups and No Relationships (H0):**
    -   This scenario often emerges when our ANOVA or regression analysis yields non-significant results.
    -   **What We Report:** The grand mean and overall variance or standard deviation. Alternatively, we can use a confidence interval to encapsulate this information.
-   **Two or More Categories with Significantly Different Means (t-test, ANOVA):**
    -   Here, we delve into data where group means are distinct and significant.
    -   **What We Report:** Group means. In classical ANOVA, we might report a common variance or standard deviation, or opt for group-specific measures (potentially through confidence intervals), ensuring to note that variances are not significantly different. With Welch's t-test, we focus on group-specific variance or standard deviation.
-   **Two Continuous Variables with a Significant Linear or Monotonic Relationship (Regression):**
    -   This model applies when there's a significant linear relationship between two continuous variables.
    -   **What We Report:** The equation of the regression line, along with its confidence limits. We can also select specific x-values and provide confidence intervals for the predicted y-values at those points.

### Model Statements:

-   **No Groups/Relationships:** $y \sim N(\mu, \sigma^2)$
-   **Categories with Different Means:** $y_{i,j} = \mu + \tau_i + \varepsilon$ (t-test allows for unequal variance: $y_{i,j} = \mu + \tau_i + \varepsilon_i$)
-   **Linear Relationship:** $y = \beta_0 + \beta_1x + \varepsilon$

Linear models handle continuous $y$; for counts or non-Gaussian $y$, we use generalized linear models (not covered here).

### Flexibility and Applicability of LMs

LMs enable us to model more complex relationships beyond the basic categorical X with continuous Y (as in t-tests and ANOVA) or continuous X with continuous Y (as in regression). They can handle diverse data types and model relationships that aren't strictly linear, making them crucial for studying environmental systems.

So far, in our exploration of parametric tests, we have primarily focused on two types of relationships:

-   **A categorical X and a continuous Y:** This includes tests like the t-test and ANOVA.
-   **A continuous X and a continuous Y:** This is typically analyzed using linear regression.

### Understanding ANOVA as a linear model:

ANOVA, or Analysis of Variance, is traditionally viewed as a technique to compare means across multiple groups. However, at its core, ANOVA is a linear model. It's a special case where the predictors are categorical, not continuous. This distinction is important but subtle.

Example: Imagine a researcher is exploring the effects of metal contamination on the species richness of sessile marine invertebrates. They're particularly interested in the impact of copper and the orientation of the substrate on which these organisms live. To investigate this, they conduct a factorial experiment, measuring species richness across different levels of copper enrichment (None, Low, High) and substrate orientation (Vertical, Horizontal). This setup allows us to not only consider each factor separately but also examine their potential interaction.

The ANOVA framework provides three null hypotheses to test:

1.  There are no differences in species richness across the copper levels.

2.  There are no differences in species richness between substrate orientations.

3.  There is no interaction effect between copper levels and substrate orientation.

These hypotheses can be represented in a linear model as follows:

$y_{ijk} = \beta_0 + \beta_{Copper_i} + \beta_{Orientation_j} + \beta_{Copper \times Orientation_{ij}} + \varepsilon_{ijk}$

Here, $\beta_0$ is the grand mean (intercept), $\beta_{Copper_i}$ and $\beta_{Orientation_j}$ are the main effects of the copper levels and orientation, respectively, and $\beta_{Copper \times Orientation_{ij}}$ represents the interaction between these factors. The $\varepsilon_{ijk}$ term captures the residual variance, or the random deviation of each observation from the model prediction. The significance of each beta coefficient is tested to determine if it makes a meaningful contribution to the model.

When conducting an ANOVA, we're essentially fitting a linear model with categorical predictors. These predictors are represented by beta coefficients in our model, which show the unique contribution of each factor (like copper levels or substrate orientation) to the outcome variable (such as species richness). The critical question we ask is whether each beta coefficient is significant---does it make a meaningful difference to our model?

We test each term with an $F$-test. The F-statistic is a ratio that compares the amount of variance explained by a particular factor to the variance not explained by the model (within-group variance or error). A higher F-statistic indicates that the factor explains a significant portion of the variability in the outcome variable, while a lower F-statistic suggests that the factor does not contribute much to our understanding of the outcome variable.

$F = \frac{MS_{Between}}{MS_{Within}}$.

In an ANOVA with two factors, we calculate the F-statistic for each factor and their interaction. Significance is tested with $F$; interpretation focuses on planned factors and interactions. If $F$ is significant, we keep the beta coefficient in our model. If not, we may consider omitting it for a simpler model.

It turns out that ANOVAs are just a type of linear model in which the predictor variable is categorical. In practice, we can perform ANOVA using the `lm()` function in R, treating the categorical predictors as factors. This allows us to use the familiar beta notation and interpret ANOVA as a linear regression model with categorical predictors.

## Dummy (Indicator) Variables and How They Work

Categorical predictors can be included in a linear model by coding them as **indicator variables**.

For a factor with $k$ groups, include $k-1$ indicators; the omitted group is the **reference**, and coefficients are interpreted relative to it.

**Example:** modeling **algal biomass** across three nutrient treatments---Control, Low N, and High N.

Use two indicators, `LowN` and `HighN`:

| Treatment | LowN | HighN |
|-----------|-----:|------:|
| Control   |    0 |     0 |
| Low N     |    1 |     0 |
| High N    |    0 |     1 |

The model is $y=\beta_0+\beta_1,\text{LowN}+\beta_2,\text{HighN}+\varepsilon$, where:

-   $\beta_0$ is the mean for the Control group (reference),
-   $\beta_1$ is Low N minus Control,
-   $\beta_2$ is High N minus Control.

In R, `lm(y ~ Treatment)` creates these indicators automatically.

### Indicator Variables in Action

When an indicator variable is turned on (set to 1), it activates the terms multiplied by that variable. This can change the intercept, the slope, or both, depending on where the indicator appears in the model.

Consider two predictors: a continuous variable $x$ and an indicator $G$ (0 or 1). Use this model:

$y = 2x + 1(x G) - 7G + 3$

**When** $G = 0$ (indicator off), only baseline terms are active: $y = 2x + 3$ Slope = 2, intercept = 3.

**When** $G = 1$ (indicator on), new terms activate: $y = (2x + 3) + (1x - 7) = 3x - 4$ Slope = 3, intercept = −4.

Turning on the indicator adds **+1 to the slope** and **--7 to the intercept**.

This is how linear models represent different groups: the indicator ($G$) and its interaction with $x$ let the equation shift the intercept and slope for each group.

Later, ANCOVA will use this same logic to formally test:

-   whether the **slopes differ** between groups (the $xG$ term), and
-   if not, whether the **adjusted means** differ (the $G$ term).

#### Visualizing the Effect

The R plots below demonstrate the change in the regression line when the indicator variable is toggled between being active (`xc = 1`) and inactive (`xc = 0`).

**When `xc` = 0:**

*Equation Simplified:* `y = 2xl + 3`

**When `xc` = 1:**

*Equation Modified:* `y = (2xl + 3) + (1xl - 7)`

The left plot shows the line when the indicator is **off** ($G = 0$). The categorical variable has no influence here, so we see only the baseline slope and intercept.

The right plot shows the line when the indicator is **on** ($G = 1$). The extra terms activate, and the categorical variable now changes both the slope and intercept, producing a different line for the second group.

```{r}
# Graph for when xc = 0
# Equation becomes: y = 2xl + 3

# Example data frame
data_xc0 <- data.frame(
  xl = c(1, 2, 3, 4, 5))
data_xc0$y= (2 * data_xc0$xl + 3)

# Plot for xc = 0
p1 <- ggplot(data_xc0, aes(x = xl, y = y)) +
  geom_line() +
  labs(title = "Regression Line when xc = 0",
       x = "Continuous Variable (xl)",
       y = "Dependent Variable (y)")+
  theme_classic()+
  ylim(-1,15)+
  ggtitle("Plot for xc=0")

# Graph for when xc = 1
# Equation becomes: y = (2xl + 3) + (1xl - 7)
data_xc1 <- data.frame(
  xl = c(1, 2, 3, 4, 5))

data_xc1$y <- ((2 * data_xc1$xl + 3) + (1 * data_xc1$xl - 7))


# Plot for xc = 1
p2 <- ggplot(data_xc1, aes(x = xl, y = y)) +
  geom_line() +
  labs(title = "Regression Line when xc = 1",
       x = "Continuous Variable (xl)",
       y = "Dependent Variable (y)")+
  theme_classic()+
  ylim(-1,15)+
  ggtitle("Plot for xc=1")

ggpubr::ggarrange(p1,p2, ncol=2)
```

**Key idea:** to change only intercepts, include $G$; to change slopes, include $x G$; to change both, include both. The corresponding parameters in a general model $y=\beta_0+\beta_1 x+\beta_2 G+\beta_3(x G)+\varepsilon$ are $\beta_2$ (intercept shift) and $\beta_3$ (slope difference).

### Interaction Term in Linear Models

**Interaction terms** tell us whether the relationship between one predictor and the response changes across levels of another predictor.

In other words: does the slope of $x$ depend on $G$?"

**Model Equation Explained:**

Let's break down a typical equation:

$$
y = \beta_0 + \beta_1x + \beta_2G + \beta_3xG + \varepsilon
$$

the term $x G$ is the **interaction**, and $\beta_3$ is its coefficient.

| Term              | Meaning                                                                    |
|----------------|-------------------------------------------------------|
| $y$               | response variable (e.g., plant growth)                                     |
| $\beta_0$         | intercept: expected value when $x$ and $G$ are zero                    |
| $\beta_1 x$     | effect of the continuous predictor (main effect of $x$)                  |
| $\beta_2 G$     | effect of the categorical predictor (main effect of $G$)                 |
| $\beta_3 x G$ | **interaction:** how the slope of $x$ changes across categories of $G$ |
| $\varepsilon$     | residual variation not explained by the model                              |

If $\beta_3$ is significant, the slope of $x$ differs across groups.

If $\beta_3$ is not significant, the relationship between $x$ and $y$ is consistent across groups, and the lines are parallel.

### Five Versions of the Linear Model

This is our most complex version of the linear model so far. Remember what we just learned in case #1, and how each simpler version (#2--5) nests within it.

1.  **Full model with interaction (**$\beta_0, \beta_1, \beta_2, \beta_3$)

    -   When both the categorical ($G$) and continuous ($x$) predictors are included, and the interaction term ($\beta_3$) is significant, the model produces **different slopes and intercepts** for each group.

    -   Equations:

        -   For $G = 0$: $y = \beta_0 + \beta_1x$
        -   For $G = 1$: $y = (\beta_0 + \beta_2) + (\beta_1 + \beta_3)x$

2.  **Model without interaction (**$\beta_0, \beta_1, \beta_2$)

    -   If both predictors are significant but the interaction is not, the model simplifies to **parallel lines**---the slope of $x$ is the same across categories of $G$, but intercepts differ.

    -   Equations:

        -   For $G = 0$: $y = \beta_0 + \beta_1x$
        -   For $G = 1$: $y = (\beta_0 + \beta_2) + \beta_1x$

3.  **Simple linear regression (**$\beta_0, \beta_1$)

    -   If only the continuous variable ($x$) matters, the model reduces to a single regression line.
    -   Equation: $y = \beta_0 + \beta_1x$

4.  **Two different means (t-test equivalent) (**$\beta_0, \beta_2$)

    -   If only the categorical variable ($G$) matters, the model compares two means---just like a t-test.

    -   Equations:

        -   For $G = 0$: $y = \beta_0$
        -   For $G = 1$: $y = \beta_0 + \beta_2$

5.  **Single mean (one-sample model) (**$\beta_0$)

    -   If neither variable contributes, the model collapses to a single overall mean.
    -   Equation: $y = \beta_0$

## Chapter Summary

Every test you've learned this semester --- one-sample and paired t-tests, independent-samples t-tests, one-way ANOVA, simple regression, and ANCOVA --- is the same linear model, with terms added or dropped depending on what your predictors look like. Indicator (dummy) variables are the mechanism that lets a single equation represent different groups, different slopes, and different intercepts all at once. Once that clicks, "which test do I use" becomes a much smaller question than "what does my model need to represent" --- which is the mindset a second course in applied regression / GLMs is built around.
